### 2019-02-07 outline/outline.Rmd
added outline

---
title: "Quantifying acute and chronic oak decline severity using phenotypic descriptors"
author: "Jasen Finch"
date: "07/02/2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Authors:** Jasen Finch, Nathan Brown, John Draper, Sandra Denman

**Proposed journals:**

**Summary:**

**Keywords:** phenotyping, tree-decline, machine-learning, quantification

## Introduction

## Methods

## Results/Discussion

### Phenotypic descriptors

### Extended descriptors

### Decline Indexes

###

### Applications

## Conclusion

### 2019-02-08 outline/outline.Rmd
added summary

@@
 ---
-title: "Quantifying acute and chronic oak decline severity using phenotypic descriptors"
+title: "Indexes for measuring acute and chronic oak decline severity using phenotypic descriptors"
 author: "Jasen Finch"
 date: "07/02/2019"
-output: word_document
+output: pdf_document
 ---
@@
 
-**Authors:** Jasen Finch, Nathan Brown, John Draper, Sandra Denman
+**Authors:** Jasen Finch, Nathan Brown, Sandra Denman & John Draper
 
-**Proposed journals:**
+**Proposed journal:**
 
 **Summary:** 
+Two indexes, the phenotypic decline index (PDI) and the decline acuteness index (DAI), were developed to describe acute and chronic oak decline severity using a multitude of easily measureable phenotypic descriptors.
+Measured phenotypic descriptors include aspects of tree shape and size, crown condition and the presence of biotic agents.
+These were collected from 174 trees at 9 sites from across the south, east and west of England.
+Between sites differences were identified as a results of aspects of tree size which were corrected for using a mean centering adjustment.
+The decline indexes were then calculated using proximity values from an unsupervised random forest of the phenotypic descriptors, multidimensional scaling and min-max scaling.
+These indexes are simple measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a useful tool for forest monitoring and management.
 
-**Keywords:** phenotyping, tree-decline, machine-learning, quantification
+**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning, quantification
 
@@
 ## Methods
+
+### Phenotypic data collection
+
+###
 
@@
 
-### Decline Indexes
+### Cross site differences and correction
 
-###
+### Oak decline indexes
+
+### Comparison to assigned decline status labels
 
@@
 
-## Conclusion
+## Conclusions
+
+## Figures/Tables

### 2019-02-08 outline/outline.Rmd
added points to sections

@@
 title: "Indexes for measuring acute and chronic oak decline severity using phenotypic descriptors"
-author: "Jasen Finch"
-date: "07/02/2019"
 output: pdf_document
@@
 **Summary:** 
-Two indexes, the phenotypic decline index (PDI) and the decline acuteness index (DAI), were developed to describe acute and chronic oak decline severity using a multitude of easily measureable phenotypic descriptors.
+Two indexes, the Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to describe acute and chronic oak decline severity using a multitude of easily measureable phenotypic descriptors.
 Measured phenotypic descriptors include aspects of tree shape and size, crown condition and the presence of biotic agents.
@@
 
+- Acute and chronic decline associated phenotypes
+
 ## Methods
@@
 
-###
+### Data preparation and analysis
 
@@
 
-### Phenotypic descriptors
+### Phenotypic descriptors for oak decline
 
-### Extended descriptors
+- Descriptions of collected phenotypic desciptors
+
+- Definitions of additionally calculated features (crown condition, crown volume, bleed prevalence, *Agrillus* exit hole density)
 
@@
 
+- Analysis of cross site differences
+- Outline of cross site correction approach using mean centering adjustment
+
 ### Oak decline indexes
 
-### Comparison to assigned decline status labels
+- Association of phenotypic descriptors with manually assigned decline status
+- Decline index calculation
+- Comparison of indexes with manually assigned decline status labels
 
 ### Applications
+
+- Disscusion of applications to forestry monitoring and management
 

### 2019-02-08 outline/outline.Rmd
added to sections

@@
 
-**Proposed journal:**
+**Proposed journal:** Still undecided! ??Plant Methods, Forestry??
 
 **Summary:** 
-Two indexes, the Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to describe acute and chronic oak decline severity using a multitude of easily measureable phenotypic descriptors.
+Two indexes, the Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to describe acute and chronic oak decline severity using a multitude of easily measurable phenotypic descriptors.
 Measured phenotypic descriptors include aspects of tree shape and size, crown condition and the presence of biotic agents.
-These were collected from 174 trees at 9 sites from across the south, east and west of England.
-Between sites differences were identified as a results of aspects of tree size which were corrected for using a mean centering adjustment.
+These were collected from 174 trees exhibiting a spectrum of decline severity at 9 sites from across the south, east and west of England.
+Inter site differences were identified as a results of aspects of tree size which were corrected for using a mean centering adjustment.
 The decline indexes were then calculated using proximity values from an unsupervised random forest of the phenotypic descriptors, multidimensional scaling and min-max scaling.
-These indexes are simple measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a useful tool for forest monitoring and management.
+These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a useful tool for forest monitoring and management.
 
-**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning, quantification
+**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning
 
@@
 - Acute and chronic decline associated phenotypes
+- Manual assignment of decline status difficult prone to error for due to contribution of many factors.
 
@@
 
+- Field phenotyping protocol
+
 ### Data preparation and analysis
+
+- Aggregation of cardinal observations.
+- Random forest and MDS analyses
 
@@
 
-- Descriptions of collected phenotypic desciptors
+- Descriptions of collected phenotypic descriptors and manual status assignment
 
@@
 
-- Association of phenotypic descriptors with manually assigned decline status
+- Suitability of collected phenotypic descriptors for describing decline status
 - Decline index calculation
 - Comparison of indexes with manually assigned decline status labels
+- Index thresholds for category definitions (healthy, declining, severe decline etc.)
 
@@
 
-- Disscusion of applications to forestry monitoring and management
+- Discussion of applications to forestry monitoring and management
+- Sensitivity for integration with molecular data compared to categorical assignments
 
@@
 
+- Crown condition descriptors most important for describing overall decline severity in the PDI.
+- Bacterial bleed related descriptors are most important for describing decline acuteness.
+- Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment
+
 ## Figures/Tables
+
+**Figure 1:** Site differences MDS plot and feature importance (top ranked trends)
+
+**Figure 2:** Example of site correction effect on a descriptor
+
+**Figure 3:** Decline index scales (graphical representations of the indexes)
+
+**Figure 4:** Decline index descriptor contributions and trends of top ranked
+
+**Figure 5:** Decline indexes versus manual status assignments
+
+**Supplementary:**
+
+  - Descriptor associations with PDI
+  - Descriptor associations with DAI

### 2019-02-08 outline/outline.Rmd
small addition
compile to word

@@
 title: "Indexes for measuring acute and chronic oak decline severity using phenotypic descriptors"
-output: pdf_document
+output: word_document
 ---
@@
 Inter site differences were identified as a results of aspects of tree size which were corrected for using a mean centering adjustment.
-The decline indexes were then calculated using proximity values from an unsupervised random forest of the phenotypic descriptors, multidimensional scaling and min-max scaling.
+The decline indexes were then calculated using proximity values from an unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
 These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a useful tool for forest monitoring and management.
@@
 
+- Site descriptions
 - Field phenotyping protocol
@@
 
+**Table 1:** Overview of phenotypic descriptors
+
 **Figure 1:** Site differences MDS plot and feature importance (top ranked trends)

### 2019-07-03 outline/outline.Rmd
alterations to outline

@@
 ---
-title: "Indexes for measuring acute and chronic oak decline severity using phenotypic descriptors"
+title: "Indexes measures for oak decline severity using phenotypic descriptors"
 output: word_document
@@
 - Comparison of indexes with manually assigned decline status labels
+- Top ranked descriptor index response surfaces
 - Index thresholds for category definitions (healthy, declining, severe decline etc.)
@@
 
-**Figure 1:** Site differences MDS plot and feature importance (top ranked trends)
+**Figure 1:** 
+    - a) Site differences MDS plot
+    - b) Site difference feature importance
+    - c) Example of site correction effect on a descriptor
 
-**Figure 2:** Example of site correction effect on a descriptor
+**Figure 2:** Decline index scales (graphical representations of the indexes)
 
-**Figure 3:** Decline index scales (graphical representations of the indexes)
+**Figure 3:** Decline index descriptor contributions
 
-**Figure 4:** Decline index descriptor contributions and trends of top ranked
+**Figure 4:** Top ranked descriptor response surfaces
 

### 2019-07-03 outline/outline.Rmd
adjustment to outline

@@
     - b) Site difference feature importance
-    - c) Example of site correction effect on a descriptor
 
-**Figure 2:** Decline index scales (graphical representations of the indexes)
+**Figure 2:** Example of site correction effect on a descriptor
 
-**Figure 3:** Decline index descriptor contributions
+**Figure 3:** Decline index scales (graphical representations of the indexes)
 
-**Figure 4:** Top ranked descriptor response surfaces
+**Figure 4:** Decline index descriptor contributions
 
-**Figure 5:** Decline indexes versus manual status assignments
+**Figure 5:** Top ranked descriptor response surfaces
+
+**Figure 6:** Decline indexes versus manual status assignments
 

### 2019-07-12 manuscript/manuscript.Rmd
added manuscript

---
title: "Indexes measures for oak decline severity using phenotypic descriptors"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,cache = TRUE)
```

```{r library_load}
source('../code/library_load.R')
```

```{r prep}
source('../code/data_prep.R')
```

```{r figure_1}
source('../code/Figure_1_site_differences.R')

figure1 %>%
  wrap_plots()
```

### 2019-07-12 code/data_prep.R
added descriptor adjustment figure

@@ -28,3 +28,18 @@ phenoData <- {
 }
 
 DIs <- calcDIs()
+
+correctedPhenoData <- siteCorrection(phenoData) %>%
+  mutate(`Crown condition (%)` = crownCondition(`Missing crown (%)`,
+                                                `Crown transparency (%)`),
+         `Crown volume (m^3)` = crownVolume(`Crown radius (m)`,
+                                            `Total height (m)`,
+                                            `Lower crown height (m)`,
+                                            `Crown condition (%)`),
+         `Bleed prevalence (%)` = bleedPrevalence(`Active bleed size (cm)`,
+                                                  `Active bleeds`,
+                                                  `Black staining size (cm)`,
+                                                  `Black staining`,
+                                                  `Diameter at breast height (cm)`),
+         `Agrilus exit hole density (m^-2)` = agrilusExitHoleDensity(`Agrilus exit holes`,`Diameter at breast height (cm)`)
+  )

### 2019-07-12 code/descriptor_adjustment.R

@@ -1,0 +1,49 @@
+varSummary <- function(dat){
+  dat %>%
+    select(Location,ID,`Diameter at breast height (cm)`) %>%
+    group_by(Location) %>%
+    summarise(Mean = mean(`Diameter at breast height (cm)`),
+              SE = sd(`Diameter at breast height (cm)`)/ sqrt(n()))
+}
+
+dbh <- varSummary(phenoData)
+
+corrected_dbh <- varSummary(correctedPhenoData)
+
+overallMean <- mean(phenoData$`Diameter at breast height (cm)`)
+
+ylims <- c(
+  dbh %>% 
+    filter(Mean == min(Mean)) %>%
+    {.$Mean[1] - .$SE[1] - 1},
+  dbh %>% 
+    filter(Mean == max(Mean)) %>%
+    {.$Mean[1] + .$SE[1] + 1}
+)
+
+descriptor_adjustment <- list(
+  a = ggplot(dbh,aes(x = Location,y = Mean,)) +
+    geom_hline(yintercept = overallMean,linetype = 2,colour = 'red') +
+    geom_errorbar(aes(ymin = Mean - SE,ymax = Mean + SE),width = 0.5) +
+    geom_point(colour = 'steelblue') +
+    theme_bw(base_size = 11) +
+    theme(plot.title = element_text(face = 'bold'),
+          axis.text.x = element_text(angle = 20,hjust = 1),
+          axis.title = element_text(face = 'bold')) +
+    labs(title = 'Unadjusted',
+         y = 'Diameter at\nbreast height (cm)',
+         x = '') +
+    ylim(ylims[1],ylims[2]),
+  b = ggplot(corrected_dbh,aes(x = Location,y = Mean,)) +
+    geom_hline(yintercept = overallMean,linetype = 2,colour = 'red') +
+    geom_errorbar(aes(ymin = Mean - SE,ymax = Mean + SE),width = 0.5) +
+    geom_point(colour = 'steelblue') +
+    theme_bw(base_size = 11) +
+    theme(plot.title = element_text(face = 'bold'),
+          axis.text.x = element_text(angle = 20,hjust = 1),
+          axis.title = element_text(face = 'bold')) +
+    labs(title = 'Adjusted',
+         y = 'Adjusted diameter at\nbreast height (cm)',
+         x = '') +
+    ylim(ylims[1],ylims[2])
+)

### 2019-07-12 manuscript/manuscript.Rmd
added descriptor adjustment figure
added DI figure
added descriptor contributions figure

@@
 
-```{r figure_1}
-source('../code/Figure_1_site_differences.R')
+```{r site_differences}
+source('../code/site_differences.R')
 
-figure1 %>%
+site_differences %>%
   wrap_plots()
 ```
+
+```{r descriptor_adjustment,fig.cap='Error bars show ±1SE of the mean.\nDashed red line shows the overall mean.'}
+source('../code/descriptor_adjustment.R')
+
+descriptor_adjustment %>%
+  wrap_plots()
+```
+
+```{r DIspace}
+DIspace()
+```
+
+```{r}
+source('../code/descriptor_contributions.R')
+descriptor_contributions %>%
+  wrap_plots() + 
+  plot_annotation(title = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures',caption = 'Gini importance scores averaged across 100 repetitions')
+```

### 2019-07-12 code/data_prep.R

@@ -43,3 +43,9 @@ correctedPhenoData <- siteCorrection(phenoData) %>%
                                                   `Diameter at breast height (cm)`),
          `Agrilus exit hole density (m^-2)` = agrilusExitHoleDensity(`Agrilus exit holes`,`Diameter at breast height (cm)`)
   )
+
+analysisTable <- makeAnalysisTable(correctedPhenoData)
+
+set.seed(1234)
+PDIrf <- randomForest(analysisTable,DIs$DIs$PDI,ntree = 10000)
+DAIrf <- randomForest(analysisTable,DIs$DIs$DAI,ntree = 10000,mtry = 11)

### 2019-08-27 manuscript/manuscript.Rmd
added equations

--- 2019-07-12 manuscript/manuscript.Rmd
@@
 output: pdf_document
+keep_tex: yes
 ---
@@
 
-```{r figure_1}
-source('../code/Figure_1_site_differences.R')
+# Introduction
 
-figure1 %>%
+# Materials and methods
+
+# Results
+
+## Additional Descriptors
+
+**Crown condition (%)** - percentage of present crown proportional to the percentage crown density.
+
+$$c = (100 - m)(1-\frac{t}{100})$$
+
+$m$ = missing crown (%), $t$ = crown transparency (%)
+
+**Estimated crown volume (m<sup>3</sup>)** - estimated volume of crown as an ellipsoid proportional to the crown condition.
+
+$$v = \frac{c(h - l)}{150}\pi r^2$$
+
+$r$ = crown radius (m), $h$ = total height (m), $l$ = lower crown height (m), $c$ = crown condition (%)
+
+**Estimated bleed prevalence (%)** - estimated percentage of the surveyed trunk area affected by bleeds or black staining.
+
+$$p = \frac{a^2A + b^2B}{3d\pi}$$
+
+$a$ = average active bleed size (cm), $A$ = number of active bleeds, $b$ = average black stain size (cm), $B$ = black stain number, $d$ = diameter at breast height (cm)
+
+***Agrilus* exit hole density (m<sup>-2</sup>)** - density of observed *Agrilus* exit holes across the surveyed trunk area.
+$$e = \frac{h}{\frac{2d}{100}\pi}$$
+$h$ = number of *Agrilus* exit holes, $d$ = diameter at breast height (cm)
+```{r site_differences}
+source('../code/site_differences.R')
+site_differences %>%
   wrap_plots()
 ```
+```{r descriptor_adjustment,fig.cap='Error bars show ±1SE of the mean.\nDashed red line shows the overall mean.'}
+source('../code/descriptor_adjustment.R')
+descriptor_adjustment %>%
+  wrap_plots()
+```
+```{r DIspace}
+DIspace()
+```
+```{r}
+source('../code/descriptor_contributions.R')
+descriptor_contributions %>%
+  wrap_plots() + 
+  plot_annotation(title = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures',caption = 'Gini importance scores averaged across 100 repetitions')
+```
+# Discussion
+# Conclusions
+# References

### 2019-08-28 manuscript/figures.Rmd
moved figures into own file. Moved adjustment figure to supplementary file. Adjusted and finished site differences figure

---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,cache = TRUE)
```

```{r library_load}
source('../code/library_load.R')
```

```{r prep}
source('../code/data_prep.R')
```

# Figures

```{r site_differences,fig.cap='Differences between sites in phenotypic descriptors analysed by supervised Random Forest. a) Multidimensional scaling (MDS) plot of random forest proximity values of individual trees scaled to two dimensions. Mean margin value shown. b) Relative descriptor Gini importances measures.',fig.height=5}
source('../code/site_differences.R')
grid.arrange(site_differences$a,site_differences$b,nrow = 1)
```

```{r DI_space,fig.cap=''}
DIspace()
```

```{r descriptor_contributions, fig.cap = ''}
source('../code/descriptor_contributions.R')
descriptor_contributions %>%
  wrap_plots() + 
  plot_annotation(title = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures',caption = 'Gini importance scores averaged across 100 repetitions')
```

### 2019-08-28 manuscript/manuscript.Rmd
moved figures into own file. Moved adjustment figure to supplementary file. Adjusted and finished site differences figure

@@
 $h$ = number of *Agrilus* exit holes, $d$ = diameter at breast height (cm)
-```{r site_differences}
+
+# Discussion
+
+# Conclusions
+
+# References
+
+# Figures
+
+```{r site_differences,fig.cap='Site differences',fig.height=6}
 source('../code/site_differences.R')
-site_differences %>%
-  wrap_plots()
+grid.arrange(site_differences$a,site_differences$b,nrow = 1)
 ```
+
 ```{r descriptor_adjustment,fig.cap='Error bars show ±1SE of the mean.\nDashed red line shows the overall mean.'}
@@
 ```
-```{r DIspace}
+
+```{r DI_space,fig.cap=''}
 DIspace()
 ```
-```{r}
+
+```{r descriptor_contributions, fig.cap = ''}
 source('../code/descriptor_contributions.R')
@@
 ```
-# Discussion
-# Conclusions
-# References

### 2019-08-28 manuscript/supplementary.Rmd
moved figures into own file. Moved adjustment figure to supplementary file. Adjusted and finished site differences figure

--- 2019-07-12 manuscript/manuscript.Rmd
@@
----
-title: "Indexes measures for oak decline severity using phenotypic descriptors"
-output: pdf_document
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE,cache = TRUE)
-```
-
-```{r library_load}
-source('../code/library_load.R')
-```
-
-```{r prep}
-source('../code/data_prep.R')
-```
-
-```{r site_differences}
-source('../code/site_differences.R')
-
-site_differences %>%
-  wrap_plots()
-```
-
 ```{r descriptor_adjustment,fig.cap='Error bars show ±1SE of the mean.\nDashed red line shows the overall mean.'}
 source('../code/descriptor_adjustment.R')
-
 descriptor_adjustment %>%
@@
 ```
-
-```{r DIspace}
-DIspace()
-```
-
-```{r}
-source('../code/descriptor_contributions.R')
-descriptor_contributions %>%
-  wrap_plots() + 
-  plot_annotation(title = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures',caption = 'Gini importance scores averaged across 100 repetitions')
-```

### 2019-08-28 manuscript/figures.Rmd
added site location map figure

@@
 
-```{r library_load}
+```{r library_load,cache=FALSE}
 source('../code/library_load.R')
@@
 # Figures
+
+```{r siteLocations,fig.cap='Map showing locations of the nine sites used for collection of phenotypic descriptors.',fig.width=3,fig.height=5}
+source('../code/site_locations.R')
+site_locations_map
+```
 

### 2019-08-28 code/library_load.R

@@ -1,1 +1,1 @@
-pacman::p_load(tidyverse,pdi,ggthemes,randomForest,patchwork,gridExtra)
+pacman::p_load(maps,mapdata,pdi,ggthemes,randomForest,patchwork,gridExtra,ggrepel,tidyverse)

### 2019-08-28 code/site_differences.R

@@ -39,7 +39,12 @@ descriptorLabels <- locationImportance$Descriptor %>%
 
 site_differences <- list(
   a = ggplot() +
-    geom_point(data = locationMDS,aes(x = `Dimension 1`, y = `Dimension 2`,fill = Location),shape = 21,size = 3) +
+    geom_point(data = locationMDS,
+               aes(x = `Dimension 1`, 
+                   y = `Dimension 2`,
+                   fill = Location),
+               shape = 21,
+               size = 3) +
     geom_text(data = locationMargin,aes(x = x,y = y,label = label)) +
     scale_fill_ptol() +
     theme_bw() +

### 2019-08-28 manuscript/figures.Rmd
added PDI response surfaces

@@
 
-```{r DI_space,fig.cap=''}
+```{r DI_space,fig.cap='Two dimensional continuum of the phenotypic decline indexes for oak decline severity.',fig.height=3,fig.width=5}
 DIspace()
@@
 
-```{r descriptor_contributions, fig.cap = ''}
+```{r descriptor_contributions, fig.height = 5,fig.cap = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures.
+a) Phenotypic decline index (PDI).
+b) Decline acuteness index (DAI)'}
 source('../code/descriptor_contributions.R')
 descriptor_contributions %>%
-  wrap_plots() + 
-  plot_annotation(title = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures',caption = 'Gini importance scores averaged across 100 repetitions')
+  wrap_plots()
 ```
+
+```{r PDI_response_surfaces,fig.cap='PDI response surfaces for simulated descriptor values from trained Random Forest models.
+All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models',cache=FALSE}
+source('../code/response_surfaces.R')
+grid.arrange(PDI_response_plot$`a)`,PDI_response_plot$`b)`,PDI_response_plot$`c)`,PDI_response_plot$legend,nrow = 2)
+```

### 2019-08-28 code/library_load.R

@@ -1,1 +1,1 @@
-pacman::p_load(maps,mapdata,pdi,ggthemes,randomForest,patchwork,gridExtra,ggrepel,tidyverse)
+pacman::p_load(maps,mapdata,pdi,ggthemes,randomForest,patchwork,gridExtra,ggrepel,metR,tidyverse,ggpubr)

### 2019-08-28 code/response_surfaces.R

@@ -1,0 +1,90 @@
+spectrumTrees_PDI <- list(
+  `a)` = DIs$data[which(DIs$DIs$PDI == min(DIs$DIs$PDI)),],
+  `b)` = DIs$data[which(DIs$DIs$PDI == min(DIs$DIs$PDI)),],
+  `c)` = DIs$data[which(DIs$DIs$PDI == min(DIs$DIs$PDI)),]
+)
+
+variables <- names(DIs$data)[map_lgl(DIs$data,is.numeric)]
+
+ranges <- variables %>%
+  map(~{
+    DIs$data[,.] %>%
+      unlist() %>%
+      {tibble(min = min(.),
+              max = max(.))} %>%
+      mutate(center = min + (max - min)/2)
+  }) %>%
+  set_names(variables) %>%
+  bind_rows(.id = 'Descriptor')
+
+spectrumTrees_PDI$`a)`$`Crown radius (m)` <- ranges$max[ranges$Descriptor == 'Crown radius (m)']
+spectrumTrees_PDI$`b)`$`Crown radius (m)` <- ranges$center[ranges$Descriptor == 'Crown radius (m)']
+spectrumTrees_PDI$`c)`$`Crown radius (m)` <- ranges$min[ranges$Descriptor == 'Crown radius (m)']
+
+spectrumTrees_PDI$`a)`$`Total height (m)` <- ranges$max[ranges$Descriptor == 'Total height (m)']
+spectrumTrees_PDI$`b)`$`Total height (m)` <- ranges$center[ranges$Descriptor == 'Total height (m)']
+spectrumTrees_PDI$`c)`$`Total height (m)` <- ranges$min[ranges$Descriptor == 'Total height (m)']
+
+spectrumTrees_PDI$`a)`$`Lower crown height (m)` <- ranges$max[ranges$Descriptor == 'Lower crown height (m)']
+spectrumTrees_PDI$`b)`$`Lower crown height (m)` <- ranges$center[ranges$Descriptor == 'Lower crown height (m)']
+spectrumTrees_PDI$`c)`$`Lower crown height (m)` <- ranges$min[ranges$Descriptor == 'Lower crown height (m)']
+
+spectrumTrees_PDI$`a)`$`Diameter at breast height (cm)` <- ranges$max[ranges$Descriptor == 'Diameter at breast height (cm)']
+spectrumTrees_PDI$`b)`$`Diameter at breast height (cm)` <- ranges$center[ranges$Descriptor == 'Diameter at breast height (cm)']
[…]

### 2019-08-29 manuscript/figures.Rmd
added DAI response surfaces

@@
 ```{r PDI_response_surfaces,fig.cap='PDI response surfaces for simulated descriptor values from trained Random Forest models.
-All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models',cache=FALSE}
-source('../code/response_surfaces.R')
+All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models',cache=TRUE}
+source('../code/PDI_response_surfaces.R')
 grid.arrange(PDI_response_plot$`a)`,PDI_response_plot$`b)`,PDI_response_plot$`c)`,PDI_response_plot$legend,nrow = 2)
 ```
+
+```{r DAI_response_surfaces,fig.height = 8,fig.cap='DAI response surfaces for simulated descriptor values from trained Random Forest models.
+All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models',cache=FALSE}
+source('../code/DAI_response_surfaces.R')
+grid.arrange(DAI_response_plot$`a)`,DAI_response_plot$`b)`,DAI_response_plot$`c)`,DAI_response_plot$`d)`,DAI_response_plot$`e)`,DAI_response_plot$`f)`,DAI_response_plot$legend,ncol = 3,layout_matrix = rbind(c(1,4,7),c(2,5,7),c(3,6,7)),widths = c(4,4,1))
+```

### 2019-08-29 code/DAI_response_surfaces.R

@@ -1,0 +1,186 @@
+spectrumTrees_DAI <- list(
+  `a)` = DIs$data[which(DIs$DIs$DAI == min(abs(DIs$DIs$DAI))),],
+  `b)` = DIs$data[which(abs(DIs$DIs$DAI - 0.5) == min(abs(DIs$DIs$DAI - 0.5))),],
+  `c)` = DIs$data[which(DIs$DIs$DAI == max(DIs$DIs$DAI)),],
+  `d)` = DIs$data[which(DIs$DIs$DAI == min(abs(DIs$DIs$DAI))),],
+  `e)` = DIs$data[which(abs(DIs$DIs$DAI + 0.5) == min(abs(DIs$DIs$DAI + 0.5))),],
+  `f)` = DIs$data[which(DIs$DIs$DAI == min(DIs$DIs$DAI)),]
+)
+
+variables <- names(DIs$data)[map_lgl(DIs$data,is.numeric)]
+
+ranges <- variables %>%
+  map(~{
+    DIs$data[,.] %>%
+      unlist() %>%
+      {tibble(min = min(.),
+              max = max(.))} %>%
+      mutate(center = min + (max - min)/2)
+  }) %>%
+  set_names(variables) %>%
+  bind_rows(.id = 'Descriptor')
+
+spectrumTrees_DAI$`a)`$`Active bleeds` <- ranges$min[ranges$Descriptor == 'Active bleeds']
+spectrumTrees_DAI$`b)`$`Active bleeds` <- ranges$center[ranges$Descriptor == 'Active bleeds'] / 16
+spectrumTrees_DAI$`c)`$`Active bleeds` <- ranges$max[ranges$Descriptor == 'Active bleeds'] / 16
+
+spectrumTrees_DAI$`a)`$`Black staining` <- ranges$min[ranges$Descriptor == 'Black staining']
+spectrumTrees_DAI$`b)`$`Black staining` <- ranges$center[ranges$Descriptor == 'Black staining'] / 16
+spectrumTrees_DAI$`c)`$`Black staining` <- ranges$max[ranges$Descriptor == 'Black staining'] / 16
+
+spectrumTrees_DAI$`d)`$`Crown radius (m)` <- ranges$max[ranges$Descriptor == 'Crown radius (m)']
+spectrumTrees_DAI$`e)`$`Crown radius (m)` <- ranges$center[ranges$Descriptor == 'Crown radius (m)']
[…]

### 2019-08-29 code/scatter_plots.R

@@ -1,0 +1,61 @@
+tuneRF(DIs$data,DIs$DIs$DAI,ntreeTry = 500)
+
+mtry <- seq(2,26,by = 4)
+ntree <- c(10,100,1000,10000,100000)
+
+a <- ntree %>%
+  map(~{
+    n <- .
+    map(mtry,~{
+      m <- .
+      randomForest(DIs$data,y = DIs$DIs$DAI,ntree = n,mtry = m) %>%
+        .$predicted %>%
+        {abs(. - DIs$DIs$DAI)} %>%
+        mean() %>%
+        tibble(MAE = .)
+    }) %>%
+      set_names(mtry) %>%
+      bind_rows(.id = 'mtry')
+  }) %>%
+  set_names(ntree) %>%
+  bind_rows(.id = 'ntree')
+
+DIs$data %>%
+  rowid_to_column(var = 'ID') %>%
+  bind_cols(select(DIs$DIs,DAI)) %>%
+  select_if(is.numeric) %>%
+  gather(Descriptor,Value,-ID,-DAI) %>%
+  ggplot(aes(x = Value,y = DAI)) +
+  geom_point(shape = 21,fill = ptol_pal()(1)) +
+  theme_bw() +
+  facet_wrap(~Descriptor,scales = 'free')
+
+DIs$data %>%
+  rowid_to_column(var = 'ID') %>%
+  bind_cols(select(DIs$DIs,PDI)) %>%
+  select_if(is.numeric) %>%
+  gather(Descriptor,Value,-ID,-PDI) %>%
+  ggplot(aes(x = Value,y = PDI)) +
+  geom_point(shape = 21,fill = ptol_pal()(1)) +
+  theme_bw() +
+  facet_wrap(~Descriptor,scales = 'free')
+
+DIs$data %>%
+  select_if(is.factor) %>%
+  rowid_to_column(var = 'ID') %>%
+  bind_cols(select(DIs$DIs,DAI)) %>%
+  gather(Descriptor,Value,-ID,-DAI) %>%
+  ggplot(aes(x = Value,y = DAI)) +
+  geom_point(shape = 21,fill = ptol_pal()(1)) +
+  theme_bw() +
+  facet_wrap(~Descriptor,scales = 'free')
+
+DIs$data %>%
+  select_if(is.factor) %>%
+  rowid_to_column(var = 'ID') %>%
+  bind_cols(select(DIs$DIs,PDI)) %>%
[…]

### 2019-09-12 manuscript/figures.Rmd
fixed errors from addition of crown contact

--- 2019-08-28 manuscript/figures.Rmd
@@
 ---
+title: Figures
 output: pdf_document
@@
 ```
-
-# Figures
 
@@
 
-```{r PDI_response_surfaces,fig.cap='PDI response surfaces for simulated descriptor values from trained Random Forest models.
+```{r PDI_response_surfaces,fig.height = 6,fig.cap='PDI response surfaces for simulated descriptor values from trained Random Forest models.
+All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models',cache=TRUE}
+source('../code/PDI_response_surfaces.R')
+grid.arrange(PDI_response_plot$`a)`,PDI_response_plot$`b)`,PDI_response_plot$`c)`,PDI_response_plot$legend,nrow = 2,newpage = F)
+```
+
+```{r DAI_response_surfaces,fig.height = 8,fig.cap='DAI response surfaces for simulated descriptor values from trained Random Forest models.
 All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models',cache=FALSE}
-source('../code/response_surfaces.R')
-grid.arrange(PDI_response_plot$`a)`,PDI_response_plot$`b)`,PDI_response_plot$`c)`,PDI_response_plot$legend,nrow = 2)
+source('../code/DAI_response_surfaces.R')
+grid.arrange(DAI_response_plot$`a)`,DAI_response_plot$`b)`,DAI_response_plot$`c)`,DAI_response_plot$`d)`,DAI_response_plot$`e)`,DAI_response_plot$`f)`,DAI_response_plot$legend,ncol = 3,layout_matrix = rbind(c(1,4,7),c(2,5,7),c(3,6,7)),widths = c(4,4,1),newpage = F)
 ```

### 2019-09-12 code/DAI_response_surfaces.R

@@ -1,8 +1,8 @@
 spectrumTrees_DAI <- list(
-  `a)` = DIs$data[which(DIs$DIs$DAI == min(abs(DIs$DIs$DAI))),],
+  `a)` = DIs$data[which(DIs$DIs$DAI == -min(abs(DIs$DIs$DAI))),],
   `b)` = DIs$data[which(abs(DIs$DIs$DAI - 0.5) == min(abs(DIs$DIs$DAI - 0.5))),],
   `c)` = DIs$data[which(DIs$DIs$DAI == max(DIs$DIs$DAI)),],
-  `d)` = DIs$data[which(DIs$DIs$DAI == min(abs(DIs$DIs$DAI))),],
+  `d)` = DIs$data[which(DIs$DIs$DAI == -min(abs(DIs$DIs$DAI))),],
   `e)` = DIs$data[which(abs(DIs$DIs$DAI + 0.5) == min(abs(DIs$DIs$DAI + 0.5))),],
   `f)` = DIs$data[which(DIs$DIs$DAI == min(DIs$DIs$DAI)),]
 )
@@ -20,13 +20,23 @@ ranges <- variables %>%
   set_names(variables) %>%
   bind_rows(.id = 'Descriptor')
 
-spectrumTrees_DAI$`a)`$`Active bleeds` <- ranges$min[ranges$Descriptor == 'Active bleeds']
-spectrumTrees_DAI$`b)`$`Active bleeds` <- ranges$center[ranges$Descriptor == 'Active bleeds'] / 16
-spectrumTrees_DAI$`c)`$`Active bleeds` <- ranges$max[ranges$Descriptor == 'Active bleeds'] / 16
+# spectrumTrees_DAI$`a)`$`Active bleeds` <- ranges$min[ranges$Descriptor == 'Active bleeds']
+# spectrumTrees_DAI$`b)`$`Active bleeds` <- ranges$center[ranges$Descriptor == 'Active bleeds'] / 16
+# spectrumTrees_DAI$`c)`$`Active bleeds` <- ranges$max[ranges$Descriptor == 'Active bleeds'] / 16
 
-spectrumTrees_DAI$`a)`$`Black staining` <- ranges$min[ranges$Descriptor == 'Black staining']
-spectrumTrees_DAI$`b)`$`Black staining` <- ranges$center[ranges$Descriptor == 'Black staining'] / 16
[…]

### 2019-09-12 code/descriptor_contributions.R

@@ -15,7 +15,7 @@ descriptor_contributions <- list(
       as.character() %>%
       {c(.[1:5],
          expression(Agrillus~exit~hole~density ( m^-2 ) ),
-         .[7:33],
+         .[7:34],
          expression(Crown~volume ( m^3 ) )
       )}
     
@@ -43,11 +43,11 @@ descriptor_contributions <- list(
     
     descriptorLabels <- dat$Feature %>%
       as.character() %>%
-      {c(.[1:10],
+      {c(.[1:11],
          expression(Agrillus~exit~hole~density ( m^-2 ) ),
-         .[12:30],
+         .[13:31],
          expression(Crown~volume ( m^3 ) ),
-         .[32:34]
+         .[33:35]
       )}
     
     ggplot(dat,aes(x = Feature,y = `Relative Importance`)) +

### 2019-09-12 code/site_differences.R

@@ -34,7 +34,7 @@ descriptorLabels <- locationImportance$Descriptor %>%
                       expression(Agrillus~exit~hole~density ( m^-2 ) ),
                       .[8:25],
                       expression(Crown~volume ( m^3 ) ),
-                      .[27:34])
+                      .[27:35])
   }
 
 site_differences <- list(

### 2019-09-12 manuscript/manuscript.Rmd
added word count, spelling and grammar checks

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE,cache = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
@@
 source('../code/library_load.R')
-```
-
-```{r prep}
-source('../code/data_prep.R')
 ```
@@
 
-# Figures
-
-```{r site_differences,fig.cap='Site differences',fig.height=6}
-source('../code/site_differences.R')
-grid.arrange(site_differences$a,site_differences$b,nrow = 1)
+```{r check}
+source('../code/check.R')
+check()
 ```
-
-```{r descriptor_adjustment,fig.cap='Error bars show ±1SE of the mean.\nDashed red line shows the overall mean.'}
-source('../code/descriptor_adjustment.R')
-descriptor_adjustment %>%
-  wrap_plots()
-```
-
-```{r DI_space,fig.cap=''}
-DIspace()
-```
-
-```{r descriptor_contributions, fig.cap = ''}
-source('../code/descriptor_contributions.R')
-descriptor_contributions %>%
-  wrap_plots() + 
-  plot_annotation(title = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures',caption = 'Gini importance scores averaged across 100 repetitions')
-```

### 2019-09-12 manuscript/manuscript.Rmd
added abstract and more section headings

@@
 
-# Introduction
+## Abstract
 
-# Materials and methods
+Two indexes, the Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to describe acute and chronic oak decline severity using a multitude of easily measurable phenotypic descriptors.
+Measured phenotypic descriptors include aspects of tree shape and size, crown condition and the presence of biotic agents.
+These were collected from 174 trees exhibiting a spectrum of decline severity at 9 sites from across the south, east and west of England.
+Inter site differences were identified as a results of aspects of tree size which were corrected for using a mean centring adjustment.
+The decline indexes were then calculated using proximity values from an unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
+These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a useful tool for forest monitoring and management.
 
-# Results
+## Introduction
 
-## Additional Descriptors
+## Materials and methods
+
+### Phenotypic data collection
+
+### Phenotypic descriptor preparation and analysis
+
+## Results
+
+### Phenotypic descriptors
+
+The following four additional descriptors were calculated:
 
@@
 
-**Estimated crown volume (m<sup>3</sup>)** - estimated volume of crown as an ellipsoid proportional to the crown condition.
+**Estimated crown volume (m^3^)** - estimated volume of crown as an ellipsoid proportional to the crown condition.
 
@@
 
-***Agrilus* exit hole density (m<sup>-2</sup>)** - density of observed *Agrilus* exit holes across the surveyed trunk area.
+***Agrilus* exit hole density (m^-2^)** - density of observed *Agrilus* exit holes across the surveyed trunk area.
 $$e = \frac{h}{\frac{2d}{100}\pi}$$
@@
 
-# Discussion
+### Tree size descriptor adjustment by site
 
-# Conclusions
+### Phenotypic decline and decline acuteness indexes
 
-# References
+## Discussion
 
-```{r check}
+### Applications
+
+## Conclusions
+
+## References
+
+```{r check,comment=''}
 source('../code/check.R')
-check()
+manuscript_check()
 ```

### 2019-09-12 code/check.R

@@ -1,6 +1,7 @@
 
-check <- function(file = 'manuscript.Rmd'){
+manuscript_check <- function(file = 'manuscript.Rmd'){
   
+  cat('######### Manuscript Checks #########\n')
   cat('\nWord count')
   
   suppressMessages({

### 2019-11-11 manuscript/figures.Rmd
warnings and messages now not included. Sorted library loading errors
figures included within manuscript

@@
----
-title: Figures
-output: pdf_document
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE,cache = TRUE)
-```
-
-```{r library_load,cache=FALSE}
-source('../code/library_load.R')
-```
-
-```{r prep}
-source('../code/data_prep.R')
-```
-
 ```{r siteLocations,fig.cap='Map showing locations of the nine sites used for collection of phenotypic descriptors.',fig.width=3,fig.height=5}
@@
 descriptor_contributions %>%
-  wrap_plots()
+  patchwork::wrap_plots()
 ```

### 2019-11-11 code/library_load.R

@@ -1,3 +1,3 @@
 pacman::p_load(maps,mapdata,pdi,ggthemes,randomForest,
-               thomasp85/patchwork,gridExtra,ggrepel,metR,tidyverse,
-               ggpubr,wordcountaddin,spelling,ropenscilabs/gramr)
+               patchwork,gridExtra,ggrepel,metR,tidyverse,
+               ggpubr,wordcountaddin,spelling,gramr)

### 2019-11-11 manuscript/figures.Rmd
updated DAI response surface plots

@@
 ```{r PDI_response_surfaces,fig.height = 6,fig.cap='PDI response surfaces for simulated descriptor values from trained Random Forest models.
-All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models',cache=TRUE}
+All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models'}
 source('../code/PDI_response_surfaces.R')

### 2019-11-11 manuscript/manuscript.Rmd
warnings and messages now not included. Sorted library loading errors
figures included within manuscript
updated DAI response surface plots

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE,cache = TRUE,warning = FALSE,message = FALSE)
 ```
 
-```{r library_load}
+```{r library_load,cache=FALSE}
 source('../code/library_load.R')
+```
+
+```{r prep}
+source('../code/data_prep.R')
 ```
@@
 
-## References
+## Figures
+
+```{r figures,child='figures.Rmd'}
+```
+
+## Checks
 
@@
 ```
+
+## References

### 2019-11-11 code/DAI_response_surfaces.R

@@ -1,8 +1,8 @@
 spectrumTrees_DAI <- list(
-  `a)` = DIs$data[which(DIs$DIs$DAI == -min(abs(DIs$DIs$DAI))),],
+  `a)` = DIs$data[which(abs(DIs$DIs$DAI) == min(abs(DIs$DIs$DAI))),],
   `b)` = DIs$data[which(abs(DIs$DIs$DAI - 0.5) == min(abs(DIs$DIs$DAI - 0.5))),],
   `c)` = DIs$data[which(DIs$DIs$DAI == max(DIs$DIs$DAI)),],
-  `d)` = DIs$data[which(DIs$DIs$DAI == -min(abs(DIs$DIs$DAI))),],
+  `d)` = DIs$data[which(abs(DIs$DIs$DAI) == min(abs(DIs$DIs$DAI))),],
   `e)` = DIs$data[which(abs(DIs$DIs$DAI + 0.5) == min(abs(DIs$DIs$DAI + 0.5))),],
   `f)` = DIs$data[which(DIs$DIs$DAI == min(DIs$DIs$DAI)),]
 )
@@ -20,23 +20,17 @@ ranges <- variables %>%
   set_names(variables) %>%
   bind_rows(.id = 'Descriptor')
 
-# spectrumTrees_DAI$`a)`$`Active bleeds` <- ranges$min[ranges$Descriptor == 'Active bleeds']
-# spectrumTrees_DAI$`b)`$`Active bleeds` <- ranges$center[ranges$Descriptor == 'Active bleeds'] / 16
-# spectrumTrees_DAI$`c)`$`Active bleeds` <- ranges$max[ranges$Descriptor == 'Active bleeds'] / 16
+spectrumTrees_DAI$`a)`$`Active bleeds` <- ranges$min[ranges$Descriptor == 'Active bleeds']
+spectrumTrees_DAI$`b)`$`Active bleeds` <- {ranges$center[ranges$Descriptor == 'Active bleeds'] / 16} %>%
+  round()
+spectrumTrees_DAI$`c)`$`Active bleeds` <- {ranges$max[ranges$Descriptor == 'Active bleeds'] / 16} %>%
+  round()
 
-# spectrumTrees_DAI$`a)`$`Active bleed size (cm)` <- ranges$min[ranges$Descriptor == 'Active bleed size (cm)']
-spectrumTrees_DAI$`a)`$`Active bleed size (cm)` <- 1 / 4
[…]

### 2019-11-11 manuscript/figures.Rmd
added phenotypic descriptor table

@@
 ```{r DAI_response_surfaces,fig.height = 8,fig.cap='DAI response surfaces for simulated descriptor values from trained Random Forest models.
-All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models',cache=FALSE}
+All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models'}
 source('../code/DAI_response_surfaces.R')

### 2019-11-11 manuscript/tables.Rmd
added phenotypic descriptor table

```{r phenotypic_descriptors,cache=FALSE}
source('../code/phenotypic_descriptors.R')
pt %>%
  kable()
```

```{r resampled_results}

```

### 2019-11-11 outline/outline.Rmd
added phenotypic descriptor table

@@
 title: "Indexes measures for oak decline severity using phenotypic descriptors"
-output: word_document
+output: pdf_document
 ---
@@
 
+**Table 2:** Resampled index performance results
+
 **Figure 1:**
+
   - a) Site differences MDS plot

### 2019-11-11 code/library_load.R

@@ -1,3 +1,3 @@
 pacman::p_load(maps,mapdata,pdi,ggthemes,randomForest,
                patchwork,gridExtra,ggrepel,metR,tidyverse,
-               ggpubr,wordcountaddin,spelling,gramr)
+               ggpubr,wordcountaddin,spelling,gramr,knitr)

### 2019-11-11 code/phenotypic_descriptors.R

@@ -1,0 +1,48 @@
+
+pt <- analysisTable %>%
+  {tibble(Descriptor = names(.),`Data Type` = map_chr(.,class),Description = NA,`Categories / units` = NA,Reference = NA)} %>%
+  filter(!(Descriptor %in% c('Crown condition (%)','Crown volume (m^3)','Bleed prevalence (%)','Agrilus exit hole density (m^-2)')))
+
+type <- tibble(`Descriptor Type` = c(rep('tree size',5),
+                                     rep('crown',13),
+                                     rep('trunk',13)),
+               Descriptor = c('Diameter at breast height (cm)',
+                              'Crown radius (m)',
+                              'Total height (m)',
+                              'Lower crown height (m)',
+                              'Timber height (m)',
+                              'Missing crown (%)',
+                              'Crown transparency (%)',
+                              'Crown contact (%)',
+                              'Crown mildew',
+                              'Branch epicormics',
+                              'Crown fruiting bodies',
+                              'Dieback location',
+                              'Dieback type','Insect defoliation',
+                              'Insect defoliation type',
+                              'Pruning / branch loss',
+                              'Canopy closure',
+                              'Social class',
+                              'Agrilus exit holes',
+                              'Stem epicormics',
+                              'Dead stem tissue',
+                              'Ground level fruiting bodies',
+                              'Stem fruiting bodies',
+                              'Oval shaped exit holes',
+                              'Small circular shaped exit holes',
+                              'Active bleeds',
+                              'Active bleed size (cm)',
+                              'Black staining',
+                              'Black staining size (cm)',
+                              'Calloused wound',
+                              'Calloused wound size (cm)')
+)
+
+pt <- pt %>%
+  left_join(type, by = "Descriptor") %>%
+  mutate(Descriptor = str_to_lower(Descriptor)) %>%
+  select(`Descriptor Type`,everything()) %>%
+  arrange(`Descriptor Type`,Descriptor)
+
+pt$`Data Type`[pt$`Data Type` == 'numeric'] <- 'continous'
+pt$`Data Type`[pt$`Data Type` == 'factor'] <- 'categorical'

### 2019-11-11 manuscript/manuscript.Rmd
added phenotypic descriptor table
output format change to pdf_document2

@@
 ---
-title: "Indexes measures for oak decline severity using phenotypic descriptors"
-output: pdf_document
+title: "Index measures for oak decline severity using phenotypic descriptors"
+output: 
+  bookdown::pdf_document2:
+   toc: false
+   number_sections: false
 keep_tex: yes
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE,cache = TRUE,warning = FALSE,message = FALSE)
+knitr::opts_chunk$set(echo = FALSE,
+                      cache = TRUE,
+                      fig.pos = 'H',
+                      warning = FALSE,
+                      message = FALSE)
 ```
@@
 
+Oak decline syndromes are facilitated by a variety of both biotic and abiotic factors
+
 ## Materials and methods
@@
 ### Phenotypic descriptor preparation and analysis
+
+R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and analysis of phenotypic data.
 
@@
 
+Large phenotypic differences were found between the sites by supervised random forest (Figure \@ref(fig:siteLocations))
+
 ### Phenotypic decline and decline acuteness indexes
@@
 
+\newpage
+## Tables
+
+```{r tables,child='tables.Rmd'}
+```
+
+\newpage
 ## Figures
@@
 
+\newpage
 ## Checks
 
-```{r check,comment=''}
+## Figures
+
+```{r figures,child='figures.Rmd'}
+
+```
+
+```{r check,comment='',cache=FALSE}
 source('../code/check.R')
@@
 
+\newpage
 ## References

### 2019-11-11 code/figures/check.R

@@ -1,6 +1,8 @@
 
 manuscript_check <- function(file = 'manuscript.Rmd'){
   
+  options(knitr.table.format = 'markdown')
+  
   cat('######### Manuscript Checks #########\n')
   cat('\nWord count')
   
@@ -9,12 +11,12 @@ manuscript_check <- function(file = 'manuscript.Rmd'){
       print()
   })
   
-  cat('\nSpelling')
+  cat('\nSpelling\n\n')
   
   spelling::spell_check_files(file,lang = 'en_GB') %>%
     print()
   
-  cat('\nGrammar')
+  cat('\nStyle')
   
   gramr::write_good_file(file) %>%
     print()

### 2019-11-11 manuscript/figures.Rmd
big rearrange

@@
 ```{r siteLocations,fig.cap='Map showing locations of the nine sites used for collection of phenotypic descriptors.',fig.width=3,fig.height=5}
-source('../code/site_locations.R')
+source('../code/figures/site_locations.R')
 site_locations_map
@@
 b) Relative descriptor Gini importances measures.',fig.height=5}
-source('../code/site_differences.R')
+source('../code/figures/site_differences.R')
 grid.arrange(site_differences$a,site_differences$b,nrow = 1)
@@
 b) Decline acuteness index (DAI)'}
-source('../code/descriptor_contributions.R')
+source('../code/figures/descriptor_contributions.R')
 descriptor_contributions %>%
@@
 All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models'}
-source('../code/PDI_response_surfaces.R')
+source('../code/figures/PDI_response_surfaces.R')
 grid.arrange(PDI_response_plot$`a)`,PDI_response_plot$`b)`,PDI_response_plot$`c)`,PDI_response_plot$legend,nrow = 2,newpage = F)
@@
 All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models'}
-source('../code/DAI_response_surfaces.R')
+source('../code/figures/DAI_response_surfaces.R')
 grid.arrange(DAI_response_plot$`a)`,DAI_response_plot$`b)`,DAI_response_plot$`c)`,DAI_response_plot$`d)`,DAI_response_plot$`e)`,DAI_response_plot$`f)`,DAI_response_plot$legend,ncol = 3,layout_matrix = rbind(c(1,4,7),c(2,5,7),c(3,6,7)),widths = c(4,4,1),newpage = F)

### 2019-11-11 manuscript/manuscript.Rmd
big rearrange

@@
 ```{r library_load,cache=FALSE}
-source('../code/library_load.R')
+source('../code/figures/library_load.R')
 ```
@@
 ```{r prep}
-source('../code/data_prep.R')
+source('../code/figures/data_prep.R')
 ```
@@
 ```{r check,comment='',cache=FALSE}
-source('../code/check.R')
+source('../code/figures/check.R')
 manuscript_check()

### 2019-11-11 manuscript/supplementary.Rmd
big rearrange

@@
 ```{r descriptor_adjustment,fig.cap='Error bars show ±1SE of the mean.\nDashed red line shows the overall mean.'}
-source('../code/descriptor_adjustment.R')
+source('../code/figures/descriptor_adjustment.R')
 descriptor_adjustment %>%

### 2019-11-11 R/R/DIspace.R

@@ -1,0 +1,55 @@
+#' @export
+
+DIspace <- function(){
+  labels <- tibble(
+    label = c('Healthy','Acute\nDecline','Chronic\nDecline','Declining','Severe\nDecline','Early\nAcute Decline'),
+    x = c(0.17,0.75,0.66,0.5,0.83,0.25),
+    y = c(0,0.66,-0.66,0,0,0.66)
+  )
+
+  segs <- 3000
+  rectangle <- map(0:segs,~{
+    f <- . /segs
+    if (f < 0.5) {
+      yend <- f * -1.34 - 0.33
+      # yend <- f * -2 - 0
+    } else {
+      yend = -1
+    }
+    tibble(
+      x = f,
+      xend = f,
+      y = 1,
+      yend = yend
+    )
+  }) %>%
+    bind_rows()
+
+  hlines <- tibble(
+    x = c(0.33,1,0,1),
+    y = c(-0.33,-0.33,0.33,0.33),
+    groups = c(1,1,2,2)
+  )
+
+  vlines <- tibble(
+    x = c(0.33,0.33,0.5,0.5,0.66,0.66),
+    y = c(0.33 * -1.34 - 0.33,0.33,0.33,1,-0.33,0.33),
+    groups = c(1,1,2,2,3,3)
+  )
+
+  ggplot() +
+    geom_segment(data = rectangle,aes(x = x,xend = xend,y = y,yend = yend,colour = x)) +
+    scale_colour_gradient2(low = 'green',mid = 'orange',high = 'red',midpoint = 0.5) +
+    scale_x_continuous(limits = c(0,1), expand = c(0, 0)) +
+    scale_y_continuous(limits = c(-1,1), expand = c(0, 0)) +
+    geom_line(data = hlines,aes(x = x,y = y,group = groups),linetype = 2,colour = 'white') +
+    geom_line(data = vlines,aes(x = x,y = y,group = groups),linetype = 2,colour = 'white') +
+    geom_text(data = labels,aes(x = x,y = y,label = label)) +
+    labs(x = 'Phenotypic Decline Index (PDI)',
+         y = 'Decline Acuteness Index (DAI)') +
+    theme_classic() +
+    theme(plot.title = element_text(face = 'bold'),
+          axis.title = element_text(face = 'bold'),
+          panel.background = element_rect(fill = 'black')) +
+    guides(colour = FALSE)
+}

### 2019-11-11 R/R/MDSplot.R

@@ -1,0 +1,60 @@
+#' MDSplot
+#' @description Plot mulitidimensional scaling plot of random forest proximities.
+#' @param rfModels list containing random forest models
+#' @param phenoData tibble containing phenotype data to use for plotting
+#' @param title plot title
+#' @param colour phenoData column to use to map point colour aesthetics
+#' @param shape phenoData column to use to map point shape aesthetics
+#' @importFrom stats cmdscale
+#' @importFrom tibble as_tibble
+#' @importFrom dplyr bind_cols
+#' @importFrom ggplot2 ggplot aes element_text geom_point aes_string theme_bw theme ggtitle guides guide_legend
+#' @importFrom ggthemes scale_colour_ptol
+#' @export
+
+MDSplot <- function(rfModels,phenoData,title = '', colour = NULL, shape = NULL){
+
+   proxMDS <- rfModels %>%
+     mds() %>%
+    bind_cols(select(phenoData,Location,`Tree No`,Status))
+
+  ggplot(proxMDS,aes(x = `Dimension 1`,y = `Dimension 2`)) +
+    geom_point(aes_string(colour = colour,shape = shape)) +
+    scale_colour_ptol() +
+    theme_bw(base_size = 9) +
+    theme(legend.position = 'bottom',
+          legend.title = element_text(face = 'bold'),
+          plot.title = element_text(face = 'bold'),
+          axis.title = element_text(face = 'bold')) +
+    labs(title = title,
+         x = 'Dimension 1',
+         y = 'Dimension 2') +
+    guides(colour = guide_legend(title.position = 'top',ncol = 2),
+           shape = guide_legend(title.position = 'top',ncol = 1)) +
+    coord_fixed()
+}
+
+#' mds
+#' @description perform multidimensional scaling of random forest proximities
+#' @param rfModels list containing random forest models
+#' @export
+
+mds <- function(rfModels){
+  rfModels %>%
+    map(~{.$proximity %>%
+        as_tibble() %>%
+        rowid_to_column(var = 'Sample1') %>%
+        gather('Sample2','Proximity',-Sample1)}) %>%
+    bind_rows(.id = 'Iteration') %>%
+    mutate(Sample2 = as.numeric(Sample2)) %>%
+    group_by(Sample1,Sample2) %>%
+    summarise(Proximity = mean(Proximity)) %>%
+    spread(Sample2,Proximity) %>%
[…]

### 2019-11-11 R/R/calcDIs.R

@@ -1,0 +1,72 @@
+#' @importFrom purrr map_chr
+#' @importFrom readr read_csv
+#' @export
+
+calcDIs <- function(seed = 1234){
+  files <- list.files(system.file('phenotypeDataCollectionSheets',package = 'pdi'),full.names = T)
+
+  phenoData <- files %>%
+    map(readPhenotypeSheet)
+
+  locations <- phenoData %>%
+    map_chr(~{.$Location})
+
+  names(phenoData) <- locations
+
+  phenoData <- dataCorrections(phenoData)
+
+  phenoData <- phenoData %>%
+    mutate(`Crown condition (%)` = crownCondition(`Missing crown (%)`,
+                                                  `Crown transparency (%)`),
+           `Crown volume (m^3)` = crownVolume(`Crown radius (m)`,
+                                              `Total height (m)`,
+                                              `Lower crown height (m)`,
+                                              `Crown condition (%)`),
+           `Bleed prevalence (%)` = bleedPrevalence(`Active bleed size (cm)`,
+                                                    `Active bleeds`,
+                                                    `Black staining size (cm)`,
+                                                    `Black staining`,
+                                                    `Diameter at breast height (cm)`),
+           `Agrilus exit hole density (m^-2)` = agrilusExitHoleDensity(`Agrilus exit holes`,`Diameter at breast height (cm)`)
+    )
+  analysisTable <- makeAnalysisTable(phenoData)
+
+  correctedPhenoData <- siteCorrection(phenoData) %>%
+    mutate(`Crown condition (%)` = crownCondition(`Missing crown (%)`,
+                                                  `Crown transparency (%)`),
+           `Crown volume (m^3)` = crownVolume(`Crown radius (m)`,
+                                              `Total height (m)`,
+                                              `Lower crown height (m)`,
+                                              `Crown condition (%)`),
+           `Bleed prevalence (%)` = bleedPrevalence(`Active bleed size (cm)`,
+                                                    `Active bleeds`,
+                                                    `Black staining size (cm)`,
+                                                    `Black staining`,
+                                                    `Diameter at breast height (cm)`),
[…]

### 2019-11-11 R/R/functions.R

@@ -1,0 +1,50 @@
+#' dataCorrections
+#' @description a specific function to correct phenotype data inaccuracies
+#' @param phenoData tibble containing phenotype data
+#' @importFrom magrittr %>%
+#' @importFrom stringr str_detect
+#' @importFrom purrr map
+#' @importFrom dplyr bind_rows
+#' @export
+
+dataCorrections <- function(phenoData){
+  phenoData$`Big Wood`$Description$Value[
+    phenoData$`Big Wood`$Description$Value[phenoData$`Big Wood`$Description$Descriptor == 'Tree No'] %>% str_detect('CONTROL') &
+      phenoData$`Big Wood`$Description$Descriptor == 'Symptomatic'
+    ] <- 'N'
+
+  phenoData$`Big Wood`$Description$Value[
+    !(phenoData$`Big Wood`$Description$Value[phenoData$`Big Wood`$Description$Descriptor == 'Tree No'] %>% str_detect('CONTROL')) &
+      phenoData$`Big Wood`$Description$Descriptor == 'Symptomatic'
+    ] <- 'Y'
+
+  phenoData <- phenoData %>%
+    map(preparePhenotypeData) %>%
+    bind_rows(.id = 'Location')
+
+  phenoData$`Crown transparency (%)`[phenoData$Location %in% c('Attingham','Langdale')] <- 100 - phenoData$`Crown transparency (%)`[phenoData$Location %in% c('Attingham','Langdale')]
+
+  return(phenoData)
+}
+
+#' makeAnalysisTable
+#' @description prepare phenoData table ready for random forest analysis
+#' @param phenoData tibble containing phenotype data
+#' @importFrom dplyr select
+#' @importFrom purrr map_dfc
+#' @export
+
+makeAnalysisTable <- function(phenoData){
+  analysisTable <- phenoData %>%
+    select(-Location,-ID,-GPS,-Status,-`Tree No`)
+
+  if ('ChosenGroup' %in% names(analysisTable)) {
+    analysisTable <- analysisTable %>%
+      select(-ChosenGroup)
+  }
+
+  analysisTable[,{map_dfc(analysisTable,is.character) %>% as.logical()}] <- analysisTable[, {map_dfc(analysisTable,is.character) %>% as.logical()}] %>%
+    map_dfc(factor)
+
+  return(analysisTable)
+}

### 2019-11-11 R/R/importancePlot.R

@@ -1,0 +1,37 @@
+#' importancePlot
+#' @description plot random forest Gini importance values
+#' @param rfModels list containing random forest models
+#' @param phenoData tibble containing phenotype data to use for plotting
+#' @param title plot title
+#' @importFrom randomForest importance
+#' @importFrom dplyr arrange
+#' @importFrom ggplot2 coord_flip xlab labs
+#' @importFrom ggthemes ptol_pal
+#' @importFrom stringr str_c
+#' @export
+
+importancePlot <- function(rfModels,phenoData,title = ''){
+  imp <- rfModels %>%
+    map(~{
+      imp <- importance(.)
+      feat <- rownames(imp)
+      imp <- imp[,1]
+      tibble(Feature = feat,`Gini Importance` = imp)
+    }) %>%
+    bind_rows(.id = 'Repeat') %>%
+    group_by(Feature) %>%
+    summarise(`Gini Importance` = mean(`Gini Importance`)) %>%
+    arrange(`Gini Importance`) %>%
+    mutate(Feature = factor(Feature,levels = Feature))
+
+  ggplot(imp,aes(x = Feature,y = `Gini Importance`)) +
+    geom_point(colour = ptol_pal()(1)) +
+    coord_flip() +
+    theme_bw() +
+    ggtitle(title) +
+    theme(plot.title = element_text(face = 'bold',size = 9),
+          axis.title = element_text(face = 'bold')) +
+    labs(caption = str_c('Scores averaged across ',length(rfModels),' repetitions'),
+         y = 'Gini importance',
+         x = 'Descriptor')
+}

### 2019-11-11 R/R/margins.R

@@ -1,0 +1,20 @@
+#' margins
+#' @description calculate class margins
+#' @param rfModels list containing random forest models
+#' @importFrom randomForest margin
+#' @export
+
+margins <- function(rfModels){
+  rfModels %>%
+    map(~{
+      d <- .
+      margin(d) %>%
+        tibble(Class = d$y, Margin = .) %>%
+        group_by(Class) %>%
+        summarise(Margin = mean(Margin))
+    }) %>%
+    bind_rows(.id = 'Repeat') %>%
+    group_by(Class) %>%
+    summarise(Margin = mean(Margin) %>%
+                round(3))
+}

### 2019-11-11 R/R/minMaxScale.R

@@ -1,0 +1,5 @@
+#' @export
+
+minMaxScale <- function(vec){
+  (vec - min(vec)) / (max(vec) - min(vec))
+}

### 2019-11-11 R/R/permute.R

@@ -1,0 +1,36 @@
+#' @importFrom parallel makeCluster parLapply stopCluster detectCores
+#' @importFrom purrr map_dbl
+#' @export
+
+permute <- function(dat,cls, nReps = 100, nPerm = 1000, nCores = detectCores() * 0.75, clusterType = 'PSOCK'){
+
+  permCls <- 1:nPerm %>%
+    map(~{
+      sample_n(cls,nrow(cls),replace = F)
+    })
+
+  clus <- makeCluster(nCores,clusterType)
+
+  permDist <- permCls %>%
+    parLapply(cl = clus,.,function(classes,dat,nReps){
+      rf(dat,classes,nReps) %>%
+        map(randomForest::margin(.)) %>% mean()
+    },dat = dat,nReps = nReps) %>%
+    unlist()
+
+  stopCluster(clus)
+
+  return(permDist)
+}
+
+#' @export
+
+permuteSignif <- function(margin,permDist){
+  list(
+    margin = margin,
+    distribution = permDist,
+    pvalue = pnorm(mean(margin),mean(permDist),sd(permDist),lower.tail = F),
+    threshold = qnorm(0.05,mean(permDist),sd(permDist),lower.tail = F)
+  )
+
+}

### 2019-11-11 R/R/preparePhenotypeData.R

@@ -1,0 +1,132 @@
+#' preparePhenotypeData
+#' @description process parsed phenotype data sheets into a tibble suitable for random forest analysis
+#' @param phenotypeData parsed phenotype data collection sheet returned from \code{readPhenotypeSheet}
+#' @importFrom tidyr spread
+#' @importFrom dplyr filter group_by mutate summarise tbl_df rename left_join
+#' @export
+
+preparePhenotypeData <- function(phenotypeData){
+  description <- phenotypeData %>%
+      .$Description %>%
+    spread(Descriptor,Value)
+
+  cardinalAssessments <- phenotypeData %>%
+      .$CardinalAssessments %>%
+    filter(!(Descriptor %in% c('Whole stem photo ref number',
+                               '3m section photo ref number')
+    )) %>%
+    spread(Descriptor,Value)
+
+  averageCrownRadius <- cardinalAssessments %>%
+    group_by(ID) %>%
+    mutate(`Crown radius` = `Crown radius` %>% as.numeric()) %>%
+    summarise(`Average Crown radius` = mean(`Crown radius`))
+
+  deadTissue <- cardinalAssessments %>%
+    group_by(ID) %>%
+    summarise(`Dead Stem Tissue` = 'Hollow' %in% `Tap test`)
+  deadTissue$`Dead Stem Tissue`[deadTissue$`Dead Stem Tissue` == T] <- 'present'
+  deadTissue$`Dead Stem Tissue`[deadTissue$`Dead Stem Tissue` == F] <- 'absent'
+
+  canopyClosure <- cardinalAssessments %>%
+    group_by(ID) %>%
+    summarise(`Canopy Closure` = 'Y' %in% `Canopy closure`)
+  canopyClosure$`Canopy Closure`[canopyClosure$`Canopy Closure` == T] <- 'present'
+  canopyClosure$`Canopy Closure`[canopyClosure$`Canopy Closure` == F] <- 'absent'
+
+  symptoms <- phenotypeData %>%
+      .$Symptoms
+  symptoms$Size[is.na(symptoms$Size)] <- 0
+
+  crackFrequencies <- symptoms
+  crackFrequencies$Size[crackFrequencies$Size > 0] <- 1
+
+  crackFrequencies <- crackFrequencies %>%
+    group_by(ID,`Symptom Type`) %>%
+    summarise(Count = sum(Size)) %>%
[…]

### 2019-11-11 R/R/rf.R

@@ -1,0 +1,14 @@
+#' rf
+#' @description perform random forest repetitions
+#' @param analysisTable tibble of phenotype data suitable for random forest analysis as returned by \code{preparePhenotypeData}
+#' @param cls analysisTable column to use as response vector. NULL for unsupervised analyses.
+#' @param nreps number of repetitions
+#' @importFrom randomForest randomForest
+#' @export
+
+rf <- function(analysisTable,cls,nreps = 100,seed = 1234){
+  set.seed(seed)
+  map(1:nreps,~{
+    randomForest(analysisTable,cls,proximity = T)
+  })
+}

### 2019-11-11 R/R/siteCorrectionFactors.R

@@ -1,0 +1,35 @@
+#' siteCorrectionFactors
+#' @description return site correction correction factors of selected descriptors
+#' @param phenoData phenoData tibble containing phenotype data
+#' @param descriptors columns of phenoData on which calculate site correction factors
+#' @export
+
+
+siteCorrectionFactors <- function(phenoData,descriptors = c("Diameter at breast height (cm)",
+                                                            "Lower crown height (m)",
+                                                            "Timber height (m)",
+                                                            "Total height (m)",
+                                                            "Crown radius (m)")){
+
+  siteCorrect <- phenoData %>%
+    select(Location,ID,descriptors) %>%
+    gather('Descriptor','Value',-Location,-ID)
+
+  overallMeans <- siteCorrect %>%
+    group_by(Descriptor) %>%
+    summarise(Mean = mean(Value))
+
+  siteCorrections <- siteCorrect %>%
+    group_by(Location,Descriptor) %>%
+    summarise(Mean = mean(Value)) %>%
+    tbl_df() %>%
+    split(.$Descriptor) %>%
+    map(~{
+      d <- .
+      d %>%
+      mutate(Correction = Mean - ({overallMeans %>% filter(Descriptor == d$Descriptor[1]) %>% .$Mean}))
+    }) %>%
+    bind_rows() %>%
+    select(Descriptor,Location,Mean,Correction)
+  return(siteCorrections)
+}

### 2019-11-11 R/figures/library_load.R

@@ -1,3 +1,3 @@
-pacman::p_load(maps,mapdata,pdi,ggthemes,randomForest,
+pacman::p_load(maps,mapdata,ggthemes,randomForest,
                patchwork,gridExtra,ggrepel,metR,tidyverse,
                ggpubr,wordcountaddin,spelling,gramr,knitr)

### 2019-11-11 manuscript/figures.Rmd
fixed file path changes

--- 2019-11-11 manuscript/figures.Rmd
@@
 ```{r siteLocations,fig.cap='Map showing locations of the nine sites used for collection of phenotypic descriptors.',fig.width=3,fig.height=5}
-source('../code/site_locations.R')
+source('../R/figures/site_locations.R')
 site_locations_map
@@
 b) Relative descriptor Gini importances measures.',fig.height=5}
-source('../code/site_differences.R')
+source('../R/figures/site_differences.R')
 grid.arrange(site_differences$a,site_differences$b,nrow = 1)
@@
 b) Decline acuteness index (DAI)'}
-source('../code/descriptor_contributions.R')
+source('../R/figures/descriptor_contributions.R')
 descriptor_contributions %>%
@@
 All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models'}
-source('../code/PDI_response_surfaces.R')
+source('../R/figures/PDI_response_surfaces.R')
 grid.arrange(PDI_response_plot$`a)`,PDI_response_plot$`b)`,PDI_response_plot$`c)`,PDI_response_plot$legend,nrow = 2,newpage = F)
@@
 All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models'}
-source('../code/DAI_response_surfaces.R')
+source('../R/figures/DAI_response_surfaces.R')
 grid.arrange(DAI_response_plot$`a)`,DAI_response_plot$`b)`,DAI_response_plot$`c)`,DAI_response_plot$`d)`,DAI_response_plot$`e)`,DAI_response_plot$`f)`,DAI_response_plot$legend,ncol = 3,layout_matrix = rbind(c(1,4,7),c(2,5,7),c(3,6,7)),widths = c(4,4,1),newpage = F)

### 2019-11-11 manuscript/manuscript.Rmd
fixed file path changes

--- 2019-11-11 manuscript/manuscript.Rmd
@@
 ```{r library_load,cache=FALSE}
-source('../code/library_load.R')
+source('../R/figures/library_load.R')
+
+'../R/R/' %>%
+  list.files(full.names = T) %>%
+  walk(source)
 ```
@@
 ```{r prep}
-source('../code/data_prep.R')
+source('../R/figures/data_prep.R')
 ```
@@
 
-## Figures
-
-```{r figures,child='figures.Rmd'}
-
-```
-
 ```{r check,comment='',cache=FALSE}
-source('../code/check.R')
+source('../R/figures/check.R')
 manuscript_check()

### 2019-11-12 manuscript/figures.Rmd
phenotypic descriptor info table added to plan

@@
 ```{r siteLocations,fig.cap='Map showing locations of the nine sites used for collection of phenotypic descriptors.',fig.width=3,fig.height=5}
-source('../R/figures/site_locations.R')
+source('../code/figures/site_locations.R')
 site_locations_map

### 2019-11-12 code/functions/phenotypic_descriptors.R

@@ -1,0 +1,55 @@
+#' descriptorInfo
+#' @description collate descriptor information table
+#' @param analysisTable analysis suitable phenotype data tibble
+
+descriptorInfo <- function(analysisTable){
+  pt <- analysisTable %>%
+    {tibble(Descriptor = names(.),`Data Type` = map_chr(.,class),Description = NA,`Categories / units` = NA,Reference = NA)} %>%
+    filter(!(Descriptor %in% c('Crown condition (%)','Crown volume (m^3)','Bleed prevalence (%)','Agrilus exit hole density (m^-2)')))
+  
+  type <- tibble(`Descriptor Type` = c(rep('tree size',5),
+                                       rep('crown',13),
+                                       rep('trunk',13)),
+                 Descriptor = c('Diameter at breast height (cm)',
+                                'Crown radius (m)',
+                                'Total height (m)',
+                                'Lower crown height (m)',
+                                'Timber height (m)',
+                                'Missing crown (%)',
+                                'Crown transparency (%)',
+                                'Crown contact (%)',
+                                'Crown mildew',
+                                'Branch epicormics',
+                                'Crown fruiting bodies',
+                                'Dieback location',
+                                'Dieback type','Insect defoliation',
+                                'Insect defoliation type',
+                                'Pruning / branch loss',
+                                'Canopy closure',
+                                'Social class',
+                                'Agrilus exit holes',
+                                'Stem epicormics',
+                                'Dead stem tissue',
+                                'Ground level fruiting bodies',
+                                'Stem fruiting bodies',
+                                'Oval shaped exit holes',
+                                'Small circular shaped exit holes',
+                                'Active bleeds',
+                                'Active bleed size (cm)',
+                                'Black staining',
+                                'Black staining size (cm)',
+                                'Calloused wound',
+                                'Calloused wound size (cm)')
+  )
+  
+  pt <- pt %>%
+    left_join(type, by = "Descriptor") %>%
+    mutate(Descriptor = str_to_lower(Descriptor)) %>%
+    select(`Descriptor Type`,everything()) %>%
+    arrange(`Descriptor Type`,Descriptor)
+  
+  pt$`Data Type`[pt$`Data Type` == 'numeric'] <- 'continous'
+  pt$`Data Type`[pt$`Data Type` == 'factor'] <- 'categorical'
+  
+  return(pt)
+}

### 2019-11-12 manuscript/figures.Rmd
fixed descriptor table, site difference rf plot and DI space plot

--- 2019-11-11 manuscript/figures.Rmd
@@
 ```{r siteLocations,fig.cap='Map showing locations of the nine sites used for collection of phenotypic descriptors.',fig.width=3,fig.height=5}
-source('../R/figures/site_locations.R')
-site_locations_map
+loadd(site_location_map)
+
+print(site_location_map)
 ```
@@
 b) Relative descriptor Gini importances measures.',fig.height=5}
-source('../R/figures/site_differences.R')
-grid.arrange(site_differences$a,site_differences$b,nrow = 1)
+loadd(site_differences_mds_plot)
+grid.arrange(site_differences_mds_plot$a,site_differences_mds_plot$b,nrow = 1)
 ```

### 2019-11-12 code/figures/site_differences.R

@@ -1,68 +1,0 @@
-
-locationRF <- rf(makeAnalysisTable(phenoData),
-                 phenoData$Location %>%
-                   factor(),
-                 100)
-
-locationMDS <- mds(locationRF) %>%
-  bind_cols(phenoData %>%
-              select(Location))
-
-locationMargin <- margins(locationRF) %>%
-  summarise(Margin = mean(Margin)) %>%
-  mutate(x = 0.4,y = 0.5,label = str_c('Margin = ',Margin %>% round(2)))
-
-locationImportance <- locationRF %>%
-  map(~{
-    imp <- importance(.)
-    desc <- rownames(imp)
-    imp <- imp %>%
-      as_tibble() %>%
-      mutate(Descriptor = desc) %>%
-      select(Descriptor,MeanDecreaseGini)
-    }) %>%
-  bind_rows(.id = 'Rep') %>%
-  group_by(Descriptor) %>%
-  summarise(Importance = mean(MeanDecreaseGini)) %>%
-  mutate(`Relative Importance` = Importance / max(Importance)) %>%
-  arrange(`Relative Importance`) %>%
-  mutate(Descriptor = factor(Descriptor,levels = Descriptor))
-
-descriptorLabels <- locationImportance$Descriptor %>%
-  as.character() %>%
-  {c(.[1:6],
-                      expression(Agrillus~exit~hole~density ( m^-2 ) ),
-                      .[8:25],
-                      expression(Crown~volume ( m^3 ) ),
-                      .[27:35])
-  }
-
-site_differences <- list(
-  a = ggplot() +
-    geom_point(data = locationMDS,
-               aes(x = `Dimension 1`, 
-                   y = `Dimension 2`,
-                   fill = Location),
-               shape = 21,
-               size = 3) +
-    geom_text(data = locationMargin,aes(x = x,y = y,label = label)) +
-    scale_fill_ptol() +
-    theme_bw() +
-    theme(plot.title = element_text(face = 'bold'),
-          axis.title = element_text(face = 'bold'),
-          legend.title = element_text(face = 'bold'),
-          legend.position = 'bottom') +
-    coord_fixed() +
-    guides(fill = guide_legend(ncol = 2,title.position = "top")) +
-    labs(title = 'a)',
-         fill = 'Site'),
-  b = ggplot(locationImportance,aes(x = `Relative Importance`,y = Descriptor)) +
[…]

### 2019-11-12 code/functions/DIspace.R

@@ -1,0 +1,56 @@
+#' DIspace
+#' @description Plot the 2 dimensional decline index continuum.
+
+DIspace <- function(){
+  labels <- tibble(
+    label = c('Healthy','Acute\nDecline','Chronic\nDecline','Declining','Severe\nDecline','Early\nAcute Decline'),
+    x = c(0.17,0.75,0.66,0.5,0.83,0.25),
+    y = c(0,0.66,-0.66,0,0,0.66)
+  )
+  
+  segs <- 3000
+  rectangle <- map(0:segs,~{
+    f <- . /segs
+    if (f < 0.5) {
+      yend <- f * -1.34 - 0.33
+      # yend <- f * -2 - 0
+    } else {
+      yend = -1
+    }
+    tibble(
+      x = f,
+      xend = f,
+      y = 1,
+      yend = yend
+    )
+  }) %>%
+    bind_rows()
+  
+  hlines <- tibble(
+    x = c(0.33,1,0,1),
+    y = c(-0.33,-0.33,0.33,0.33),
+    groups = c(1,1,2,2)
+  )
+  
+  vlines <- tibble(
+    x = c(0.33,0.33,0.5,0.5,0.66,0.66),
+    y = c(0.33 * -1.34 - 0.33,0.33,0.33,1,-0.33,0.33),
+    groups = c(1,1,2,2,3,3)
+  )
+  
+  ggplot() +
+    geom_segment(data = rectangle,aes(x = x,xend = xend,y = y,yend = yend,colour = x)) +
+    scale_colour_gradient2(low = 'green',mid = 'orange',high = 'red',midpoint = 0.5) +
+    scale_x_continuous(limits = c(0,1), expand = c(0, 0)) +
+    scale_y_continuous(limits = c(-1,1), expand = c(0, 0)) +
+    geom_line(data = hlines,aes(x = x,y = y,group = groups),linetype = 2,colour = 'white') +
+    geom_line(data = vlines,aes(x = x,y = y,group = groups),linetype = 2,colour = 'white') +
+    geom_text(data = labels,aes(x = x,y = y,label = label)) +
+    labs(x = 'Phenotypic Decline Index (PDI)',
+         y = 'Decline Acuteness Index (DAI)') +
+    theme_classic() +
+    theme(plot.title = element_text(face = 'bold'),
+          axis.title = element_text(face = 'bold'),
+          panel.background = element_rect(fill = 'black')) +
+    guides(colour = FALSE)
+}

### 2019-11-12 code/functions/margins.R

@@ -1,0 +1,18 @@
+#' margins
+#' @description calculate class margins
+#' @param rfModels list containing random forest models
+
+margins <- function(rfModels){
+  rfModels %>%
+    map(~{
+      d <- .
+      margin(d) %>%
+        tibble(Class = d$y, Margin = .) %>%
+        group_by(Class) %>%
+        summarise(Margin = mean(Margin))
+    }) %>%
+    bind_rows(.id = 'Repeat') %>%
+    group_by(Class) %>%
+    summarise(Margin = mean(Margin) %>%
+                round(3))
+}

### 2019-11-12 manuscript/figures.Rmd
manuscript now compiles

@@
 b) Decline acuteness index (DAI)'}
-source('../R/figures/descriptor_contributions.R')
-descriptor_contributions %>%
-  patchwork::wrap_plots()
+loadd(descriptor_contribution_plots)
+
+descriptor_contribution_plots %>%
+  wrap_plots()
 ```
@@
 All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models'}
-source('../R/figures/PDI_response_surfaces.R')
-grid.arrange(PDI_response_plot$`a)`,PDI_response_plot$`b)`,PDI_response_plot$`c)`,PDI_response_plot$legend,nrow = 2,newpage = F)
+loadd(PDI_response_surfaces)
+
+grid.arrange(PDI_response_surfaces$`a)`,PDI_response_surfaces$`b)`,PDI_response_surfaces$`c)`,PDI_response_surfaces$legend,nrow = 2,newpage = F)
 ```
@@
 All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models'}
-source('../R/figures/DAI_response_surfaces.R')
-grid.arrange(DAI_response_plot$`a)`,DAI_response_plot$`b)`,DAI_response_plot$`c)`,DAI_response_plot$`d)`,DAI_response_plot$`e)`,DAI_response_plot$`f)`,DAI_response_plot$legend,ncol = 3,layout_matrix = rbind(c(1,4,7),c(2,5,7),c(3,6,7)),widths = c(4,4,1),newpage = F)
+loadd(DAI_response_surfaces)
+
+grid.arrange(DAI_response_surfaces$`a)`,DAI_response_surfaces$`b)`,DAI_response_surfaces$`c)`,DAI_response_surfaces$`d)`,DAI_response_surfaces$`e)`,DAI_response_surfaces$`f)`,DAI_response_surfaces$legend,ncol = 3,layout_matrix = rbind(c(1,4,7),c(2,5,7),c(3,6,7)),widths = c(4,4,1),newpage = F)
 ```

### 2019-11-12 code/figures/DAI_response_surfaces.R

@@ -1,190 +1,0 @@
-spectrumTrees_DAI <- list(
-  `a)` = DIs$data[which(abs(DIs$DIs$DAI) == min(abs(DIs$DIs$DAI))),],
-  `b)` = DIs$data[which(abs(DIs$DIs$DAI - 0.5) == min(abs(DIs$DIs$DAI - 0.5))),],
-  `c)` = DIs$data[which(DIs$DIs$DAI == max(DIs$DIs$DAI)),],
-  `d)` = DIs$data[which(abs(DIs$DIs$DAI) == min(abs(DIs$DIs$DAI))),],
-  `e)` = DIs$data[which(abs(DIs$DIs$DAI + 0.5) == min(abs(DIs$DIs$DAI + 0.5))),],
-  `f)` = DIs$data[which(DIs$DIs$DAI == min(DIs$DIs$DAI)),]
-)
-
-variables <- names(DIs$data)[map_lgl(DIs$data,is.numeric)]
-
-ranges <- variables %>%
-  map(~{
-    DIs$data[,.] %>%
-      unlist() %>%
-      {tibble(min = min(.),
-              max = max(.))} %>%
-      mutate(center = min + (max - min)/2)
-  }) %>%
-  set_names(variables) %>%
-  bind_rows(.id = 'Descriptor')
-
-spectrumTrees_DAI$`a)`$`Active bleeds` <- ranges$min[ranges$Descriptor == 'Active bleeds']
-spectrumTrees_DAI$`b)`$`Active bleeds` <- {ranges$center[ranges$Descriptor == 'Active bleeds'] / 16} %>%
-  round()
-spectrumTrees_DAI$`c)`$`Active bleeds` <- {ranges$max[ranges$Descriptor == 'Active bleeds'] / 16} %>%
-  round()
-
-spectrumTrees_DAI$`a)`$`Black staining` <- ranges$min[ranges$Descriptor == 'Black staining']
-spectrumTrees_DAI$`b)`$`Black staining` <- {ranges$center[ranges$Descriptor == 'Black staining'] / 16} %>%
-  round()
-spectrumTrees_DAI$`c)`$`Black staining` <- {ranges$max[ranges$Descriptor == 'Black staining'] / 16} %>%
-  round()
-
-spectrumTrees_DAI$`d)`$`Crown radius (m)` <- ranges$max[ranges$Descriptor == 'Crown radius (m)']
[…]

### 2019-11-12 code/figures/data_prep.R

@@ -1,51 +1,0 @@
-phenoData <- {
-  files <- list.files(system.file('phenotypeDataCollectionSheets',package = 'phi'),full.names = T)
-  
-  phenoData <- files %>%
-    map(readPhenotypeSheet) 
-  
-  locations <- phenoData %>%
-    map_chr(~{.$Location})
-  
-  names(phenoData) <- locations
-  
-  phenoData <- dataCorrections(phenoData)
-  
-  phenoData %>%
-    mutate(`Crown condition (%)` = crownCondition(`Missing crown (%)`,
-                                                  `Crown transparency (%)`),
-           `Crown volume (m^3)` = crownVolume(`Crown radius (m)`,
-                                              `Total height (m)`,
-                                              `Lower crown height (m)`,
-                                              `Crown condition (%)`),
-           `Bleed prevalence (%)` = bleedPrevalence(`Active bleed size (cm)`,
-                                                    `Active bleeds`,
-                                                    `Black staining size (cm)`,
-                                                    `Black staining`,
-                                                    `Diameter at breast height (cm)`),
-           `Agrilus exit hole density (m^-2)` = agrilusExitHoleDensity(`Agrilus exit holes`,`Diameter at breast height (cm)`)
-    )
-}
-
-DIs <- calcDIs()
-
-correctedPhenoData <- siteCorrection(phenoData) %>%
-  mutate(`Crown condition (%)` = crownCondition(`Missing crown (%)`,
-                                                `Crown transparency (%)`),
-         `Crown volume (m^3)` = crownVolume(`Crown radius (m)`,
-                                            `Total height (m)`,
-                                            `Lower crown height (m)`,
-                                            `Crown condition (%)`),
-         `Bleed prevalence (%)` = bleedPrevalence(`Active bleed size (cm)`,
-                                                  `Active bleeds`,
-                                                  `Black staining size (cm)`,
-                                                  `Black staining`,
-                                                  `Diameter at breast height (cm)`),
-         `Agrilus exit hole density (m^-2)` = agrilusExitHoleDensity(`Agrilus exit holes`,`Diameter at breast height (cm)`)
-  )
-
-analysisTable <- makeAnalysisTable(correctedPhenoData)
-
-set.seed(1234)
-PDIrf <- randomForest(analysisTable,DIs$DIs$PDI,ntree = 10000)
-DAIrf <- randomForest(analysisTable,DIs$DIs$DAI,ntree = 10000,mtry = 11)

### 2019-11-12 code/functions/siteDifferencesRFplot.R

@@ -1,9 +1,9 @@
-#' site_differences_rf_plot
+#' siteDifferencesRFplot
 #' @description create MDS scatter and random forest variable importance plots for site phenotypic differences
 #' @param site_differences_rf list of site supervised random forest models
 #' @param phenoData tibble containing phenotype data
 
-site_differences_rf_plot <- function(site_differences_rf,phenoData){
+siteDifferencesRFplot <- function(site_differences_rf,phenoData){
   locationMDS <- site_differences_rf %>%
     mds() %>%
     bind_cols(phenoData %>%

### 2019-11-12 code/plan.R

@@ -39,7 +39,7 @@ plan <- drake_plan(
   ## site differences MDS and importance plots
   
   site_differences_mds_plot = site_differences_rf %>%
-    site_differences_rf_plot(pheno_data_with_additional_descriptors),
+    siteDifferencesRFplot(pheno_data_with_additional_descriptors),
     
   ## apply site corrections and recalculate additional descriptors
   site_corrected_pheno_data = pheno_data %>%
@@ -57,6 +57,32 @@ plan <- drake_plan(
   ## calculate decline indexes
   decline_indexes = unsupervised_rf %>%
     calcDIs(site_corrected_pheno_data),
+  
+  ## generate PDI predictive random forest model
+  PDI_rf_model = site_corrected_analysis_suitable_data %>%
+    {set.seed(1234)
+      randomForest(.,y = decline_indexes$PDI,
+                 ntree = 10000)},
+  
+  ## generate DAI predictive random forest model
+  DAI_rf_model = site_corrected_analysis_suitable_data %>%
+    {set.seed(1234)
+      randomForest(.,y = decline_indexes$DAI,
+                   ntree = 10000,
+                   mtry = 11)},
+  
+  ## create descriptor contribution plots
+  descriptor_contribution_plots = descriptorImportancePlots(PDI_rf_model,DAI_rf_model),
+  
+  ## PDI rf model response surfaces
+  PDI_response_surfaces = PDIresponseSurfaces(PDI_rf_model,
+                                              decline_indexes,
+                                              site_corrected_analysis_suitable_data),
+  
+  ## DAI rf model response surfaces
+  DAI_response_surfaces = DAIresponseSurfaces(DAI_rf_model,
+                                              decline_indexes,
+                                              site_corrected_analysis_suitable_data),
   
   ## render manuscript
   manuscript = render(knitr_in('manuscript/manuscript.Rmd'),quiet = T)

### 2019-11-12 manuscript/manuscript.Rmd
phenotypic descriptor info table added to plan
fixed descriptor table, site difference rf plot and DI space plot
manuscript now compiles
fixed DAI inversion

--- 2019-11-11 manuscript/manuscript.Rmd
@@
 knitr::opts_chunk$set(echo = FALSE,
-                      cache = TRUE,
+                      #cache = TRUE,
                       fig.pos = 'H',
@@
                       message = FALSE)
-```
-
-```{r library_load,cache=FALSE}
-source('../code/library_load.R')
-```
-
-```{r prep}
-source('../code/data_prep.R')
 ```
@@
 
-## Figures
-
-```{r figures,child='figures.Rmd'}
-
-```
-
 ```{r check,comment='',cache=FALSE}
-source('../code/check.R')
 manuscript_check()

### 2019-11-12 manuscript/supplementary.Rmd
supplementary now compiles

--- 2019-08-28 manuscript/supplementary.Rmd
@@
+---
+output: pdf_document
+---
+
 ```{r descriptor_adjustment,fig.cap='Error bars show ±1SE of the mean.\nDashed red line shows the overall mean.'}
-source('../code/descriptor_adjustment.R')
-descriptor_adjustment %>%
+loadd(descriptor_adjustment_example)
+
+descriptor_adjustment_example %>%
   wrap_plots()

### 2019-11-12 code/plan.R
tex no longer kept

@@ -13,7 +13,6 @@ plan <- drake_plan(
     dataCorrections(), 
   
   ## calculate additional desecriptors
-  
   pheno_data_with_additional_descriptors = pheno_data %>%
     calcAdditionalDescriptors(),
   
@@ -22,28 +21,23 @@ plan <- drake_plan(
     makeAnalysisTable(),
   
   ## collect phenotypic descriptor information into table
-  
   phenotypic_descriptor_info = analysis_suitable_data %>%
     descriptorInfo(),
   
   ## create site location map
-  
   site_location_map = locationMap(),
   
   ## Analyse site differences using supervised random forest
-  
   site_differences_rf = analysis_suitable_data %>%
     rf(cls = pheno_data_with_additional_descriptors$Location %>% factor(),
        nreps = 100),
   
   ## site differences MDS and importance plots
-  
   site_differences_mds_plot = site_differences_rf %>%
     siteDifferencesRFplot(pheno_data_with_additional_descriptors),
     
   ## calculate descriptor correction factors
-  
-  site_correction_factors <- pheno_data_with_additional_descriptors %>%
+  site_correction_factors = pheno_data_with_additional_descriptors %>%
     siteCorrectionFactors(),
   
   ## apply site corrections and recalculate additional descriptors
@@ -52,7 +46,6 @@ plan <- drake_plan(
     calcAdditionalDescriptors(),
   
   ## plot descriptor adjustment example using dbh
-  
   descriptor_adjustment_example = pheno_data_with_additional_descriptors %>%
     descriptorAdjustmentPlot(site_corrected_pheno_data),
   

### 2019-11-12 manuscript/manuscript.Rmd

@@
    number_sections: false
-keep_tex: yes
+   keep_tex: no
 ---

### 2019-11-12 manuscript/supplementary.Rmd
tex no longer kept

@@
 ---
-output: pdf_document
+output: 
+  pdf_document:
+    keep_tex: no
 ---
+
+## Supplementary information
 
@@
 ```
+
+```{r descriptor_adjustment_factors}
+loadd(site_correction_factors)
+
+site_correction_factors %>%
+  kable(caption = 'Descriptor correction factors at each site')
+```
+
+```{r PDInumeric,fig.height=8}
+loadd(descriptor_scatter_plots)
+
+descriptor_scatter_plots$PDI_numeric
+```
+
+```{r PDIfactor,fig.height=8}
+descriptor_scatter_plots$PDI_factor
+```
+
+```{r DAInumeric,fig.height=8}
+descriptor_scatter_plots$DAI_numeric
+```
+
+```{r DAIfactor,fig.height=8}
+descriptor_scatter_plots$DAI_factor
+```

### 2019-11-14 manuscript/manuscript.Rmd
tables and figures temporarily rendered separate from the manuscript

@@
 knitr::opts_chunk$set(echo = FALSE,
-                      #cache = TRUE,
-                      fig.pos = 'H',
                       warning = FALSE,
@@
 
-Large phenotypic differences were found between the sites by supervised random forest (Figure \@ref(fig:siteLocations))
+Large phenotypic differences were found between the sites by supervised random forest (Figure #)
 
@@
 \newpage
-## Tables
-
-```{r tables,child='tables.Rmd'}
-```
-
-\newpage
-## Figures
-
-```{r figures,child='figures.Rmd'}
-```
-
-\newpage
 ## Checks

### 2019-11-14 manuscript/tables.Rmd
big rearrange
fixed file path changes
phenotypic descriptor info table added to plan
fixed descriptor table, site difference rf plot and DI space plot
supplementary now compiles
tables and figures temporarily rendered separate from the manuscript

@@
+---
+title: Tables
+output: pdf_document
+---
+
+## Tables
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE,
+                      warning = FALSE,
+                      message = FALSE)
+```
+
 ```{r phenotypic_descriptors,cache=FALSE}
-source('../code/phenotypic_descriptors.R')
-pt %>%
-  kable()
+loadd(phenotypic_descriptor_info)
+
+phenotypic_descriptor_info %>%
+  kable(format = 'markdown')
 ```

### 2019-11-14 code/plan.R

@@ -91,6 +91,12 @@ plan <- drake_plan(
                                               decline_indexes,
                                               site_corrected_analysis_suitable_data),
   
+  ## render tables
+  tables = render(knitr_in('manuscript/tables.Rmd'),quiet = T),
+  
+  ## render figures
+  figures = render(knitr_in('manuscript/figures.Rmd'),quiet = T),
+  
   ## render manuscript
   manuscript = render(knitr_in('manuscript/manuscript.Rmd'),quiet = T),
   

### 2019-11-15 code/functions/optimalParams.R
added rf model tuning

@@ -1,0 +1,6 @@
+
+optimalParams <- function(tuning_results){
+  tuning_results %>%
+    filter(MAE == min(MAE)) %>%
+    {list(mtry = .$mtry[1],ntree = .$ntree[1])}
+}

### 2019-11-15 code/functions/tuneModel.R

@@ -1,0 +1,25 @@
+
+tuneModel <- function(site_corrected_analysis_suitable_data,index){
+  mtry <- seq(2,26,by = 4)
+  ntree <- seq(1,4,by = 1) %>%
+    {10 ^ .}
+
+  comb <- expand.grid(mtry,ntree) %>%
+    as_tibble() %>%
+    set_names(c('mtry','ntree'))
+  
+  res <- comb %>%
+    split(1:nrow(.)) %>%
+    map(~{
+      n <- .
+      set.seed(1234)
+        randomForest(site_corrected_analysis_suitable_data,y = index,ntree = n$ntree,mtry = n$mtry) %>%
+          .$predicted %>%
+          {abs(. - index)} %>%
+          mean(na.rm = TRUE) %>%
+          tibble(mtry = n$mtry,ntree = n$ntree,MAE = .)
+    }) %>%
+    bind_rows()
+  
+  return(res)
+}

### 2019-11-15 code/functions/tuneResultsPlot.R

@@ -1,0 +1,15 @@
+
+tuneResultsPlot <- function(PDI_rf_tune_results,DAI_rf_tune_results) {
+  list(PDI = PDI_rf_tune_results,
+       DAI = DAI_rf_tune_results) %>%
+    bind_rows(.id = 'Index') %>%
+    mutate(ntree = factor(ntree)) %>%
+    ggplot(aes(x = mtry,y = MAE,colour = ntree)) +
+    geom_line() +
+    scale_colour_ptol() +
+    theme_bw() +
+    theme(axis.title = element_text(face = 'bold'),
+          legend.title = element_text(face = 'bold')) +
+    labs(y = 'Mean absolute error (MAE)') +
+    facet_wrap(~Index,scales = 'free')
+}

### 2019-11-15 code/plan.R

@@ -65,18 +65,30 @@ plan <- drake_plan(
   descriptor_scatter_plots = site_corrected_analysis_suitable_data %>%
     descriptorScatterPlots(decline_indexes),
   
+  ## find optimal random forest parameters for PDI
+  PDI_rf_tune_results = tuneModel(site_corrected_analysis_suitable_data,decline_indexes$PDI),
+  PDI_rf_tune_params = optimalParams(PDI_rf_tune_results),
+  
+  ## find optimal random forest parameters for PDI
+  DAI_rf_tune_results = tuneModel(site_corrected_analysis_suitable_data,decline_indexes$DAI),
+  DAI_rf_tune_params = optimalParams(DAI_rf_tune_results),
+  
+  ## plot tuning results
+  rf_tune_plot = tuneResultsPlot(PDI_rf_tune_results,DAI_rf_tune_results),
+  
   ## generate PDI predictive random forest model
   PDI_rf_model = site_corrected_analysis_suitable_data %>%
     {set.seed(1234)
       randomForest(.,y = decline_indexes$PDI,
-                 ntree = 10000)},
+                 ntree = PDI_rf_tune_params$ntree,
+                 mtry = PDI_rf_tune_params$mtry)},
   
   ## generate DAI predictive random forest model
   DAI_rf_model = site_corrected_analysis_suitable_data %>%
     {set.seed(1234)
       randomForest(.,y = decline_indexes$DAI,
-                   ntree = 10000,
-                   mtry = 11)},
+                   ntree = DAI_rf_tune_params$ntree,
+                   mtry = DAI_rf_tune_params$mtry)},
   
   ## create descriptor contribution plots
   descriptor_contribution_plots = descriptorImportancePlots(PDI_rf_model,DAI_rf_model),

### 2019-11-18 code/functions/declineIndexesPlot.R
added decline index plot

@@ -1,0 +1,15 @@
+
+declineIndexesPlot <- function(decline_indexes){
+  decline_indexes %>%
+    filter(ChosenGroup != 'Extra') %>%
+    ggplot(aes(x = PDI,y = DAI)) +
+    geom_point(aes(fill = ChosenGroup),shape = 21,size = 3) +
+    scale_fill_manual(values = ptol_pal()(4)[c(4,1,2,3)]) +
+    theme_bw() +
+    #facet_wrap(~Location) +
+    theme(legend.position = 'bottom',
+          axis.title = element_text(face = 'bold'),
+          legend.title = element_text(face = 'bold')) +
+    labs(fill = 'Assigned Decline Status') +
+    coord_fixed()
+}

### 2019-11-18 code/plan.R

@@ -61,6 +61,9 @@ plan <- drake_plan(
   decline_indexes = unsupervised_rf %>%
     calcDIs(site_corrected_pheno_data),
   
+  ## plot decline indexes
+  decline_indexes_plot = declineIndexesPlot(decline_indexes),
+  
   ## plot descriptors against decline indexes 
   descriptor_scatter_plots = site_corrected_analysis_suitable_data %>%
     descriptorScatterPlots(decline_indexes),

### 2019-11-18 manuscript/supplementary.Rmd
tables and figures temporarily rendered separate from the manuscript
added rf model tuning
added decline index plot
removed code chunks from supplementary
added PDI lime analysis plot to supplementary

@@
 ---
+title: Supplementary information
 output: 
   pdf_document:
-    keep_tex: no
+    keep_tex: false
 ---
 
-## Supplementary information
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE,
+                      warning = FALSE,
+                      message = FALSE)
+```
 
@@
   kable(caption = 'Descriptor correction factors at each site')
+```
+
+```{r decline_index_plot}
+loadd(decline_indexes_plot)
+
+decline_indexes_plot
 ```
@@
 ```
+
+```{r modelTuningResultsPlot}
+loadd(rf_tune_plot)
+
+rf_tune_plot
+```
+
+```{r PDIlimeAnalysisPlot}
+loadd(PDI_lime_analysis_plot)
+
+PDI_lime_analysis_plot
+```

### 2019-11-18 code/functions/DAIlimeAnalysis.R
added DAI lime analysis plot to supplementary

@@ -2,15 +2,12 @@
 DAIlimeAnalysis <- function(site_corrected_analysis_suitable_data,DAI_rf_model,decline_indexes){
   DAI_explainer <- lime(site_corrected_analysis_suitable_data, DAI_rf_model, bin_continuous = TRUE, quantile_bins = FALSE)
   
-  DAI_explainer <- lime(site_corrected_analysis_suitable_data, DAI_rf_model, bin_continuous = TRUE, quantile_bins = FALSE)
-  
   spectrumTrees_DAI <- list(
-    `a)` = site_corrected_analysis_suitable_data[which(abs(decline_indexes$DAI) == min(abs(decline_indexes$DAI))),],
-    `b)` = site_corrected_analysis_suitable_data[which(abs(decline_indexes$DAI - 0.5) == min(abs(decline_indexes$DAI - 0.5))),],
-    `c)` = site_corrected_analysis_suitable_data[which(decline_indexes$DAI == max(decline_indexes$DAI)),],
-    `d)` = site_corrected_analysis_suitable_data[which(abs(decline_indexes$DAI) == min(abs(decline_indexes$DAI))),],
-    `e)` = site_corrected_analysis_suitable_data[which(abs(decline_indexes$DAI + 0.5) == min(abs(decline_indexes$DAI + 0.5))),],
-    `f)` = site_corrected_analysis_suitable_data[which(decline_indexes$DAI == min(decline_indexes$DAI)),]
+    neutral = site_corrected_analysis_suitable_data %>% filter(abs(decline_indexes$DAI) == min(abs(decline_indexes$DAI))),
+    moderateAOD = site_corrected_analysis_suitable_data %>% filter(abs(decline_indexes$DAI - 0.5) == min(abs(decline_indexes$DAI - 0.5))),
+    severeAOD = site_corrected_analysis_suitable_data %>% filter(decline_indexes$DAI == max(decline_indexes$DAI)),
+    moderateCOD = site_corrected_analysis_suitable_data %>% filter(abs(decline_indexes$DAI + 0.5) == min(abs(decline_indexes$DAI + 0.5))),
+    severeCOD = site_corrected_analysis_suitable_data %>% filter(decline_indexes$DAI == min(decline_indexes$DAI))
   ) %>%
     bind_rows()
   

### 2019-11-25 code/functions/descriptorInfo.R
added to descriptor table

@@ -4,47 +4,52 @@
 
 descriptorInfo <- function(analysisTable){
   pt <- analysisTable %>%
-    {tibble(Descriptor = names(.),`Data Type` = map_chr(.,class),Description = NA,`Categories / units` = NA,Reference = NA)} %>%
+    {tibble(Descriptor = names(.),`Data Type` = map_chr(.,class))} %>%
     filter(!(Descriptor %in% c('Crown condition (%)','Crown volume (m^3)','Bleed prevalence (%)','Agrilus exit hole density (m^-2)')))
   
-  type <- tibble(`Descriptor Type` = c(rep('tree size',5),
-                                       rep('crown',13),
-                                       rep('trunk',13)),
-                 Descriptor = c('Diameter at breast height (cm)',
-                                'Crown radius (m)',
-                                'Total height (m)',
-                                'Lower crown height (m)',
-                                'Timber height (m)',
-                                'Missing crown (%)',
-                                'Crown transparency (%)',
-                                'Crown contact (%)',
-                                'Crown mildew',
-                                'Branch epicormics',
-                                'Crown fruiting bodies',
-                                'Dieback location',
-                                'Dieback type','Insect defoliation',
-                                'Insect defoliation type',
-                                'Pruning / branch loss',
-                                'Canopy closure',
-                                'Social class',
-                                'Agrilus exit holes',
-                                'Stem epicormics',
-                                'Dead stem tissue',
-                                'Ground level fruiting bodies',
-                                'Stem fruiting bodies',
-                                'Oval shaped exit holes',
-                                'Small circular shaped exit holes',
-                                'Active bleeds',
-                                'Active bleed size (cm)',
-                                'Black staining',
-                                'Black staining size (cm)',
-                                'Calloused wound',
-                                'Calloused wound size (cm)')
-  )
+  type <- tribble(
+    ~Descriptor,~Description,~`Categories / units`,~Reference,
+    'Diameter at breast height (mm)','','mm','',
+    'Crown radius (m)','','m','',
+    'Total height (m)','','m','',
+    'Lower crown height (m)','','m','',
+    'Timber height (m)','','m','',
[…]

### 2019-11-25 manuscript/manuscript.Rmd
report text added to manuscript

@@
                       message = FALSE)
+```
+
+```{r loadData, include = FALSE}
+loadd(site_rf_margin)
+loadd(site_rf_post_correction_margin)
 ```
@@
 
-Large phenotypic differences were found between the sites by supervised random forest (Figure #)
+Large phenotypic differences were found between the sites by supervised random forest (Figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
+This was as a results of differences in descriptors relating tree size and likely the ages of the trees between the sites. 
+To account for these differences and to make the trees across the sites comparable in the context of decline, an adjustment of site means to the overall mean was applied to each of the size descriptors (diameter at breast height, lower crown height, timber height, total height, crown radius).
+An example of this adjustment is shown for diameter at breast height in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.  
+This adjustment removed inter-site variability but maintained intra-site variability, allowing comparisons to be made between the trees at the different sites in the context of decline severity.
+Post adjustment the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
 
 ### Phenotypic decline and decline acuteness indexes
+
+Using unsupervised random forest proximities and the phenotypic descriptors as reported previously,multidimensional scaling (MDS) to two dimensions was performed to yield two phenotypic indexes:
+
+* **Phenotypic decline index (PDI)** - the inverse of the phenotypic health index reported previously providing a measure of how severely a tree is declining. 
+Min-max scaling performed on the first MDS dimension giving score between 0 and 1, with more severely declining trees having a score closer to 1.
+
+* **Decline acuteness index (DAI)** - a measure that allows the differentiation between chronically and acutely declining oaks. 
+The second MDS dimension was min-max scaled to give scores between -1 and 1 with chronically declining trees having a score closer to -1 and acutely declining trees having a score closer to 1. 
+Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.
+
+The two dimensional space provided by these phenotypic indexes, for measuring decline severity isshown in Figure 1. 
+The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
+They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations, which could have great utility in long term forest health monitoring for both forestry management and research.
+
+Crown related descriptors were found to contribute most to the PDI (Figure 2a) with more severely declining trees having reduced crown volume and condition. 
+The DAI was more complex in descriptor contribution with trunk staining and bleeding descriptors contributing highly with bleed prevalence contributing the most.
+Trees with any black staining or bleeding had scores greater than 0.
+Trees with high bleed prevalence and therefore a larger area of their trunk covered with black staining or active bleeds had much higher DAI scores.
+
+The phenotypic indexes of all 171 trees (including the 20 trees Chestnuts) are shown in Figure 3. 
+It can be seen that across the 9 sites, purely chronic oak decline sites such as Chestnuts and Speculation are easily distinguishable using the DAI compared to more AOD dominant sites such as Langdale or mixed sites such as Great Monks Wood.
+
+This plot highlights the inaccuracies in manual field classifications of decline status with several trees classified as healthy controls that, when compared quantitatively to other trees, are in fact
+in a decline state. 
+Most notably this has occurred in a number of trees at Attingham, Hatchlands and Winding Wood. 
+Also notable from Figure 3 is the disparity between the relative health of the healthier trees across the sites. 
+The healthier trees at sites such as Attingham and Hatchlands were found to be in a more declined state than those at sites such as Langdale and Chestnuts.
+
+### Trends of key phenotypic descriptors within the PDI and DAI measures
+
+The Phenotypic Decline Index (PDI) and Decline Acuteness Index (DAI) were previously defined in the 30 month report. 
+A machine learning approach was used to derive these measures as it allows the complex interactions and trends between phenotypic descriptors to be modelled from which the decline indexes are calculated.
+This would likely be oversimplified using conventional modelling strategies. 
+These complex interactions can be investigated by simulating phenotypic data and generating the index values from random forest models that have been trained for each of the indexes, supervised using phenotypic descriptors from the 171 trees.
+The index values generated by the models from these simulated phenotypic data can then be visualised as response surface plots, from which descriptor interactions within the indexes can be interpreted.
+Figures 1 and 2 show the simulated index value response surfaces under a variety of different circumstances for key phenotypic descriptors that contribute highly to the PDI and DAI.
+
+#### PDI
+
+Crown condition descriptors such as crown volume and crown condition contributed highly to the PDI. 
+The phenotypic measures central to this were percentage missing crown and percentage crowntransparency from which crown condition and crown volume are calculated. 
+It can be seen in Figure 1 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and estimated crown volume decrease, the PDI increases. 
+A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g.
+greater crown radius, total height, lower crown height and diameter at breast height; Figures 1a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.
+
+#### DAI
+
+The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, diameter at breast height and crown condition descriptors. 
+Simulated response surfaces for the DAI are shown in Figure 2. 
+These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 2a, b & c) and crown condition descriptors (Figures 2d, e & f) change for non-symptomatic AOD or COD trees (Figures 2a & d respectively), moderate AOD or COD (Figures 2b & e respectively) and severe AOD or COD (Figures 2c & f respectively).
+Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
+Also higher DAI values could be obtained from trees with greater stature (Figure 2c). 
+Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 2d, e & f) and therefore more severe COD.
+The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflict. 
+AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.
 

### 2019-11-26 code/functions/descriptorInfo.R
added to descriptor table

@@ -9,10 +9,10 @@ descriptorInfo <- function(analysisTable){
   
   type <- tribble(
     ~Descriptor,~Description,~`Categories / units`,~Reference,
-    'Diameter at breast height (mm)','','mm','',
+    'Diameter at breast height (mm)','Diameter of the trunk to within 1mm at a height of 1.3m.','mm','',
     'Crown radius (m)','','m','',
-    'Total height (m)','','m','',
-    'Lower crown height (m)','','m','',
+    'Total height (m)','Total tree height.','m','',
+    'Lower crown height (m)','Lowest live crown height to within 0.1m.','m','',
     'Timber height (m)','','m','',
     'Missing crown (%)','','%','',
     'Crown transparency (%)','','%','',
@@ -26,7 +26,7 @@ descriptorInfo <- function(analysisTable){
     'Insect defoliation type','','','',
     'Pruning / branch loss','','','',
     'Canopy closure','','','',
-    'Social class','','','',
+    'Social class','','1 = dominant; 2 = codominant; 3 = subdominant; 4 = suppressed','',
     'Agrilus exit holes','Number of *Agrilus* exit holes.','frequency','',
     'Stem epicormics','','','',
     'Dead stem tissue','','','',

### 2019-11-29 manuscript/manuscript.Rmd
added to descriptor table
added to introduction

@@
 
+**Authors:** Jasen Finch, Nathan Brown, Sandra Denman & John Draper
+
+**Proposed journal:** Still undecided! ??Plant Methods, Forestry??
+
+**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning
+
 ## Abstract
@@
 
-Oak decline syndromes are facilitated by a variety of both biotic and abiotic factors
+- define phenotype and descriptor
+- Acute and chronic decline associated phenotypes
+- The oak decline spectrum move away from simplistic binary categorisations
+- Manual assignment of decline status difficult prone to error for due to contribution of many factors.
+
+Oak decline syndromes are facilitated by a variety of both abiotic and biotic factors.
+
+Abiotic factors can include...
+Biotic factors can include...
+
+As a results of this, there are many visual cues measures that contribute to what can be defined as a declining oak tree.
+A severely declining oak is prominantly characterised by very poor crown condition and numerous dead branches.
+Compounding this complexity is the addition of different types of oak decline (AOD) that are often categorised based on the duration over which a tree has been declining or the biotic agents/factors associated with a declining tree.
+For instance a chronically declining oak can subsequently begin to also develop the stem bleeds associated with AOD.
+Accurately diagnosing in the field the severity and type of oak decline present is difficult and can requires much experience, expertise and training.
+
+Here, we propose a framework for quantitatively measuring the severity of oak decline
+
+* The phenotypic decline index (PDI) - a measure of decline severity scoring trees between 0 and 1.
+More severly declining oak trees have a score closer to 1.
+
+* The decline acuteness index (DAI) - a measure differentiating between chronically and acutely declining oak trees with a score between -1 and 1. 
+Acutely declining trees have a score closer to 1 and chronically declining trees have a score closer to -1.
 
@@
 
+- Site descriptions
+- Field phenotyping protocol
+
+Phenotypic data were collected from a total of 174 trees from 9 sites from across the south, eas and west of England.
+
 ### Phenotypic descriptor preparation and analysis
+
+- Aggregation of cardinal observations.
+- Random forest and MDS analyses
 
@@
 
-### Phenotypic descriptors
+### Phenotypic descriptors for oak decline
+
+- Descriptions of collected phenotypic descriptors and manual status assignment
 
@@
 
+- Analysis of cross site differences
+- Outline of cross site correction approach using mean centering adjustment
+
 Large phenotypic differences were found between the sites by supervised random forest (Figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
@@
 
-### Phenotypic decline and decline acuteness indexes
+### Phenotypic decline and decline acuteness indexes for oak decline severity
+
+- Decline index calculation
+- Comparison of indexes with manually assigned decline status labels
+- Top ranked descriptor index response surfaces
 
@@
 
+- Accessability of collecting descriptors
+- Suitability of collected phenotypic descriptors for describing decline status
+- Concession of model interpretability to predictive capability by using a machine learning approach
+- Trends of key phenotypic descriptors withing the PDI and DAI mesures
+- Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
+- Index thresholds for category definitions (healthy, declining, severe decline etc.)
+- When viewed in the 2D continuum, the categorisation of AOD to only include bleeds of certain bacteria seems likely to be too specific and by definition should only describe the period of time under which the tree have been declining.
+Perhaps subcategories of AOD would be more suitable such as Agrillus associated AOD or Armillaria associated AOD.
+
 ### Applications
 
+- Discussion of applications to forestry monitoring and management
+- Sensitivity for integration with molecular data compared to categorical assignments
+
 ## Conclusions
+
+- Crown condition descriptors most important for describing overall decline severity in the PDI.
+- Bacterial bleed related descriptors are most important for describing decline acuteness.
+- Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment
 
@@
 
-\newpage
 ## References

### 2019-12-06 manuscript/manuscript.Rmd
added a few notes

@@
 - Manual assignment of decline status difficult prone to error for due to contribution of many factors. 
+- Random forest
+- Unsupervised learning for complex patterns between descriptors.
 
@@
 
-Here, we propose a framework for quantitatively measuring the severity of oak decline
+Random forest is an ensemble machine learning
+
+Unsupervised learning ...
+
+Here, we propose a framework for quantitatively measuring the phenotypic severity of oak decline that is based on unsupervised random forest machine learning
 
@@
 - Field phenotyping protocol
+- Crown assessments
+- Cardinal assessments
+- Stem assessments - 2m base, 3m base
 
-Phenotypic data were collected from a total of 174 trees from 9 sites from across the south, eas and west of England.
+Phenotypic data were collected from a total of 174 trees from 9 sites from across the south, east and west of England (Figure #).
+A total of 30 phenotypic descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
 
@@
 - Random forest and MDS analyses
+- Descriptions of collected phenotypic descriptors and manual status assignment
 
-R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and analysis of phenotypic data.
+R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and analysis of phenotypic data (ref).
+
+Cardinal crown radius measurements were averaged.
 
@@
 ### Phenotypic descriptors for oak decline
-
-- Descriptions of collected phenotypic descriptors and manual status assignment
 
@@
 
-- Analysis of cross site differences
-- Outline of cross site correction approach using mean centering adjustment
-
 Large phenotypic differences were found between the sites by supervised random forest (Figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
@@
 
-### Phenotypic decline and decline acuteness indexes for oak decline severity
+### Calculation of indexes for oak decline severity
 
@@
 - Top ranked descriptor index response surfaces
+- Random forest able to assess categorical and continuous descriptors simultaneously
 
-Using unsupervised random forest proximities and the phenotypic descriptors as reported previously,multidimensional scaling (MDS) to two dimensions was performed to yield two phenotypic indexes:
+The phenotypic descriptors were analysed using 100 repeatitions of unsupervised random forest.
+The proximity 
+The relative proximities and the phenotypic descriptors multidimensional scaling (MDS) to two dimensions was performed to yield two phenotypic indexes:
 
-* **Phenotypic decline index (PDI)** - the inverse of the phenotypic health index reported previously providing a measure of how severely a tree is declining. 
-Min-max scaling performed on the first MDS dimension giving score between 0 and 1, with more severely declining trees having a score closer to 1.
+Proximity matrix transformed to a disimilarity (1- P) then multidimensional scaling (MDS) to 2 dimensions applied.
+The first dimension was found to be related to the overall severity of decline. 
+It was min-max scaled to give the PDI a scale between 0 and 1 with more severly declining trees having a score closer to 1.
 
@@
 
-The two dimensional space provided by these phenotypic indexes, for measuring decline severity isshown in Figure 1. 
-The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
-They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations, which could have great utility in long term forest health monitoring for both forestry management and research.
+### Descriptor contributions and trends
 
-Crown related descriptors were found to contribute most to the PDI (Figure 2a) with more severely declining trees having reduced crown volume and condition. 
-The DAI was more complex in descriptor contribution with trunk staining and bleeding descriptors contributing highly with bleed prevalence contributing the most.
+To understand the descriptors contributing to each of the indexes, individual random forest regression models of the phenotypic descriptors were trained, supervised using each decline index. 
+Crown related descriptors were found to contribute most to the PDI (Figure #) with more severely declining trees having reduced crown volume and condition. 
+The DAI was a more complex in descriptor contribution with trunk staining and bleeding descriptors contributing highly with bleed prevalence contributing the most.
 Trees with any black staining or bleeding had scores greater than 0.
@@
 The healthier trees at sites such as Attingham and Hatchlands were found to be in a more declined state than those at sites such as Langdale and Chestnuts.
-
-### Trends of key phenotypic descriptors within the PDI and DAI measures
 
@@
 
+The two dimensional space provided by these phenotypic indexes, for measuring decline severity isshown in Figure 1. 
+The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
+They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations, which could have great utility in long term forest health monitoring for both forestry management and research.
+
 ### Applications

### 2019-12-09 manuscript/plan.Rmd
added manucript plan

---
title: "Plan"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2019-12-09 R/plan.R

@@ -147,6 +147,9 @@ plan <- drake_plan(
   ## render figures
   figures = render(knitr_in('manuscript/figures.Rmd'),quiet = T),
   
+  ## render plan
+  plan = render(knit_in('manuscript/plan.Rmd'),quiet = TRUE),
+  
   ## render manuscript
   manuscript = render(knitr_in('manuscript/manuscript.Rmd'),quiet = T),
   

### 2019-12-10 manuscript/plan.Rmd
plan compiles

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
 
-## R Markdown
+\newpage
+## Checks
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
-
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
-
-```{r cars}
-summary(cars)
+```{r check,comment='',cache=FALSE}
+manuscript_check(file = 'plan.Rmd')
 ```
-
-## Including Plots
-
-You can also embed plots, for example:
-
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
-
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2019-12-10 R/plan.R

@@ -148,7 +148,7 @@ plan <- drake_plan(
   figures = render(knitr_in('manuscript/figures.Rmd'),quiet = T),
   
   ## render plan
-  plan = render(knit_in('manuscript/plan.Rmd'),quiet = TRUE),
+  plan = render(knitr_in('manuscript/plan.Rmd'),quiet = TRUE),
   
   ## render manuscript
   manuscript = render(knitr_in('manuscript/manuscript.Rmd'),quiet = T),

### 2019-12-10 manuscript/manuscript.Rmd
manuscript now rendered as a word document. Added to introduction. Added references

@@
 output: 
-  bookdown::pdf_document2:
+  bookdown::word_document2:
    toc: false
-   number_sections: false
-   keep_tex: no
+bibliography: references.bib
 ---
@@
 
-Random forest is an ensemble machine learning
+Machine learning is finding increasing utility in biological and geographical applications due to it's...
+A popular algroith is that of random forest due to it's versatility, relatively trivial training requirements and resiliance against overfitting [@breiman2001] (refs). 
+It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag (OOB) sample prediction accuracies and variable importance.
+Unsupervised learning is also possible
 
-Unsupervised learning ...
-
-Here, we propose a framework for quantitatively measuring the phenotypic severity of oak decline that is based on unsupervised random forest machine learning
+Here, we propose a framework to enable quantitative visible oak decline severity measurement that utilises a range of targeted phenotypic descriptors and unsupervised random forest machine learning.
+These measures include:
 

### 2019-12-10 manuscript/manuscript.Rmd
added to introduction

@@
 
-- define phenotype and descriptor
+- Increasing importance of monitoring 
 - Acute and chronic decline associated phenotypes
@@
 
-Oak decline syndromes are facilitated by a variety of both abiotic and biotic factors.
+Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
+Abiotic factors can include both predisposing and inciting factors such as drought and are often
 
-Abiotic factors can include...
 Biotic factors can include...
 
-As a results of this, there are many visual cues measures that contribute to what can be defined as a declining oak tree.
-A severely declining oak is prominantly characterised by very poor crown condition and numerous dead branches.
+As a results of this, there are many visual phenotypic descriptors encompassing the entirety of an oak tree that contribute to what can be defined as a declining oak tree.
+A severely declining oak is most prominantly characterised by very poor crown condition and numerous dead branches (ref).
+
 Compounding this complexity is the addition of different types of oak decline (AOD) that are often categorised based on the duration over which a tree has been declining or the biotic agents/factors associated with a declining tree.
@@
 
-Machine learning is finding increasing utility in biological and geographical applications due to it's...
+Machine learning is finding increasing utility in biological and geographical applications due to it's ability to learn complex patterns withing and between 
 A popular algroith is that of random forest due to it's versatility, relatively trivial training requirements and resiliance against overfitting [@breiman2001] (refs). 
 It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag (OOB) sample prediction accuracies and variable importance.
-Unsupervised learning is also possible
+Unsupervised learning is also possible [@shi2006]
 
@@
 
-### Phenotypic descriptor preparation and analysis
+### Phenotypic descriptor preparation
 
@@
 Cardinal crown radius measurements were averaged.
+
+### Machine learning analyses
+
+The code used to generate the decline indexes is available as an R package at [https://github.com/jasenfinch/pdi].
+All raw data and code used to generate this manuscript are available at [https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors]
 

### 2019-12-10 manuscript/manuscript.Rmd
manuscript check types can be selected
spell check

@@
 As a results of this, there are many visual phenotypic descriptors encompassing the entirety of an oak tree that contribute to what can be defined as a declining oak tree.
-A severely declining oak is most prominantly characterised by very poor crown condition and numerous dead branches (ref).
+A severely declining oak is most prominently characterised by very poor crown condition and numerous dead branches (ref).
 
@@
 Machine learning is finding increasing utility in biological and geographical applications due to it's ability to learn complex patterns withing and between 
-A popular algroith is that of random forest due to it's versatility, relatively trivial training requirements and resiliance against overfitting [@breiman2001] (refs). 
-It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag (OOB) sample prediction accuracies and variable importance.
+A popular algorithm is that of random forest due to it's versatility, relatively trivial training requirements and resilience against over fitting [@breiman2001] (refs). 
+It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag sample prediction accuracies and variable importance.
 Unsupervised learning is also possible [@shi2006]
@@
 * The phenotypic decline index (PDI) - a measure of decline severity scoring trees between 0 and 1.
-More severly declining oak trees have a score closer to 1.
+More severely declining oak trees have a score closer to 1.
 
@@
 
-The phenotypic descriptors were analysed using 100 repeatitions of unsupervised random forest.
+The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest.
 The proximity 
@@
 
-Proximity matrix transformed to a disimilarity (1- P) then multidimensional scaling (MDS) to 2 dimensions applied.
+Proximity matrix transformed to a dissimilarity (1- P) then multidimensional scaling (MDS) to 2 dimensions applied.
 The first dimension was found to be related to the overall severity of decline. 
-It was min-max scaled to give the PDI a scale between 0 and 1 with more severly declining trees having a score closer to 1.
+It was min-max scaled to give the PDI a scale between 0 and 1 with more severely declining trees having a score closer to 1.
 
@@
 - Concession of model interpretability to predictive capability by using a machine learning approach
-- Trends of key phenotypic descriptors withing the PDI and DAI mesures
+- Trends of key phenotypic descriptors withing the PDI and DAI measures
 - Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
@@
 - When viewed in the 2D continuum, the categorisation of AOD to only include bleeds of certain bacteria seems likely to be too specific and by definition should only describe the period of time under which the tree have been declining.
-Perhaps subcategories of AOD would be more suitable such as Agrillus associated AOD or Armillaria associated AOD.
+Perhaps subcategories of AOD would be more suitable such as Agrilus associated AOD or Armillaria associated AOD.
 
-The two dimensional space provided by these phenotypic indexes, for measuring decline severity isshown in Figure 1. 
+The two dimensional space provided by these phenotypic indexes, for measuring decline severity is shown in Figure 1. 
 The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
@@
 ```{r check,comment='',cache=FALSE}
-manuscript_check()
+manuscript_check(checks = c('word count','spelling'))
 ```

### 2019-12-10 manuscript/figures.Rmd
tables and figures temporarily rendered separate from the manuscript
fixed descriptor contribution plots with patchwork updates
removed site location map

@@
-```{r siteLocations,fig.cap='Map showing locations of the nine sites used for collection of phenotypic descriptors.',fig.width=3,fig.height=5}
-loadd(site_location_map)
+---
+title: Figures
+output: 
+  pdf_document:
+    keep_tex: false
+---
 
-print(site_location_map)
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE,
+                      warning = FALSE,
+                      message = FALSE)
 ```
@@
 
-descriptor_contribution_plots %>%
-  wrap_plots()
+descriptor_contribution_plots$a + descriptor_contribution_plots$b
 ```

### 2019-12-10 manuscript/manuscript.Rmd
added to introduction and methods

@@
 
-- Increasing importance of monitoring 
+- Monitoring currently focusses on crown assessments
 - Acute and chronic decline associated phenotypes
@@
 - Manual assignment of decline status difficult prone to error for due to contribution of many factors. 
-- Random forest
 - Unsupervised learning for complex patterns between descriptors.
@@
 Abiotic factors can include both predisposing and inciting factors such as drought and are often
-
-Biotic factors can include...
+Biotic agents can include insect infestation such as the larvae of bark boring bettle *Agrilus biguttatus* associated
 
@@
 A severely declining oak is most prominently characterised by very poor crown condition and numerous dead branches (ref).
+It is these characteristics that many studies use as the primary indicator of tree decline severity [@zarnoch2004].
 
 Compounding this complexity is the addition of different types of oak decline (AOD) that are often categorised based on the duration over which a tree has been declining or the biotic agents/factors associated with a declining tree.
-For instance a chronically declining oak can subsequently begin to also develop the stem bleeds associated with AOD.
+For instance a chronically declining oak can subsequently begin to also develop the stem bleeds associated with AOD [@denman2014].
 Accurately diagnosing in the field the severity and type of oak decline present is difficult and can requires much experience, expertise and training.
+
+It is becoming increasingly important that oak tree health within woodlands is accurately and sensitively monitoried to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices (ref).
+Monitoring methods need to take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.
 
@@
 Here, we propose a framework to enable quantitative visible oak decline severity measurement that utilises a range of targeted phenotypic descriptors and unsupervised random forest machine learning.
-These measures include:
+The two resulting measures are:
 
@@
 
-- Site descriptions
+- Site types
+- Descriptor types
 - Field phenotyping protocol
 - Crown assessments
-- Cardinal assessments
 - Stem assessments - 2m base, 3m base
 
-Phenotypic data were collected from a total of 174 trees from 9 sites from across the south, east and west of England (Figure #).
+Phenotypic data were collected from a total of 174 trees from 9 sites from across the south, east and west of England.
+
 A total of 30 phenotypic descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
+These descriptors are shown in Table #.
+Crown radius, canopy closure and tap tests for dead stem tissue were measured for each cardinal point.
+AOD stem symptoms (active bleeds, black staining, callused wounds) were collected from the basel 3m of the stem.
+Each individual
 
@@
 
+- Descriptors converted to presence and absence
 - Aggregation of cardinal observations.
+- Symptom type sizes average and frequency taken
 - Random forest and MDS analyses
@@
 
-Cardinal crown radius measurements were averaged.
+Cardinal assessments were averaged to give a single value.
 
@@
 
-The code used to generate the decline indexes is available as an R package at [https://github.com/jasenfinch/pdi].
-All raw data and code used to generate this manuscript are available at [https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors]
+The code used to generate the decline indexes is available as an R package pdi (v) at https://github.com/jasenfinch/pdi.
+All raw data and code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.
 
@@
 
-**Estimated crown volume (m^3^)** - estimated volume of crown as an ellipsoid proportional to the crown condition.
+**Estimated crown volume (m^3^)** - estimated volume of crown as an paraboloid proportional to the crown condition. Based on the composite crown volumne from @zarnoch2004 .
 
-$$v = \frac{c(h - l)}{150}\pi r^2$$
+$$v = \frac{\pi r^2c(h-l)}{200}$$
 
@@
 
-**Estimated bleed prevalence (%)** - estimated percentage of the surveyed trunk area affected by bleeds or black staining.
+**Estimated bleed prevalence (%)** - estimated percentage of the surveyed 3m basal trunk area affected by bleeds or black staining.
 
@@
 
-***Agrilus* exit hole density (m^-2^)** - density of observed *Agrilus* exit holes across the surveyed trunk area.
+***Agrilus* exit hole density (m^-2^)** - density of observed *Agrilus* exit holes across the surveyed 2m basal trunk area.
 $$e = \frac{h}{\frac{2d}{100}\pi}$$
@@
 - Index thresholds for category definitions (healthy, declining, severe decline etc.)
+- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
 - When viewed in the 2D continuum, the categorisation of AOD to only include bleeds of certain bacteria seems likely to be too specific and by definition should only describe the period of time under which the tree have been declining.

### 2019-12-11 manuscript/manuscript.Rmd
pdi package now used

@@
 
-$a$ = average active bleed size (cm), $A$ = number of active bleeds, $b$ = average black stain size (cm), $B$ = black stain number, $d$ = diameter at breast height (cm)
+$a$ = average active bleed size (mm), $A$ = number of active bleeds, $b$ = average black stain size (mm), $B$ = black stain number, $d$ = diameter at breast height (mm)
 

### 2019-12-11 R/functions/additionalDescriptors.R

@@ -1,16 +1,0 @@
-
-crownCondition <- function(m,t) {
-  (100 - m) * (1 - t/100)
-}
-
-crownVolume <- function(r,h,l,c) {
-  (pi * r ^ 2 * c * (h - l)) / 200
-}
-
-bleedPrevalence <- function(a,A,b,B,d) {
-  (a^2 * A + b^2 * B)/(3 * d * pi)
-}
-
-agrilusExitHoleDensity <- function(h,d) {
-  h/((2 * d)/1000 * pi)
-}

### 2019-12-11 R/functions/readPhenotypeSheet.R

@@ -1,92 +1,0 @@
-#' readPhenotypeSheet
-#' @description parse .xls phenotype data collection sheets
-#' @param file file path to excel file to parse
-
-readPhenotypeSheet <- function(file){
-  suppressWarnings(suppressMessages(description <- read_excel(file,sheet = 'Description')))
-  suppressWarnings(suppressMessages(symptoms <- read_excel(file,sheet = 'Symptoms')))
-
-  location <- description[1,2] %>%
-    unlist(use.names = F)
-  surveyor <- description[2,2] %>%
-    unlist(use.names = F)
-  date <- description[1,6] %>%
-    unlist(use.names = F)
-
-  emptyRows <- is.na(description$Description)
-
-  description <- description[!emptyRows,]
-
-  description <- description[-1,]
-
-  directionObservations <- description[,9:28]
-  description <- description[-1,-(9:28)]
-  colnames(description) <- description[1,]
-  description <- description[-1,]
-
-  description <- description %>%
-    rowid_to_column(var = 'ID') %>%
-    gather('Descriptor','Value',-ID,)
-
-  description <- description %>%
-    filter(!(Descriptor == "NA" | is.na(Descriptor)))
-
-  description$Descriptor[description$Descriptor == "Symptomatic (Y/N)"] <- 'Symptomatic'
-  description$Descriptor[description$Descriptor == "Social Class (1-5)"] <- 'Social Class'
-  description$Descriptor[description$Descriptor == "Missing crown    %"] <- 'Missing crown'
-  description$Descriptor[description$Descriptor == "Crown density     %"] <- 'Crown density'
-  description$Descriptor[description$Descriptor == "Agrillus exit holes (1,2...50, 50+)"] <- 'Agrillus exit holes'
-
-  description$Value[is.na(description$Value) & description$Descriptor == 'Insect defoliation type'] <- 'None'
-  description$Value[description$Descriptor == 'Symptomatic'] <- description$Value[description$Descriptor == 'Symptomatic'] %>% toupper()
-
-  directions <- directionObservations[1,] %>%
-    unlist(use.names = F) %>%
-    na.omit() %>%
-    toupper()
-
-  directionObservations <- directionObservations[-1,]
-
-  directionObservations <- split(1:20,ceiling(seq_along(1:20)/5)) %>%
-    map(~{
[…]

### 2019-12-11 R/packages.R

@@ -23,7 +23,8 @@ pacman::p_load(drake,
 
 # Load dependant GitHub libraries
 
-pacman::p_load_gh('ropenscilabs/gramr',
+pacman::p_load_gh('jasenfinch/pdi@devel',
+                  'ropenscilabs/gramr',
                   'benmarwick/wordcountaddin')
 
 # Resolve conflicts
@@ -33,3 +34,4 @@ conflict_prefer('gather','tidyr',quiet = T)
 conflict_prefer('filter','dplyr',quiet = T)
 conflict_prefer('margin','randomForest',quiet = T)
 conflict_prefer("explain", "lime",quiet = T)
+conflict_prefer("rf", "pdi",quiet = T)

### 2019-12-11 manuscript/figures.Rmd
changed figure order
swapped index scatter plot and site differences plot from supplementary to figures

@@
 
-```{r site_differences,fig.cap='Differences between sites in phenotypic descriptors analysed by supervised Random Forest.
-a) Multidimensional scaling (MDS) plot of random forest proximity values of individual trees scaled to two dimensions.
-Mean margin value shown.
-b) Relative descriptor Gini importances measures.',fig.height=5}
-loadd(site_differences_mds_plot)
-grid.arrange(site_differences_mds_plot$a,site_differences_mds_plot$b,nrow = 1)
-```
+```{r decline_index_plot}
+loadd(decline_indexes_plot)
 
-```{r DI_space,fig.cap='Two dimensional continuum of the phenotypic decline indexes for oak decline severity.',fig.height=3,fig.width=5}
-DIspace()
+decline_indexes_plot
 ```
@@
 ```
+
+```{r DI_space,fig.cap='Two dimensional continuum of the phenotypic decline indexes for oak decline severity.',fig.height=3,fig.width=5}
+DIspace()
+```

### 2019-12-11 manuscript/supplementary.Rmd
added DAI lime analysis plot to supplementary
swapped index scatter plot and site differences plot from supplementary to figures

@@
                       message = FALSE)
+```
+
+```{r site_differences,fig.cap='Differences between sites in phenotypic descriptors analysed by supervised Random Forest.
+a) Multidimensional scaling (MDS) plot of random forest proximity values of individual trees scaled to two dimensions.
+Mean margin value shown.
+b) Relative descriptor Gini importances measures.',fig.height=5}
+loadd(site_differences_mds_plot)
+grid.arrange(site_differences_mds_plot$a,site_differences_mds_plot$b,nrow = 1)
 ```
@@
   kable(caption = 'Descriptor correction factors at each site')
-```
-
-```{r decline_index_plot}
-loadd(decline_indexes_plot)
-
-decline_indexes_plot
 ```
@@
 ```
+
+```{r DAIlimeAnalysisPlot,fig.height=8}
+loadd(DAI_lime_analysis_plot)
+
+DAI_lime_analysis_plot
+```

### 2019-12-11 manuscript/manuscript.Rmd
added to introduction and methods

@@
 It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag sample prediction accuracies and variable importance.
+It is also able to simultaneously handle both continuous and categorical variables.
 Unsupervised learning is also possible [@shi2006]
@@
 Each individual 
+All trees were manually assigned a decline status category; healthy, AOD, COD or remission.
 
@@
 
-R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and analysis of phenotypic data (ref).
+R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and analysis of phenotypic data [@R2019].
 
@@
 
-The following four additional descriptors were calculated:
+Four additional composite descriptors were calculated to account for the three dimensional attributes of the tree and covariate nature of certain descriptors as suggested by @@zarnoch2004.
+These included:
 
@@
 
-Large phenotypic differences were found between the sites by supervised random forest (Figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
+Large phenotypic differences were found between the sites by supervised random forest (Supplementary  figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
 This was as a results of differences in descriptors relating tree size and likely the ages of the trees between the sites. 
@@
 - Decline index calculation
-- Comparison of indexes with manually assigned decline status labels
-- Top ranked descriptor index response surfaces
-- Random forest able to assess categorical and continuous descriptors simultaneously
 
@@
 
-Proximity matrix transformed to a dissimilarity (1- P) then multidimensional scaling (MDS) to 2 dimensions applied.
+Proximity matrix ($$P$$) was transformed to a dissimilarity ($$1 - P$$) then multidimensional scaling (MDS) to 2 dimensions applied.
 The first dimension was found to be related to the overall severity of decline.
 It was min-max scaled to give the PDI a scale between 0 and 1 with more severely declining trees having a score closer to 1.
-
-* **Decline acuteness index (DAI)** - a measure that allows the differentiation between chronically and acutely declining oaks. 
-The second MDS dimension was min-max scaled to give scores between -1 and 1 with chronically declining trees having a score closer to -1 and acutely declining trees having a score closer to 1. 
+The second MDS dimension was found to be associated with  min-max scaled to give scores between -1 and 1 with chronically declining trees having a score closer to -1 and acutely declining trees having a score closer to 1.
 Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.
+The PDI and DAI indexes are plotted for all 174 trees in Figure # showing the relationship with the
 
 ### Descriptor contributions and trends
+
+- LIME analysis to identify locally important descriptors under circumstances of interest
+- Top ranked descriptor index response surfaces
 
@@
 The DAI was a more complex in descriptor contribution with trunk staining and bleeding descriptors contributing highly with bleed prevalence contributing the most. 
-Trees with any black staining or bleeding had scores greater than 0.
 Trees with high bleed prevalence and therefore a larger area of their trunk covered with black staining or active bleeds had much higher DAI scores.
@@
 It can be seen that across the 9 sites, purely chronic oak decline sites such as Chestnuts and Speculation are easily distinguishable using the DAI compared to more AOD dominant sites such as Langdale or mixed sites such as Great Monks Wood.
-
-This plot highlights the inaccuracies in manual field classifications of decline status with several trees classified as healthy controls that, when compared quantitatively to other trees, are in fact
-in a decline state. 
-Most notably this has occurred in a number of trees at Attingham, Hatchlands and Winding Wood. 
-Also notable from Figure 3 is the disparity between the relative health of the healthier trees across the sites. 
-The healthier trees at sites such as Attingham and Hatchlands were found to be in a more declined state than those at sites such as Langdale and Chestnuts.
 
@@
 
-#### PDI
-
 Crown condition descriptors such as crown volume and crown condition contributed highly to the PDI. 
@@
 greater crown radius, total height, lower crown height and diameter at breast height; Figures 1a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.
-
-#### DAI
 
@@
 - Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
+- Scalability of method for wider application
 - When viewed in the 2D continuum, the categorisation of AOD to only include bleeds of certain bacteria seems likely to be too specific and by definition should only describe the period of time under which the tree have been declining.
 Perhaps subcategories of AOD would be more suitable such as Agrilus associated AOD or Armillaria associated AOD. 
+- Thresholds for decline groupings within index ranges
+- Validation of decline scores - chicken and paradox
+- Refined definition of remission
+
+This plot highlights the inaccuracies in manual field classifications of decline status with several trees classified as healthy controls that, when compared quantitatively to other trees, are in fact
+in a decline state. 
+Most notably this has occurred in a number of trees at Attingham, Hatchlands and Winding Wood. 
+Also notable from Figure 3 is the disparity between the relative health of the healthier trees across the sites. 
+The healthier trees at sites such as Attingham and Hatchlands were found to be in a more declined state than those at sites such as Langdale and Chestnuts.
 

### 2019-12-11 manuscript/manuscript.Rmd
added discussion point

@@
 - Refined definition of remission
+- use of unsupervised over supervised analysis
 

### 2019-12-11 manuscript/manuscript.Rmd

@@
 
+**Live crown ratio** -
+
 **Crown condition (%)** - percentage of present crown proportional to the percentage crown density.
@@
 
-**Estimated crown volume (m^3^)** - estimated volume of crown as an paraboloid proportional to the crown condition. Based on the composite crown volumne from @zarnoch2004 .
+**Composite crown volume (m^3^)** - Estimated volume of crown as an paraboloid proportional to the crown condition.
+Based on the composite crown volumne from @zarnoch2004 but uses crown condition instead of crown density.
 
@@
 $r$ = crown radius (m), $h$ = total height (m), $l$ = lower crown height (m), $c$ = crown condition (%)
+
+**Composite crown surface area (m^2^)** -
+
+**Crown production efficiency** -
 

### 2019-12-12 manuscript/manuscript.Rmd
added t-test results

@@
 ```{r loadData, include = FALSE}
-loadd(site_rf_margin)
-loadd(site_rf_post_correction_margin)
+loadd(site_rf_margin,
+      site_rf_post_correction_margin,
+      PDI_status_ttest,
+      DAI_groups_ttest)
 ```
@@
 Each individual 
-All trees were manually assigned a decline status category; healthy, AOD, COD or remission.
+All trees were manually assigned a decline status category; control, AOD, COD or remission.
 
@@
 R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and analysis of phenotypic data [@R2019].
-
-Cardinal assessments were averaged to give a single value.
-
-### Machine learning analyses
-
 The code used to generate the decline indexes is available as an R package pdi (v) at https://github.com/jasenfinch/pdi.
@@
 
-## Results
-
-### Phenotypic descriptors for oak decline
+Cardinal assessments were averaged to give a single value.
 
@@
 
+### Machine learning analyses
+
+The randomForest R package
+
+## Results
+
+### Phenotypic descriptors for oak decline
+
 ### Tree size descriptor adjustment by site
@@
 
-Proximity matrix ($$P$$) was transformed to a dissimilarity ($$1 - P$$) then multidimensional scaling (MDS) to 2 dimensions applied.
+Proximity matrix (*P*) was transformed to a dissimilarity (1 - *P*) then multidimensional scaling (MDS) to 2 dimensions applied.
 The first dimension was found to be related to the overall severity of decline.
@@
 Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.
-The PDI and DAI indexes are plotted for all 174 trees in Figure # showing the relationship with the
+A significant difference was found in the PDI between non-symptomatic (control) and symptomatic (AOD and COD) decline statuses using an independant two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
+Significance was also found in the DAI between the AOD and COD groups (`r glue("t({DAI_groups_ttest$parameter}) = {signif(DAI_groups_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(DAI_groups_ttest$estimate1,digits = 3)` and `r signif(DAI_groups_ttest$estimate2,digits = 3)` for the AOD and COD groups respectively.
+A scatter plot of PDI against DAI is shown in Figure # for all 174 trees.
+It can clearly be seen that the
 

### 2019-12-12 R/plan.R

@@ -87,6 +87,30 @@ plan <- drake_plan(
   ## plot decline indexes
   decline_indexes_plot = declineIndexesPlot(decline_indexes),
   
+  ## plot PDI against symptomatic and non-sympotmatic groups
+  PDI_status_plot = decline_indexes %>% 
+    ggplot(aes(x = Status, y = PDI)) + 
+    geom_boxplot() + 
+    theme_bw(),
+  
+  ## t-test for PDI and status
+  PDI_status_ttest = decline_indexes %>%
+    t.test(PDI~Status,data = .,var.equal = TRUE) %>%
+    tidy(),
+  
+  ## plot DAI against AOD and COD groups
+  DAI_groups_plot = decline_indexes %>% 
+    filter(ChosenGroup %in% c('AOD','COD')) %>% 
+    ggplot(aes(x = ChosenGroup, y = DAI)) + 
+    geom_boxplot() + 
+    theme_bw(),
+  
+  ## t-test for DAI and AOD and COD groups
+  DAI_groups_ttest = decline_indexes %>% 
+    filter(ChosenGroup %in% c('AOD','COD')) %>%
+    t.test(DAI~ChosenGroup,data = .,var.equal = TRUE) %>%
+    tidy(),
+  
   ## plot descriptors against decline indexes 
   descriptor_scatter_plots = site_corrected_analysis_suitable_data %>%
     descriptorScatterPlots(decline_indexes),

### 2019-12-12 manuscript/manuscript.Rmd
added equations for newly added additional descriptors

@@
 - Site types
-- Descriptor types
 - Field phenotyping protocol
@@
 Crown radius, canopy closure and tap tests for dead stem tissue were measured for each cardinal point.
-AOD stem symptoms (active bleeds, black staining, callused wounds) were collected from the basel 3m of the stem.
-Each individual 
+The crown radius was
+The AOD stem symptoms (active bleeds, black staining, callused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem (ref).
+The frequency of *A.
+biguttatus* exit holes were counted along with the presence or absence of each of oval shaped exit holes, small circular shaped exit holes and ground level fungal fruiting bodies were recorded from the basal 2m of the stem (ref).
 All trees were manually assigned a decline status category; control, AOD, COD or remission.
@@
 
+$$R = \frac{(h-l)}{h} \times 100$$
+
+$h$ = total height (m), $l$ = lower crown height (m)
+
 **Crown condition (%)** - percentage of present crown proportional to the percentage crown density.
@@
 
+$$S = \frac{4\pi(h-l)}{3r^2} \left[ \left( r^2 + \frac{r^4}{4(h-l)^2} \right) ^{1.5} - \left( \frac{r^4}{4(h-l)^2} \right) ^{1.5} \right] \times c $$
+
+$r$ = crown radius (m), $h$ = total height (m), $l$ = lower crown height (m), c = crown condition (%)
+
 **Crown production efficiency** -
+
+$$e = \frac{S}{v}$$
+
+$S$ = composite crown surface area (m^2^), $v$ = composite crown volume (m^3^)
 

### 2019-12-12 manuscript/supplementary.Rmd
added new composite crown descriptor names to manuscript
fixed importance plots

@@
 Mean margin value shown.
-b) Relative descriptor Gini importances measures.',fig.height=5}
+b) Relative descriptor Gini importances measures.',fig.height=6}
 loadd(site_differences_mds_plot)
@@
 ```
+
+\newpage
 

### 2019-12-12 R/functions/descriptorImportancePlots.R

@@ -16,7 +16,8 @@ descriptorImportancePlots <- function(PDIrf,DAIrf){
         as.character() %>%
         {c(.[1:(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) - 1)],
            expression(Agrillus~exit~hole~density ( m^-2 ) ),
-           .[(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) + 1):34],
+           .[(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) + 1):36],
+           expression(Crown~surface~area ( m^2 )),
            expression(Crown~volume ( m^3 ) )
         )}
       
@@ -26,7 +27,7 @@ descriptorImportancePlots <- function(PDIrf,DAIrf){
         theme_bw() +
         theme(plot.title = element_text(face = 'bold'),
               axis.title = element_text(face = 'bold',size = 10)) +
-        labs(title = 'a)',
+        labs(title = 'a) PDI',
              y = 'Relative Gini\nimportance',
              x = NULL) +
         scale_x_discrete(labels = descriptorLabels)
@@ -46,9 +47,10 @@ descriptorImportancePlots <- function(PDIrf,DAIrf){
         as.character() %>%
         {c(.[1:(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) - 1)],
            expression(Agrillus~exit~hole~density ( m^-2 ) ),
-           .[(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) + 1):(which(str_detect(.,coll('Crown volume (m^3)'))) - 1)],
+           .[(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) + 1):(which(str_detect(.,coll('Crown surface area (m^2)'))) - 1)],
+           expression(Crown~surface~area ( m^2 )),
            expression(Crown~volume ( m^3 ) ),
-           .[(which(str_detect(.,coll('Crown volume (m^3)'))) + 1):35]
+           .[(which(str_detect(.,coll('Crown volume (m^3)'))) + 1):38]
         )}
       
       ggplot(dat,aes(x = Feature,y = `Relative Importance`)) +
@@ -57,7 +59,7 @@ descriptorImportancePlots <- function(PDIrf,DAIrf){
         theme_bw() +
         theme(plot.title = element_text(face = 'bold'),
               axis.title = element_text(face = 'bold',size = 10)) +
-        labs(title = 'b)',
[…]

### 2019-12-12 manuscript/manuscript.Rmd
Methods and results rearrange. Added site details

@@
 - Crown assessments
-- Stem assessments - 2m base, 3m base
+- Other assessments
 
-Phenotypic data were collected from a total of 174 trees from 9 sites from across the south, east and west of England.
+Phenotypic data were collected from a total of 174 trees from nine sites from across the south, east and west of England between June and September in 2016 or 2017.
+The occurance of AOD has previously been reported and studied at six of the sites [@brown2016; @brown2017].
 
@@
 
-### Machine learning analyses
-
-The randomForest R package
-
-## Results
-
-### Phenotypic descriptors for oak decline
-
-### Tree size descriptor adjustment by site
+### Tree size descriptor adjustment for age and inter site variance
 
@@
 
+The randomForest R package
+
 The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest.
@@
 Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.
+
+###
+
+## Results
+
+### Site differences
+
+### Oak decline indexes
+
 A significant difference was found in the PDI between non-symptomatic (control) and symptomatic (AOD and COD) decline statuses using an independant two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.

### 2019-12-12 manuscript/manuscript.Rmd
added to methods section

@@
 output: 
-  bookdown::word_document2:
-   toc: false
+  #bookdown::word_document2:
+  # toc: false
+  bookdown::pdf_document2: default
 bibliography: references.bib
@@
       PDI_status_ttest,
-      DAI_groups_ttest)
+      DAI_groups_ttest,
+      PDI_rf_tune_params,
+      DAI_rf_tune_params)
 ```
@@
 
-R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and analysis of phenotypic data [@R2019].
-The code used to generate the decline indexes is available as an R package pdi (v) at https://github.com/jasenfinch/pdi.
+R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and subsequent analysis of phenotypic data [@R2019].
+The code used to generate the decline indexes is available as an R package pdi v`r packageVersion('pdi') %>% as.character()` at https://github.com/jasenfinch/pdi.
 All raw data and code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.
@@
 
-Large phenotypic differences were found between the sites by supervised random forest (Supplementary  figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
-This was as a results of differences in descriptors relating tree size and likely the ages of the trees between the sites. 
 To account for these differences and to make the trees across the sites comparable in the context of decline, an adjustment of site means to the overall mean was applied to each of the size descriptors (diameter at breast height, lower crown height, timber height, total height, crown radius).
@@
 This adjustment removed inter-site variability but maintained intra-site variability, allowing comparisons to be made between the trees at the different sites in the context of decline severity.
-Post adjustment the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
 
@@
 
-- Decline index calculation
-
-The randomForest R package
-
-The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest.
+The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest using the randomForest R package v`r packageVersion('randomForest') %>% as.character()` [@liaw2002].
 The proximity 
@@
 
-###
+### Overall and local descriptor importance
+
+- LIME analysis to identify locally important descriptors under circumstances of interest
+- simulated descriptor response surfaces
+
+Individual random forest regression models of the phenotypic descriptors were trained, supervised using each decline index.
+Optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training. 
+The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and as follows: PDI - `r glue(ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)})`; DAI - ntree = , mtry = , PMAE = (Supplementary figure #)
+To understand the descriptors contributing to each of the indexes, Local Interpretable Model-Agnostic Explanations (LIME) using the lime R package v`r packageVersion('lime') %>% as.character()` [@ribeiro2016; @pedersen2019]
 
@@
 ### Site differences
+
+Large phenotypic differences were found between the sites by supervised random forest (Supplementary  figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
+This was as a results of differences in descriptors relating tree size and likely the ages of the trees between the sites. 
+Post adjustment the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
 
@@
 
-- LIME analysis to identify locally important descriptors under circumstances of interest
+- Accuracies of retrained models
+- Lime results
 - Top ranked descriptor index response surfaces
 
-To understand the descriptors contributing to each of the indexes, individual random forest regression models of the phenotypic descriptors were trained, supervised using each decline index. 
 Crown related descriptors were found to contribute most to the PDI (Figure #) with more severely declining trees having reduced crown volume and condition. 

### 2019-12-12 manuscript/supplementary.Rmd
figure rearrange
lime figures converted to tables

@@
     keep_tex: false
+header-includes:
+  - \usepackage{booktabs}
+  - \usepackage{longtable}
+  - \usepackage{array}
+  - \usepackage{multirow}
+  - \usepackage{wrapfig}
+  - \usepackage{float}
+  - \usepackage{colortbl}
+  - \usepackage{pdflscape}
+  - \usepackage{tabu}
+  - \usepackage{threeparttable}
+  - \usepackage{threeparttablex}
+  - \usepackage[normalem]{ulem}
+  - \usepackage{makecell}
+  - \usepackage{xcolor}
 ---
@@
 
-```{r site_differences,fig.cap='Differences between sites in phenotypic descriptors analysed by supervised Random Forest.
-a) Multidimensional scaling (MDS) plot of random forest proximity values of individual trees scaled to two dimensions.
-Mean margin value shown.
-b) Relative descriptor Gini importances measures.',fig.height=6}
-loadd(site_differences_mds_plot)
-grid.arrange(site_differences_mds_plot$a,site_differences_mds_plot$b,nrow = 1)
+```{r phenotypic_descriptors,cache=FALSE}
+loadd(phenotypic_descriptor_info)
+
+phenotypic_descriptor_info %>%
+  kable(format = 'latex',booktabs = TRUE)
 ```
@@
 site_correction_factors %>%
+  mutate_if(is.double,signif,digits = 3) %>%
   kable(caption = 'Descriptor correction factors at each site')
@@
 
-\newpage
+```{r modelTuningResultsPlot}
+loadd(rf_tune_plot)
+
+rf_tune_plot
+```
+
+```{r site_differences,fig.cap='Differences between sites in phenotypic descriptors analysed by supervised Random Forest.
+a) Multidimensional scaling (MDS) plot of random forest proximity values of individual trees scaled to two dimensions.
+Mean margin value shown.
+b) Relative descriptor Gini importances measures.',fig.height=6}
+loadd(site_differences_mds_plot)
+grid.arrange(site_differences_mds_plot$a,site_differences_mds_plot$b,nrow = 1)
+```
 
@@
 ```
-
-```{r modelTuningResultsPlot}
-loadd(rf_tune_plot)
-
-rf_tune_plot
-```
-
-```{r PDIlimeAnalysisPlot}
-loadd(PDI_lime_analysis_plot)
-
-PDI_lime_analysis_plot
-```
-
-```{r DAIlimeAnalysisPlot,fig.height=8}
-loadd(DAI_lime_analysis_plot)
-
-DAI_lime_analysis_plot
-```

### 2019-12-12 manuscript/tables.Rmd
added to descriptor table
lime figures converted to tables

@@
 title: Tables
-output: pdf_document
+output: 
+  pdf_document:
+    includes:
+header-includes:
+  - \usepackage{booktabs}
+  - \usepackage{longtable}
+  - \usepackage{array}
+  - \usepackage{multirow}
+  - \usepackage{wrapfig}
+  - \usepackage{float}
+  - \usepackage{colortbl}
+  - \usepackage{pdflscape}
+  - \usepackage{tabu}
+  - \usepackage{threeparttable}
+  - \usepackage{threeparttablex}
+  - \usepackage[normalem]{ulem}
+  - \usepackage{makecell}
+  - \usepackage{xcolor}
 ---
-
-## Tables
 
@@
 
-```{r phenotypic_descriptors,cache=FALSE}
-loadd(phenotypic_descriptor_info)
+```{r PDI_lime_results}
+loadd(PDI_lime_analysis_table)
 
-phenotypic_descriptor_info %>%
-  kable(format = 'markdown')
+PDI_lime_analysis_table %>%
+  kable(format = 'latex',booktabs = TRUE,escape = FALSE,caption = ' ') %>%
+  collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
 ```
 
-```{r resampled_results}
+```{r DAI_lime_results}
+loadd(DAI_lime_analysis_table)
 
+DAI_lime_analysis_table %>%
+  kable(format = 'latex',booktabs = TRUE,escape = FALSE,caption = ' ') %>%
+  collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
 ```

### 2019-12-12 R/functions/PDIlimeAnalysisTable.R

@@ -1,0 +1,22 @@
+
+PDIlimeAnalysisTable <- function(PDI_lime_analysis,decline_indexes){
+  cases <- tibble(case = 1:3 %>% as.character(),
+                  name = c('Healthy','Moderate decline','Severe decline'),
+                  PDI = c(min(decline_indexes$PDI),
+                          decline_indexes$PDI[abs(decline_indexes$PDI - 0.5) == min(abs(decline_indexes$PDI - 0.5))],
+                          max(decline_indexes$PDI)))
+  
+  PDI_lime_analysis %>%
+    left_join(cases,by = 'case') %>%
+    select(Status = name,PDI,`Predicted PDI` = prediction,Descriptor = feature,Value = feature_value,`Descriptor (\\textit{d}) range` = feature_desc,Weight = feature_weight) %>%
+    {
+      for (i in unique(.$Descriptor)) {
+        .$`Descriptor (\\textit{d}) range` <- str_replace_all(.$`Descriptor (\\textit{d}) range`,coll(i),'\\textit{d}')
+      }
+      return(.)
+    } %>%
+    mutate_if(is.double,signif,digits = 3) %>%
+    mutate(Descriptor = str_replace_all(Descriptor,coll('m^3'),'m\\textsuperscript{3}'),
+           Descriptor = str_replace_all(Descriptor,coll('m^2'),'m\\textsuperscript{2}'),
+           Descriptor = str_replace_all(Descriptor,coll('%'),'\\%'))
+}

### 2019-12-12 R/packages.R

@@ -21,7 +21,8 @@ pacman::p_load(drake,
                lime,
                patchwork,
                broom,
-               glue)
+               glue,
+               kableExtra)
 
 # Load dependant GitHub libraries
 

### 2019-12-12 R/plan.R

@@ -152,8 +152,14 @@ plan <- drake_plan(
   ## DAI lime analysis
   DAI_lime_analysis = DAIlimeAnalysis(site_corrected_analysis_suitable_data,DAI_rf_model,decline_indexes),
   
+  ## PDI lime analysis table
+  PDI_lime_analysis_table = PDIlimeAnalysisTable(PDI_lime_analysis,decline_indexes),
+  
   ## PDI lime analysis plot
   PDI_lime_analysis_plot = PDIlimeAnalysisPlot(PDI_lime_analysis,decline_indexes),
+  
+  ## PDI lime analysis table
+  DAI_lime_analysis_table = DAIlimeAnalysisTable(DAI_lime_analysis,decline_indexes),
   
   ## DAI lime analysis plot
   DAI_lime_analysis_plot = DAIlimeAnalysisPlot(DAI_lime_analysis,decline_indexes),

### 2019-12-12 manuscript/manuscript.Rmd
added to methods

@@
   # toc: false
-  bookdown::pdf_document2: default
+  bookdown::pdf_document2: 
+    toc: false
+    number_sections: false
 bibliography: references.bib
@@
 
-**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning
+**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning, unsupervised learning, random forest
 
@@
 - Other assessments
+- Descriptions of collected phenotypic descriptors
 
@@
 
-A total of 30 phenotypic descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
+Descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
 These descriptors are shown in Table #.
@@
 
+All raw data and analysis code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.
+
 ### Phenotypic descriptor preparation
@@
 - Symptom type sizes average and frequency taken
-- Random forest and MDS analyses
-- Descriptions of collected phenotypic descriptors and manual status assignment
+- DBH converted to m and symptom sizes to mm
 
 R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and subsequent analysis of phenotypic data [@R2019].
-The code used to generate the decline indexes is available as an R package pdi v`r packageVersion('pdi') %>% as.character()` at https://github.com/jasenfinch/pdi.
-All raw data and code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.
 
@@
 
-Four additional composite descriptors were calculated to account for the three dimensional attributes of the tree and covariate nature of certain descriptors as suggested by @@zarnoch2004.
-These included:
+A total of 31 raw descriptors collected
+
+### Composite descriptor calculation
+
+Seven additional, composite descriptors were calculated to account for the three dimensional attributes of the tree and covariate nature of certain descriptors as suggested by @zarnoch2004.
+This then gave a total of 38 raw and composite phenotypic descriptors.
+The additional descriptors are defined as follows:
 
@@
 ***Agrilus* exit hole density (m^-2^)** - density of observed *Agrilus* exit holes across the surveyed 2m basal trunk area.
+
 $$e = \frac{h}{\frac{2d}{100}\pi}$$
+
 $h$ = number of *Agrilus* exit holes, $d$ = diameter at breast height (cm)
@@
 An example of this adjustment is shown for diameter at breast height in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.
-This adjustment removed inter-site variability but maintained intra-site variability, allowing comparisons to be made between the trees at the different sites in the context of decline severity.
 
@@
 The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest using the randomForest R package v`r packageVersion('randomForest') %>% as.character()` [@liaw2002].
-The proximity 
-The relative proximities and the phenotypic descriptors multidimensional scaling (MDS) to two dimensions was performed to yield two phenotypic indexes:
+The proximity matrices from each repetition were averaged to a single proximity matrix ($P$).
+This was then transformed to a dissimilarity matrix ($1 - P$) and scaled to two dimensions using  multidimensional scaling (MDS).
+The first MDS dimension was min-max scaled to give a scale between 0 and 1 with more severely declining trees having a score closer to 1.
+This dimension was named the PDI.
+The second MDS dimension ($D$) was first inverted ($1 - D$) to ensure the correct dimension orientation relative to decline type.
+The dimension was then min-max scaled and then transformed ($2D - 1$) to give a scale between -1 and 1 with COD trees having a score closer to -1 and AOD having a score closer to 1. 
+This dimension was named the DAI.
+With different data, it is likely that the MDS dimensions may require inversion prior to min-max scaling to ensure their correct orientation.
 
-Proximity matrix (*P*) was transformed to a dissimilarity (1 - *P*) then multidimensional scaling (MDS) to 2 dimensions applied.
-The first dimension was found to be related to the overall severity of decline.
-It was min-max scaled to give the PDI a scale between 0 and 1 with more severely declining trees having a score closer to 1.
-The second MDS dimension was found to be associated with  min-max scaled to give scores between -1 and 1 with chronically declining trees having a score closer to -1 and acutely declining trees having a score closer to 1.
-Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.
+The code used to generate the decline indexes is available as an R package pdi v`r packageVersion('pdi') %>% as.character()` at https://github.com/jasenfinch/pdi.
 
@@
 Optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training. 
-The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and as follows: PDI - `r glue(ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)})`; DAI - ntree = , mtry = , PMAE = (Supplementary figure #)
+The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and as follows: PDI - `r glue('ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; DAI - `r glue('ntree = {DAI_rf_tune_params$ntree}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure #)
+
 To understand the descriptors contributing to each of the indexes, Local Interpretable Model-Agnostic Explanations (LIME) using the lime R package v`r packageVersion('lime') %>% as.character()` [@ribeiro2016; @pedersen2019]
@@
 This was as a results of differences in descriptors relating tree size and likely the ages of the trees between the sites. 
+This adjustment removed inter-site variability but maintained intra-site variability, allowing comparisons to be made between the trees at the different sites in the context of decline severity.
 Post adjustment the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
@@
 
+Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.
 A significant difference was found in the PDI between non-symptomatic (control) and symptomatic (AOD and COD) decline statuses using an independant two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
@@
 - Scalability of method for wider application
-- When viewed in the 2D continuum, the categorisation of AOD to only include bleeds of certain bacteria seems likely to be too specific and by definition should only describe the period of time under which the tree have been declining.
-Perhaps subcategories of AOD would be more suitable such as Agrilus associated AOD or Armillaria associated AOD. 
 - Thresholds for decline groupings within index ranges
@@
 - use of unsupervised over supervised analysis
+- Scale inversion likely need with new data
 
@@
 - Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment
+- Encouraged to use this approach in other contexts
 

### 2019-12-12 manuscript/supplementary.Rmd
improved descriptor info table

@@
 phenotypic_descriptor_info %>%
-  kable(format = 'latex',booktabs = TRUE)
+  kable(format = 'latex',booktabs = TRUE,longtable = TRUE) %>%
+  kable_styling(latex_options = 'repeat_header',repeat_header_continued = TRUE) %>%
+  column_spec(1,width = '1.5cm') %>%
+  column_spec(2,width = '3cm') %>%
+  column_spec(4:5,width = '7cm') %>%
+  collapse_rows(columns = 1,latex_hline = "major", valign = "middle") %>%
+  landscape()
 ```
@@
   mutate_if(is.double,signif,digits = 3) %>%
-  kable(caption = 'Descriptor correction factors at each site')
+  kable(format = 'latex',booktabs = TRUE,caption = 'Descriptor correction factors at each site') %>%
+  collapse_rows(columns = 1,latex_hline = "major", valign = "middle")
 ```

### 2019-12-12 R/functions/descriptorInfo.R

@@ -5,7 +5,7 @@
 descriptorInfo <- function(analysisTable){
   pt <- analysisTable %>%
     {tibble(Descriptor = names(.),`Data Type` = map_chr(.,class))} %>%
-    filter(!(Descriptor %in% c('Crown condition (%)','Crown volume (m^3)','Bleed prevalence (%)','Agrilus exit hole density (m^-2)')))
+    filter(!(Descriptor %in% c('Crown condition (%)','Crown volume (m^3)','Bleed prevalence (%)','Agrilus exit hole density (m^-2)',"Live crown ratio (%)","Crown surface area (m^2)","Crown production efficiency")))
   
   type <- tribble(
     ~Descriptor,~Description,~`Categories / units`,~Reference,

### 2019-12-12 manuscript/figures.Rmd

@@
 
-```{r descriptor_contributions, fig.height = 5,fig.cap = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures.
+```{r descriptor_contributions, fig.height = 6,fig.cap = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures.
 a) Phenotypic decline index (PDI).

### 2019-12-12 manuscript/manuscript.Rmd

@@
 
-**Live crown ratio** -
+**Live crown ratio (%)** - percentage of the whole tree height that supports live crown.
 
@@
 **Composite crown volume (m^3^)** - Estimated volume of crown as an paraboloid proportional to the crown condition.
-Based on the composite crown volumne from @zarnoch2004 but uses crown condition instead of crown density.
+Based on the composite crown volumne from @zarnoch2004 but uses crown condition instead of crown density alone.
 
@@
 
-$r$ = crown radius (m), $h$ = total height (m), $l$ = lower crown height (m), $c$ = crown condition (%)
+$r$ = crown radius (m)
 
@@
 
-$r$ = crown radius (m), $h$ = total height (m), $l$ = lower crown height (m), c = crown condition (%)
-
 **Crown production efficiency** -
@@
 $$e = \frac{S}{v}$$
-
-$S$ = composite crown surface area (m^2^), $v$ = composite crown volume (m^3^)
 
@@
 
-$h$ = number of *Agrilus* exit holes, $d$ = diameter at breast height (cm)
+$h$ = number of *Agrilus* exit holes
 

### 2019-12-13 manuscript/figures.Rmd

@@
 
-grid.arrange(PDI_response_surfaces$`a)`,PDI_response_surfaces$`b)`,PDI_response_surfaces$`c)`,PDI_response_surfaces$legend,nrow = 2,newpage = F)
+grid.arrange(PDI_response_surfaces$`a)`,
+             PDI_response_surfaces$`b)`,
+             PDI_response_surfaces$`c)`,
+             PDI_response_surfaces$legend,
+             nrow = 2,newpage = F)
 ```
@@
 
-grid.arrange(DAI_response_surfaces$`a)`,DAI_response_surfaces$`b)`,DAI_response_surfaces$`c)`,DAI_response_surfaces$`d)`,DAI_response_surfaces$`e)`,DAI_response_surfaces$`f)`,DAI_response_surfaces$legend,ncol = 3,layout_matrix = rbind(c(1,4,7),c(2,5,7),c(3,6,7)),widths = c(4,4,1),newpage = F)
+grid.arrange(textGrob(expression(bold('AOD')),hjust = -0.1),
+             textGrob(expression(bold('COD')),hjust = -0.1),
+             textGrob(expression(bold('Neutral')),rot = 90,hjust = -0.1),
+             textGrob(expression(bold('Moderate')),rot = 90,hjust = -0.1),
+             textGrob(expression(bold('Severe')),rot = 90,hjust = -0.1),
+             DAI_response_surfaces$`a)`,
+             DAI_response_surfaces$`b)`,
+             DAI_response_surfaces$`c)`,
+             DAI_response_surfaces$`d)`,
+             DAI_response_surfaces$`e)`,
+             DAI_response_surfaces$`f)`,
+             DAI_response_surfaces$legend,
+             ncol = 4,
+             layout_matrix = rbind(c(NA,1,2,NA),c(3,6,9,12),c(4,7,10,12),c(5,8,11,12)),
+             widths = c(1,10,10,2),
+             heights = c(1,10,10,10),
+             newpage = F)
 ```

### 2019-12-13 manuscript/manuscript.Rmd
composite descriptor descriptions complete

@@
 
-Machine learning is finding increasing utility in biological and geographical applications due to it's ability to learn complex patterns withing and between 
+Machine learning is finding increasing utility in biological and geographical applications due to it's ability to learn complex patterns within and between variables.
 A popular algorithm is that of random forest due to it's versatility, relatively trivial training requirements and resilience against over fitting [@breiman2001] (refs). 
@@
 Descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
-These descriptors are shown in Table #.
 Crown radius, canopy closure and tap tests for dead stem tissue were measured for each cardinal point.
@@
 biguttatus* exit holes were counted along with the presence or absence of each of oval shaped exit holes, small circular shaped exit holes and ground level fungal fruiting bodies were recorded from the basal 2m of the stem (ref).
+These descriptors are shown in Supplementary table #.
 All trees were manually assigned a decline status category; control, AOD, COD or remission.
-
 All raw data and analysis code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.
@@
 **Composite crown volume (m^3^)** - Estimated volume of crown as an paraboloid proportional to the crown condition.
-Based on the composite crown volumne from @zarnoch2004 but uses crown condition instead of crown density alone.
+Adapted from the composite crown volume from @zarnoch2004 using crown condition instead of crown density alone.
 
@@
 
-**Composite crown surface area (m^2^)** -
+**Composite crown surface area (m^2^)** - Estimated crown surface area proportional to crown condition. Adapted from @laroque1994.
 
@@
 
-**Crown production efficiency** -
+**Crown production efficiency** - Estimate of crown production effieceny as a ratio between the composite crown surface area and the composite crown volumne [@zarnoch2004].
 

### 2019-12-13 manuscript/manuscript.Rmd
added to methods section

@@
 
+### Tree size descriptor adjustment for age and inter site variance
+
+To account age, size and site specific tree difference, an adjustment of site means to the overall mean was applied to each of the size descriptors (diameter at breast height, lower crown height, timber height, total height, crown radius).
+An example of this adjustment is shown for diameter at breast height in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.
+
 ### Composite descriptor calculation
@@
 
-### Tree size descriptor adjustment for age and inter site variance
-
-To account for these differences and to make the trees across the sites comparable in the context of decline, an adjustment of site means to the overall mean was applied to each of the size descriptors (diameter at breast height, lower crown height, timber height, total height, crown radius).
-An example of this adjustment is shown for diameter at breast height in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.
-
 ### Calculation of indexes for oak decline severity
@@
 
-### Overall and local descriptor importance
+### Identifying important descriptors and their trends within the decline indexes
 
-- LIME analysis to identify locally important descriptors under circumstances of interest
 - simulated descriptor response surfaces
@@
 Individual random forest regression models of the phenotypic descriptors were trained, supervised using each decline index.
-Optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training. 
-The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and as follows: PDI - `r glue('ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; DAI - `r glue('ntree = {DAI_rf_tune_params$ntree}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure #)
+The optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training. 
+The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: PDI - `r glue('ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; DAI - `r glue('ntree = {DAI_rf_tune_params$ntree}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure #)
 
-To understand the descriptors contributing to each of the indexes, Local Interpretable Model-Agnostic Explanations (LIME) using the lime R package v`r packageVersion('lime') %>% as.character()` [@ribeiro2016; @pedersen2019]
+Local Interpretable Model-Agnostic Explanations (LIME) To understand the descriptors contributing to case specific examples of each of the indexes,  were computed for a range of specific cases [@ribeiro2016].
+The lime R package v`r packageVersion('lime') %>% as.character()` was used with the default parameters [@pedersen2019].
+A low, a moderate and a severe decline cases was used for the PDI. For each case, a tree with a score closest to 0, 0.5 and 1 respectively was selected.
+For the DAI, a decline neautral case was selected with a score closes to 0.
+Then for each of the AOD and COD syndrome types a moderatefor the DAI .
 
@@
 
-### Descriptor contributions and trends
+### Descriptor contributions and trends to the decline indexes
 

### 2019-12-13 manuscript/tables.Rmd
adjustment to DAI LIME results table

@@
   kable(format = 'latex',booktabs = TRUE,escape = FALSE,caption = ' ') %>%
-  collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
+  collapse_rows(columns = 1:4, latex_hline = "custom", valign = "middle",custom_latex_hline = 1:4)
 ```

### 2019-12-13 R/functions/DAIlimeAnalysisTable.R

@@ -1,7 +1,8 @@
 
 DAIlimeAnalysisTable <- function(DAI_lime_analysis,decline_indexes){
   cases <- tibble(case = 1:5 %>% as.character(),
-                  name = c('Neutral','Moderate AOD','Severe AOD','Moderate COD','Severe COD'),
+                  syndrome = c('',rep('AOD',2),rep('COD',2)),
+                  name = c('Neutral','Moderate','Severe','Moderate','Severe'),
                   DAI = c(decline_indexes %>% filter(abs(DAI) == min(abs(DAI))) %>% .$DAI,
                           decline_indexes %>% filter(abs(DAI - 0.5) == min(abs(DAI - 0.5))) %>% .$DAI,
                           decline_indexes %>% filter(DAI == max(DAI)) %>% .$DAI,
@@ -11,7 +12,7 @@ DAIlimeAnalysisTable <- function(DAI_lime_analysis,decline_indexes){
   
   DAI_lime_analysis %>%
     left_join(cases,by = 'case') %>%
-    select(Status = name,DAI,`Predicted DAI` = prediction,Descriptor = feature,Value = feature_value,`Descriptor (\\textit{d}) range` = feature_desc,Weight = feature_weight) %>%
+    select(Syndrome = syndrome,Status = name,DAI,`Predicted DAI` = prediction,Descriptor = feature,Value = feature_value,`Descriptor (\\textit{d}) range` = feature_desc,Weight = feature_weight) %>%
     {
       for (i in unique(.$Descriptor)) {
         .$`Descriptor (\\textit{d}) range` <- str_replace_all(.$`Descriptor (\\textit{d}) range`,coll(i),'\\textit{d}')

### 2019-12-14 manuscript/manuscript.Rmd
fixed feature importance plots

@@
 Abiotic factors can include both predisposing and inciting factors such as drought and are often
-Biotic agents can include insect infestation such as the larvae of bark boring bettle *Agrilus biguttatus* associated
+Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* associated
 
@@
 
-It is becoming increasingly important that oak tree health within woodlands is accurately and sensitively monitoried to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices (ref).
+It is becoming increasingly important that oak tree health within woodlands is accurately and sensitively monitored to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices (ref).
 Monitoring methods need to take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.
@@
 Phenotypic data were collected from a total of 174 trees from nine sites from across the south, east and west of England between June and September in 2016 or 2017.
-The occurance of AOD has previously been reported and studied at six of the sites [@brown2016; @brown2017].
+The occurrence of AOD has previously been reported and studied at six of the sites [@brown2016; @brown2017].
 
@@
 
-**Crown production efficiency** - Estimate of crown production effieceny as a ratio between the composite crown surface area and the composite crown volumne [@zarnoch2004].
+**Crown production efficiency** - Estimate of crown production efficiency as a ratio between the composite crown surface area and the composite crown volume [@zarnoch2004].
 
@@
 This was then transformed to a dissimilarity matrix ($1 - P$) and scaled to two dimensions using  multidimensional scaling (MDS).
-The first MDS dimension was min-max scaled to give a scale between 0 and 1 with more severely declining trees having a score closer to 1.
+The first MDS dimension ($D$) was first inverted ($1 - D$) to ensure the correct dimension orientation relative to decline severity.
+It was then min-max scaled to give a scale between 0 and 1 with more severely declining trees having a score closer to 1.
 This dimension was named the PDI.
-The second MDS dimension ($D$) was first inverted ($1 - D$) to ensure the correct dimension orientation relative to decline type.
+The second MDS dimension was similarly inverted to ensure the correct dimension orientation relative to decline type.
 The dimension was then min-max scaled and then transformed ($2D - 1$) to give a scale between -1 and 1 with COD trees having a score closer to -1 and AOD having a score closer to 1. 
@@
 
+- descriptor importance for each index
 - simulated descriptor response surfaces
@@
 
-Local Interpretable Model-Agnostic Explanations (LIME) To understand the descriptors contributing to case specific examples of each of the indexes,  were computed for a range of specific cases [@ribeiro2016].
+Overall descriptor importance contributions to each of the indexes was retrieved from these models using gini impurity (ref).
+The assoication of key raw descriptors identified from the overall importance were investigated by simulating
+
+Local Interpretable Model-Agnostic Explanations (LIME) were computed to identify descriptors contributing locally to case specific examples of each of the indexes [@ribeiro2016]. 
 The lime R package v`r packageVersion('lime') %>% as.character()` was used with the default parameters [@pedersen2019].
-A low, a moderate and a severe decline cases was used for the PDI. For each case, a tree with a score closest to 0, 0.5 and 1 respectively was selected.
-For the DAI, a decline neautral case was selected with a score closes to 0.
-Then for each of the AOD and COD syndrome types a moderatefor the DAI .
+A low, a moderate and a severe decline cases was used for the PDI. 
+For each case, a tree with a score closest to 0, 0.5 and 1 respectively was selected.
+For the DAI, a decline neutral case was selected with a score closes to 0.
+Then for each of the AOD and COD syndromes, moderate and severe cases were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD.
 
@@
 Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.
-A significant difference was found in the PDI between non-symptomatic (control) and symptomatic (AOD and COD) decline statuses using an independant two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
+A significant difference was found in the PDI between non-symptomatic (control) and symptomatic (AOD and COD) decline statuses using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
 Significance was also found in the DAI between the AOD and COD groups (`r glue("t({DAI_groups_ttest$parameter}) = {signif(DAI_groups_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(DAI_groups_ttest$estimate1,digits = 3)` and `r signif(DAI_groups_ttest$estimate2,digits = 3)` for the AOD and COD groups respectively.
@@
 
-- Accuracies of retrained models
+- Trends of key phenotypic descriptors withing the PDI and DAI measures
+- Agrilus exit hole density had no contribution
 - Lime results
@@
 Crown condition descriptors such as crown volume and crown condition contributed highly to the PDI. 
-The phenotypic measures central to this were percentage missing crown and percentage crowntransparency from which crown condition and crown volume are calculated. 
+The phenotypic measures central to this were percentage missing crown and percentage crown transparency from which crown condition and crown volume are calculated. 
 It can be seen in Figure 1 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and estimated crown volume decrease, the PDI increases. 
@@
 
+- reasons for selecting machine learing approach
+- use of unsupervised over supervised analysis
+- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
+- For given context, descriptors not contributing to the indexes could be removed
+
 - Accessability of collecting descriptors
@@
 - Concession of model interpretability to predictive capability by using a machine learning approach
-- Trends of key phenotypic descriptors withing the PDI and DAI measures
-- Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
 - Index thresholds for category definitions (healthy, declining, severe decline etc.)
-- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
 - Scalability of method for wider application
@@
 - Validation of decline scores - chicken and paradox
-- Refined definition of remission
-- use of unsupervised over supervised analysis
-- Scale inversion likely need with new data
 
@@
 - Sensitivity for integration with molecular data compared to categorical assignments
+- Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
+- Refined definition of remission - repeat measurement
 

### 2019-12-14 R/functions/calcDIs.R

@@ -4,6 +4,7 @@ calcDIs <- function(rfModels,phenoData){
     mds() %>%
     rename(PDI = `Dimension 1`,DAI = `Dimension 2`) %>%
     mutate(PDI = PDI %>%
+             {. * -1} %>%
              minMaxScale(),
            DAI = DAI %>%
              {. * -1} %>%

### 2019-12-14 R/functions/descriptorImportancePlots.R

@@ -6,29 +6,30 @@ descriptorImportancePlots <- function(PDIrf,DAIrf){
         {
           imp <- importance(.)
           feat <- rownames(imp)
-          imp <- imp[,1]
-          tibble(Feature = feat,`Gini Importance` = imp)
+          tibble(Feature = feat,`%IncMSE` = imp[,'%IncMSE'],
+                 IncNodePurity = imp[,'IncNodePurity'])
         } %>%
-        arrange(`Gini Importance`) %>%
-        mutate(Feature = factor(Feature,levels = Feature), `Relative Importance` = `Gini Importance` / max(`Gini Importance`)) 
+        arrange(`%IncMSE`) %>%
+        mutate(Feature = factor(Feature,levels = Feature), 
+               `Relative Importance` = `%IncMSE` / max(`%IncMSE`)) 
       
       descriptorLabels <- dat$Feature %>%
         as.character() %>%
-        {c(.[1:(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) - 1)],
-           expression(Agrillus~exit~hole~density ( m^-2 ) ),
-           .[(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) + 1):36],
-           expression(Crown~surface~area ( m^2 )),
-           expression(Crown~volume ( m^3 ) )
-        )}
+      {c(.[1:(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) - 1)],
+         expression(Agrillus~exit~hole~density ( m^-2 ) ),
+         .[(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) + 1):36],
+         expression(Crown~volume ( m^3 ) ),
+         expression(Crown~surface~area ( m^2 ))
+      )}
       
-      ggplot(dat,aes(x = Feature,y = `Relative Importance`)) +
+      ggplot(dat,aes(x = Feature,y = `%IncMSE`)) +
         geom_point(fill = ptol_pal()(1),shape = 21) +
         coord_flip() +
         theme_bw() +
         theme(plot.title = element_text(face = 'bold'),
               axis.title = element_text(face = 'bold',size = 10)) +
         labs(title = 'a) PDI',
-             y = 'Relative Gini\nimportance',
+             y = 'Mean decrease in\naccuracy',
              x = NULL) +
         scale_x_discrete(labels = descriptorLabels)
     },
[…]

### 2019-12-14 manuscript/manuscript.Rmd
added to descriptor simulation section

@@
       PDI_rf_tune_params,
-      DAI_rf_tune_params)
+      DAI_rf_tune_params,
+      site_corrected_analysis_suitable_data)
 ```
@@
 
-R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and subsequent analysis of phenotypic data [@R2019].
+R version `r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and subsequent analysis of phenotypic data [@R2019].
 
@@
 
-A total of 31 raw descriptors collected
+After preparation there were a total of 31 raw descriptors for analysis.
 
@@
 
-### Identifying important descriptors and their trends within the decline indexes
-
-- descriptor importance for each index
-- simulated descriptor response surfaces
+### Descriptor contribution decline indexes
 
@@
 
-Overall descriptor importance contributions to each of the indexes was retrieved from these models using gini impurity (ref).
-The assoication of key raw descriptors identified from the overall importance were investigated by simulating
+Overall descriptor importance contributions to each of the indexes was retrieved from these models using the mean decrease in accuracy measure (ref).
+
+### Index model simulations to investigate interactions of key descriptors
+
+The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model responses surfaces for specific cases.
+This approach was similar to response surface methodology without the need for probabalistic approximation due to the low computation time required for index calculation (ref).
+
+For the PDI, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severly tree.
+The example cases were identified by selecting the descriptor values for trees with PDI values closes to the 25, 50 and 75% quantiles for the healthy, moderate and severely declining trees respectively.
+The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors divided into 100 increments.
+For a given case, each combination of theses simulated descriptor values were used to calculate a PDI value from the supervised random forest regression model, building a model response surface.
+
+A similar process was used for the DAI using also a comparison between active bleed size and black staining size as well as missing crown and crown transparency.
+Cases included neutral, moderate and severe AOD and COD decline types.
+
+- reduced by a factor of 16 for clearer visualisation
+
+### Local descriptor contribution to index models
 
@@
 The lime R package v`r packageVersion('lime') %>% as.character()` was used with the default parameters [@pedersen2019].
-A low, a moderate and a severe decline cases was used for the PDI. 
+A healthy, a moderate and a severe decline cases was used for the PDI. 
 For each case, a tree with a score closest to 0, 0.5 and 1 respectively was selected.

### 2019-12-14 R/functions/PDIresponseSurfaces.R

@@ -33,22 +33,6 @@ PDIresponseSurfaces <- function(PDIrf,decline_indexes,site_corrected_analysis_su
     set_names(variables) %>%
     bind_rows(.id = 'Descriptor')
   
-  # spectrumTrees_PDI$`a)`$`Crown radius (m)` <- ranges$max[ranges$Descriptor == 'Crown radius (m)']
-  # spectrumTrees_PDI$`b)`$`Crown radius (m)` <- ranges$center[ranges$Descriptor == 'Crown radius (m)']
-  # spectrumTrees_PDI$`c)`$`Crown radius (m)` <- ranges$min[ranges$Descriptor == 'Crown radius (m)']
-  # 
-  # spectrumTrees_PDI$`a)`$`Total height (m)` <- ranges$max[ranges$Descriptor == 'Total height (m)']
-  # spectrumTrees_PDI$`b)`$`Total height (m)` <- ranges$center[ranges$Descriptor == 'Total height (m)']
-  # spectrumTrees_PDI$`c)`$`Total height (m)` <- ranges$min[ranges$Descriptor == 'Total height (m)']
-  # 
-  # spectrumTrees_PDI$`a)`$`Lower crown height (m)` <- ranges$max[ranges$Descriptor == 'Lower crown height (m)']
-  # spectrumTrees_PDI$`b)`$`Lower crown height (m)` <- ranges$center[ranges$Descriptor == 'Lower crown height (m)']
-  # spectrumTrees_PDI$`c)`$`Lower crown height (m)` <- ranges$min[ranges$Descriptor == 'Lower crown height (m)']
-  # 
-  # spectrumTrees_PDI$`a)`$`Diameter at breast height (mm)` <- ranges$max[ranges$Descriptor == 'Diameter at breast height (mm)']
-  # spectrumTrees_PDI$`b)`$`Diameter at breast height (mm)` <- ranges$center[ranges$Descriptor == 'Diameter at breast height (mm)']
-  # spectrumTrees_PDI$`c)`$`Diameter at breast height (mm)` <- ranges$min[ranges$Descriptor == 'Diameter at breast height (mm)']
-  
   plotRanges_PDI <- spectrumTrees_PDI %>%
     names() %>%
     map(~{

### 2019-12-14 R/functions/descriptorImportancePlots.R

@@ -23,7 +23,7 @@ descriptorImportancePlots <- function(PDIrf,DAIrf){
       )}
       
       ggplot(dat,aes(x = Feature,y = `%IncMSE`)) +
-        geom_point(fill = ptol_pal()(1),shape = 21) +
+        geom_point(fill = ptol_pal()(1),shape = 21,size = 2) +
         coord_flip() +
         theme_bw() +
         theme(plot.title = element_text(face = 'bold'),
@@ -55,7 +55,7 @@ descriptorImportancePlots <- function(PDIrf,DAIrf){
         )}
       
       ggplot(dat,aes(x = Feature,y = `%IncMSE`)) +
-        geom_point(fill = ptol_pal()(1),shape = 21) +
+        geom_point(fill = ptol_pal()(1),shape = 21,size = 2) +
         coord_flip() +
         theme_bw() +
         theme(plot.title = element_text(face = 'bold'),

### 2019-12-14 manuscript/manuscript.Rmd
DAI response surface method added

@@
 
-A similar process was used for the DAI using also a comparison between active bleed size and black staining size as well as missing crown and crown transparency.
-Cases included neutral, moderate and severe AOD and COD decline types.
+```{r stem_symptom_ranges}
+symptom_ranges <- site_corrected_analysis_suitable_data %>%
+  select(`Active bleed size (mm)`,`Black staining size (mm)`) %>%
+  gather(descriptor,value) %>%
+  group_by(descriptor) %>%
+  summarise(minimum = min(value),
+             maximum = max(value),
+             center = minimum + (maximum - minimum) / 2
+             )
+```
 
-- reduced by a factor of 16 for clearer visualisation
+The same approach was used for the DAI using also a comparison between active bleed size and black staining size () as well as missing crown and crown transparency.
+The upper boundary for the ranges of active bleed size and black stain size was taken as the center value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.
+Cases included neutral, moderate and severe with DAI scores closest to 0, 0.5 and 1 respectively for AOD and 0, -0.5 and -1 respectively for COD.
 

### 2019-12-14 manuscript/manuscript.Rmd
added to phenotyping section

@@
 A popular algorithm is that of random forest due to it's versatility, relatively trivial training requirements and resilience against over fitting [@breiman2001] (refs).
+
+- Random forest has previously been used in a forest condition context [@vitale2014].
+
 It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag sample prediction accuracies and variable importance.
@@
 
-- Site types
-- Field phenotyping protocol
-- Crown assessments
-- Other assessments
-- Descriptions of collected phenotypic descriptors
-
-Phenotypic data were collected from a total of 174 trees from nine sites from across the south, east and west of England between June and September in 2016 or 2017.
+The phenotypic data were collected from a total of 174 trees from nine sites from across the south (Hatchlands and Richmond Park), east (Big Wood, Great Monks Wood and Winding Wood) and west (Attingham, Chestnuts, Langdale and Speculation) of England between June and September in 2016 or 2017.
 The occurrence of AOD has previously been reported and studied at six of the sites [@brown2016; @brown2017].
+These sites included a variety of woodland types, management types, soil types and tree sizes/ages.
 
 Descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
+All trees were manually assigned a decline status category; control, AOD, COD or remission prior to collection of phenotypic descriptors.
+
+- 12 figure grid reference for each tree
+- DBH to nearest 0.1cm
+- Total height to nearest 0.1m
+- Timber height - either 7cm top diameter or where no main stem can be distinguished
+- Lowest live crown to nearest 0.1m
+- Social class
+
 Crown radius, canopy closure and tap tests for dead stem tissue were measured for each cardinal point.
-The crown radius was
+The crown radius was measured using a vertex transponder attached to the relevant side of the stem at a height of 1.3m.
+
+- Canopy closure recorded as yes or no depending on whether any other crowns are touching the sampling tree.
+- High resolution photo collected from the vase to the spring of the crown on each cardinal point
+- High resolution photo of stem from the base to a height of 3m
+
+- Canopy closure percentage
+- Presence or absence of pruning damage or branch loss from main stem
+- % missing crown excluded from crown density assessment
+- absolute crown density in 5% classes
+- dieback type
+- main crown dieback location
+- stem epicormic growth
+- branch epicormic growth
+- insect damage on leaves and type
+- crown mildew
+
 The AOD stem symptoms (active bleeds, black staining, callused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem (ref).
@@
 These descriptors are shown in Supplementary table #.
-All trees were manually assigned a decline status category; control, AOD, COD or remission.
-All raw data and analysis code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.
 
@@
 
-- Descriptors converted to presence and absence
-- Aggregation of cardinal observations.
-- Symptom type sizes average and frequency taken
-- DBH converted to m and symptom sizes to mm
+R version `r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and subsequent analysis of phenotypic data [@R2019].
+All raw data, data preparation routines and analysis code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.
 
-R version `r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and subsequent analysis of phenotypic data [@R2019].
-
-Cardinal assessments were averaged to give a single value.
-
+All cardinal assessments were averaged to give a single value per tree.
+The crown density ($CD$) values erranously collected at Attingham and Langdale were converted to crown transparency ($t = 100 - CD$)
+All descriptors collected in cm were converted to mm.
+The tap tests that were positive for hollow stem tissue were converted to present and absent renamed dead stem tissue.
+The canopy closure descriptor was converted from yes and now to present and absent.
+Finally, the stem symptoms sizes for active bleeds, black staining and callused wounds were averaged and a frequency per tree collected.
 After preparation there were a total of 31 raw descriptors for analysis.

### 2019-12-14 manuscript/manuscript.Rmd
removed crown surface area and crown production efficiency

@@
       DAI_rf_tune_params,
+      analysis_suitable_data,
       site_corrected_analysis_suitable_data)
@@
 
-Seven additional, composite descriptors were calculated to account for the three dimensional attributes of the tree and covariate nature of certain descriptors as suggested by @zarnoch2004.
-This then gave a total of 38 raw and composite phenotypic descriptors.
+Five additional, composite descriptors were calculated to account for the three dimensional attributes of the tree and covariate nature of certain descriptors as suggested by @zarnoch2004.
+This then gave a total of `r ncol(analysis_suitable_data)` raw and composite phenotypic descriptors.
 The additional descriptors are defined as follows:
@@
 $r$ = crown radius (m)
-
-**Composite crown surface area (m^2^)** - Estimated crown surface area proportional to crown condition. Adapted from @laroque1994.
-
-$$S = \frac{4\pi(h-l)}{3r^2} \left[ \left( r^2 + \frac{r^4}{4(h-l)^2} \right) ^{1.5} - \left( \frac{r^4}{4(h-l)^2} \right) ^{1.5} \right] \times c $$
-
-**Crown production efficiency** - Estimate of crown production efficiency as a ratio between the composite crown surface area and the composite crown volume [@zarnoch2004].
-
-$$e = \frac{S}{v}$$
 

### 2019-12-14 manuscript/manuscript.Rmd
fixed descriptor importance plots

@@
 This dimension was named the PDI.
-The second MDS dimension was similarly inverted to ensure the correct dimension orientation relative to decline type.
-The dimension was then min-max scaled and then transformed ($2D - 1$) to give a scale between -1 and 1 with COD trees having a score closer to -1 and AOD having a score closer to 1. 
+The second MDS dimension was min-max scaled and then transformed ($2D - 1$) to give a scale between -1 and 1 with COD trees having a score closer to -1 and AOD having a score closer to 1. 
 This dimension was named the DAI.

### 2019-12-14 R/functions/calcAdditionalDescriptors.R

@@ -12,12 +12,6 @@ calcAdditionalDescriptors <- function(phenoData){
                                               `Total height (m)`,
                                               `Lower crown height (m)`,
                                               `Crown condition (%)`),
-           # `Crown surface area (m^2)` = crownSurfaceArea(`Crown radius (m)`,
-           #                                               `Total height (m)`,
-           #                                               `Lower crown height (m)`,
-           #                                               `Crown condition (%)`),
-           # `Crown production efficiency` = crownProductionEfficiency(`Crown surface area (m^2)`,
-           #                                                            `Crown volume (m^3)`),
            `Bleed prevalence (%)` = bleedPrevalence(`Active bleed size (mm)`,
                                                     `Active bleeds`,
                                                     `Black staining size (mm)`,

### 2019-12-14 R/functions/descriptorImportancePlots.R

@@ -14,13 +14,12 @@ descriptorImportancePlots <- function(PDIrf,DAIrf){
                `Relative Importance` = `%IncMSE` / max(`%IncMSE`)) 
       
       descriptorLabels <- dat$Feature %>%
-        as.character() #%>%
-      # {c(.[1:(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) - 1)],
-      #    expression(Agrillus~exit~hole~density ( m^-2 ) ),
-      #    .[(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) + 1):36],
-      #    expression(Crown~volume ( m^3 ) ),
-      #    expression(Crown~surface~area ( m^2 ))
-      # )}
+        as.character() %>%
+        {
+          .[str_detect(.,coll('Agrilus exit hole density (m^-2)'))] <- expression(Agrillus~exit~hole~density ( m^-2 ) )
+          .[str_detect(.,coll('Crown volume (m^3)'))] <- expression(Crown~volume ( m^3 ) )
+          return(.)
+        }
       
       ggplot(dat,aes(x = Feature,y = `%IncMSE`)) +
         geom_point(fill = ptol_pal()(1),shape = 21,size = 2) +
@@ -46,13 +45,12 @@ descriptorImportancePlots <- function(PDIrf,DAIrf){
                `Relative Importance` = `%IncMSE` / max(`%IncMSE`)) 
       
       descriptorLabels <- dat$Feature %>%
-        as.character() #%>%
-        # {c(.[1:(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) - 1)],
-        #    expression(Agrillus~exit~hole~density ( m^-2 ) ),
-        #    .[(which(str_detect(.,coll('Agrilus exit hole density (m^-2)'))) + 1):(which(str_detect(.,coll('Crown surface area (m^2)'))) - 1)],
-        #    expression(Crown~surface~area ( m^2 )),
-        #    expression(Crown~volume ( m^3 ) )
-        # )}
+        as.character() %>%
+        {
+          .[str_detect(.,coll('Agrilus exit hole density (m^-2)'))] <- expression(Agrillus~exit~hole~density ( m^-2 ) )
+          .[str_detect(.,coll('Crown volume (m^3)'))] <- expression(Crown~volume ( m^3 ) )
+          return(.)
+        }
       
[…]

### 2019-12-14 R/functions/siteDifferencesRFplot.R

@@ -31,16 +31,12 @@ siteDifferencesRFplot <- function(site_differences_rf,phenoData){
     mutate(Descriptor = factor(Descriptor,levels = Descriptor))
   
   descriptorLabels <- locationImportance$Descriptor %>%
-    as.character() #%>%
-    # {c(.[1:2],
-    #    expression(Agrillus~exit~hole~density ( m^-2 ) ),
-    #    .[4:16],
-    #    expression(Crown~surface~area ( m^2 )),
-    #    .[18:22],
-    #    expression(Crown~volume ( m^3 ) ),
-    #    .[24:38]
-    #   )
-    # }
+    as.character() %>%
+    {
+      .[str_detect(.,coll('Agrilus exit hole density (m^-2)'))] <- expression(Agrillus~exit~hole~density ( m^-2 ) )
+      .[str_detect(.,coll('Crown volume (m^3)'))] <- expression(Crown~volume ( m^3 ) )
+      return(.)
+    }
   
   site_differences <- list(
     a = ggplot() +

### 2019-12-14 manuscript/manuscript.Rmd
fixed DAI response surface figure

@@
 
-The same approach was used for the DAI using also a comparison between active bleed size and black staining size () as well as missing crown and crown transparency.
+The same approach was used for the DAI using also a comparison between active bleed size and black staining size as well as missing crown and crown transparency.
 The upper boundary for the ranges of active bleed size and black stain size was taken as the center value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.
-Cases included neutral, moderate and severe with DAI scores closest to 0, 0.5 and 1 respectively for AOD and 0, -0.5 and -1 respectively for COD.
+Cases included neutral, moderate and severe with PDI scores greater than 0.5 and DAI scores closest to 0, 0.5 and 1 respectively for AOD and 0, -0.5 and -1 respectively for COD.
 

### 2019-12-14 R/functions/DAIresponseSurfaces.R

@@ -1,13 +1,26 @@
 
 DAIresponseSurfaces <- function(DAIrf,decline_indexes,site_corrected_analysis_suitable_data){
+  
+  dat <- site_corrected_analysis_suitable_data %>%
+    bind_cols(decline_indexes %>%
+                select(PDI,DAI)) %>%
+    filter(PDI > 0.5)
+  
   spectrumTrees_DAI <- list(
-    `a)` = site_corrected_analysis_suitable_data[which(abs(decline_indexes$DAI) == min(abs(decline_indexes$DAI))),],
-    `b)` = site_corrected_analysis_suitable_data[which(abs(decline_indexes$DAI - 0.5) == min(abs(decline_indexes$DAI - 0.5))),],
-    `c)` = site_corrected_analysis_suitable_data[which(decline_indexes$DAI == max(decline_indexes$DAI)),],
-    `d)` = site_corrected_analysis_suitable_data[which(abs(decline_indexes$DAI) == min(abs(decline_indexes$DAI))),],
-    `e)` = site_corrected_analysis_suitable_data[which(abs(decline_indexes$DAI + 0.5) == min(abs(decline_indexes$DAI + 0.5))),],
-    `f)` = site_corrected_analysis_suitable_data[which(decline_indexes$DAI == min(decline_indexes$DAI)),]
-  )
+    `a)` = dat %>%
+      filter(abs(DAI) == min(abs(DAI))),
+    `b)` = dat %>%
+      filter(abs(DAI - 0.5) == min(abs(DAI - 0.5))),
+    `c)` = dat %>%
+      filter(DAI == max(DAI)),
+    `d)` = dat %>%
+      filter(abs(DAI) == min(abs(DAI))),
+    `e)` = dat %>%
+      filter(abs(DAI + 0.5) == min(abs(DAI + 0.5))),
+    `f)` = dat %>%
+      filter(DAI == min(DAI))
+  ) %>%
+    map(select,-PDI,-DAI)
   
   variables <- names(site_corrected_analysis_suitable_data)[map_lgl(site_corrected_analysis_suitable_data,is.numeric)]
   

### 2019-12-14 manuscript/manuscript.Rmd
finished methods section

@@
 All trees were manually assigned a decline status category; control, AOD, COD or remission prior to collection of phenotypic descriptors.
+Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
+Total height, timber hight and lowest live crown were measured using an inclinometer to the nearest 0.1m
+The timber height was measured at either 7cm top stem diameter or where no main stem could be distinguished.
+For the crown, canopy closure, missing crown and absolute crown density percentages were visually estimated within 5% classes
 
-- 12 figure grid reference for each tree
-- DBH to nearest 0.1cm
-- Total height to nearest 0.1m
-- Timber height - either 7cm top diameter or where no main stem can be distinguished
-- Lowest live crown to nearest 0.1m
-- Social class
+Other categorical descriptors included social class, diaback type, main crown dieback location, stem epicormic growth, branch epicormic growth, extent of insect damage, insect damage type, extent of crown mildew and the presence or absence of pruning damage or branch loss from the main stem.
+See supplementary table # for category details.
 
 Crown radius, canopy closure and tap tests for dead stem tissue were measured for each cardinal point.
-The crown radius was measured using a vertex transponder attached to the relevant side of the stem at a height of 1.3m.
-
-- Canopy closure recorded as yes or no depending on whether any other crowns are touching the sampling tree.
-- High resolution photo collected from the vase to the spring of the crown on each cardinal point
-- High resolution photo of stem from the base to a height of 3m
-
-- Canopy closure percentage
-- Presence or absence of pruning damage or branch loss from main stem
-- % missing crown excluded from crown density assessment
-- absolute crown density in 5% classes
-- dieback type
-- main crown dieback location
-- stem epicormic growth
-- branch epicormic growth
-- insect damage on leaves and type
-- crown mildew
+The crown radius was measured using a vertex transponder attached to the relevant side of the stem at a height of 1.3m.Canopy closure recorded as yes or no depending on whether any other crowns are touching the sampling tree.
+A high resolution photo was taken of the vase to the spring of the crown and of the stem from the base to a height of 3m for each cardinal point for later reference.
 

### 2019-12-14 manuscript/manuscript.Rmd
Finished site differences results section

@@
 
-Large phenotypic differences were found between the sites by supervised random forest (Supplementary  figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
-This was as a results of differences in descriptors relating tree size and likely the ages of the trees between the sites. 
-This adjustment removed inter-site variability but maintained intra-site variability, allowing comparisons to be made between the trees at the different sites in the context of decline severity.
-Post adjustment the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
+High discrimination was found in the phenotypic descriptors between between the sites using supervised random forest classification (Supplementary  figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`).
+The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density acrosses the nine sites (Supplementary figure #b).
+After the trees size adjustment was applied to removed inter-site variability but maintain intra-site variability, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
+This enabled comparisons to be made between the trees at the different sites in the context of decline severity.
 

### 2019-12-14 manuscript/manuscript.Rmd
group decline index section finished

@@
       analysis_suitable_data,
-      site_corrected_analysis_suitable_data)
+      site_corrected_analysis_suitable_data,
+      remission_PDI_group_ttest,
+      remission_DAI_group_ttest)
 ```
@@
 All trees were manually assigned a decline status category; control, AOD, COD or remission prior to collection of phenotypic descriptors.
+The remission group included trees that had shown stem bleeds associatied with AOD in the past but were not showing signs of stem bleeds at the time of assessment.
+
 Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
@@
 
-### Oak decline indexes
+### Oak decline severity indexes of manually assigned decline statuses
 
-Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.
-A significant difference was found in the PDI between non-symptomatic (control) and symptomatic (AOD and COD) decline statuses using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
-Significance was also found in the DAI between the AOD and COD groups (`r glue("t({DAI_groups_ttest$parameter}) = {signif(DAI_groups_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(DAI_groups_ttest$estimate1,digits = 3)` and `r signif(DAI_groups_ttest$estimate2,digits = 3)` for the AOD and COD groups respectively.
-A scatter plot of PDI against DAI is shown in Figure # for all 174 trees.
-It can clearly be seen that the
+A scatter plot of the calculated PDI and DAI scores is shown in Figure # for all 174 trees for which phenotypic descriptor data was collected.
+It can clearly be seen that the control trees group towards the lower end of the PDI scale and the sympotmatic COD and AOD groups towards the higher end.
+This gave a significant difference when PDI of the control and symptomatic (AOD and COD) decline statuses were compared using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
 
-### Descriptor contributions and trends to the decline indexes
+On the DAI scale, the AOD trees group at the upper end of the scale (> 0) and the COD trees at the lower end (< 0).
+This was confirmed with significance found in the DAI between these groups (`r glue("t({DAI_groups_ttest$parameter}) = {signif(DAI_groups_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(DAI_groups_ttest$estimate1,digits = 3)` and `r signif(DAI_groups_ttest$estimate2,digits = 3)` for the AOD and COD groups respectively.
 
+The remission trees were found to be significantly different in their PDI scores from both the control (`r glue("t({remission_PDI_group_ttest$parameter[1]}) = {signif(remission_PDI_group_ttest$statistic[1],digits = 3) %>% abs()}, p < 0.001")`) and symptomatic (AOD and COD; `r glue("t({remission_PDI_group_ttest$parameter[2]}) = {signif(remission_PDI_group_ttest$statistic[2],digits = 3) %>% abs()}, p < 0.001")`) groups with means of `r signif(remission_PDI_group_ttest$estimate1[1],digits = 3)`, `r signif(remission_PDI_group_ttest$estimate2[1],digits = 3)` and `r signif(remission_PDI_group_ttest$estimate2[2],digits = 3)` for the control, remission and symptomatic groups respectively.
+As a group these trees are showing moderate overall decline.
+
+The remission trees were also found to be significantly different to the AOD (`r glue("t({remission_DAI_group_ttest$parameter[1]}) = {signif(remission_DAI_group_ttest$statistic[1],digits = 3) %>% abs()}, p = < 0.001")`) and COD (`r glue("t({remission_DAI_group_ttest$parameter[2]}) = {signif(remission_DAI_group_ttest$statistic[2],digits = 3) %>% abs()}, p = {round(remission_DAI_group_ttest$p.value[2],3)}")`) groups on the DAI scale with group means of `r signif(remission_DAI_group_ttest$estimate1[1],digits = 3)`, `r signif(remission_DAI_group_ttest$estimate2[1],digits = 3)` and `r signif(remission_DAI_group_ttest$estimate1[2],digits = 3)` for the AOD, remission and COD groups respectively.
+As a group the remission trees are of neutral decline type.
+
+### Overall descriptor contributions and trends to the decline indexes
+
+- Top ranked descriptors contributing to the decline indexes
 - Trends of key phenotypic descriptors withing the PDI and DAI measures
 - Agrilus exit hole density had no contribution
-- Lime results
 - Top ranked descriptor index response surfaces
 
-Crown related descriptors were found to contribute most to the PDI (Figure #) with more severely declining trees having reduced crown volume and condition. 
+Crown related descriptors were found to contribute most to the PDI (Figure #a) with more severely declining trees having reduced crown volume and condition. 
 The DAI was a more complex in descriptor contribution with trunk staining and bleeding descriptors contributing highly with bleed prevalence contributing the most. 
@@
 AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.
+
+### Localised descriptor contributions and trends within the PDI and DAI models
+
+- Lime results
 

### 2019-12-14 R/plan.R

@@ -44,7 +44,7 @@ plan <- drake_plan(
   ## site differences MDS and importance plots
   site_differences_mds_plot = site_differences_rf %>%
     siteDifferencesRFplot(pheno_data_with_additional_descriptors),
-    
+  
   ## calculate descriptor correction factors
   site_correction_factors = pheno_data_with_additional_descriptors %>%
     siteCorrectionFactors(),
@@ -95,6 +95,7 @@ plan <- drake_plan(
   
   ## t-test for PDI and status
   PDI_status_ttest = decline_indexes %>%
+    filter(ChosenGroup %in% c('Control','AOD','COD')) %>%
     t.test(PDI~Status,data = .,var.equal = TRUE) %>%
     tidy(),
   
@@ -104,6 +105,48 @@ plan <- drake_plan(
     group_by(ChosenGroup) %>% 
     summarise(median = median(DAI),
               mean = mean(DAI)),
+  
+  ## Remission PDI t-test
+  remission_PDI_group_ttest = decline_indexes %>%
+    filter(ChosenGroup %in% c('Control','AOD','COD','Remission')) %>%
+    {
+      d <- .
+      c_r <- d %>%
+        filter(ChosenGroup %in% c('Control','Remission')) %>%
+        t.test(PDI~ChosenGroup,data = .,var.equal = TRUE) %>%
+        tidy()
+      s_r <- d %>%
+        filter(ChosenGroup %in% c('Remission','AOD','COD')) %>%
+        mutate(new_status = ChosenGroup) %>%
+        {
+          .$new_status[.$ChosenGroup == 'AOD'] <- 'Symptomatic'
+          .$new_status[.$ChosenGroup == 'COD'] <- 'Symptomatic'
+          return(.)
+        } %>%
+        t.test(PDI~new_status,data = .,var.equal = TRUE) %>%
+        tidy()
+      list(`control~remission` = c_r,
+           `symptomatic~remission` = s_r) %>%
+        bind_rows(.id = 'comparison')
+    },
+  
+  ## Remission DAI ttest
+  remission_DAI_group_ttest = decline_indexes %>%
+    filter(ChosenGroup %in% c('AOD','COD','Remission')) %>%
+    {
+      d <- .
+      r_a <- d %>%
+        filter(ChosenGroup %in% c('AOD','Remission')) %>%
+        t.test(DAI~ChosenGroup,data = .,var.equal = TRUE) %>%
+        tidy()
+      r_c <- d %>%
[…]

### 2019-12-14 manuscript/manuscript.Rmd
update PDI response surfaces method. Re-arrange of methods and results sections

@@
 
-### Descriptor contribution decline indexes
+### Descriptor contribution to decline indexes
 
@@
 
+### Local descriptor contribution to index models
+
+Local Interpretable Model-Agnostic Explanations (LIME) were computed to identify descriptors contributing locally to case specific examples of each of the indexes [@ribeiro2016]. 
+The lime R package v`r packageVersion('lime') %>% as.character()` was used with the default parameters [@pedersen2019].
+A healthy, a moderate and a severe decline cases was used for the PDI. 
+For each case, a tree with a score closest to 0, 0.5 and 1 respectively was selected.
+For the DAI, a decline neutral case was selected with a score closes to 0.
+Then for each of the AOD and COD syndromes, moderate and severe cases were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD.
+
 ### Index model simulations to investigate interactions of key descriptors
@@
 For the PDI, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severly tree.
-The example cases were identified by selecting the descriptor values for trees with PDI values closes to the 25, 50 and 75% quantiles for the healthy, moderate and severely declining trees respectively.
+The example cases included the trees with the minimum and maximum PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
 The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors divided into 100 increments.
@@
 Cases included neutral, moderate and severe with PDI scores greater than 0.5 and DAI scores closest to 0, 0.5 and 1 respectively for AOD and 0, -0.5 and -1 respectively for COD.
-
-### Local descriptor contribution to index models
-
-Local Interpretable Model-Agnostic Explanations (LIME) were computed to identify descriptors contributing locally to case specific examples of each of the indexes [@ribeiro2016]. 
-The lime R package v`r packageVersion('lime') %>% as.character()` was used with the default parameters [@pedersen2019].
-A healthy, a moderate and a severe decline cases was used for the PDI. 
-For each case, a tree with a score closest to 0, 0.5 and 1 respectively was selected.
-For the DAI, a decline neutral case was selected with a score closes to 0.
-Then for each of the AOD and COD syndromes, moderate and severe cases were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD.
 
@@
 
-### Overall descriptor contributions and trends to the decline indexes
+### Key descriptors contributing to the decline indexes
 
 - Top ranked descriptors contributing to the decline indexes
-- Trends of key phenotypic descriptors withing the PDI and DAI measures
-- Agrilus exit hole density had no contribution
-- Top ranked descriptor index response surfaces
 
-Crown related descriptors were found to contribute most to the PDI (Figure #a) with more severely declining trees having reduced crown volume and condition. 
+The top ranked descriptors found to contr crown related descriptors were found to contribute most to the PDI (Figure #a) with more severely declining trees having reduced crown volume and condition.
+
+- Plots of the individual descriptors available
+
 The DAI was a more complex in descriptor contribution with trunk staining and bleeding descriptors contributing highly with bleed prevalence contributing the most. 
@@
 
-The phenotypic indexes of all 171 trees (including the 20 trees Chestnuts) are shown in Figure 3. 
-It can be seen that across the 9 sites, purely chronic oak decline sites such as Chestnuts and Speculation are easily distinguishable using the DAI compared to more AOD dominant sites such as Langdale or mixed sites such as Great Monks Wood.
+- Agrilus exit hole density had no contribution
 
-The Phenotypic Decline Index (PDI) and Decline Acuteness Index (DAI) were previously defined in the 30 month report. 
-A machine learning approach was used to derive these measures as it allows the complex interactions and trends between phenotypic descriptors to be modelled from which the decline indexes are calculated.
-This would likely be oversimplified using conventional modelling strategies. 
+### Localised descriptor contributions and trends within the PDI and DAI models
+
+- Lime results
+
+### Key descriptor interactions within the decline index models
+
+- Top ranked descriptor index response surfaces
+- Trends of key phenotypic descriptors within the PDI and DAI measures
+- Relatively smaller trees associated with more severe decline.
+
 These complex interactions can be investigated by simulating phenotypic data and generating the index values from random forest models that have been trained for each of the indexes, supervised using phenotypic descriptors from the 171 trees.
@@
 AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.
-
-### Localised descriptor contributions and trends within the PDI and DAI models
-
-- Lime results
 
@@
 The healthier trees at sites such as Attingham and Hatchlands were found to be in a more declined state than those at sites such as Langdale and Chestnuts.
+
+A machine learning approach was used to derive these measures as it allows the complex interactions and trends between phenotypic descriptors to be modelled from which the decline indexes are calculated.
+This would likely be oversimplified using conventional modelling strategies.
 

### 2019-12-15 manuscript/manuscript.Rmd
updated lime DAI method. Response surface results complete. Added points to lime results section

@@
 For each case, a tree with a score closest to 0, 0.5 and 1 respectively was selected.
-For the DAI, a decline neutral case was selected with a score closes to 0.
+For the DAI cases were selected from trees with a PDI score greater than 0.5, 
+A decline neutral case was selected with a score closest to 0.
 Then for each of the AOD and COD syndromes, moderate and severe cases were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD.
@@
 
-- Lime results
+- Important descriptors identified model-wide aren't necessarily relevant to localised cases.
+
+- Top descriptors for each case same as top descriptors in overall model importance lists for PDI.
+- Decline is associated with a reduction in the size and condition of the crown
+
+- Different descriptors locally compared to overall importance for DAI.
+- COD trees have greater crown contact than AOD trees.
+- Bleed prevalence not included in any local importances
+- Descriptors relevant to context
 
@@
 
-- Top ranked descriptor index response surfaces
-- Trends of key phenotypic descriptors within the PDI and DAI measures
-- Relatively smaller trees associated with more severe decline.
+The complex descriptor interactions within the index models can be investigated by simulating phenotypic data and generating the index values from the supervised index random forest regression models.
+The index values generated by the models from these simulated phenotypic data can then be visualised as response surface plots, from which descriptor interactions within the indexes can be interpreted.
+Figures # and # show the simulated index value response surfaces under a variety of different circumstances for key phenotypic descriptors that contribute highly to the PDI and DAI.
 
-These complex interactions can be investigated by simulating phenotypic data and generating the index values from random forest models that have been trained for each of the indexes, supervised using phenotypic descriptors from the 171 trees.
-The index values generated by the models from these simulated phenotypic data can then be visualised as response surface plots, from which descriptor interactions within the indexes can be interpreted.
-Figures 1 and 2 show the simulated index value response surfaces under a variety of different circumstances for key phenotypic descriptors that contribute highly to the PDI and DAI.
-
-Crown condition descriptors such as crown volume and crown condition contributed highly to the PDI. 
+Crown condition descriptors such as crown volume and crown condition contributed highly to the PDI (Figure #). 
 The phenotypic measures central to this were percentage missing crown and percentage crown transparency from which crown condition and crown volume are calculated. 
-It can be seen in Figure 1 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and estimated crown volume decrease, the PDI increases. 
+It can be seen in Figure # that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and estimated crown volume decrease, the PDI increases. 
 A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g.
-greater crown radius, total height, lower crown height and diameter at breast height; Figures 1a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.
+greater crown radius, total height, lower crown height and diameter at breast height; Figures #a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.
 
 The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, diameter at breast height and crown condition descriptors. 
-Simulated response surfaces for the DAI are shown in Figure 2. 
-These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 2a, b & c) and crown condition descriptors (Figures 2d, e & f) change for non-symptomatic AOD or COD trees (Figures 2a & d respectively), moderate AOD or COD (Figures 2b & e respectively) and severe AOD or COD (Figures 2c & f respectively).
+Simulated response surfaces for the DAI are shown in Figure c. 
+These response surfaces show circumstances under which bleed prevalence related descriptors (Figures #a, b & c) and crown condition descriptors (Figures dd, e & f) change for non-symptomatic AOD or COD trees (Figures #a & d respectively), moderate AOD or COD (Figures #b & e respectively) and severe AOD or COD (Figures #c & f respectively).
 Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
-Also higher DAI values could be obtained from trees with greater stature (Figure 2c). 
-Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 2d, e & f) and therefore more severe COD.
+Also higher DAI values could be obtained from trees with greater stature (Figure #c). 
+Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure #d).
+Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures #e & f) and therefore more severe COD.
 The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflict. 

### 2019-12-15 manuscript/manuscript.Rmd
finished descriptor contribution results section

@@
       remission_PDI_group_ttest,
-      remission_DAI_group_ttest)
+      remission_DAI_group_ttest,
+      DAI_descriptor_importance)
 ```
@@
 Machine learning is finding increasing utility in biological and geographical applications due to it's ability to learn complex patterns within and between variables.
-A popular algorithm is that of random forest due to it's versatility, relatively trivial training requirements and resilience against over fitting [@breiman2001] (refs).
+A popular algorithm is that of random forest due to it's versatility, relatively trivial training requirements and resilience against over fitting [@breiman2001a] (refs).
 
@@
 
-- Top ranked descriptors contributing to the decline indexes
+The top four ranked descriptors found to contribute to the PDI index model were all crown size and condition related descriptors with crown volume the top ranked descriptor overall (Figure #a).
+Other tree size related descriptors such as and stem bleeding descriptors such as contributed secondarily.
 
-The top ranked descriptors found to contr crown related descriptors were found to contribute most to the PDI (Figure #a) with more severely declining trees having reduced crown volume and condition.
+The DAI was more complex in descriptor contribution many more descriptors contributing more equally than the PDI (Figure #b).
+The overall tree size descriptors live crown ratio, DBH and crown volume contributed almost equally to the DAI index model.
+These were followed closely by the stem staining and bleeding descriptors such as bleed prevalence. 
+The presence of *A.
+biguttatus* exit holes had very little contribution to the DAI with exit hole density ranked only `r which(DAI_descriptor_importance$Feature == 'Agrilus exit hole density (m^-2)')` th.
 
-- Plots of the individual descriptors available
-
-The DAI was a more complex in descriptor contribution with trunk staining and bleeding descriptors contributing highly with bleed prevalence contributing the most. 
-Trees with high bleed prevalence and therefore a larger area of their trunk covered with black staining or active bleeds had much higher DAI scores.
-
-- Agrilus exit hole density had no contribution
+Plots of the individual descriptors against the decline indexes are available in Supplementary figures #-#. 
+The following sections identify how the descriptors contribute and interact to the decline index models in specific decline cases.
 
@@
 
-- Important descriptors identified model-wide aren't necessarily relevant to localised cases.
+The LIME eas used to identify theImportant descriptors identified model-wide aren't necessarily relevant to localised cases.
 
@@
 
-- reasons for selecting machine learing approach
-- use of unsupervised over supervised analysis
-- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
-- For given context, descriptors not contributing to the indexes could be removed
+### Suitablity of the decline indexes for describing decline severity and type
 
-- Accessability of collecting descriptors
-- Suitability of collected phenotypic descriptors for describing decline status
-- Concession of model interpretability to predictive capability by using a machine learning approach
+- Suitability of collected phenotypic descriptors and decline indexes for describing decline status
 - Index thresholds for category definitions (healthy, declining, severe decline etc.)
-- Scalability of method for wider application
-- Thresholds for decline groupings within index ranges
 - Validation of decline scores - chicken and paradox
+- Refined definition of remission - repeat measurement
 
@@
 Also notable from Figure 3 is the disparity between the relative health of the healthier trees across the sites. 
+The two dimensional space provided by these phenotypic indexes, for measuring decline severity is shown in Figure 1. 
 The healthier trees at sites such as Attingham and Hatchlands were found to be in a more declined state than those at sites such as Langdale and Chestnuts.
-
-A machine learning approach was used to derive these measures as it allows the complex interactions and trends between phenotypic descriptors to be modelled from which the decline indexes are calculated.
-This would likely be oversimplified using conventional modelling strategies.
-
-The two dimensional space provided by these phenotypic indexes, for measuring decline severity is shown in Figure 1. 
 The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
@@
 
-### Applications
+### The machine learning approach
 
+- reasons for selecting machine learing approach
+- use of unsupervised over supervised analysis
+- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
+- For given context, descriptors not contributing to the indexes could be removed
+- Concession of model interpretability to predictive capability by using a machine learning approach
+- seen as black box but LIME and response surfaces enable interpretation and therefore trust
+
+A machine learning approach was used to derive these measures as it allows the complex interactions and trends between phenotypic descriptors to be modelled from which the decline indexes are calculated.
+This would likely be oversimplified using conventional modelling strategies [@breiman2001b].
+
+### Applications and wider scaleablity
+
+- Scalability of method for wider application
+- Accessability of collecting descriptors
 - Discussion of applications to forestry monitoring and management
@@
 - Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
-- Refined definition of remission - repeat measurement
+- Refinement of protocol for large scale surveying
 

### 2019-12-15 R/plan.R

@@ -187,6 +187,9 @@ plan <- drake_plan(
                    mtry = PDI_rf_tune_params$mtry,
                    importance = TRUE)},
   
+  ## PDI descriptor importance
+  PDI_descriptor_importance = descriptorImportance(PDI_rf_model),
+  
   ## generate DAI predictive random forest model
   DAI_rf_model = site_corrected_analysis_suitable_data %>%
     {set.seed(1234)
@@ -195,8 +198,11 @@ plan <- drake_plan(
                    mtry = DAI_rf_tune_params$mtry,
                    importance = TRUE)},
   
+  ## DAI descriptor importance
+  DAI_descriptor_importance = descriptorImportance(DAI_rf_model),
+  
   ## create descriptor contribution plots
-  descriptor_contribution_plots = descriptorImportancePlots(PDI_rf_model,DAI_rf_model),
+  descriptor_contribution_plots = descriptorImportancePlots(PDI_descriptor_importance,DAI_descriptor_importance),
   
   ## PDI lime analysis
   PDI_lime_analysis = PDIlimeAnalysis(site_corrected_analysis_suitable_data,PDI_rf_model,decline_indexes),

### 2019-12-15 manuscript/manuscript.Rmd
added disscussion point

@@
 - seen as black box but LIME and response surfaces enable interpretation and therefore trust
+- LIME can be applied to out of place cases
 

### 2019-12-15 manuscript/supplementary.Rmd
increased size of numeric descriptor plots
finished descriptor table

@@
 phenotypic_descriptor_info %>%
+  select(-`Descriptor Type`,-`Data Type`) %>%
   kable(format = 'latex',booktabs = TRUE,longtable = TRUE) %>%
-  kable_styling(latex_options = 'repeat_header',repeat_header_continued = TRUE) %>%
-  column_spec(1,width = '1.5cm') %>%
-  column_spec(2,width = '3cm') %>%
-  column_spec(4:5,width = '7cm') %>%
-  collapse_rows(columns = 1,latex_hline = "major", valign = "middle") %>%
-  landscape()
+  kable_styling(latex_options = c('striped','repeat_header'),repeat_header_continued = TRUE) %>%
+  # column_spec(1,width = '1.5cm') %>%
+  column_spec(1,width = '3cm') %>%
+  column_spec(2:3,width = '6cm') #%>%
+  # collapse_rows(columns = 1,latex_hline = "major", valign = "middle")
 ```
@@
 
-```{r PDInumeric,fig.height=8}
+```{r PDInumeric,fig.height=9}
 loadd(descriptor_scatter_plots)
@@
 
-```{r DAInumeric,fig.height=8}
+```{r DAInumeric,fig.height=9}
 descriptor_scatter_plots$DAI_numeric

### 2019-12-15 manuscript/manuscript.Rmd
finished LIME results section

--- 2019-12-15 manuscript/manuscript.Rmd
@@
 High discrimination was found in the phenotypic descriptors between between the sites using supervised random forest classification (Supplementary  figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`).
-The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density acrosses the nine sites (Supplementary figure #b).
+The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary figure #b).
 After the trees size adjustment was applied to removed inter-site variability but maintain intra-site variability, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
@@
 The presence of *A.
-biguttatus* exit holes had very little contribution to the DAI with exit hole density ranked only `r which(DAI_descriptor_importance$Feature == 'Agrilus exit hole density (m^-2)')` th.
+biguttatus* exit holes had very little contribution to the DAI with exit hole density ranked only `r which(DAI_descriptor_importance$Feature == 'Agrilus exit hole density (m^-2)')`^th^.
 
@@
 
-The LIME eas used to identify theImportant descriptors identified model-wide aren't necessarily relevant to localised cases.
+LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to the PDI in each of the healthy, moderate and severe decline cases using lime (Table #).
+These descriptors were also the same top descriptors in the overall model importance lists for the PDI.
+Increasing decline severity was associated with a reduction in the size and condition of the crown.
+For instance, declining trees were found to have an estimated crown volume below 841m^3^.
+Healthy trees were found to have a crown condition greater than 71.3%, moderately declining trees between 47.6% and 71.3% and severly declining trees a condition below 23.9%.
 
-- Top descriptors for each case same as top descriptors in overall model importance lists for PDI.
-- Decline is associated with a reduction in the size and condition of the crown
-
-- Different descriptors locally compared to overall importance for DAI.
-- COD trees have greater crown contact than AOD trees.
-- Bleed prevalence not included in any local importances
-- Descriptors relevant to context
+For the DAI, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table #).
+This likely reflects the greater number of descriptors contributing to the DAI with different descriptors relevant to particular decline contexts.
+The important descriptors identified were mainly associated with tree size such as live crown ratio and lower crown radius.
+Bleed prevalence and stem symptom descriptors were not found to be important to any of the decline type cases.
+Live crown ratio greater than 61.6% in the severe AOD and moderate COD cases.
+COD trees were found to have greater crown contact (greater than 50%) compared to AOD trees.
 
@@
 - seen as black box but LIME and response surfaces enable interpretation and therefore trust
+- LIME can be applied to out of place cases
+- Important descriptors identified model-wide aren't necessarily relevant to localised cases.
 

### 2019-12-15 manuscript/manuscript.Rmd
couple of additions to discussion

@@
 The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
-They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations, which could have great utility in long term forest health monitoring for both forestry management and research.
+They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations.
 
@@
 
+which could have great utility in long term forest health monitoring for both forestry management and research.
+
 ## Conclusions

### 2019-12-15 manuscript/manuscript.Rmd
added a couple of disscussion points. Spell check

@@
 
-**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning, unsupervised learning, random forest
+**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning, unsupervised learning, random forest, decline spectrum
 
@@
 All trees were manually assigned a decline status category; control, AOD, COD or remission prior to collection of phenotypic descriptors.
-The remission group included trees that had shown stem bleeds associatied with AOD in the past but were not showing signs of stem bleeds at the time of assessment.
+The remission group included trees that had shown stem bleeds associated with AOD in the past but were not showing signs of stem bleeds at the time of assessment.
 
@@
 
-Other categorical descriptors included social class, diaback type, main crown dieback location, stem epicormic growth, branch epicormic growth, extent of insect damage, insect damage type, extent of crown mildew and the presence or absence of pruning damage or branch loss from the main stem.
+Other categorical descriptors included social class, dieback type, main crown dieback location, stem epicormic growth, branch epicormic growth, extent of insect damage, insect damage type, extent of crown mildew and the presence or absence of pruning damage or branch loss from the main stem.
 See supplementary table # for category details.
@@
 All cardinal assessments were averaged to give a single value per tree.
-The crown density ($CD$) values erranously collected at Attingham and Langdale were converted to crown transparency ($t = 100 - CD$)
+The crown density ($CD$) values erroneously collected at Attingham and Langdale were converted to crown transparency ($t = 100 - CD$)
 All descriptors collected in cm were converted to mm.
@@
 The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model responses surfaces for specific cases.
-This approach was similar to response surface methodology without the need for probabalistic approximation due to the low computation time required for index calculation (ref).
+This approach was similar to response surface methodology without the need for probabilistic approximation due to the low computation time required for index calculation (ref).
 
-For the PDI, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severly tree.
+For the PDI, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severely tree.
 The example cases included the trees with the minimum and maximum PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
@@
 The same approach was used for the DAI using also a comparison between active bleed size and black staining size as well as missing crown and crown transparency.
-The upper boundary for the ranges of active bleed size and black stain size was taken as the center value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.
+The upper boundary for the ranges of active bleed size and black stain size was taken as the centre value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.
 Cases included neutral, moderate and severe with PDI scores greater than 0.5 and DAI scores closest to 0, 0.5 and 1 respectively for AOD and 0, -0.5 and -1 respectively for COD.
@@
 A scatter plot of the calculated PDI and DAI scores is shown in Figure # for all 174 trees for which phenotypic descriptor data was collected.
-It can clearly be seen that the control trees group towards the lower end of the PDI scale and the sympotmatic COD and AOD groups towards the higher end.
+It can clearly be seen that the control trees group towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
 This gave a significant difference when PDI of the control and symptomatic (AOD and COD) decline statuses were compared using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
@@
 For instance, declining trees were found to have an estimated crown volume below 841m^3^.
-Healthy trees were found to have a crown condition greater than 71.3%, moderately declining trees between 47.6% and 71.3% and severly declining trees a condition below 23.9%.
+Healthy trees were found to have a crown condition greater than 71.3%, moderately declining trees between 47.6% and 71.3% and severely declining trees a condition below 23.9%.
 
@@
 Simulated response surfaces for the DAI are shown in Figure c. 
-These response surfaces show circumstances under which bleed prevalence related descriptors (Figures #a, b & c) and crown condition descriptors (Figures dd, e & f) change for non-symptomatic AOD or COD trees (Figures #a & d respectively), moderate AOD or COD (Figures #b & e respectively) and severe AOD or COD (Figures #c & f respectively).
+These response surfaces show circumstances under which bleed prevalence related descriptors (Figures #a, b & c) and crown condition descriptors (Figures #d, e & f) change for non-symptomatic AOD or COD trees (Figures #a & d respectively), moderate AOD or COD (Figures #b & e respectively) and severe AOD or COD (Figures #c & f respectively).
 Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
@@
 
-### Suitablity of the decline indexes for describing decline severity and type
+### Suitability of the decline indexes for describing decline severity and type
+
+The phenotypic descriptors were selected for use in this study purposefully
 
@@
 - Validation of decline scores - chicken and paradox
+- Direct validation difficult 
 - Refined definition of remission - repeat measurement
@@
 in a decline state.
-Most notably this has occurred in a number of trees at Attingham, Hatchlands and Winding Wood. 
-Also notable from Figure 3 is the disparity between the relative health of the healthier trees across the sites. 
-The two dimensional space provided by these phenotypic indexes, for measuring decline severity is shown in Figure 1. 
-The healthier trees at sites such as Attingham and Hatchlands were found to be in a more declined state than those at sites such as Langdale and Chestnuts.
-The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
-They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations.
+
+These two indexes provide a two dimensional continuum under which the spectrum of decline severity and type can be assessed.
+
+The two dimensional space provided by these phenotypic indexes, for measuring decline severity is shown in Figure #.
 
@@
 
-- reasons for selecting machine learing approach
+A machine learning approach was used to derive the phenotypic index measures as it allows the moulded to the complex interactions and trends between phenotypic descriptors from which the decline indexes are calculated.
+This would likely be oversimplified using conventional modelling strategies [@breiman2001b].
+
+- reasons for selecting machine learning approach
 - use of unsupervised over supervised analysis
@@
 - Concession of model interpretability to predictive capability by using a machine learning approach
+
+Machine learning approaches are usually seen as black box 
 - seen as black box but LIME and response surfaces enable interpretation and therefore trust
@@
 
-A machine learning approach was used to derive these measures as it allows the complex interactions and trends between phenotypic descriptors to be modelled from which the decline indexes are calculated.
-This would likely be oversimplified using conventional modelling strategies [@breiman2001b].
-
-### Applications and wider scaleablity
+### Applications and wider scalability
 
@@
 
+The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
+They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations.
 which could have great utility in long term forest health monitoring for both forestry management and research.

### 2019-12-15 manuscript/supplementary.Rmd
added table captions

@@
   select(-`Descriptor Type`,-`Data Type`) %>%
-  kable(format = 'latex',booktabs = TRUE,longtable = TRUE) %>%
+  kable(format = 'latex',
+        booktabs = TRUE,
+        longtable = TRUE,
+        caption = 'Phenotypic descriptors collected for oak decline severity.') %>%
   kable_styling(latex_options = c('striped','repeat_header'),repeat_header_continued = TRUE) %>%

### 2019-12-15 manuscript/manuscript.Rmd
added acknowledgements

@@
 Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
-Total height, timber hight and lowest live crown were measured using an inclinometer to the nearest 0.1m
+Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m
 The timber height was measured at either 7cm top stem diameter or where no main stem could be distinguished.
@@
 - Scalability of method for wider application
-- Accessability of collecting descriptors
+- Accessibility of collecting descriptors
 - Discussion of applications to forestry monitoring and management
@@
 
-\newpage
+## Acknowledgements
+
+We would like to thank the Forest Research TSU for collecting the phenotypic data presented in this article especially Liz Richardson and Mark Oram for coordinating surveying.
+This work was supported by Woodland Heritage and the Future Proofing Plant Health (FPPH) Oak Resilience project.
+
 ## Checks

### 2019-12-16 manuscript/manuscript.Rmd
converted DBH to m

@@
 
-Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
+DBH (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
 Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m
@@
 
-To account age, size and site specific tree difference, an adjustment of site means to the overall mean was applied to each of the size descriptors (diameter at breast height, lower crown height, timber height, total height, crown radius).
-An example of this adjustment is shown for diameter at breast height in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.
+To account age, size and site specific tree difference, an adjustment of site means to the overall mean was applied to each of the size descriptors (DBH, lower crown height, timber height, total height, crown radius).
+An example of this adjustment is shown for DBH in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.
 
@@
 
-$$p = \frac{a^2A + b^2B}{3d\pi}$$
+$$p = \frac{a^2A + b^2B}{3000d\pi} \times 100$$
 
-$a$ = average active bleed size (mm), $A$ = number of active bleeds, $b$ = average black stain size (mm), $B$ = black stain number, $d$ = diameter at breast height (mm)
+$a$ = average active bleed size (mm), $A$ = number of active bleeds, $b$ = average black stain size (mm), $B$ = black stain number, $d$ = $DBH (m) \times 1000$
 
@@
 
-$$e = \frac{h}{\frac{2d}{100}\pi}$$
+$$e = \frac{h}{2d\pi}$$
 
-$h$ = number of *Agrilus* exit holes
+$h$ = number of *Agrilus* exit holes, d = DBH (m)
 
@@
 A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g.
-greater crown radius, total height, lower crown height and diameter at breast height; Figures #a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.
+greater crown radius, total height, lower crown height and DBH; Figures #a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.
 
-The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, diameter at breast height and crown condition descriptors. 
+The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, DBH and crown condition descriptors. 
 Simulated response surfaces for the DAI are shown in Figure c. 

### 2019-12-16 R/functions/DAIresponseSurfaces.R

@@ -63,7 +63,7 @@ DAIresponseSurfaces <- function(DAIrf,decline_indexes,site_corrected_analysis_su
                                                    `Active bleeds`,
                                                    `Black staining size (mm)`,
                                                    `Black staining`,
-                                                   `Diameter at breast height (mm)`)
+                                                   `Diameter at breast height (m)`)
         ) %>%
         mutate(DAI = predict(DAIrf,newdata = .))
     }) %>%
@@ -96,7 +96,7 @@ DAIresponseSurfaces <- function(DAIrf,decline_indexes,site_corrected_analysis_su
                                                    `Active bleeds`,
                                                    `Black staining size (mm)`,
                                                    `Black staining`,
-                                                   `Diameter at breast height (mm)`)
+                                                   `Diameter at breast height (m)`)
         ) %>%
         mutate(DAI = predict(DAIrf,newdata = .))
     }) %>%
@@ -124,7 +124,7 @@ DAIresponseSurfaces <- function(DAIrf,decline_indexes,site_corrected_analysis_su
              caption = str_c(
                'Active bleeds = ',plotRanges_DAI_abc[[type]]$`Active bleeds`[1] %>% round(3),'\n',
                'Black staining = ',plotRanges_DAI_abc[[type]]$`Black staining`[1] %>% round(3),'\n',
-               'Diameter at breast height (mm) = ',plotRanges_DAI_abc[[type]]$`Diameter at breast height (mm)`[1] %>% round(3)
+               'Diameter at breast height (m) = ',plotRanges_DAI_abc[[type]]$`Diameter at breast height (m)`[1] %>% round(3)
              )
         )
       return(pl)
@@ -153,7 +153,7 @@ DAIresponseSurfaces <- function(DAIrf,decline_indexes,site_corrected_analysis_su
                'Crown radius (m) = ',plotRanges_DAI_def[[type]]$`Crown radius (m)`[1] %>% round(3),'\n',
                'Total height (m) = ',plotRanges_DAI_def[[type]]$`Total height (m)`[1] %>% round(3),'\n',
                'Lower crown height (m) = ',plotRanges_DAI_def[[type]]$`Lower crown height (m)`[1] %>% round(3),'\n',
-               'Diameter at breast height (mm) = ',plotRanges_DAI_def[[type]]$`Diameter at breast height (mm)`[1] %>% round(3)
[…]

### 2019-12-16 R/functions/PDIresponseSurfaces.R

@@ -11,7 +11,7 @@ PDIresponseSurfaces <- function(PDIrf,decline_indexes,site_corrected_analysis_su
       a <- decline_indexes$PDI %>%
         {. - .x} %>%
         abs()
-      decline_indexes$PDI[which(a == min(a))]
+      decline_indexes$PDI[which(a == min(a))[1]]
       })
   
   spectrumTrees_PDI <- list(
@@ -73,7 +73,7 @@ PDIresponseSurfaces <- function(PDIrf,decline_indexes,site_corrected_analysis_su
              caption = str_c('Crown radius (m) = ',plotRanges_PDI[[type]]$`Crown radius (m)`[1] %>% round(3),'\n',
                              'Total height (m) = ',plotRanges_PDI[[type]]$`Total height (m)`[1] %>% round(3),'\n',
                              'Lower crown height (m) = ',plotRanges_PDI[[type]]$`Lower crown height (m)`[1] %>% round(3),'\n',
-                             'Diameter at breast height (mm) = ',plotRanges_PDI[[type]]$`Diameter at breast height (mm)`[1] %>% round(3)
+                             'Diameter at breast height (m) = ',plotRanges_PDI[[type]]$`Diameter at breast height (m)`[1] %>% round(3)
              )
         )
       return(pl)

### 2019-12-16 R/functions/calcAdditionalDescriptors.R

@@ -16,7 +16,7 @@ calcAdditionalDescriptors <- function(phenoData){
                                                     `Active bleeds`,
                                                     `Black staining size (mm)`,
                                                     `Black staining`,
-                                                    `Diameter at breast height (mm)`),
-           `Agrilus exit hole density (m^-2)` = agrilusExitHoleDensity(`Agrilus exit holes`,`Diameter at breast height (mm)`)
+                                                    `Diameter at breast height (m)`),
+           `Agrilus exit hole density (m^-2)` = agrilusExitHoleDensity(`Agrilus exit holes`,`Diameter at breast height (m)`)
     )
 }

### 2019-12-16 manuscript/figures.Rmd
added figure numbers

@@
 
-```{r decline_index_plot}
+```{r decline_index_plot,fig.cap='Decline indexes for all 174 trees.
+The chosen group categories were selected in the field prior to data collection.
+AOD = acute oak decline; COD = chronic oak decline.
+Remisson tress were those that were known to have had active bleeds within the previous 3 years.'}
 loadd(decline_indexes_plot)

### 2019-12-16 manuscript/manuscript.Rmd
added figure numbers

@@
 Other categorical descriptors included social class, dieback type, main crown dieback location, stem epicormic growth, branch epicormic growth, extent of insect damage, insect damage type, extent of crown mildew and the presence or absence of pruning damage or branch loss from the main stem.
-See supplementary table # for category details.
+See Supplementary Table 1 for category details.
 
@@
 biguttatus* exit holes were counted along with the presence or absence of each of oval shaped exit holes, small circular shaped exit holes and ground level fungal fruiting bodies were recorded from the basal 2m of the stem (ref).
-These descriptors are shown in Supplementary table #.
+These descriptors are shown in Supplementary Table 1.
 
@@
 To account age, size and site specific tree difference, an adjustment of site means to the overall mean was applied to each of the size descriptors (DBH, lower crown height, timber height, total height, crown radius).
-An example of this adjustment is shown for DBH in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.
+An example of this adjustment is shown for DBH in Supplementary Figure 1 and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table 2.
 
@@
 The optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training. 
-The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: PDI - `r glue('ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; DAI - `r glue('ntree = {DAI_rf_tune_params$ntree}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure #)
+The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: PDI - `r glue('ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; DAI - `r glue('ntree = {DAI_rf_tune_params$ntree}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2)
 
@@
 
-High discrimination was found in the phenotypic descriptors between between the sites using supervised random forest classification (Supplementary  figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`).
-The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary figure #b).
+High discrimination was found in the phenotypic descriptors between between the sites using supervised random forest classification (Supplementary  figure 3a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`).
+The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary figure 3b).
 After the trees size adjustment was applied to removed inter-site variability but maintain intra-site variability, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
@@
 
-A scatter plot of the calculated PDI and DAI scores is shown in Figure # for all 174 trees for which phenotypic descriptor data was collected.
+A scatter plot of the calculated PDI and DAI scores is shown in Figure 1 for all `r nrow(site_corrected_analysis_suitable_data)` trees for which phenotypic descriptor data was collected.
 It can clearly be seen that the control trees group towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
@@
 
-The top four ranked descriptors found to contribute to the PDI index model were all crown size and condition related descriptors with crown volume the top ranked descriptor overall (Figure #a).
+The top four ranked descriptors found to contribute to the PDI index model were all crown size and condition related descriptors with crown volume the top ranked descriptor overall (Figure 2a).
 Other tree size related descriptors such as and stem bleeding descriptors such as contributed secondarily.
 
-The DAI was more complex in descriptor contribution many more descriptors contributing more equally than the PDI (Figure #b).
+The DAI was more complex in descriptor contribution many more descriptors contributing more equally than the PDI (Figure 2b).
 The overall tree size descriptors live crown ratio, DBH and crown volume contributed almost equally to the DAI index model.
@@
 
-Plots of the individual descriptors against the decline indexes are available in Supplementary figures #-#. 
+Plots of the individual descriptors against the decline indexes are available in Supplementary Figures 4-7. 
 The following sections identify how the descriptors contribute and interact to the decline index models in specific decline cases.
@@
 
-LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to the PDI in each of the healthy, moderate and severe decline cases using lime (Table #).
+LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to the PDI in each of the healthy, moderate and severe decline cases using lime (Table 1).
 These descriptors were also the same top descriptors in the overall model importance lists for the PDI.
@@
 
-For the DAI, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table #).
+For the DAI, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table 2).
 This likely reflects the greater number of descriptors contributing to the DAI with different descriptors relevant to particular decline contexts.
@@
 The index values generated by the models from these simulated phenotypic data can then be visualised as response surface plots, from which descriptor interactions within the indexes can be interpreted.
-Figures # and # show the simulated index value response surfaces under a variety of different circumstances for key phenotypic descriptors that contribute highly to the PDI and DAI.
+Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for key phenotypic descriptors that contribute highly to the PDI and DAI.
 
-Crown condition descriptors such as crown volume and crown condition contributed highly to the PDI (Figure #).
+Crown condition descriptors such as crown volume and crown condition contributed highly to the PDI (Figure 2a).
 The phenotypic measures central to this were percentage missing crown and percentage crown transparency from which crown condition and crown volume are calculated. 
-It can be seen in Figure # that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and estimated crown volume decrease, the PDI increases. 
+It can be seen in Figure 3 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and estimated crown volume decrease, the PDI increases. 
 A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g.
-greater crown radius, total height, lower crown height and DBH; Figures #a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.
+greater crown radius, total height, lower crown height and DBH; Figures 3a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.
 
@@
 Simulated response surfaces for the DAI are shown in Figure c. 
-These response surfaces show circumstances under which bleed prevalence related descriptors (Figures #a, b & c) and crown condition descriptors (Figures #d, e & f) change for non-symptomatic AOD or COD trees (Figures #a & d respectively), moderate AOD or COD (Figures #b & e respectively) and severe AOD or COD (Figures #c & f respectively).
+These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and crown condition descriptors (Figures 4d, e & f) change for non-symptomatic AOD or COD trees (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively).
 Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
-Also higher DAI values could be obtained from trees with greater stature (Figure #c). 
-Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure #d).
-Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures #e & f) and therefore more severe COD.
+Also higher DAI values could be obtained from trees with greater stature (Figure 4c). 
+Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure 4d).
+Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD.
 The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflict. 
@@
 
-The two dimensional space provided by these phenotypic indexes, for measuring decline severity is shown in Figure #.
+The two dimensional space provided by these phenotypic indexes, for measuring decline severity is shown in Figure 5.
 

### 2019-12-16 manuscript/supplementary.Rmd
added figure numbers

@@
 
-```{r descriptor_adjustment,fig.cap='Error bars show ±1SE of the mean.\nDashed red line shows the overall mean.'}
+```{r descriptor_adjustment,fig.cap='Exapmle of overall mean adjustment applied to the diameter at breast height descriptor. Error bars show ±1SE of the mean. Dashed red line shows the overall mean.'}
 loadd(descriptor_adjustment_example)
@@
   mutate_if(is.double,signif,digits = 3) %>%
-  kable(format = 'latex',booktabs = TRUE,caption = 'Descriptor correction factors at each site') %>%
+  kable(format = 'latex',booktabs = TRUE,caption = 'Descriptor adjustment factors at each site.') %>%
   collapse_rows(columns = 1,latex_hline = "major", valign = "middle")
@@
 
-```{r modelTuningResultsPlot}
+```{r modelTuningResultsPlot,fig.cap='Random forest parameter tuning results for decline index models.'}
 loadd(rf_tune_plot)
@@
 
-```{r PDInumeric,fig.height=9}
+```{r PDInumeric,fig.height=9,fig.cap='Continuous descriptors plotted against the PDI.'}
 loadd(descriptor_scatter_plots)
@@
 
-```{r PDIfactor,fig.height=8}
+```{r PDIfactor,fig.height=8,fig.cap='Categorical descriptors plotted against the PDI.'}
 descriptor_scatter_plots$PDI_factor
@@
 
-```{r DAInumeric,fig.height=9}
+```{r DAInumeric,fig.height=9,fig.cap='Continuous descriptors plotted against the DAI.'}
 descriptor_scatter_plots$DAI_numeric
@@
 
-```{r DAIfactor,fig.height=8}
+```{r DAIfactor,fig.height=8,fig.cap='Categorical descriptors plotted against the DAI.'}
 descriptor_scatter_plots$DAI_factor

### 2019-12-16 manuscript/manuscript.Rmd
alteration to abstract

@@
 
-Two indexes, the Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to describe acute and chronic oak decline severity using a multitude of easily measurable phenotypic descriptors.
+Oak decline is becoming increasingly prevalent along with the need for accurate quantitative oak health monitoring.
+The Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to quantitatively describe and differentiate acute and chronic oak decline severity using a multitude of easily measurable phenotypic descriptors.
 Measured phenotypic descriptors include aspects of tree shape and size, crown condition and the presence of biotic agents.
-These were collected from 174 trees exhibiting a spectrum of decline severity at 9 sites from across the south, east and west of England.
-Inter site differences were identified as a results of aspects of tree size which were corrected for using a mean centring adjustment.
 The decline indexes were then calculated using proximity values from an unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.

### 2019-12-16 manuscript/manuscript.Rmd
added to introduction

@@
 Measured phenotypic descriptors include aspects of tree shape and size, crown condition and the presence of biotic agents.
-The decline indexes were then calculated using proximity values from an unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
-These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a useful tool for forest monitoring and management.
+The decline indexes were derived using proximity values from n unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
+These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a potentially useful tool for forest monitoring and management.
 
@@
 
-- Monitoring currently focusses on crown assessments
-- Acute and chronic decline associated phenotypes
-- The oak decline spectrum move away from simplistic binary categorisations
-- Manual assignment of decline status difficult prone to error for due to contribution of many factors. 
-- Unsupervised learning for complex patterns between descriptors.
-
 Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
 Abiotic factors can include both predisposing and inciting factors such as drought and are often
-Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* associated
+Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with acute oak decline (AOD) [@brown2017; @denman2014].
 
@@
 
-Compounding this complexity is the addition of different types of oak decline (AOD) that are often categorised based on the duration over which a tree has been declining or the biotic agents/factors associated with a declining tree.
-For instance a chronically declining oak can subsequently begin to also develop the stem bleeds associated with AOD [@denman2014].
-Accurately diagnosing in the field the severity and type of oak decline present is difficult and can requires much experience, expertise and training.
+Compounding the complexity of oak decline syndromes is the addition of different types of oak decline that are often categorised based on the duration over which a tree has been declining and the biotic agents/factors associated with a declining tree.
+Mortality in trees suffering from chronic oak decline (COD) can occur over a period of decades where as AOD can occur in 5-10 years [@denman2009].
+Also a chronically declining oak (COD) can subsequently begin to develop the stem bleeds associated with AOD.
+The visual, manual assignment of decline status and severity in the field is variable and error prone due to the contribution of many factors.
+Doing this accurately requires much expertise, experience, and training (ref).
 
-It is becoming increasingly important that oak tree health within woodlands is accurately and sensitively monitored to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices (ref).
-Monitoring methods need to take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.
+Assigning discrete decline status labels neglects the true spectrum of decline severity encountered in the field.
+There is a growning need for sensitive, continuous measures of tree decline that accuratly reflect the decline spectrum [@pontius2014].
+This has the potential to greatly imporove monitoring to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices (ref).
+Along with this, monitoring methods need to take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.
+
+- Monitoring currently focusses on crown assessments
 
@@
 Unsupervised learning is also possible [@shi2006]
+
+- Unsupervised learning for complex patterns between descriptors.
 

### 2019-12-16 manuscript/manuscript.Rmd
finished introduction

@@
 
+Oak decline is becoming/predicted to become increasingly prevalent in the UK and across Europe with the onset of climate change (ref).
 Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
@@
 
-- Monitoring currently focusses on crown assessments
-
-Machine learning is finding increasing utility in biological and geographical applications due to it's ability to learn complex patterns within and between variables.
+Machine learning is finding increasing utility in ecological and environmental applications due to it's ability to learn complex patterns within and between variables [@cutler2007;@belgiu2016].
 A popular algorithm is that of random forest due to it's versatility, relatively trivial training requirements and resilience against over fitting [@breiman2001a] (refs).
-
-- Random forest has previously been used in a forest condition context [@vitale2014].
 
@@
 It is also able to simultaneously handle both continuous and categorical variables.
-Unsupervised learning is also possible [@shi2006]
-
-- Unsupervised learning for complex patterns between descriptors.
+Unsupervised learning is also possible using the assumption that if the data contains any underlying structure, this should be distinguishable from a randomly generated version of itself [@shi2006].
+An observation proximity matrix can be derived from this process on which additional unsupervised methods can be applied such as multidimensional scaling (ref).
 
@@
 - Refinement of protocol for large scale surveying
+- Aid long term monitoring by providing comparibility over time.
+This was highlighted in the recently published Action Oak Knowledge Review as a key requirment in the future for safeguarding oak health.
+[@quine2019],
+- Applied to long established monitoring data sets such as The International Co-operative Programme on Assessment and Monitoring of Air Pollution Effects on Forests (ICP Forests) [@lorenz1995].
 

### 2019-12-16 manuscript/manuscript.Rmd
added to discussion

@@
 
-$$v = \frac{\pi r^2c(h-l)}{200}$$
+$$v = \frac{1}{2} \pi r^2(h-l) \times \frac{c}{100}$$
 
@@
 
+- use of indexes derived from machine learning are context agnostic
+
 ### Applications and wider scalability
 
-- Scalability of method for wider application
-- Accessibility of collecting descriptors
+The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment and are based on well established tree condition assessment protocols (refs).
+The representation of the decline severity spectrum provided by these decline indexes provides greatly enhanced sensitivity over categorical decline status assignments that could enhance integration with large scale data such as those from genome wide association studies or remote sensing (refs).
+
+The current context in which these decline indexes are presented is based on a very focus set of descriptors on sites with a high proportion of declining trees.
+
+Scaling this technique to enable generalised naive compuation of decline indexes for any tree sampled in the UK would require a substantially larger training data set that incorporates the descirptive diversity of oak decline in the population as whole (ref).
+Scaling to this level would also require an approprite method of tree size correction that could be applied on an individual tree basis.
+A potential solution to this would be to correct trees based on estimated age from DBH as proposed by @white1998.
+This would enable the development of web based tools that can then provide acces to
+
 - Discussion of applications to forestry monitoring and management
-- Sensitivity for integration with molecular data compared to categorical assignments
-- Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
 - Refinement of protocol for large scale surveying
+
+Alternative phenotyping protocols could be developed compare across species that would incorporate appropriate tree size corrections
 - Aid long term monitoring by providing comparibility over time.
@@
 
-The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment. 
 They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations.
@@
 - Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment
-- Encouraged to use this approach in other contexts
+- Encouraged to apply this approach in other contexts
+
+The decline indexes presented here provided intuitive measures 
+The approach presented here is easily adaptable to incorporate any phenot
 

### 2019-12-17 manuscript/manuscript.Rmd
proofing

@@
 
-Oak decline is becoming/predicted to become increasingly prevalent in the UK and across Europe with the onset of climate change (ref).
+Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change (ref).
 Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
-Abiotic factors can include both predisposing and inciting factors such as drought and are often
+Abiotic factors can include both predisposing and inciting factors such as drought.
 Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with acute oak decline (AOD) [@brown2017; @denman2014].
@@
 Mortality in trees suffering from chronic oak decline (COD) can occur over a period of decades where as AOD can occur in 5-10 years [@denman2009].
-Also a chronically declining oak (COD) can subsequently begin to develop the stem bleeds associated with AOD.
+A chronically declining oak (COD) can also subsequently begin to develop the stem bleeds associated with AOD.
 The visual, manual assignment of decline status and severity in the field is variable and error prone due to the contribution of many factors.
@@
 Assigning discrete decline status labels neglects the true spectrum of decline severity encountered in the field.
-There is a growning need for sensitive, continuous measures of tree decline that accuratly reflect the decline spectrum [@pontius2014].
-This has the potential to greatly imporove monitoring to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices (ref).
+There is a growing need for sensitive, continuous measures of tree decline that accurately reflect the decline spectrum [@pontius2014].
+This has the potential to greatly improve monitoring to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices (ref).
 Along with this, monitoring methods need to take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.
@@
 All trees were manually assigned a decline status category; control, AOD, COD or remission prior to collection of phenotypic descriptors.
-The remission group included trees that had shown stem bleeds associated with AOD in the past but were not showing signs of stem bleeds at the time of assessment.
+The remission group included trees that had shown stem bleeds associated with AOD within the previous three years but were not showing signs of stem bleeds at the time of assessment.
 
-DBH (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
+Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
 Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m
 The timber height was measured at either 7cm top stem diameter or where no main stem could be distinguished.
-For the crown, canopy closure, missing crown and absolute crown density percentages were visually estimated within 5% classes
+For the crown, crown contact, missing crown and absolute crown density percentages were visually estimated within 5% classes
 
@@
 Crown radius, canopy closure and tap tests for dead stem tissue were measured for each cardinal point.
-The crown radius was measured using a vertex transponder attached to the relevant side of the stem at a height of 1.3m.Canopy closure recorded as yes or no depending on whether any other crowns are touching the sampling tree.
+The crown radius was measured using a vertex transponder attached to the relevant side of the stem at a height of 1.3m.
+Canopy closure recorded as yes or no depending on whether any other crowns are touching the sampling tree.
 A high resolution photo was taken of the vase to the spring of the crown and of the stem from the base to a height of 3m for each cardinal point for later reference.
 
-The AOD stem symptoms (active bleeds, black staining, callused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem (ref).
+The AOD stem symptoms (active bleeds, black staining, callused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem.
 The frequency of *A.
-biguttatus* exit holes were counted along with the presence or absence of each of oval shaped exit holes, small circular shaped exit holes and ground level fungal fruiting bodies were recorded from the basal 2m of the stem (ref).
-These descriptors are shown in Supplementary Table 1.
+biguttatus* exit holes were counted along with the presence or absence of each of oval shaped exit holes, small circular shaped exit holes and ground level fungal fruiting bodies were recorded from the basal 2m of the stem.
+The collected descriptors are all included in Supplementary Table 1.
 
@@
 The crown density ($CD$) values erroneously collected at Attingham and Langdale were converted to crown transparency ($t = 100 - CD$)
-All descriptors collected in cm were converted to mm.
+All descriptors collected in cm were converted to mm except for DBH which was converted to m.
 The tap tests that were positive for hollow stem tissue were converted to present and absent renamed dead stem tissue.
-The canopy closure descriptor was converted from yes and now to present and absent.
+The canopy closure descriptor was converted from yes and no to present and absent.
 Finally, the stem symptoms sizes for active bleeds, black staining and callused wounds were averaged and a frequency per tree collected.
@@
 
-To account age, size and site specific tree difference, an adjustment of site means to the overall mean was applied to each of the size descriptors (DBH, lower crown height, timber height, total height, crown radius).
+To account for age, size and site specific tree difference, an adjustment of site means to the overall mean was applied to each of the size descriptors (DBH, lower crown height, timber height, total height, crown radius).
 An example of this adjustment is shown for DBH in Supplementary Figure 1 and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table 2.
@@
 
-The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest using the randomForest R package v`r packageVersion('randomForest') %>% as.character()` [@liaw2002].
+The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest using the randomForest R package version `r packageVersion('randomForest') %>% as.character()` [@liaw2002].
 The proximity matrices from each repetition were averaged to a single proximity matrix ($P$).
@@
 This dimension was named the DAI.
-With different data, it is likely that the MDS dimensions may require inversion prior to min-max scaling to ensure their correct orientation.
 
-The code used to generate the decline indexes is available as an R package pdi v`r packageVersion('pdi') %>% as.character()` at https://github.com/jasenfinch/pdi.
+The code used to generate the decline indexes is available as an R package pdi version `r packageVersion('pdi') %>% as.character()` at https://github.com/jasenfinch/pdi.
 
@@
 Individual random forest regression models of the phenotypic descriptors were trained, supervised using each decline index.
-The optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training. 
-The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: PDI - `r glue('ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; DAI - `r glue('ntree = {DAI_rf_tune_params$ntree}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2)
-
-Overall descriptor importance contributions to each of the indexes was retrieved from these models using the mean decrease in accuracy measure (ref).
+The optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
+The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: PDI - `r glue('ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; DAI - `r glue('ntree = {DAI_rf_tune_params$ntree}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
+Overall descriptor importance contributions to each of the indexes was retrieved from these models using the mean decrease in accuracy measure.
 
@@
 Local Interpretable Model-Agnostic Explanations (LIME) were computed to identify descriptors contributing locally to case specific examples of each of the indexes [@ribeiro2016]. 
-The lime R package v`r packageVersion('lime') %>% as.character()` was used with the default parameters [@pedersen2019].
+The lime R package version `r packageVersion('lime') %>% as.character()` was used with the default parameters [@pedersen2019].
 A healthy, a moderate and a severe decline cases was used for the PDI. 
@@
 
-High discrimination was found in the phenotypic descriptors between between the sites using supervised random forest classification (Supplementary  figure 3a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`).
+High discrimination was found in the phenotypic descriptors between the sites was investigated using supervised random forest classification (Supplementary  figure 3a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`).
 The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary figure 3b).
-After the trees size adjustment was applied to removed inter-site variability but maintain intra-site variability, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
-This enabled comparisons to be made between the trees at the different sites in the context of decline severity.
+After the application of trees size adjustment to removed inter-site variability but maintain intra-site variability, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
+This enabled subsequent comparisons to be made between the trees at the different sites in the context of decline severity.
 
@@
 The remission trees were found to be significantly different in their PDI scores from both the control (`r glue("t({remission_PDI_group_ttest$parameter[1]}) = {signif(remission_PDI_group_ttest$statistic[1],digits = 3) %>% abs()}, p < 0.001")`) and symptomatic (AOD and COD; `r glue("t({remission_PDI_group_ttest$parameter[2]}) = {signif(remission_PDI_group_ttest$statistic[2],digits = 3) %>% abs()}, p < 0.001")`) groups with means of `r signif(remission_PDI_group_ttest$estimate1[1],digits = 3)`, `r signif(remission_PDI_group_ttest$estimate2[1],digits = 3)` and `r signif(remission_PDI_group_ttest$estimate2[2],digits = 3)` for the control, remission and symptomatic groups respectively.
-As a group these trees are showing moderate overall decline.
+As a group, these trees are showing moderate overall decline.
 
 The remission trees were also found to be significantly different to the AOD (`r glue("t({remission_DAI_group_ttest$parameter[1]}) = {signif(remission_DAI_group_ttest$statistic[1],digits = 3) %>% abs()}, p = < 0.001")`) and COD (`r glue("t({remission_DAI_group_ttest$parameter[2]}) = {signif(remission_DAI_group_ttest$statistic[2],digits = 3) %>% abs()}, p = {round(remission_DAI_group_ttest$p.value[2],3)}")`) groups on the DAI scale with group means of `r signif(remission_DAI_group_ttest$estimate1[1],digits = 3)`, `r signif(remission_DAI_group_ttest$estimate2[1],digits = 3)` and `r signif(remission_DAI_group_ttest$estimate1[2],digits = 3)` for the AOD, remission and COD groups respectively.
-As a group the remission trees are of neutral decline type.
+As a group, the remission trees were found to be neither AOD or COD.
 
@@
 
-The top four ranked descriptors found to contribute to the PDI index model were all crown size and condition related descriptors with crown volume the top ranked descriptor overall (Figure 2a).
-Other tree size related descriptors such as and stem bleeding descriptors such as contributed secondarily.
+The top four ranked descriptors found to contribute to the PDI index model were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure 2a).
+Other tree size related descriptors such as and stem bleeding descriptors contributed secondarily.
 
@@
 Increasing decline severity was associated with a reduction in the size and condition of the crown.
-For instance, declining trees were found to have an estimated crown volume below 841m^3^.
+For instance, declining trees were found to have a composite crown volume below 841m^3^.
 Healthy trees were found to have a crown condition greater than 71.3%, moderately declining trees between 47.6% and 71.3% and severely declining trees a condition below 23.9%.
@@
 The important descriptors identified were mainly associated with tree size such as live crown ratio and lower crown radius.
-Bleed prevalence and stem symptom descriptors were not found to be important to any of the decline type cases.
+Bleed prevalence and stem symptom descriptors were not found to be locally important to any of the decline type cases.
 Live crown ratio greater than 61.6% in the severe AOD and moderate COD cases.
@@
 
-Crown condition descriptors such as crown volume and crown condition contributed highly to the PDI (Figure 2a).
+Crown condition descriptors such as composite crown volume and crown condition contributed highly to the PDI (Figure 2a).
 The phenotypic measures central to this were percentage missing crown and percentage crown transparency from which crown condition and crown volume are calculated. 
-It can be seen in Figure 3 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and estimated crown volume decrease, the PDI increases. 
+It can be seen in Figure 3 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and composite crown volume decrease, the PDI increases. 
 A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g.
@@
 The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, DBH and crown condition descriptors. 
-Simulated response surfaces for the DAI are shown in Figure c. 
+Simulated response surfaces for the DAI are shown in Figure 4. 
 These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and crown condition descriptors (Figures 4d, e & f) change for non-symptomatic AOD or COD trees (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively).
@@
 Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD.
-The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflict. 
+The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflicted. 
 AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.
@@
 - Refined definition of remission - repeat measurement
+- With different data, it is likely that the MDS dimensions may require inversion prior to min-max scaling to ensure their correct orientation.
 
@@
 
-Scaling this technique to enable generalised naive compuation of decline indexes for any tree sampled in the UK would require a substantially larger training data set that incorporates the descirptive diversity of oak decline in the population as whole (ref).
-Scaling to this level would also require an approprite method of tree size correction that could be applied on an individual tree basis.
+Scaling this technique to enable generalised naive computation of decline indexes for any tree sampled in the UK would require a substantially larger training data set that incorporates the descriptive diversity of oak decline in the population as whole (ref).
+Scaling to this level would also require an appropriate method of tree size correction that could be applied on an individual tree basis.
 A potential solution to this would be to correct trees based on estimated age from DBH as proposed by @white1998.
-This would enable the development of web based tools that can then provide acces to
+This would enable the development of web based tools that can then provide access to
 
@@
 Alternative phenotyping protocols could be developed compare across species that would incorporate appropriate tree size corrections
-- Aid long term monitoring by providing comparibility over time.
-This was highlighted in the recently published Action Oak Knowledge Review as a key requirment in the future for safeguarding oak health.
+- Aid long term monitoring by providing comparability over time.
+This was highlighted in the recently published Action Oak Knowledge Review as a key requirement in the future for safeguarding oak health.
 [@quine2019],
@@
 The decline indexes presented here provided intuitive measures 
-The approach presented here is easily adaptable to incorporate any phenot
+The approach presented here is easily adaptable to incorporate any phenotype
 

### 2019-12-17 manuscript/manuscript.Rmd
finished machine learning discussions section

@@
 output: 
-  #bookdown::word_document2:
-  # toc: false
-  bookdown::pdf_document2: 
-    toc: false
-    number_sections: false
+  word_document: default
+  # pdf_document: default
 bibliography: references.bib
@@
 
+The phenotypic descriptors collected were based on @innes1990.
 Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
@@
 
-The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model responses surfaces for specific cases.
-This approach was similar to response surface methodology without the need for probabilistic approximation due to the low computation time required for index calculation (ref).
+The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model response surfaces for specific cases.
+This approach was similar to response surface methodology without the need for probabilistic approximation due to the low computation time required for index calculation [@abbasi2012].
 
@@
 Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD.
-The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflicted. 
-AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.
 
@@
 
-The phenotypic descriptors were selected for use in this study purposefully
+The decline indexes presented here show 
+These two indexes provide a two dimensional continuum under which the spectrum of decline severity and type can be assessed (Figure 5).
 
+- Indexes place the descriptors into single intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity
 - Suitability of collected phenotypic descriptors and decline indexes for describing decline status
-- Index thresholds for category definitions (healthy, declining, severe decline etc.)
-- Validation of decline scores - chicken and paradox
-- Direct validation difficult 
-- Refined definition of remission - repeat measurement
-- With different data, it is likely that the MDS dimensions may require inversion prior to min-max scaling to ensure their correct orientation.
 
@@
 
-These two indexes provide a two dimensional continuum under which the spectrum of decline severity and type can be assessed.
+- Index thresholds for category definitions (healthy, declining, severe decline etc.)
 
-The two dimensional space provided by these phenotypic indexes, for measuring decline severity is shown in Figure 5.
+- Association of PDI and overall oak decline with crown condition well documented
+- Association of crown condition and AOD severity. Could not previously be shown by @denman2014.
+- Bleed prevalence and AOD
+- Agrillus exit hole occurance with stem bleeding - smaller area surveyed compared to bleeding. Trees with bleeds do not always have stem bleeds [@brown2017;@vansteenkiste2004]
+- Size differences between AOD and COD trees
+
+The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflict. 
+AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.
+
+- Refined definition of remission - repeat measurement
 
@@
 
-- reasons for selecting machine learning approach
-- use of unsupervised over supervised analysis
-- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
-- For given context, descriptors not contributing to the indexes could be removed
-- Concession of model interpretability to predictive capability by using a machine learning approach
+An unsupervised approach was favoured in this context due to the concession that there know difficulties and inaccuracy of manual field decline status classification (ref).
+This is highlighted in Figure 1 where a numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (eg.
+COD trees with DAI values > 0.5).
+The unsupervised approach has worked in this context in aptly describing decline severity as the phenotypic descriptors selected were highly specific and well suited to describing the decline severity problem.
+In situations were descriptors may well be less suited, perhaps as a result of technical limitations, the use of a semi-supervised approach may be favourable (ref).
+The unsupervised approach to derive the decline indexes would be applied to a subset of the observations on which the descriptors are known to perform well and that would be considered benchmark observations by the investigators.
+Indexes could then be attained for the rest of the observations by prediction based on supervised models of the benchmark subset (ref).
+This approach would be more suitable than a fully supervised approach as it would limit the inaccuacies that would be introduced by manual classification and still allow the data driving flexibility of the unsupervised approach.
 
-Machine learning approaches are usually seen as black box 
-- seen as black box but LIME and response surfaces enable interpretation and therefore trust
-- LIME can be applied to out of place cases
-- Important descriptors identified model-wide aren't necessarily relevant to localised cases.
+Machine learning approaches are usually seen as black box and a lack of model interpretability is a concession in this approach however the are substantial gains in terms of predictive capability for model application in real world situations (ref).
+The assessment descriptor contributions to the index models are an integral part of this approach.
+Inout of knowledge of the biological problem by the investigator at this stage is essential in ensuring that the models suitably learn the patterns relevant biological problem to which they have been applied and that the descriptors contirbuting to these are sensible and that the algorithm (random forest) is not learning spurious or artificial patterns within the data (ref).
+Approaches such as LIME and model response surfaces enable investigators to query the contributions that phenotypic descriptors have in particular cases of interest.
+This inturn promotes trust in the indexes derived from these models and reduces their black box nature (ref).
 
-- use of indexes derived from machine learning are context agnostic
+### Applications
 
-### Applications and wider scalability
+- Explain the wider application of the approach
 
@@
 
-The current context in which these decline indexes are presented is based on a very focus set of descriptors on sites with a high proportion of declining trees.
-
-Scaling this technique to enable generalised naive computation of decline indexes for any tree sampled in the UK would require a substantially larger training data set that incorporates the descriptive diversity of oak decline in the population as whole (ref).
-Scaling to this level would also require an appropriate method of tree size correction that could be applied on an individual tree basis.
-A potential solution to this would be to correct trees based on estimated age from DBH as proposed by @white1998.
 This would enable the development of web based tools that can then provide access to
@@
 - Discussion of applications to forestry monitoring and management
-- Refinement of protocol for large scale surveying
+- Refinement of protocol for large scale surveying or alternative applications
 
@@
 - Applied to long established monitoring data sets such as The International Co-operative Programme on Assessment and Monitoring of Air Pollution Effects on Forests (ICP Forests) [@lorenz1995].
+
+The current context in which these decline indexes are presented is based on a very focus set of descriptors on sites with a high proportion of declining trees. 
+Scaling this technique to enable generalised naive computation of decline indexes for any tree sampled in the UK would require a substantially larger training data set that incorporates the descriptive diversity of oak decline in the population as whole (ref).
+Scaling to this level would also require an appropriate method of tree size correction that could be applied on an individual tree basis.
+A potential solution to this would be to correct trees based on estimated age from DBH as proposed by @white1998.
 
@@
 - Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment
-- Encouraged to apply this approach in other contexts
 
@@
 The approach presented here is easily adaptable to incorporate any phenotype
+It is encouraged that this approach of deriving descriptive index measurse in other tree health phenotyping contexts.
 

### 2019-12-17 manuscript/manuscript.Rmd
finished applications discussion

@@
 output: 
-  word_document: default
-  # pdf_document: default
+  #word_document: default
+  pdf_document: default
 bibliography: references.bib
@@
 
-**Authors:** Jasen Finch, Nathan Brown, Sandra Denman & John Draper
+**Authors:** Jasen Finch, Nathan Brown, John Draper & Sandra Denman
 
@@
 Abiotic factors can include both predisposing and inciting factors such as drought.
-Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with acute oak decline (AOD) [@brown2017; @denman2014].
+Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with acute oak decline (AOD) [@brown2017a; @denman2014].
 
@@
 The phenotypic data were collected from a total of 174 trees from nine sites from across the south (Hatchlands and Richmond Park), east (Big Wood, Great Monks Wood and Winding Wood) and west (Attingham, Chestnuts, Langdale and Speculation) of England between June and September in 2016 or 2017.
-The occurrence of AOD has previously been reported and studied at six of the sites [@brown2016; @brown2017].
-These sites included a variety of woodland types, management types, soil types and tree sizes/ages.
+The occurrence of AOD has previously been reported and studied at six of the sites [@brown2016; @brown2017a].
+These sites included a variety of woodland types, management types, soil types and tree sizes/ages and were made up mainly *Quercus robur*.
 
@@
 - Bleed prevalence and AOD
-- Agrillus exit hole occurance with stem bleeding - smaller area surveyed compared to bleeding. Trees with bleeds do not always have stem bleeds [@brown2017;@vansteenkiste2004]
+- Agrillus exit hole occurance with stem bleeding - smaller area surveyed compared to bleeding. Trees with bleeds do not always have stem bleeds [@brown2017a;@vansteenkiste2004]
 - Size differences between AOD and COD trees
@@
 
-- Explain the wider application of the approach
+The current context in which these decline indexes are presented is study specific and is based on a relatively small sample size that may not be generally representative of the UK oak population as a whole.
+Subsequent studies could incorporate the data from this study with additional data to train random forest models if comparable phenotyping protocols are used or generate study specific decline index values using the unsupervised learning approach.
 
-The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment and are based on well established tree condition assessment protocols (refs).
-The representation of the decline severity spectrum provided by these decline indexes provides greatly enhanced sensitivity over categorical decline status assignments that could enhance integration with large scale data such as those from genome wide association studies or remote sensing (refs).
+The representation of the decline severity spectrum provided by these decline indexes provides greatly enhanced sensitivity over categorical decline status assignments.
+This could enhance integration with large scale data such as those from genome wide association studies or remote sensing [@stocks2019] (ref).
+These index measures allow sensitive comparisons of visual health to be made for the same tree between years as well as comparisons of trees between different locations.
 
-This would enable the development of web based tools that can then provide access to
-
-- Discussion of applications to forestry monitoring and management
-- Refinement of protocol for large scale surveying or alternative applications
-
-Alternative phenotyping protocols could be developed compare across species that would incorporate appropriate tree size corrections
-- Aid long term monitoring by providing comparability over time.
-This was highlighted in the recently published Action Oak Knowledge Review as a key requirement in the future for safeguarding oak health.
-[@quine2019],
-- Applied to long established monitoring data sets such as The International Co-operative Programme on Assessment and Monitoring of Air Pollution Effects on Forests (ICP Forests) [@lorenz1995].
-
-The current context in which these decline indexes are presented is based on a very focus set of descriptors on sites with a high proportion of declining trees. 
-Scaling this technique to enable generalised naive computation of decline indexes for any tree sampled in the UK would require a substantially larger training data set that incorporates the descriptive diversity of oak decline in the population as whole (ref).
+Scaling this technique to enable generalised naive computation of decline indexes for any tree sampled in the UK would require index models trained using a substantially larger data set that incorporates the descriptive diversity of oak decline in the population as whole (ref).
 Scaling to this level would also require an appropriate method of tree size correction that could be applied on an individual tree basis.
 A potential solution to this would be to correct trees based on estimated age from DBH as proposed by @white1998.
+These indexes could have great utility in long term forest health monitoring by providing comparability over time that could be utilised for both forestry management and research (ref).
+The development of tools to aid long term monitoring was highlighted as a key requirement in the future for safeguarding UK oak health in the recently published Action Oak Knowledge Review [@quine2019].
 
-They allow sensitive comparisons to be made for the same tree between years as well as comparisons of trees between different locations.
-which could have great utility in long term forest health monitoring for both forestry management and research.
+The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment and are based on well established tree condition assessment protocols [@innes1990].
+Refinement of the protocols used here could make could reduce survey time and complexity and potentially be incorporated into citizen science programmes as a resource to build general UK oak training set [@brown2017b]
+
+Alternatively, the machine learning approach used to derive these decline indexes could also be applied to long established oak health monitoring programmes to aid sensitivity in detecting overall trends in oak health.
+An example programme includes The International Co-operative Programme on Assessment and Monitoring of Air Pollution Effects on Forests (ICP Forests) [@lorenz1995].
 

### 2019-12-17 manuscript/manuscript.Rmd
added to discussion

@@
 - Association of PDI and overall oak decline with crown condition well documented
-- Association of crown condition and AOD severity. Could not previously be shown by @denman2014.
+- Association of crown condition and AOD severity in trees  Could not previously be shown by @denman2014.
 - Bleed prevalence and AOD
-- Agrillus exit hole occurance with stem bleeding - smaller area surveyed compared to bleeding. Trees with bleeds do not always have stem bleeds [@brown2017a;@vansteenkiste2004]
+
+There was no contribution of *A.
+biguttatus* exit hole occurance with with the DAI and therefore stem bleeding - smaller area surveyed compared to bleeding.
+It has also been found previouly that trees with bleeds do not always have exit holes as  [@brown2017a;@vansteenkiste2004].
+
 - Size differences between AOD and COD trees

### 2019-12-18 manuscript/manuscript.Rmd
Intro and methods proofing

@@
 
-Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change (ref).
+Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change [@sturrock2011.
 Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
 Abiotic factors can include both predisposing and inciting factors such as drought.
-Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with acute oak decline (AOD) [@brown2017a; @denman2014].
+Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014].
 
 As a results of this, there are many visual phenotypic descriptors encompassing the entirety of an oak tree that contribute to what can be defined as a declining oak tree.
-A severely declining oak is most prominently characterised by very poor crown condition and numerous dead branches (ref).
+A severely declining oak is most prominently characterised by very poor crown condition and numerous dead branches.
 It is these characteristics that many studies use as the primary indicator of tree decline severity [@zarnoch2004].
 
-Compounding the complexity of oak decline syndromes is the addition of different types of oak decline that are often categorised based on the duration over which a tree has been declining and the biotic agents/factors associated with a declining tree.
-Mortality in trees suffering from chronic oak decline (COD) can occur over a period of decades where as AOD can occur in 5-10 years [@denman2009].
-A chronically declining oak (COD) can also subsequently begin to develop the stem bleeds associated with AOD.
+Compounding the complexity of oak decline syndromes is the addition of different types of oak decline associated with *Quercus robur* and *Q.
+petraea* that are often categorised based on the duration over which a tree has been declining and the biotic agents associated with a declining tree.
+Mortality in trees suffering from Chronic Oak Decline (COD) can occur over a period of decades where as AOD can occur in 5-10 years [@denman2009].
+A chronically declining oak can also subsequently begin to develop the stem bleeds associated with AOD.
 The visual, manual assignment of decline status and severity in the field is variable and error prone due to the contribution of many factors.
-Doing this accurately requires much expertise, experience, and training (ref).
+Doing this accurately requires much expertise, experience, and training [@ferretti1997].
 
@@
 There is a growing need for sensitive, continuous measures of tree decline that accurately reflect the decline spectrum [@pontius2014].
-This has the potential to greatly improve monitoring to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices (ref).
-Along with this, monitoring methods need to take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.
+This has the potential to greatly improve monitoring to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices.
+The recently published Action Oak Knowledge Review highlighted monitoring long term consistently as a key requirement for safeguarding UK oak health in the future [@quine2019].
+Coupled with this, monitoring methods should take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.
 
 Machine learning is finding increasing utility in ecological and environmental applications due to it's ability to learn complex patterns within and between variables [@cutler2007;@belgiu2016].
-A popular algorithm is that of random forest due to it's versatility, relatively trivial training requirements and resilience against over fitting [@breiman2001a] (refs).
+A popular algorithm is that of random forest due to it's versatility, relatively trivial training requirements and ease of use [@breiman2001a;@biau2016].
 
@@
 It is also able to simultaneously handle both continuous and categorical variables.
-Unsupervised learning is also possible using the assumption that if the data contains any underlying structure, this should be distinguishable from a randomly generated version of itself [@shi2006].
-An observation proximity matrix can be derived from this process on which additional unsupervised methods can be applied such as multidimensional scaling (ref).
+Unsupervised learning is possible using the assumption that if the data contains any underlying structure, this should be distinguishable from a randomly generated version of itself.
+An observation proximity matrix can be derived from this process on which additional unsupervised methods can be applied such as multidimensional scaling [@shi2006].
 
@@
 The occurrence of AOD has previously been reported and studied at six of the sites [@brown2016; @brown2017a].
-These sites included a variety of woodland types, management types, soil types and tree sizes/ages and were made up mainly *Quercus robur*.
+These sites included a variety of woodland, management and soil types as well as tree sizes/ages and were made up of mainly *Q. robur*.
 
-Descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
 All trees were manually assigned a decline status category; control, AOD, COD or remission prior to collection of phenotypic descriptors.
@@
 
-The phenotypic descriptors collected were based on @innes1990.
+The phenotypic descriptors collected related to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
+The descriptors collected were based on @innes1990.
 Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
@@
 The timber height was measured at either 7cm top stem diameter or where no main stem could be distinguished.
-For the crown, crown contact, missing crown and absolute crown density percentages were visually estimated within 5% classes
+For the crown, crown contact, missing crown and absolute crown transparency percentages were visually estimated within 5% classes
 
@@
 The frequency of *A.
-biguttatus* exit holes were counted along with the presence or absence of each of oval shaped exit holes, small circular shaped exit holes and ground level fungal fruiting bodies were recorded from the basal 2m of the stem.
+biguttatus* exit holes were counted along with the presence or absence of each of oval shaped exit holes, small circular shaped exit holes and ground level fungal fruiting bodies which were recorded from the basal 2m of the stem.
 The collected descriptors are all included in Supplementary Table 1.
@@
 All descriptors collected in cm were converted to mm except for DBH which was converted to m.
-The tap tests that were positive for hollow stem tissue were converted to present and absent renamed dead stem tissue.
+The tap tests that were positive for hollow stem tissue were converted to present and absent then renamed dead stem tissue.
 The canopy closure descriptor was converted from yes and no to present and absent.
@@
 This then gave a total of `r ncol(analysis_suitable_data)` raw and composite phenotypic descriptors.
-The additional descriptors are defined as follows:
+The composite descriptors are defined as follows:
 
-**Live crown ratio (%)** - percentage of the whole tree height that supports live crown.
+Live crown ratio (%) - percentage of the whole tree height that supports live crown.
 
@@
 
-**Crown condition (%)** - percentage of present crown proportional to the percentage crown density.
+Crown condition (%) - percentage of present crown proportional to the percentage crown density.
 
@@
 
-**Composite crown volume (m^3^)** - Estimated volume of crown as an paraboloid proportional to the crown condition.
+Composite crown volume (m^3^) - estimated volume of the crown as an paraboloid proportional to the crown condition.
 Adapted from the composite crown volume from @zarnoch2004 using crown condition instead of crown density alone.
@@
 
-**Estimated bleed prevalence (%)** - estimated percentage of the surveyed 3m basal trunk area affected by bleeds or black staining.
+Estimated bleed prevalence (%) - estimated percentage of the surveyed 3m basal trunk area affected by bleeds or black staining.
 
@@
 
-***Agrilus* exit hole density (m^-2^)** - density of observed *Agrilus* exit holes across the surveyed 2m basal trunk area.
+Agrilus exit hole density (m^-2^) - density of observed *A. biguttatus* exit holes across the surveyed 2m basal trunk area.
 
@@
 
-$h$ = number of *Agrilus* exit holes, d = DBH (m)
+$h$ = number of *A. biguttatus* exit holes, d = DBH (m)
 
@@
 The optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
-The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: PDI - `r glue('ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; DAI - `r glue('ntree = {DAI_rf_tune_params$ntree}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
-Overall descriptor importance contributions to each of the indexes was retrieved from these models using the mean decrease in accuracy measure.
+The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: PDI - `r glue('ntree = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; DAI - `r glue('ntree = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
+Overall descriptor importance contributions to each of the indexes was measured in each model using the mean decrease in accuracy measure.
 
@@
 The lime R package version `r packageVersion('lime') %>% as.character()` was used with the default parameters [@pedersen2019].
-A healthy, a moderate and a severe decline cases was used for the PDI. 
+A healthy, a moderate and a severe decline case was used for the PDI. 
 For each case, a tree with a score closest to 0, 0.5 and 1 respectively was selected.
@@
 A decline neutral case was selected with a score closest to 0.
-Then for each of the AOD and COD syndromes, moderate and severe cases were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD.
+Then for each of the AOD and COD syndromes, moderate and severe cases were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD respectively.
 
@@
 The example cases included the trees with the minimum and maximum PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
-The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors divided into 100 increments.
+The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors were divided into 100 increments.
 For a given case, each combination of theses simulated descriptor values were used to calculate a PDI value from the supervised random forest regression model, building a model response surface.
@@
 These indexes could have great utility in long term forest health monitoring by providing comparability over time that could be utilised for both forestry management and research (ref).
-The development of tools to aid long term monitoring was highlighted as a key requirement in the future for safeguarding UK oak health in the recently published Action Oak Knowledge Review [@quine2019].
 

### 2019-12-18 manuscript/manuscript.Rmd
change control to healthy. Results proofing

@@
 
-All trees were manually assigned a decline status category; control, AOD, COD or remission prior to collection of phenotypic descriptors.
+All trees were manually assigned a decline status category; healthy, AOD, COD or remission prior to collection of phenotypic descriptors.
 The remission group included trees that had shown stem bleeds associated with AOD within the previous three years but were not showing signs of stem bleeds at the time of assessment.
@@
 
-High discrimination was found in the phenotypic descriptors between the sites was investigated using supervised random forest classification (Supplementary  figure 3a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`).
+High discrimination was found in the phenotypic descriptors between the sites, which was investigated using supervised random forest classification (Supplementary  figure 3a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`).
 The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary figure 3b).
@@
 A scatter plot of the calculated PDI and DAI scores is shown in Figure 1 for all `r nrow(site_corrected_analysis_suitable_data)` trees for which phenotypic descriptor data was collected.
-It can clearly be seen that the control trees group towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
+It can clearly be seen that the healthy trees group towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
 This gave a significant difference when PDI of the control and symptomatic (AOD and COD) decline statuses were compared using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
@@
 
-The remission trees were found to be significantly different in their PDI scores from both the control (`r glue("t({remission_PDI_group_ttest$parameter[1]}) = {signif(remission_PDI_group_ttest$statistic[1],digits = 3) %>% abs()}, p < 0.001")`) and symptomatic (AOD and COD; `r glue("t({remission_PDI_group_ttest$parameter[2]}) = {signif(remission_PDI_group_ttest$statistic[2],digits = 3) %>% abs()}, p < 0.001")`) groups with means of `r signif(remission_PDI_group_ttest$estimate1[1],digits = 3)`, `r signif(remission_PDI_group_ttest$estimate2[1],digits = 3)` and `r signif(remission_PDI_group_ttest$estimate2[2],digits = 3)` for the control, remission and symptomatic groups respectively.
+The remission trees were found to be significantly different in their PDI scores from both the healthy (`r glue("t({remission_PDI_group_ttest$parameter[1]}) = {signif(remission_PDI_group_ttest$statistic[1],digits = 3) %>% abs()}, p < 0.001")`) and symptomatic (AOD and COD; `r glue("t({remission_PDI_group_ttest$parameter[2]}) = {signif(remission_PDI_group_ttest$statistic[2],digits = 3) %>% abs()}, p < 0.001")`) groups with means of `r signif(remission_PDI_group_ttest$estimate1[1],digits = 3)`, `r signif(remission_PDI_group_ttest$estimate2[1],digits = 3)` and `r signif(remission_PDI_group_ttest$estimate2[2],digits = 3)` for the healthy, remission and symptomatic groups respectively.
 As a group, these trees are showing moderate overall decline.
@@
 The top four ranked descriptors found to contribute to the PDI index model were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure 2a).
-Other tree size related descriptors such as and stem bleeding descriptors contributed secondarily.
+Other tree size related descriptors such as and live crown ratio stem bleeding descriptors contributed secondarily.
 
@@
 LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to the PDI in each of the healthy, moderate and severe decline cases using lime (Table 1).
-These descriptors were also the same top descriptors in the overall model importance lists for the PDI.
+These descriptors were also the same top descriptors in the overall model importance lists for the PDI (Figure 2a).
 Increasing decline severity was associated with a reduction in the size and condition of the crown.
@@
 A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g.
-greater crown radius, total height, lower crown height and DBH; Figures 3a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.
+greater crown radius, total height, lower crown height and DBH; Figures 3a & b), with relatively small stature trees having PDIs of 0.6 or greater even with 100% crown condition.
 

### 2019-12-18 manuscript/manuscript.Rmd
added to index discussion

@@
 
+Although significance was obtained between the decline status groups on both the PDI and DAI scales, Figure 5 highlights the inaccuracies in the manual decline status classifications when viewed in an individual tree context.
+It can be seen that a number of the trees, when compared quantitatively using these measures, could in fact be classified with a different decline state.
+
 ### Key descriptors contributing to the decline indexes
@@
 Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure 4d).
-Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD.
+Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD however only narrow ranges of DAI values could be obtained for each case when varying the ranges of these descriptors.
 
@@
 
-The decline indexes presented here show 
-These two indexes provide a two dimensional continuum under which the spectrum of decline severity and type can be assessed (Figure 5).
+The decline indexes presented here place single intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity.
+These two indexes provide a two dimensional oak decline continuum under which the spectrum of decline severity and type can be assessed (Figure 5).
+The index thresholds for category definitions (healthy, declining, severe decline etc.) shown there currently serve only as a guide to aid human interpretation as opposed to definitive thresholds of disease.
+Further work is needed that includes phenotypic data with greater spatial and temporal resolution would be needed to establishing definitive thresholds.
 
-- Indexes place the descriptors into single intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity
 - Suitability of collected phenotypic descriptors and decline indexes for describing decline status
 
-This plot highlights the inaccuracies in manual field classifications of decline status with several trees classified as healthy controls that, when compared quantitatively to other trees, are in fact
-in a decline state.
+Crown condition and volume were shown to be the key contributors to the PDI. 
+This is to be expected and hoped for when using an unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007].
+However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
+It has been shown here that firstly factors of overal crown condition and size are unlikely to contribute equally to our visual definition of oak decline severity (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly to our visual definition of oak decline severity (Figure 3a).
 
-- Index thresholds for category definitions (healthy, declining, severe decline etc.)
-
-- Association of PDI and overall oak decline with crown condition well documented
-- Association of crown condition and AOD severity in trees  Could not previously be shown by @denman2014.
-- Bleed prevalence and AOD
+The high contribution of bleed prevalence related descriptors to the DAI is expected as this is one of the defining features of AOD [@denman2014].
 
@@
 It has also been found previouly that trees with bleeds do not always have exit holes as  [@brown2017a;@vansteenkiste2004].
+
+AOD severity could also be associated with overall decline severity with trees identified that had a combination of both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure 1).
+The association of crown condition and AOD severity could not previously be shown by @denman2014 due to a low sample size.
 

### 2019-12-18 manuscript/manuscript.Rmd
finished decline index discussion

@@
 
-- Suitability of collected phenotypic descriptors and decline indexes for describing decline status
-
 Crown condition and volume were shown to be the key contributors to the PDI. 
@@
 The high contribution of bleed prevalence related descriptors to the DAI is expected as this is one of the defining features of AOD [@denman2014]. 
-
 There was no contribution of *A.
@@
 
-- Size differences between AOD and COD trees
-
-The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflict. 
+It was shown that tree size, reflected by the live crown ratio descriptor, contributes most to the DAI in descriminating between COD and AOD trees.
+The relationship of relative tree stature and the acuteness of oak decline has not previously been shown and likely reflects the period of time under which these trees have been afflicted. 
 AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.
 
-- Refined definition of remission - repeat measurement
+The remission trees as a group were shown to be intermediate in decline severity and type to the healthy, AOD and COD trees on both the PDI and DAI index scales.
+However, this was also the most varied of the decline status groups.
+It is possible that this could reflect an underly relationship of the potential for trees to recover from AOD but this is speculative and it is not possible to attribute this further with these observations.
+Using these decline indexes, remission trees can be more objectively defined in terms of their temporal trajectory within the decline index continuum.
 
@@
 
+- Suitability of collected phenotypic descriptors and decline indexes for describing decline status
 - Crown condition descriptors most important for describing overall decline severity in the PDI.

### 2019-12-18 manuscript/manuscript.Rmd
finished conclusion

@@
 output: 
-  #word_document: default
-  pdf_document: default
+  word_document: default
+  # pdf_document: default
 bibliography: references.bib
@@
 
-- Suitability of collected phenotypic descriptors and decline indexes for describing decline status
-- Crown condition descriptors most important for describing overall decline severity in the PDI.
-- Bacterial bleed related descriptors are most important for describing decline acuteness.
-- Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment
+The decline indexes presented here provide a new framework in which to quantitative describe the phenotypic oak decline spectrum using intuitive measures severity (PDI) and type (DAI).
+Their basis is on simple phenotypic descriptor measuerments and a utilisation of a data driven, unsupervised random forest based machine learning approach.
+This has shown the relative contributions that aspects of tree size, crown condition, and symptoms such as stem bleeding have on quantitatively defining our concepts of oak decline status.
 
-The decline indexes presented here provided intuitive measures 
-The approach presented here is easily adaptable to incorporate any phenotype
-It is encouraged that this approach of deriving descriptive index measurse in other tree health phenotyping contexts.
+Strategies have been presented such as the use of LIME and simulated model response surfaces that can aid the interpretation of the complex descriptor interactions that underly these simple index measures and  greatly reduce the black box nature that is often associated with these approaches.
+It is encouraged that this approach of deriving descriptive index measures using machine learning is applied in other tree health phenotyping contexts and it is hoped that this will have an impact on future the consistency and comparability of oak health monitoring.
 

### 2019-12-18 manuscript/manuscript.Rmd
proofing and spell check

@@
   word_document: default
-  # pdf_document: default
+  pdf_document: default
 bibliography: references.bib
@@
 
-Oak decline is becoming increasingly prevalent along with the need for accurate quantitative oak health monitoring.
+Oak decline is becoming increasingly prevalent along with the need for accurate and quantitative oak health monitoring.
 The Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to quantitatively describe and differentiate acute and chronic oak decline severity using a multitude of easily measurable phenotypic descriptors.
-Measured phenotypic descriptors include aspects of tree shape and size, crown condition and the presence of biotic agents.
+Measured phenotypic descriptors included aspects of tree shape and size, crown condition and the presence of biotic agents.
 The decline indexes were derived using proximity values from n unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
@@
 
-The decline indexes presented here place single intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity.
-These two indexes provide a two dimensional oak decline continuum under which the spectrum of decline severity and type can be assessed (Figure 5).
-The index thresholds for category definitions (healthy, declining, severe decline etc.) shown there currently serve only as a guide to aid human interpretation as opposed to definitive thresholds of disease.
-Further work is needed that includes phenotypic data with greater spatial and temporal resolution would be needed to establishing definitive thresholds.
+The decline indexes presented here provide single intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity.
+These two indexes can be represented as a two dimensional oak decline continuum under which the spectrum of decline severity and type can be assessed (Figure 5).
+The index thresholds for category definitions (healthy, declining, severe decline etc.) shown currently serve only as a guide to aid human interpretation as opposed to definitive thresholds of disease.
+Further work is needed that includes phenotypic data with greater spatial and temporal to establish definitive thresholds.
 
 Crown condition and volume were shown to be the key contributors to the PDI. 
-This is to be expected and hoped for when using an unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007].
+This is to be expected and an indicator of model suitablity for the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007].
 However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
-It has been shown here that firstly factors of overal crown condition and size are unlikely to contribute equally to our visual definition of oak decline severity (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly to our visual definition of oak decline severity (Figure 3a).
+It has been shown here that in our visual definitions of oak decline severity; firstly, factors of overal crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).
 
-The high contribution of bleed prevalence related descriptors to the DAI is expected as this is one of the defining features of AOD [@denman2014]. 
+The high contribution of bleed prevalence related descriptors to the DAI is also expected as this is one of the defining features of AOD [@denman2014]. 
 There was no contribution of *A.
-biguttatus* exit hole occurance with with the DAI and therefore stem bleeding - smaller area surveyed compared to bleeding.
-It has also been found previouly that trees with bleeds do not always have exit holes as  [@brown2017a;@vansteenkiste2004].
+biguttatus* exit hole occurance with with the DAI and therefore stem bleeding.
+This could partly be due to smaller area surveyed compared to the stem bleed descriptors but it has also been identified previouly that trees with bleeds do not always have exit holes, which has been attributed to the fact that not all *A.
+biguttatus* larvae complete their life cycle [@brown2017a;@vansteenkiste2004].
 
-AOD severity could also be associated with overall decline severity with trees identified that had a combination of both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure 1).
+AOD severity was able to be associated with overall decline severity with trees identified that had a combination of both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure 1).
 The association of crown condition and AOD severity could not previously be shown by @denman2014 due to a low sample size.
@@
 
-A machine learning approach was used to derive the phenotypic index measures as it allows the moulded to the complex interactions and trends between phenotypic descriptors from which the decline indexes are calculated.
+A machine learning approach was used to derive the phenotypic index measures as it allows the models to be moulded to the complex interactions and trends between phenotypic descriptors from which the decline indexes are calculated.
 This would likely be oversimplified using conventional modelling strategies [@breiman2001b].
 
-An unsupervised approach was favoured in this context due to the concession that there know difficulties and inaccuracy of manual field decline status classification (ref).
+An unsupervised approach was favoured in this context due to the concession that there know difficulties and inaccuracy of manual decline status classification in the field [@ferretti1997].
 This is highlighted in Figure 1 where a numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (eg.
@@
 The unsupervised approach has worked in this context in aptly describing decline severity as the phenotypic descriptors selected were highly specific and well suited to describing the decline severity problem.
-In situations were descriptors may well be less suited, perhaps as a result of technical limitations, the use of a semi-supervised approach may be favourable (ref).
-The unsupervised approach to derive the decline indexes would be applied to a subset of the observations on which the descriptors are known to perform well and that would be considered benchmark observations by the investigators.
-Indexes could then be attained for the rest of the observations by prediction based on supervised models of the benchmark subset (ref).
+In situations were descriptors may well be less suited, perhaps as a result of technical limitations, the use of a semi-supervised approach may be favourable.
+The unsupervised approach to dearXivrive the decline indexes would be applied to a subset of the observations on which the descriptors are known to perform well and that would be considered benchmark observations by the investigators.
+Indexes could then be attained for the rest of the observations by prediction based on supervised models of the benchmark subset [@chapelle2010].
 This approach would be more suitable than a fully supervised approach as it would limit the inaccuacies that would be introduced by manual classification and still allow the data driving flexibility of the unsupervised approach.
 
-Machine learning approaches are usually seen as black box and a lack of model interpretability is a concession in this approach however the are substantial gains in terms of predictive capability for model application in real world situations (ref).
+Machine learning approaches are usually seen as black box and difficulty in model interpretation is a concession in this approach; however, there are substantial gains in terms of predictive capability for model application in real world situations [@touw2012].
 The assessment descriptor contributions to the index models are an integral part of this approach.
-Inout of knowledge of the biological problem by the investigator at this stage is essential in ensuring that the models suitably learn the patterns relevant biological problem to which they have been applied and that the descriptors contirbuting to these are sensible and that the algorithm (random forest) is not learning spurious or artificial patterns within the data (ref).
+Input of knowledge of the biological problem by the investigator at this stage is essential in ensuring that the models suitably learn the patterns relevant biological problem to which they have been applied.
+The descriptors contirbuting to these should be sensible and that the algorithm (random forest) is not learning spurious or artificial patterns within the data.
 Approaches such as LIME and model response surfaces enable investigators to query the contributions that phenotypic descriptors have in particular cases of interest.
-This inturn promotes trust in the indexes derived from these models and reduces their black box nature (ref).
+This inturn promotes trust in the indexes derived from these models and reduces their black box nature [@ribeiro2016].
 
@@
 The current context in which these decline indexes are presented is study specific and is based on a relatively small sample size that may not be generally representative of the UK oak population as a whole.
-Subsequent studies could incorporate the data from this study with additional data to train random forest models if comparable phenotyping protocols are used or generate study specific decline index values using the unsupervised learning approach.
+Subsequent studies could incorporate the data from this study with additional data to train random forest models if comparable phenotyping protocols are used.
+Alternatively, study specific decline index values could be generated using the unsupervised random forest approach.
 
 The representation of the decline severity spectrum provided by these decline indexes provides greatly enhanced sensitivity over categorical decline status assignments.
-This could enhance integration with large scale data such as those from genome wide association studies or remote sensing [@stocks2019] (ref).
+This could enhance integration with large scale data such as those from genome wide association studies or remote sensing [@stocks2019; @evans2012].
 These index measures allow sensitive comparisons of visual health to be made for the same tree between years as well as comparisons of trees between different locations.
 
-Scaling this technique to enable generalised naive computation of decline indexes for any tree sampled in the UK would require index models trained using a substantially larger data set that incorporates the descriptive diversity of oak decline in the population as whole (ref).
+Scaling this technique to enable generalised naive computation of decline indexes for any tree sampled in the UK would require index models trained using a substantially larger data set that incorporates the descriptive diversity of oak decline in the population as whole.
 Scaling to this level would also require an appropriate method of tree size correction that could be applied on an individual tree basis.
 A potential solution to this would be to correct trees based on estimated age from DBH as proposed by @white1998.
-These indexes could have great utility in long term forest health monitoring by providing comparability over time that could be utilised for both forestry management and research (ref).
+These indexes could have great utility in long term forest health monitoring by providing comparability over time that could be utilised for both forestry management and research [@denman2010].
 
-The phenotypic descriptors used to calculate these indexes are easily measurable in the field and require no specialist equipment and are based on well established tree condition assessment protocols [@innes1990].
-Refinement of the protocols used here could make could reduce survey time and complexity and potentially be incorporated into citizen science programmes as a resource to build general UK oak training set [@brown2017b]
-
+The phenotypic descriptors used to calculate these indexes are easily measurable in the field, requiring no specialist equipment and are based on well established tree condition assessment protocols [@innes1990].
+Refinement of the protocols used here could make could reduce survey time and complexity and potentially be incorporated into citizen science programmes as a resource to build general UK oak training set [@brown2017b].
 Alternatively, the machine learning approach used to derive these decline indexes could also be applied to long established oak health monitoring programmes to aid sensitivity in detecting overall trends in oak health.
@@
 
-The decline indexes presented here provide a new framework in which to quantitative describe the phenotypic oak decline spectrum using intuitive measures severity (PDI) and type (DAI).
+The decline indexes presented here provide a new framework in which to quantitatively describe the phenotypic oak decline spectrum using intuitive measures severity (PDI) and type (DAI).
 Their basis is on simple phenotypic descriptor measuerments and a utilisation of a data driven, unsupervised random forest based machine learning approach.
-This has shown the relative contributions that aspects of tree size, crown condition, and symptoms such as stem bleeding have on quantitatively defining our concepts of oak decline status.
+This has shown the relative contributions that aspects of tree size, crown condition and symptoms such as stem bleeding have on quantitatively defining our concepts of oak decline status.
 
-Strategies have been presented such as the use of LIME and simulated model response surfaces that can aid the interpretation of the complex descriptor interactions that underly these simple index measures and  greatly reduce the black box nature that is often associated with these approaches.
+Strategies have been presented such as the use of LIME and simulated model response surfaces that can aid the interpretation of the complex descriptor interactions that underpin these simple index measures and  greatly reduce the black box nature that is often associated with these approaches.
 It is encouraged that this approach of deriving descriptive index measures using machine learning is applied in other tree health phenotyping contexts and it is hoped that this will have an impact on future the consistency and comparability of oak health monitoring.

### 2019-12-18 R/plan.R

@@ -238,8 +238,10 @@ plan <- drake_plan(
   ## render figures
   figures = render(knitr_in('manuscript/figures.Rmd'),quiet = T),
   
-  ## render manuscript
-  manuscript = render(knitr_in('manuscript/manuscript.Rmd'),quiet = T),
+  ## render manuscript 
+  manuscript = render(knitr_in('manuscript/manuscript.Rmd'),
+                      quiet = T,
+                      output_format = 'all'),
   
   ## render supplementary
   supplementary = render(knitr_in('manuscript/supplementary.Rmd'),quiet = T)

### 2019-12-18 manuscript/manuscript.Rmd
spell check

@@
 Crown condition and volume were shown to be the key contributors to the PDI. 
-This is to be expected and an indicator of model suitablity for the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007].
+This is to be expected and an indicator of model suitability for the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007].
 However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
-It has been shown here that in our visual definitions of oak decline severity; firstly, factors of overal crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).
+It has been shown here that in our visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).
 
@@
 There was no contribution of *A.
-biguttatus* exit hole occurance with with the DAI and therefore stem bleeding.
-This could partly be due to smaller area surveyed compared to the stem bleed descriptors but it has also been identified previouly that trees with bleeds do not always have exit holes, which has been attributed to the fact that not all *A.
+biguttatus* exit hole occurrence with with the DAI and therefore stem bleeding.
+This could partly be due to smaller area surveyed compared to the stem bleed descriptors but it has also been identified previously that trees with bleeds do not always have exit holes, which has been attributed to the fact that not all *A.
 biguttatus* larvae complete their life cycle [@brown2017a;@vansteenkiste2004].
@@
 
-It was shown that tree size, reflected by the live crown ratio descriptor, contributes most to the DAI in descriminating between COD and AOD trees.
+It was shown that tree size, reflected by the live crown ratio descriptor, contributes most to the DAI in discriminating between COD and AOD trees.
 The relationship of relative tree stature and the acuteness of oak decline has not previously been shown and likely reflects the period of time under which these trees have been afflicted. 
@@
 However, this was also the most varied of the decline status groups.
-It is possible that this could reflect an underly relationship of the potential for trees to recover from AOD but this is speculative and it is not possible to attribute this further with these observations.
+It is possible that this could reflect an underpin relationship of the potential for trees to recover from AOD but this is speculative and it is not possible to attribute this further with these observations.
 Using these decline indexes, remission trees can be more objectively defined in terms of their temporal trajectory within the decline index continuum.
@@
 An unsupervised approach was favoured in this context due to the concession that there know difficulties and inaccuracy of manual decline status classification in the field [@ferretti1997].
-This is highlighted in Figure 1 where a numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (eg.
+This is highlighted in Figure 1 where a numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g..
 COD trees with DAI values > 0.5).
@@
 In situations were descriptors may well be less suited, perhaps as a result of technical limitations, the use of a semi-supervised approach may be favourable.
-The unsupervised approach to dearXivrive the decline indexes would be applied to a subset of the observations on which the descriptors are known to perform well and that would be considered benchmark observations by the investigators.
+The unsupervised approach to derive the decline indexes would be applied to a subset of the observations on which the descriptors are known to perform well and that would be considered benchmark observations by the investigators.
 Indexes could then be attained for the rest of the observations by prediction based on supervised models of the benchmark subset [@chapelle2010].
-This approach would be more suitable than a fully supervised approach as it would limit the inaccuacies that would be introduced by manual classification and still allow the data driving flexibility of the unsupervised approach.
+This approach would be more suitable than a fully supervised approach as it would limit the inaccuracies that would be introduced by manual classification and still allow the data driving flexibility of the unsupervised approach.
 
@@
 Input of knowledge of the biological problem by the investigator at this stage is essential in ensuring that the models suitably learn the patterns relevant biological problem to which they have been applied.
-The descriptors contirbuting to these should be sensible and that the algorithm (random forest) is not learning spurious or artificial patterns within the data.
+The descriptors contributing to these should be sensible and that the algorithm (random forest) is not learning spurious or artificial patterns within the data.
 Approaches such as LIME and model response surfaces enable investigators to query the contributions that phenotypic descriptors have in particular cases of interest.
-This inturn promotes trust in the indexes derived from these models and reduces their black box nature [@ribeiro2016].
+This in turn promotes trust in the indexes derived from these models and reduces their black box nature [@ribeiro2016].
 
@@
 The decline indexes presented here provide a new framework in which to quantitatively describe the phenotypic oak decline spectrum using intuitive measures severity (PDI) and type (DAI).
-Their basis is on simple phenotypic descriptor measuerments and a utilisation of a data driven, unsupervised random forest based machine learning approach.
+Their basis is on simple phenotypic descriptor measurements and a utilisation of a data driven, unsupervised random forest based machine learning approach.
 This has shown the relative contributions that aspects of tree size, crown condition and symptoms such as stem bleeding have on quantitatively defining our concepts of oak decline status.

### 2019-12-18 manuscript/figures.Rmd
proof figure and table captions

@@
 
-```{r descriptor_contributions, fig.height = 6,fig.cap = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures.
-a) Phenotypic decline index (PDI).
-b) Decline acuteness index (DAI)'}
+```{r descriptor_contributions, fig.height = 6,fig.cap = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures.'}
 loadd(descriptor_contribution_plots)
@@
 
-```{r PDI_response_surfaces,fig.height = 6,fig.cap='PDI response surfaces for simulated descriptor values from trained Random Forest models.
-All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models'}
+```{r PDI_response_surfaces,fig.height = 6,fig.cap='PDI response surfaces for simulated descriptor values from trained random forest models.'}
 loadd(PDI_response_surfaces)
@@
 
-```{r DAI_response_surfaces,fig.height = 8,fig.cap='DAI response surfaces for simulated descriptor values from trained Random Forest models.
-All other descriptors were set at the values of those of the trees with the lowest PDI used for trainig the models'}
+```{r DAI_response_surfaces,fig.height = 8,fig.cap='DAI response surfaces for simulated descriptor values from trained random forest models.'}
 loadd(DAI_response_surfaces)
@@
 
-```{r DI_space,fig.cap='Two dimensional continuum of the phenotypic decline indexes for oak decline severity.',fig.height=3,fig.width=5}
+```{r DI_space,fig.cap='Two dimensional continuum of the phenotypic decline indexes for oak decline severity.
+Decline category boundaries intended as a guide to aid interpretation only.
+Grey area indicates area for which the decline indexes are currently undefined.' ,fig.height=3,fig.width=5}
 DIspace()

### 2019-12-18 manuscript/supplementary.Rmd
proof figure and table captions

@@
 
-```{r descriptor_adjustment,fig.cap='Exapmle of overall mean adjustment applied to the diameter at breast height descriptor. Error bars show ±1SE of the mean. Dashed red line shows the overall mean.'}
+```{r descriptor_adjustment,fig.cap='Exapmle of overall mean adjustment applied to the diameter at breast height (m) descriptor.
+Error bars show ±1SE of the mean.
+Dashed red line shows the overall mean.'}
 loadd(descriptor_adjustment_example)
@@
 Mean margin value shown.
-b) Relative descriptor Gini importances measures.',fig.height=6}
+b) Descriptor importances measures.',fig.height=6}
 loadd(site_differences_mds_plot)

### 2019-12-23 manuscript/tables.Rmd

--- 2019-12-12 manuscript/tables.Rmd
@@
 PDI_lime_analysis_table %>%
-  kable(format = 'latex',booktabs = TRUE,escape = FALSE,caption = ' ') %>%
+  kable(format = 'latex',
+        booktabs = TRUE,
+        escape = FALSE,
+        caption = 'LIME results for oak decline severity cases in the PDI index model. Top four important descriptors show. Descriptors ordered by descending absolute weight within each case.') %>%
   collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
@@
 DAI_lime_analysis_table %>%
-  kable(format = 'latex',booktabs = TRUE,escape = FALSE,caption = ' ') %>%
-  collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
+  kable(format = 'latex',
+        booktabs = TRUE,
+        escape = FALSE,
+        caption = 'LIME results of oak decline type and severity in the DAI index model. Top four important descriptors shown. Descriptors ordered by descending absolute weight within each case.') %>%
+  column_spec(5,width = '4cm') %>%
+  collapse_rows(columns = 1:4, latex_hline = "custom", valign = "middle",custom_latex_hline = 1:4)
 ```

### 2020-01-08 manuscript/manuscript.Rmd
Nathan additions to introduction

@@
 
-Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change [@sturrock2011.
+Trees and forests across the globe face threats from a changing climate, notably warming and an increase in drought episodes affect the health status of individual trees causing vegetation changes at the landscape scale [@choat2012; @millar2015; @allen2015].
+Within this context biotic agents (Pests and Diseases) can flourish and often drive mortality [@kautz2017; @jactel2012]. 
+Monitoring vegetation changes and understanding the underlying processes affecting tree health are important first steps to predicting the scale and extent of these threats [@hartmann2018] and plan future policies for planting and mitigation efforts [@bradford2017].
+Current ‘omics and remote sensing approaches provide powerful tools to document and understand these changes across multiple scales (from cellular to tree to landscape scales), but an important prerequisite is the ability to accurately describe tree health status, so that even subtle changes can be documented [@dungey2018; @hornero2020].
+Phenotyping, documenting the physical traits of an individual or group, is thus a crucial first step relevant to studies across disciplines and further, it is necessary to understand how these measurements combine to describe health status.
+Here, we present a method using unsupervised machine learning (Random Forest) to quantify both the extent and form of decline observed in oak forests across England.
+
+Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change [@sturrock2011; @petr2014; @brown2018].
 Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
 Abiotic factors can include both predisposing and inciting factors such as drought.
-Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014].
+Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].
 
-As a results of this, there are many visual phenotypic descriptors encompassing the entirety of an oak tree that contribute to what can be defined as a declining oak tree.
-A severely declining oak is most prominently characterised by very poor crown condition and numerous dead branches.
-It is these characteristics that many studies use as the primary indicator of tree decline severity [@zarnoch2004].
-
-Compounding the complexity of oak decline syndromes is the addition of different types of oak decline associated with *Quercus robur* and *Q.
-petraea* that are often categorised based on the duration over which a tree has been declining and the biotic agents associated with a declining tree.
-Mortality in trees suffering from Chronic Oak Decline (COD) can occur over a period of decades where as AOD can occur in 5-10 years [@denman2009].
+Oak decline syndromes are wide ranging in cause and attemps have been made to subdivide the types of decline associated with *Quercus robur* and *Q.
+petraea* based on the duration over which a tree has been declining and the biotic agents associated with a declining tree.
+Mortality in trees suffering from COD can occur over a period of decades where as AOD can occur in 5-10 years [@denman2009].
 A chronically declining oak can also subsequently begin to develop the stem bleeds associated with AOD.
@@
 
+As a result of this, there are many visual phenotypic descriptors that contribute to categorizing an oak tree as being in decline. 
+When the decline is severe there are clear symptoms visible in the crown, such as low foliage density and numerous dead branches. 
+It is these characteristics that many studies use as the primary indicator of tree decline severity [@zarnoch2004], however these symptoms can develop slowly (especially in the case of COD) and simple classes are unlikely to capture the full detail of the subtle transition from being healthy to declined.
+
 Assigning discrete decline status labels neglects the true spectrum of decline severity encountered in the field.
 There is a growing need for sensitive, continuous measures of tree decline that accurately reflect the decline spectrum [@pontius2014].
-This has the potential to greatly improve monitoring to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices.
+This has the potential to greatly improve monitoring, enable the early detection of decline for accurate forecasting and inform forestry management practices.
 The recently published Action Oak Knowledge Review highlighted monitoring long term consistently as a key requirement for safeguarding UK oak health in the future [@quine2019].
@@
 
-## Checks
-
 ```{r check,comment='',cache=FALSE}
-manuscript_check(checks = c('word count','spelling'))
+# manuscript_check(checks = c('word count','spelling'))
 ```

### 2020-01-09 manuscript/manuscript.Rmd
Nathan alterations to methods

@@
 
-It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag sample prediction accuracies and variable importance.
+Random forest is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag sample prediction accuracies and variable importance.
 It is also able to simultaneously handle both continuous and categorical variables.
@@
 
-The phenotypic data were collected from a total of 174 trees from nine sites from across the south (Hatchlands and Richmond Park), east (Big Wood, Great Monks Wood and Winding Wood) and west (Attingham, Chestnuts, Langdale and Speculation) of England between June and September in 2016 or 2017.
-The occurrence of AOD has previously been reported and studied at six of the sites [@brown2016; @brown2017a].
+The phenotypic data were collected over two summers between June and September from a total of 174 trees at nine sites across the south of England.
+In 2016 three COD sites were visited (Big Wood,Chestnuts and Speculation) and in 2017 six AOD sites were added to the dataset (Attingham, Great Monks Wood, Hatchlands, Langdale Wood, Richmond Park and Winding Wood).
+The occurrence of AOD has previously been reported and studied at these sites [@brown2016; @brown2017a].
 These sites included a variety of woodland, management and soil types as well as tree sizes/ages and were made up of mainly *Q. robur*.
@@
 All trees were manually assigned a decline status category; healthy, AOD, COD or remission prior to collection of phenotypic descriptors.
+At the COD sites trees were either COD (trees with clear signs of poor crown condition and canopy dieback, at Big Wood trees with stronger crowns were categorised as COD when *Armillaria* fruiting bodies had been observed) or healthy (dominant or co-dominant trees with full crowns).
+At the AOD sites, annual monitoring data collected since 2009 was used to guide tree selection.
+Healthy trees had no know history of AOD and fruiting bodies of root pathogens had not been observed, all trees had canopy condition of 4 or 5, see @brown2016 for details of this assessment.
+AOD trees had current stem bleeds but a variety of crown conditions. 
+COD trees either had poor crowns (category 2) or the fruiting bodies of *Gymnopus* or *Armillaria* had been observed in previous years.
 The remission group included trees that had shown stem bleeds associated with AOD within the previous three years but were not showing signs of stem bleeds at the time of assessment.
+Thus at all sites, sampling attempted to include the unhealthiest and healthiest trees.
 
 The phenotypic descriptors collected related to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
-The descriptors collected were based on @innes1990.
+The descriptors collected were based on @icpforests2016.
 Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
@@
 
-Individual random forest regression models of the phenotypic descriptors were trained, supervised using each decline index.
-The optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
-The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: PDI - `r glue('ntree = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; DAI - `r glue('ntree = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
+To investigate the descriptor contributions to the decline indexes, supervised random forest regression models were trained for each decline index using the phenotypic descriptors for all 174 trees.
+The optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest model were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
+The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: for the PDI - `r glue('ntree = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI - `r glue('ntree = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
 Overall descriptor importance contributions to each of the indexes was measured in each model using the mean decrease in accuracy measure.

### 2020-01-10 manuscript/manuscript.Rmd
Further Nathan suggestions

@@
 This has the potential to greatly improve monitoring, enable the early detection of decline for accurate forecasting and inform forestry management practices.
-The recently published Action Oak Knowledge Review highlighted monitoring long term consistently as a key requirement for safeguarding UK oak health in the future [@quine2019].
+Consistent long term monitoring has been higlited as a key requirement for understanding the threats to oak and safeguarding its future [@hartmann2018].
 Coupled with this, monitoring methods should take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.
@@
 
-Local Interpretable Model-Agnostic Explanations (LIME) were computed to identify descriptors contributing locally to case specific examples of each of the indexes [@ribeiro2016]. 
-The lime R package version `r packageVersion('lime') %>% as.character()` was used with the default parameters [@pedersen2019].
-A healthy, a moderate and a severe decline case was used for the PDI. 
-For each case, a tree with a score closest to 0, 0.5 and 1 respectively was selected.
-For the DAI cases were selected from trees with a PDI score greater than 0.5, 
-A decline neutral case was selected with a score closest to 0.
-Then for each of the AOD and COD syndromes, moderate and severe cases were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD respectively.
+To identify which phenotypic descriptors contributed locally to specific example cases of each of the decline indexes, Local Interpretable Model-Agnostic Explanations (LIME) were computed [@ribeiro2016].
+For the PDI, the example cases consisted of selecting three trees scores closest to 0 (healthy), 0.5 (moderate) and 1 (severely declining) respectively.
+For the DAI, five example cases were selected from trees with PDI scores greater than 0.5.
+These cases incuded a decline neutral case with a score closest to 0 and then for each of the AOD and COD syndromes, a moderate and a severe case were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD respectively.
+LIME analysis was conducted using the lime R package version `r packageVersion('lime') %>% as.character()` with the default parameters [@pedersen2019].
 

### 2020-01-22 manuscript/manuscript.Rmd
altered LIME methods to include the defined example cases

@@
 To identify which phenotypic descriptors contributed locally to specific example cases of each of the decline indexes, Local Interpretable Model-Agnostic Explanations (LIME) were computed [@ribeiro2016].
-For the PDI, the example cases consisted of selecting three trees scores closest to 0 (healthy), 0.5 (moderate) and 1 (severely declining) respectively.
+The example cases included the trees with the minimum (0) and maximum (1) PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
 For the DAI, five example cases were selected from trees with PDI scores greater than 0.5.

### 2020-01-26 manuscript/figures.Rmd

@@
 
-grid.arrange(PDI_response_surfaces$`a)`,
-             PDI_response_surfaces$`b)`,
-             PDI_response_surfaces$`c)`,
+grid.arrange(PDI_response_surfaces[[1]],
+             PDI_response_surfaces[[2]],
+             PDI_response_surfaces[[3]],
              PDI_response_surfaces$legend,

### 2020-01-26 manuscript/manuscript.Rmd

@@
 
-The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model response surfaces for specific cases.
+The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model response surfaces for specific example cases.
+The same example cases were used as described for the LIME analyses above.
 This approach was similar to response surface methodology without the need for probabilistic approximation due to the low computation time required for index calculation [@abbasi2012].
@@
 For the PDI, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severely tree.
-The example cases included the trees with the minimum and maximum PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
 The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors were divided into 100 increments.
@@
 The upper boundary for the ranges of active bleed size and black stain size was taken as the centre value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.
-Cases included neutral, moderate and severe with PDI scores greater than 0.5 and DAI scores closest to 0, 0.5 and 1 respectively for AOD and 0, -0.5 and -1 respectively for COD.
 

### 2020-01-26 manuscript/manuscript.Rmd
Nathan alterations to site differences results section

@@
 
-Oak decline syndromes are wide ranging in cause and attemps have been made to subdivide the types of decline associated with *Quercus robur* and *Q.
+Oak decline syndromes are wide ranging in cause and attempts have been made to subdivide the types of decline associated with *Quercus robur* and *Q.
 petraea* based on the duration over which a tree has been declining and the biotic agents associated with a declining tree.
@@
 This has the potential to greatly improve monitoring, enable the early detection of decline for accurate forecasting and inform forestry management practices.
-Consistent long term monitoring has been higlited as a key requirement for understanding the threats to oak and safeguarding its future [@hartmann2018].
+Consistent long term monitoring has been highlighted as a key requirement for understanding the threats to oak and safeguarding its future [@hartmann2018].
 Coupled with this, monitoring methods should take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.
@@
 For the DAI, five example cases were selected from trees with PDI scores greater than 0.5.
-These cases incuded a decline neutral case with a score closest to 0 and then for each of the AOD and COD syndromes, a moderate and a severe case were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD respectively.
+These cases included a decline neutral case with a score closest to 0 and then for each of the AOD and COD syndromes, a moderate and a severe case were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD respectively.
 LIME analysis was conducted using the lime R package version `r packageVersion('lime') %>% as.character()` with the default parameters [@pedersen2019].
@@
 
-High discrimination was found in the phenotypic descriptors between the sites, which was investigated using supervised random forest classification (Supplementary  figure 3a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`).
+A preliminary analysis using supervised random forest classification could separate trees based on their location (Supplementary  figure 3a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`).
 The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary figure 3b).
-After the application of trees size adjustment to removed inter-site variability but maintain intra-site variability, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
-This enabled subsequent comparisons to be made between the trees at the different sites in the context of decline severity.
+To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in the Materials and Methods.
+After the application of the tree size adjustment, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
 

### 2020-01-27 manuscript/manuscript.Rmd
added surveyors to acknowledgements

@@
       remission_DAI_group_ttest,
-      DAI_descriptor_importance)
+      DAI_descriptor_importance,
+      surveyors_names)
 ```
@@
 
-We would like to thank the Forest Research TSU for collecting the phenotypic data presented in this article especially Liz Richardson and Mark Oram for coordinating surveying.
+We would like to thank the Forest Research TSU for collecting the phenotypic data presented in this article.
+Thanks to Liz Richardson and Mark Oram for coordinating surveying and the surveyors: `r str_c(surveyors_names,collapse = ', ')`.
 This work was supported by Woodland Heritage and the Future Proofing Plant Health (FPPH) Oak Resilience project.

### 2020-01-27 R/plan.R

@@ -241,6 +241,9 @@ plan <- drake_plan(
                                               DAI_example_cases,
                                               site_corrected_analysis_suitable_data),
   
+  ## Surveyor names
+  surveyors_names = surveyors(),
+  
   ## render tables
   tables = render(knitr_in('manuscript/tables.Rmd'),quiet = T),
   

### 2020-01-27 manuscript/manuscript.Rmd
various small alterations to introduction and M&M

@@
 Measured phenotypic descriptors included aspects of tree shape and size, crown condition and the presence of biotic agents.
-The decline indexes were derived using proximity values from n unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
+The decline indexes were derived using proximity values from unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
 These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a potentially useful tool for forest monitoring and management.
@@
 
-Trees and forests across the globe face threats from a changing climate, notably warming and an increase in drought episodes affect the health status of individual trees causing vegetation changes at the landscape scale [@choat2012; @millar2015; @allen2015].
-Within this context biotic agents (Pests and Diseases) can flourish and often drive mortality [@kautz2017; @jactel2012]. 
+Trees and forests across the globe face threats from a changing climate, notably warming and an increase in drought episodes affect the health status of individual trees, causing vegetation changes at the landscape scale [@choat2012; @millar2015; @allen2015].
+Within this context biotic agents (pests and diseases) can flourish and often drive mortality [@kautz2017; @jactel2012]. 
 Monitoring vegetation changes and understanding the underlying processes affecting tree health are important first steps to predicting the scale and extent of these threats [@hartmann2018] and plan future policies for planting and mitigation efforts [@bradford2017].
@@
 Phenotyping, documenting the physical traits of an individual or group, is thus a crucial first step relevant to studies across disciplines and further, it is necessary to understand how these measurements combine to describe health status.
-Here, we present a method using unsupervised machine learning (Random Forest) to quantify both the extent and form of decline observed in oak forests across England.
 
@@
 Machine learning is finding increasing utility in ecological and environmental applications due to it's ability to learn complex patterns within and between variables [@cutler2007;@belgiu2016].
-A popular algorithm is that of random forest due to it's versatility, relatively trivial training requirements and ease of use [@breiman2001a;@biau2016].
+A popular algorithm is that of random forest due to it's versatility, relatively trivial training requirements and ease of use. 
+It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag sample prediction accuracies and variable importance [@breiman2001a;@biau2016].
 
-Random forest is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag sample prediction accuracies and variable importance.
-It is also able to simultaneously handle both continuous and categorical variables.
-Unsupervised learning is possible using the assumption that if the data contains any underlying structure, this should be distinguishable from a randomly generated version of itself.
-An observation proximity matrix can be derived from this process on which additional unsupervised methods can be applied such as multidimensional scaling [@shi2006].
+Here, we propose a framework to enable quantitative visible oak decline severity measurement that utilises a range of targeted phenotypic descriptors observed in oak forests across England and unsupervised random forest machine learning.
 
-Here, we propose a framework to enable quantitative visible oak decline severity measurement that utilises a range of targeted phenotypic descriptors and unsupervised random forest machine learning.
 The two resulting measures are:
@@
 All trees were manually assigned a decline status category; healthy, AOD, COD or remission prior to collection of phenotypic descriptors.
-At the COD sites trees were either COD (trees with clear signs of poor crown condition and canopy dieback, at Big Wood trees with stronger crowns were categorised as COD when *Armillaria* fruiting bodies had been observed) or healthy (dominant or co-dominant trees with full crowns).
+At the COD sites trees were either COD (trees with clear signs of poor crown condition and canopy dieback) or healthy (dominant or co-dominant trees with full crowns). 
+At Big Wood trees with stronger crowns were categorised as COD when *Armillaria* fruiting bodies had been observed.
 At the AOD sites, annual monitoring data collected since 2009 was used to guide tree selection.
@@
 This was then transformed to a dissimilarity matrix ($1 - P$) and scaled to two dimensions using  multidimensional scaling (MDS).
-The first MDS dimension ($D$) was first inverted ($1 - D$) to ensure the correct dimension orientation relative to decline severity.
+The first MDS dimension ($D$) was inverted ($1 - D$) to ensure the correct dimensional orientation relative to decline severity.
 It was then min-max scaled to give a scale between 0 and 1 with more severely declining trees having a score closer to 1.
@@
 To investigate the descriptor contributions to the decline indexes, supervised random forest regression models were trained for each decline index using the phenotypic descriptors for all 174 trees.
+These models will subsequently be referred to as $M_1$ and $M_2$ for the PDI and DAI respectively. 
 The optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest model were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
-The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: for the PDI - `r glue('ntree = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI - `r glue('ntree = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
+The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: for the PDI ($M_1$) - `r glue('ntree = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_2$) - `r glue('ntree = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
 Overall descriptor importance contributions to each of the indexes was measured in each model using the mean decrease in accuracy measure.
@@
 
-To identify which phenotypic descriptors contributed locally to specific example cases of each of the decline indexes, Local Interpretable Model-Agnostic Explanations (LIME) were computed [@ribeiro2016].
-The example cases included the trees with the minimum (0) and maximum (1) PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
-For the DAI, five example cases were selected from trees with PDI scores greater than 0.5.
+To identify which phenotypic descriptors contributed locally to specific example cases of each of the decline indexes, Local Interpretable Model-Agnostic Explanations (LIME) were computed for each of models $M_1$ and $M_2$ [@ribeiro2016].
+The example cases for the PDI used for $M_1$ included the trees with the minimum (0) and maximum (1) PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
+For the DAI ($M_2$), five example cases were selected from trees with PDI scores greater than 0.5.
 These cases included a decline neutral case with a score closest to 0 and then for each of the AOD and COD syndromes, a moderate and a severe case were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD respectively.

### 2020-02-03 manuscript/manuscript.Rmd
alterations to model reference notation

@@
 To investigate the descriptor contributions to the decline indexes, supervised random forest regression models were trained for each decline index using the phenotypic descriptors for all 174 trees.
-These models will subsequently be referred to as $M_1$ and $M_2$ for the PDI and DAI respectively. 
-The optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest model were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
-The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: for the PDI ($M_1$) - `r glue('ntree = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_2$) - `r glue('ntree = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
+These models will subsequently be referred to as $M_{PDI}$ and $M_{DAI}$ for the PDI and DAI respectively. 
+The optimal number of trees ($ntree$) and number randomly of variables ($mtry$) parameters for random forest model were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
+The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
 Overall descriptor importance contributions to each of the indexes was measured in each model using the mean decrease in accuracy measure.
@@
 
-To identify which phenotypic descriptors contributed locally to specific example cases of each of the decline indexes, Local Interpretable Model-Agnostic Explanations (LIME) were computed for each of models $M_1$ and $M_2$ [@ribeiro2016].
-The example cases for the PDI used for $M_1$ included the trees with the minimum (0) and maximum (1) PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
-For the DAI ($M_2$), five example cases were selected from trees with PDI scores greater than 0.5.
+To identify which phenotypic descriptors contributed locally to specific example cases of each of the decline indexes, Local Interpretable Model-Agnostic Explanations (LIME) were computed for each of models $M_{PDI}$ and $M_{DAI}$ [@ribeiro2016].
+The example cases for the PDI used for $M_{PDI}$ included the trees with the minimum (0) and maximum (1) PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
+For the DAI ($M_{DAI}$), five example cases were selected from trees with PDI scores greater than 0.5.
 These cases included a decline neutral case with a score closest to 0 and then for each of the AOD and COD syndromes, a moderate and a severe case were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD respectively.

### 2020-02-06 manuscript/manuscript.Rmd
proofing and finished alterations from Nathans comments

@@
 
-The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model response surfaces for specific example cases.
+The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model ($M_{PDI}$,$M_{DAI}$) response surfaces for specific example cases.
 The same example cases were used as described for the LIME analyses above.
@@
 
-For the PDI, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severely tree.
+For $M_{PDI}$, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severely tree.
 The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors were divided into 100 increments.
-For a given case, each combination of theses simulated descriptor values were used to calculate a PDI value from the supervised random forest regression model, building a model response surface.
+For a given case, each combination of theses simulated descriptor values were used to calculate a PDI value from $M_{PDI}$, building a model response surface.
 
@@
 
-The same approach was used for the DAI using also a comparison between active bleed size and black staining size as well as missing crown and crown transparency.
+The same approach was used for $M_{DAI}$ using a comparison between active bleed size and black staining size as well as missing crown and crown transparency.
 The upper boundary for the ranges of active bleed size and black stain size was taken as the centre value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.
@@
 
-### Site differences
+### Phenotypic differences between site locations
 
@@
 To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in the Materials and Methods.
-After the application of the tree size adjustment, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.
+After the application of the tree size adjustment, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the adjustment was able to remove the majority of the between site variance.
 
@@
 
-The top four ranked descriptors found to contribute to the PDI index model were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure 2a).
+The top four ranked descriptors found to contribute to $M_{PDI}$ were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure 2a).
 Other tree size related descriptors such as and live crown ratio stem bleeding descriptors contributed secondarily.
 
-The DAI was more complex in descriptor contribution many more descriptors contributing more equally than the PDI (Figure 2b).
-The overall tree size descriptors live crown ratio, DBH and crown volume contributed almost equally to the DAI index model.
+$M_{DAI}$ was more complex in descriptor contribution many more descriptors contributing more equally than the PDI (Figure 2b).
+The overall tree size descriptors live crown ratio, DBH and crown volume contributed almost equally to $M_{DAI}$.
 These were followed closely by the stem staining and bleeding descriptors such as bleed prevalence. 
 The presence of *A.
-biguttatus* exit holes had very little contribution to the DAI with exit hole density ranked only `r which(DAI_descriptor_importance$Feature == 'Agrilus exit hole density (m^-2)')`^th^.
+biguttatus* exit holes had very little contribution to the DAI with exit hole density ranked only `r which(DAI_descriptor_importance$Feature == 'Agrilus exit hole density (m^-2)')`th.
 
@@
 
-LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to the PDI in each of the healthy, moderate and severe decline cases using lime (Table 1).
+LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to $M_{PDI}$ in each of the healthy, moderate and severe decline cases using lime (Table 1).
 These descriptors were also the same top descriptors in the overall model importance lists for the PDI (Figure 2a).
@@
 
-For the DAI, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table 2).
+For $M_{DAI}$, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table 2).
 This likely reflects the greater number of descriptors contributing to the DAI with different descriptors relevant to particular decline contexts.
@@
 
-The complex descriptor interactions within the index models can be investigated by simulating phenotypic data and generating the index values from the supervised index random forest regression models.
-The index values generated by the models from these simulated phenotypic data can then be visualised as response surface plots, from which descriptor interactions within the indexes can be interpreted.
+The complex descriptor interactions within the index models ($M_{PDI}$,$M_{DAI}$) investigated by simulating phenotypic data and generating the index values from the supervised index random forest regression models.
+The index values generated by the models from these simulated phenotypic data were then visualised as response surface plots, from which descriptor interactions within the indexes can be interpreted. 
 Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for key phenotypic descriptors that contribute highly to the PDI and DAI.
 
-Crown condition descriptors such as composite crown volume and crown condition contributed highly to the PDI (Figure 2a).
+Crown condition descriptors such as composite crown volume and crown condition contributed highly to $M_{PDI}$ (Figure 2a).
 The phenotypic measures central to this were percentage missing crown and percentage crown transparency from which crown condition and crown volume are calculated. 
@@
 
-The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, DBH and crown condition descriptors. 
+The DAI ($M_{DAI}$) is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, DBH and crown condition descriptors. 
 Simulated response surfaces for the DAI are shown in Figure 4. 
@@
 Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
-Also higher DAI values could be obtained from trees with greater stature (Figure 4c). 
+Higher DAI values could also be obtained from trees with greater stature (Figure 4c). 
 Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure 4d).
@@
 
-Crown condition and volume were shown to be the key contributors to the PDI. 
-This is to be expected and an indicator of model suitability for the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007].
+Crown condition and volume were shown to be the key contributors to the $M_{PDI}$. 
+This is to be expected and are an indicator of model suitability for the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007].
 However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
@@
 
-The high contribution of bleed prevalence related descriptors to the DAI is also expected as this is one of the defining features of AOD [@denman2014]. 
+The high contribution of bleed prevalence related descriptors to $M_{DAI}$ is also expected as this is one of the defining features of AOD [@denman2014]. 
 There was no contribution of *A.
@@
 AOD severity was able to be associated with overall decline severity with trees identified that had a combination of both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure 1).
-The association of crown condition and AOD severity could not previously be shown by @denman2014 due to a low sample size.
+The association of crown condition and AOD severity has previously be shown by @brown2016.
 
@@
 However, this was also the most varied of the decline status groups.
-It is possible that this could reflect an underpin relationship of the potential for trees to recover from AOD but this is speculative and it is not possible to attribute this further with these observations.
-Using these decline indexes, remission trees can be more objectively defined in terms of their temporal trajectory within the decline index continuum.
+It is possible that this could reflect the potential for trees to recover from AOD but this is speculative and it is not possible to attribute this further with these observations.
+Using these decline indexes, remission trees could be more objectively defined in terms of their temporal trajectory within the decline index continuum.
 
@@
 An unsupervised approach was favoured in this context due to the concession that there know difficulties and inaccuracy of manual decline status classification in the field [@ferretti1997].
-This is highlighted in Figure 1 where a numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g..
+This is highlighted in Figure 1 where a numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g.
 COD trees with DAI values > 0.5).
@@
 Indexes could then be attained for the rest of the observations by prediction based on supervised models of the benchmark subset [@chapelle2010].
-This approach would be more suitable than a fully supervised approach as it would limit the inaccuracies that would be introduced by manual classification and still allow the data driving flexibility of the unsupervised approach.
+This approach would be more suitable than a fully supervised approach as it would limit the inaccuracies that would be introduced by manual classification and still allow the data driven flexibility of the unsupervised approach.
 

### 2020-02-07 manuscript/manuscript.Rmd
added author affiliations, line numbers and line spacing

@@
 ---
-title: "Index measures for oak decline severity using phenotypic descriptors"
 output: 
@@
   pdf_document: default
+header-includes:
+  - \linespread{2}
+  - \usepackage{lineno}
+  - \linenumbers
 bibliography: references.bib
@@
 
-**Authors:** Jasen Finch, Nathan Brown, John Draper & Sandra Denman
+**Proposed journal:** (1) Forest Ecology & Management (2) Forestry (3) European Journal of Forest Research (4) Plant Methods
 
-**Proposed journal:** Still undecided! ??Plant Methods, Forestry??
+# Title: Index measures for oak decline severity using phenotypic descriptors
+
+**Authors:** Jasen Finch^1\*^, Nathan Brown^2^, Manfred Beckmann^1^, Sandra Denman^2^ & John Draper^1^
+
+^\*^ Corresponding author
+
+**Addresses:**
+
+1. Institute of biological, environmental and rural sciences, Aberystwyth University, Aberystwyth, SY23 3DA, UK
+
+2. Centre for Forestry and Climate Change, Forest Research, Alice Holt Lodge, Farnham GU 4LH, UK
 
@@
 These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a potentially useful tool for forest monitoring and management.
+
+\newpage
 

### 2020-02-07 manuscript/manuscript.Rmd
small proofing changes

@@
 
-1. Institute of biological, environmental and rural sciences, Aberystwyth University, Aberystwyth, SY23 3DA, UK
+^1^ Institute of Biological, Environmental and Rural Sciences, Aberystwyth University, Aberystwyth, SY23 3DA, UK
 
-2. Centre for Forestry and Climate Change, Forest Research, Alice Holt Lodge, Farnham GU 4LH, UK
+^2^ Centre for Forestry and Climate Change, Forest Research, Alice Holt Lodge, Farnham GU 4LH, UK
 

### 2020-02-07 manuscript/manuscript.Rmd
small formatting additions

@@
 
-# Title: Index measures for oak decline severity using phenotypic descriptors
+# Title: Index measures for oak decline severity using phenotypic descriptors {-}
 
@@
 
-## Abstract
+## Abstract {-}
 
@@
 petraea* based on the duration over which a tree has been declining and the biotic agents associated with a declining tree.
-Mortality in trees suffering from COD can occur over a period of decades where as AOD can occur in 5-10 years [@denman2009].
+Mortality in trees suffering from COD can occur over a period of decades whereas AOD can occur in 5-10 years [@denman2009].
 A chronically declining oak can also subsequently begin to develop the stem bleeds associated with AOD.
@@
 Canopy closure recorded as yes or no depending on whether any other crowns are touching the sampling tree.
-A high resolution photo was taken of the vase to the spring of the crown and of the stem from the base to a height of 3m for each cardinal point for later reference.
+A high resolution photo was taken of the base to the top of the crown and of the stem from the base to a height of 3m for each cardinal point for later reference.
 
@@
 
-Although significance was obtained between the decline status groups on both the PDI and DAI scales, Figure 5 highlights the inaccuracies in the manual decline status classifications when viewed in an individual tree context.
+Although significance was obtained between the decline status groups on both the PDI and DAI scales, Figure 1 highlights the inaccuracies in the manual decline status classifications when viewed in an individual tree context.
 It can be seen that a number of the trees, when compared quantitatively using these measures, could in fact be classified with a different decline state.
@@
 Plots of the individual descriptors against the decline indexes are available in Supplementary Figures 4-7. 
-The following sections identify how the descriptors contribute and interact to the decline index models in specific decline cases.
+The following sections identify how the descriptors contribute and interact within the decline index models in specific decline cases.
 
-### Localised descriptor contributions and trends within the PDI and DAI models
+### Local descriptor contributions and trends within the PDI and DAI models
 
-LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to $M_{PDI}$ in each of the healthy, moderate and severe decline cases using lime (Table 1).
-These descriptors were also the same top descriptors in the overall model importance lists for the PDI (Figure 2a).
+LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to $M_{PDI}$ in each of the healthy, moderate and severe decline cases (Table 1).
+These descriptors were also among the same top descriptors in the overall model importance lists for the PDI (Figure 2a).
 Increasing decline severity was associated with a reduction in the size and condition of the crown.

### 2020-02-07 manuscript/manuscript.Rmd
alterations from Johns comments

@@
   pdf_document: default
+fontsize: 12pt
 header-includes:
@@
 Assigning discrete decline status labels neglects the true spectrum of decline severity encountered in the field.
-There is a growing need for sensitive, continuous measures of tree decline that accurately reflect the decline spectrum [@pontius2014].
+There is a growing need for sensitive, continuous measures of tree decline that accurately reflect the decline severity spectrum [@pontius2014].
 This has the potential to greatly improve monitoring, enable the early detection of decline for accurate forecasting and inform forestry management practices.
@@
 All cardinal assessments were averaged to give a single value per tree.
-The crown density ($CD$) values erroneously collected at Attingham and Langdale were converted to crown transparency ($t = 100 - CD$)
+The crown density ($CD$) values collected at Attingham and Langdale were converted to crown transparency ($t = 100 - CD$)
 All descriptors collected in cm were converted to mm except for DBH which was converted to m.
@@
 The canopy closure descriptor was converted from yes and no to present and absent.
-Finally, the stem symptoms sizes for active bleeds, black staining and callused wounds were averaged and a frequency per tree collected.
+Finally, the stem symptoms sizes for active bleeds, black staining and callused wounds were averaged and a frequency per tree totalled.
 After preparation there were a total of 31 raw descriptors for analysis.
@@
 The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
-Overall descriptor importance contributions to each of the indexes was measured in each model using the mean decrease in accuracy measure.
+Overall descriptor importance contributions to each of the indexes was measured in each model using the mean decrease in accuracy measure with higher values considered to have a higher overall contribution to the decline index.
 

### 2020-02-07 manuscript/manuscript.Rmd
active bleed, black staining and callous wound size changed to length

@@
 The canopy closure descriptor was converted from yes and no to present and absent.
-Finally, the stem symptoms sizes for active bleeds, black staining and callused wounds were averaged and a frequency per tree totalled.
+Finally, the stem symptoms lengthss for active bleeds, black staining and callused wounds were averaged and a frequency per tree totalled.
 After preparation there were a total of 31 raw descriptors for analysis.
@@
 
-$a$ = average active bleed size (mm), $A$ = number of active bleeds, $b$ = average black stain size (mm), $B$ = black stain number, $d$ = $DBH (m) \times 1000$
+$a$ = average active bleed length (mm), $A$ = number of active bleeds, $b$ = average black stain length (mm), $B$ = black stain number, $d$ = $DBH (m) \times 1000$
 
@@
 symptom_ranges <- site_corrected_analysis_suitable_data %>%
-  select(`Active bleed size (mm)`,`Black staining size (mm)`) %>%
+  select(`Active bleed length (mm)`,`Black staining length (mm)`) %>%
   gather(descriptor,value) %>%
@@
 
-The same approach was used for $M_{DAI}$ using a comparison between active bleed size and black staining size as well as missing crown and crown transparency.
-The upper boundary for the ranges of active bleed size and black stain size was taken as the centre value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.
+The same approach was used for $M_{DAI}$ using a comparison between active bleed length and black staining length as well as missing crown and crown transparency.
+The upper boundary for the ranges of active bleed length and black stain length was taken as the centre value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.
 
@@
 These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and crown condition descriptors (Figures 4d, e & f) change for non-symptomatic AOD or COD trees (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively).
-Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
+Increasing active bleed, staining length and frequency contributed to greater positive values of the DAI and more severe AOD. 
 Higher DAI values could also be obtained from trees with greater stature (Figure 4c). 

### 2020-02-07 R/functions/DAIresponseSurfaces.R

@@ -23,14 +23,14 @@ DAIresponseSurfaces <- function(DAIrf,DAI_example_cases,site_corrected_analysis_
     map(~{
       type <- .
       list(
-        `Active bleed size (mm)` = seq(ranges$min[ranges$Descriptor == 'Active bleed size (mm)'],ranges$center[ranges$Descriptor == 'Active bleed size (mm)']/8,length.out = 100),
-        `Black staining size (mm)` = seq(ranges$min[ranges$Descriptor == 'Black staining size (mm)'],ranges$center[ranges$Descriptor == 'Black staining size (mm)']/8,length.out = 100)
+        `Active bleed length (mm)` = seq(ranges$min[ranges$Descriptor == 'Active bleed length (mm)'],ranges$center[ranges$Descriptor == 'Active bleed length (mm)']/8,length.out = 100),
+        `Black staining length (mm)` = seq(ranges$min[ranges$Descriptor == 'Black staining length (mm)'],ranges$center[ranges$Descriptor == 'Black staining length (mm)']/8,length.out = 100)
       ) %>%
         expand.grid() %>%
         as_tibble() %>%
         mutate(ID = 1) %>%
         left_join(DAI_example_cases[[type]] %>%
-                    select(-`Active bleed size (mm)`,-`Black staining size (mm)`) %>%
+                    select(-`Active bleed length (mm)`,-`Black staining length (mm)`) %>%
                     mutate(ID = 1),
                   by = 'ID') %>%
         select(-ID) %>%
@@ -41,9 +41,9 @@ DAIresponseSurfaces <- function(DAIrf,DAI_example_cases,site_corrected_analysis_
                                              `Total height (m)`,
                                              `Lower crown height (m)`,
                                              `Crown condition (%)`),
-          `Bleed prevalence (%)` = bleedPrevalence(`Active bleed size (mm)`,
+          `Bleed prevalence (%)` = bleedPrevalence(`Active bleed length (mm)`,
                                                    `Active bleeds`,
-                                                   `Black staining size (mm)`,
+                                                   `Black staining length (mm)`,
                                                    `Black staining`,
                                                    `Diameter at breast height (m)`)
         ) %>%
@@ -74,9 +74,9 @@ DAIresponseSurfaces <- function(DAIrf,DAI_example_cases,site_corrected_analysis_
                                              `Total height (m)`,
                                              `Lower crown height (m)`,
                                              `Crown condition (%)`),
[…]

### 2020-02-07 R/functions/descriptorInfo.R

@@ -39,11 +39,11 @@ descriptorInfo <- function(analysisTable){
     'Oval shaped exit holes','Presence of oval shaped exit holes.','P = present; A = absent',
     'Small circular shaped exit holes','','P = present; A = absent',
     'Active bleeds','Number of active stem bleeds in 3m section from the base of the tree. Liquid running.','frequency',
-    'Active bleed size (mm)','Average active bleed size in the 3m section from the base of the tree.','mm',
+    'Active bleed length (mm)','Average active bleed length in the 3m section from the base of the tree.','mm',
     'Black staining','Number of black stains in the 3m section from the base of the tree. Inactive bleeds.','fequency',
-    'Black staining size (mm)','Average black stain size in the 3m section from the base of the tree.','mm',
+    'Black staining length (mm)','Average black stain length in the 3m section from the base of the tree.','mm',
     'Calloused wound','Number of callused wounds in the 3m section from the base of the tree. Scars.','frequency',
-    'Calloused wound size (mm)','Average callouse wound size in the 3m section from the base of the tree.','mm',
+    'Calloused wound length (mm)','Average callouse wound length in the 3m section from the base of the tree.','mm',
   ) %>%
     mutate(`Descriptor Type` = c(rep('tree size',5),
                                  rep('crown',13),

### 2020-02-07 manuscript/manuscript.Rmd
further changes from John comments

@@
 
-The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model ($M_{PDI}$,$M_{DAI}$) response surfaces for specific example cases.
+The interactions of key raw descriptors identified with high mean decrease in accuracy importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model ($M_{PDI}$,$M_{DAI}$) response surfaces for specific example cases.
 The same example cases were used as described for the LIME analyses above.
@@
 
-A scatter plot of the calculated PDI and DAI scores is shown in Figure 1 for all `r nrow(site_corrected_analysis_suitable_data)` trees for which phenotypic descriptor data was collected.
+A scatter plot of the calculated PDI and DAI scores (see Supplementary information ##) is shown in Figure 1 for all `r nrow(site_corrected_analysis_suitable_data)` trees for which phenotypic descriptor data was collected.
 It can clearly be seen that the healthy trees group towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
@@
 The complex descriptor interactions within the index models ($M_{PDI}$,$M_{DAI}$) investigated by simulating phenotypic data and generating the index values from the supervised index random forest regression models.
-The index values generated by the models from these simulated phenotypic data were then visualised as response surface plots, from which descriptor interactions within the indexes can be interpreted. 
+The index values generated by the models from these simulated phenotypic data were then visualised as model response surface plots, from which descriptor interactions within the indexes can be interpreted.
 Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for key phenotypic descriptors that contribute highly to the PDI and DAI.
@@
 
-An unsupervised approach was favoured in this context due to the concession that there know difficulties and inaccuracy of manual decline status classification in the field [@ferretti1997].
+An unsupervised approach was favoured in this context due to the concession that there known difficulties with and inaccuracy in manual decline status classification in the field [@ferretti1997].
 This is highlighted in Figure 1 where a numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g.

### 2020-02-09 manuscript/figures.Rmd
added explanation of descriptor importance plots

@@
 
-```{r descriptor_contributions, fig.height = 6,fig.cap = 'Descriptor importance scores from random forest models for relative contributions to\ndecline index measures.'}
+```{r descriptor_contributions, fig.height = 6,fig.cap = 'Descriptor mean decrease in accuracy importance scores from the random forest models $M_{PDI}$ and $M_{DAI}$ for relative contributions to the decline index measures.
+Descriptors are ordered by descending mean decrease in accuracy and therefore from highest contribution to lowest contribution to the given decline index.'}
 loadd(descriptor_contribution_plots)

### 2020-02-09 manuscript/manuscript.Rmd
added explanation of descriptor importance plots

@@
 
+The relative contributions of the phenotypic descriptors to the decline index models is shown in Figure 2.
+
 The top four ranked descriptors found to contribute to $M_{PDI}$ were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure 2a).
-Other tree size related descriptors such as and live crown ratio stem bleeding descriptors contributed secondarily.
+Other tree size related descriptors such as bleed prevalence and live crown ratio stem bleeding descriptors contributed secondarily.
 

### 2020-02-09 manuscript/manuscript.Rmd
alteration to site location results section

@@
 
-### Phenotypic differences between site locations
+### Phenotypic differences between site locations and trees size descriptor adjustments
 
-A preliminary analysis using supervised random forest classification could separate trees based on their location (Supplementary  figure 3a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`).
+A preliminary analysis using supervised random forest classification could separate trees based on their location (Supplementary  figure 3a) with an average *margin* value of `r site_rf_margin$Margin %>% signif(digits = 3)`.
 The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary figure 3b).
-To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in the Materials and Methods.
+To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in Section #.
+This ensures that the subsequent comparisons of these discriptors in the context of oak decline are not confounded by these identified site differences.
 After the application of the tree size adjustment, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the adjustment was able to remove the majority of the between site variance.
@@
 
-The complex descriptor interactions within the index models ($M_{PDI}$,$M_{DAI}$) investigated by simulating phenotypic data and generating the index values from the supervised index random forest regression models.
+The complex descriptor interactions within the index models ($M_{PDI}$, $M_{DAI}$) were investigated by simulating phenotypic data and generating the index values from the supervised index random forest regression models.
 The index values generated by the models from these simulated phenotypic data were then visualised as model response surface plots, from which descriptor interactions within the indexes can be interpreted.
-Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for key phenotypic descriptors that contribute highly to the PDI and DAI.
+Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for the key phenotypic descriptors described in Section ## that contribute highly to the PDI and DAI.
 

### 2020-02-10 manuscript/manuscript.Rmd
added methods section for RF site analysis

@@
       DAI_descriptor_importance,
-      surveyors_names)
+      surveyors_names,
+      pheno_data_with_additional_descriptors)
 ```
@@
 The canopy closure descriptor was converted from yes and no to present and absent.
-Finally, the stem symptoms lengthss for active bleeds, black staining and callused wounds were averaged and a frequency per tree totalled.
+Finally, the stem symptoms lengths for active bleeds, black staining and callused wounds were averaged and a frequency per tree totalled.
 After preparation there were a total of 31 raw descriptors for analysis.
-
-### Tree size descriptor adjustment for age and inter site variance
-
-To account for age, size and site specific tree difference, an adjustment of site means to the overall mean was applied to each of the size descriptors (DBH, lower crown height, timber height, total height, crown radius).
-An example of this adjustment is shown for DBH in Supplementary Figure 1 and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table 2.
 
@@
 
+### Random forest analysis of inter-site discrimination
+
+Supervised random forest classification was performed to assess the discrimination in the phenotypic descriptors between the site locations and identify descriptors that could potentially confound comparisons relating to oak decline severity.
+Random forest analysis was performed using the randomForest R package version `r packageVersion('randomForest') %>% as.character()` [@liaw2002] with 100 repetitions using default $mtry$ and $ntree$ parameters.
+Due to uneven class sizes, the site classes were balanced using stratified sampling with a sample size of `r pheno_data_with_additional_descriptors$Location %>% table() %>% min()` per class, the smallest class size.
+
+To assess classification performance the *margin* metric was computed for each observation using the out of bag predictions, which was averaged across all observations for each random forest repetition, and then averaged across all repetitions.
+The *margin* metric is the difference between the proportion of votes for the correct class and the maximum proportion of votes for the other classes for a given observation.
+A positive *margin* value indicates correct classification [@breiman2001a].
+
+Descriptor importance contributions were measured using the mean decrease in accuracy metric averaged for each descriptor across the repetitions.
+Higher mean decrease in accuracy values had greater overall contribution and therefore greater importance.
+
+### Tree size descriptor adjustment for age and inter site variance
+
+To account for age, size and site specific tree difference, an adjustment of site means to the overall mean was applied to each of the size descriptors (DBH, lower crown height, timber height, total height, crown radius).
+An example of this adjustment is shown for DBH in Supplementary Figure 1 and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table 2.  
+All composite descriptors utilising these adjusted descriptors were recalculated after adjustment.
+
 ### Calculation of indexes for oak decline severity
 
-The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest using the randomForest R package version `r packageVersion('randomForest') %>% as.character()` [@liaw2002].
+The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest.
 The proximity matrices from each repetition were averaged to a single proximity matrix ($P$).
@@
 The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
-Overall descriptor importance contributions to each of the indexes was measured in each model using the mean decrease in accuracy measure with higher values considered to have a higher overall contribution to the decline index.
+Overall descriptor importance contributions to each of the indexes using the mean decrease in accuracy measure as described in Section #.
 
@@
 To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in Section #.
-This ensures that the subsequent comparisons of these discriptors in the context of oak decline are not confounded by these identified site differences.
+This ensures that the subsequent comparisons of these descriptors in the context of oak decline are not confounded by these identified site differences.
 After the application of the tree size adjustment, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the adjustment was able to remove the majority of the between site variance.

### 2020-02-10 manuscript/manuscript.Rmd
alterations to site differences supplementary figure

@@
 This ensures that the subsequent comparisons of these descriptors in the context of oak decline are not confounded by these identified site differences.
-After the application of the tree size adjustment, the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the adjustment was able to remove the majority of the between site variance.
+After the application of the tree size adjustment, the average random forest margin value was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)`.
+This negative value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were on average unable to correctly classify the site locations of the trees.
 

### 2020-02-10 manuscript/supplementary.Rmd
alterations to site differences supplementary figure

@@
 
-```{r site_differences,fig.cap='Differences between sites in phenotypic descriptors analysed by supervised Random Forest.
-a) Multidimensional scaling (MDS) plot of random forest proximity values of individual trees scaled to two dimensions.
-Mean margin value shown.
-b) Descriptor importances measures.',fig.height=6}
+```{r site_differences,fig.cap='Differences between sites in phenotypic descriptors analysed by supervised random forest classification.
+a) Multidimensional scaling (MDS) scatter plot of random forest proximity values of individual trees scaled to two dimensions.
+b) Descriptor mean decrease in accuracy importances measures.
+Descriptors ordered in descending importance.',fig.height=6}
 loadd(site_differences_mds_plot)

### 2020-02-10 manuscript/manuscript.Rmd
moved discussion element from results to discussion
added word document template

@@
 output: 
-  word_document: default
+  word_document:
+    reference_docx: template.docx
   pdf_document: default
@@
 
-# Title: Index measures for oak decline severity using phenotypic descriptors {-}
+# **Title:** Index measures for oak decline severity using phenotypic descriptors {-}
 
@@
 
-Although significance was obtained between the decline status groups on both the PDI and DAI scales, Figure 1 highlights the inaccuracies in the manual decline status classifications when viewed in an individual tree context.
-It can be seen that a number of the trees, when compared quantitatively using these measures, could in fact be classified with a different decline state.
-
 ### Key descriptors contributing to the decline indexes
@@
 Further work is needed that includes phenotypic data with greater spatial and temporal to establish definitive thresholds.
+
+Although significance was obtained between the decline status groups on both the PDI and DAI scales, Figure 1 highlights the inaccuracies in the manual decline status classifications when viewed in an individual tree context.
+It can be seen that a number of the trees, when compared quantitatively using these measures, could in fact be classified with a different decline state.
 

### 2020-02-10 manuscript/manuscript.Rmd
couple of proofing changes

@@
 output: 
-  word_document:
+  bookdown::word_document2:
     reference_docx: template.docx
-  pdf_document: default
+  bookdown::pdf_document2:
+    toc: false
 fontsize: 12pt
@@
 The decline indexes were derived using proximity values from unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
-These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally providing a potentially useful tool for forest monitoring and management.
+These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally, providing a potentially useful tool for forest monitoring and management.
 
@@
 Trees and forests across the globe face threats from a changing climate, notably warming and an increase in drought episodes affect the health status of individual trees, causing vegetation changes at the landscape scale [@choat2012; @millar2015; @allen2015].
-Within this context biotic agents (pests and diseases) can flourish and often drive mortality [@kautz2017; @jactel2012]. 
+Within this context, biotic agents (pests and diseases) can flourish and often drive mortality [@kautz2017; @jactel2012]. 
 Monitoring vegetation changes and understanding the underlying processes affecting tree health are important first steps to predicting the scale and extent of these threats [@hartmann2018] and plan future policies for planting and mitigation efforts [@bradford2017].

### 2020-02-10 manuscript/manuscript.Rmd
added section references

@@
 
-## Abstract {-}
+# Abstract {-}
 
@@
 
-## Introduction
+# Introduction {#introduction}
 
@@
 
-## Materials and methods
+# Materials and methods {#materials-and-methods}
 
-### Phenotypic data collection
+## Phenotypic data collection {#data-collection}
 
@@
 
-### Phenotypic descriptor preparation
+## Phenotypic descriptor preparation {#method-decriptor-preparation}
 
@@
 
-### Composite descriptor calculation
+## Composite descriptor calculation {#method-composite-descriptors}
 
@@
 
-### Random forest analysis of inter-site discrimination
+## Random forest analysis of inter-site discrimination {#method-inter-site-differences}
 
@@
 
-### Tree size descriptor adjustment for age and inter site variance
+## Tree size descriptor adjustment for age and inter site variance {#method-adjustment}
 
@@
 
-### Calculation of indexes for oak decline severity
+## Calculation of indexes for oak decline severity {#method-index-calculation}
 
@@
 
-### Descriptor contribution to decline indexes
+## Descriptor contribution to decline indexes {#method-descriptor-contribution}
 
@@
 The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
-Overall descriptor importance contributions to each of the indexes using the mean decrease in accuracy measure as described in Section #.
+Overall descriptor importance contributions to each of the indexes using the mean decrease in accuracy measure as described in Section \@ref(method-inter-site-differences).
 
-### Local descriptor contribution to index models
+## Local descriptor contribution to index models
 
@@
 
-### Index model simulations to investigate interactions of key descriptors
+## Index model simulations to investigate interactions of key descriptors
 
@@
 
-## Results
+# Results
 
-### Phenotypic differences between site locations and trees size descriptor adjustments
+## Phenotypic differences between site locations and trees size descriptor adjustments
 
@@
 The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary figure 3b).
-To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in Section #.
+To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in Section \@ref(method-adjustment).
 This ensures that the subsequent comparisons of these descriptors in the context of oak decline are not confounded by these identified site differences.
@@
 
-### Oak decline severity indexes of manually assigned decline statuses
+## Oak decline severity indexes of manually assigned decline statuses
 
@@
 
-### Key descriptors contributing to the decline indexes
+## Key descriptors contributing to the decline indexes {#results-descriptor-contirbutions}
 
@@
 
-### Local descriptor contributions and trends within the PDI and DAI models
+## Local descriptor contributions and trends within the PDI and DAI models
 
@@
 
-### Key descriptor interactions within the decline index models
+## Key descriptor interactions within the decline index models
 
@@
 The index values generated by the models from these simulated phenotypic data were then visualised as model response surface plots, from which descriptor interactions within the indexes can be interpreted.
-Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for the key phenotypic descriptors described in Section ## that contribute highly to the PDI and DAI.
+Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for the key phenotypic descriptors described in Section \@ref(results-descriptor-contirbutions) that contribute highly to the PDI and DAI.
 
@@
 
-## Discussion
+# Discussion
 
-### Suitability of the decline indexes for describing decline severity and type
+## Suitability of the decline indexes for describing decline severity and type
 
@@
 
-### The machine learning approach
+## The machine learning approach
 
@@
 
-### Applications
+## Applications
 
@@
 
-## Conclusions
+# Conclusions
 
@@
 
-## Acknowledgements
+# Acknowledgements {-}
 
@@
 
-## References
+# References

### 2020-02-11 manuscript/manuscript.Rmd
fix typo

@@
 At the AOD sites, annual monitoring data collected since 2009 was used to guide tree selection.
-Healthy trees had no know history of AOD and fruiting bodies of root pathogens had not been observed, all trees had canopy condition of 4 or 5, see @brown2016 for details of this assessment.
+Healthy trees had no known history of AOD and fruiting bodies of root pathogens had not been observed, all trees had canopy condition of 4 or 5, see @brown2016 for details of this assessment.
 AOD trees had current stem bleeds but a variety of crown conditions. 

### 2020-02-17 manuscript/manuscript.Rmd

--- 2020-02-10 manuscript/manuscript.Rmd
@@
 At the AOD sites, annual monitoring data collected since 2009 was used to guide tree selection.
-Healthy trees had no know history of AOD and fruiting bodies of root pathogens had not been observed, all trees had canopy condition of 4 or 5, see @brown2016 for details of this assessment.
+Healthy trees had no known history of AOD and fruiting bodies of root pathogens had not been observed; all trees had canopy condition of 4 or 5, see @brown2016 for details of this assessment.
 AOD trees had current stem bleeds but a variety of crown conditions. 
@@
 It was then min-max scaled to give a scale between 0 and 1 with more severely declining trees having a score closer to 1.
-This dimension was named the PDI.
+This dimension was named the Phenotypic Decline Index, PDI.
 The second MDS dimension was min-max scaled and then transformed ($2D - 1$) to give a scale between -1 and 1 with COD trees having a score closer to -1 and AOD having a score closer to 1. 
-This dimension was named the DAI.
-
+This dimension was named the Decline Acuteness Index, DAI.
 The code used to generate the decline indexes is available as an R package pdi version `r packageVersion('pdi') %>% as.character()` at https://github.com/jasenfinch/pdi.
@@
 These models will subsequently be referred to as $M_{PDI}$ and $M_{DAI}$ for the PDI and DAI respectively. 
-The optimal number of trees ($ntree$) and number randomly of variables ($mtry$) parameters for random forest model were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
+The parameters number of trees ($ntree$) and number of randomly assigned variables ($mtry$) for random forest model were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
 The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
@@
 
-For $M_{PDI}$, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severely tree.
+For $M_{PDI}$, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severely declining tree.
 The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors were divided into 100 increments.
@@
 
-A scatter plot of the calculated PDI and DAI scores (see Supplementary information ##) is shown in Figure 1 for all `r nrow(site_corrected_analysis_suitable_data)` trees for which phenotypic descriptor data was collected.
-It can clearly be seen that the healthy trees group towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
+A scatter plot (Figure 1) of the calculated PDI and DAI scores (see Supplementary information Figures 4 to 7) for all `r nrow(site_corrected_analysis_suitable_data)` trees for which phenotypic descriptor data was collected shows that the healthy trees group towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
 This gave a significant difference when PDI of the control and symptomatic (AOD and COD) decline statuses were compared using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
@@
 
-$M_{DAI}$ was more complex in descriptor contribution many more descriptors contributing more equally than the PDI (Figure 2b).
+$M_{DAI}$ was more complex with many more descriptors contributing more equally than for the PDI (Figure 2b).
 The overall tree size descriptors live crown ratio, DBH and crown volume contributed almost equally to $M_{DAI}$.
@@
 Bleed prevalence and stem symptom descriptors were not found to be locally important to any of the decline type cases.
-Live crown ratio greater than 61.6% in the severe AOD and moderate COD cases.
+Live crown ratio was greater than 61.6% in the severe AOD and moderate COD cases.
 COD trees were found to have greater crown contact (greater than 50%) compared to AOD trees.
@@
 The complex descriptor interactions within the index models ($M_{PDI}$, $M_{DAI}$) were investigated by simulating phenotypic data and generating the index values from the supervised index random forest regression models.
-The index values generated by the models from these simulated phenotypic data were then visualised as model response surface plots, from which descriptor interactions within the indexes can be interpreted.
+The index values generated by the models from these simulated phenotypic data were then visualised as model response surface plots, to allow interpretation of descriptor interactions between the indexes.
 Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for the key phenotypic descriptors described in Section \@ref(results-descriptor-contirbutions) that contribute highly to the PDI and DAI.
@@
 A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g.
-greater crown radius, total height, lower crown height and DBH; Figures 3a & b), with relatively small stature trees having PDIs of 0.6 or greater even with 100% crown condition.
+greater crown radius, total height, lower crown height and DBH; Figures 3a & b), while relatively small stature trees have PDIs of 0.6 or greater even with 100% crown condition.
 
@@
 Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure 4d).
-Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD however only narrow ranges of DAI values could be obtained for each case when varying the ranges of these descriptors.
+Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD; however, only narrow ranges of DAI values could be obtained for each case when varying the ranges of these descriptors.
 
@@
 The index thresholds for category definitions (healthy, declining, severe decline etc.) shown currently serve only as a guide to aid human interpretation as opposed to definitive thresholds of disease.
-Further work is needed that includes phenotypic data with greater spatial and temporal to establish definitive thresholds.
+Further work is needed that includes phenotypic data with greater spatial and temporal distribution to establish definitive thresholds.
 
@@
 However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
-It has been shown here that in our visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).
+It has been shown here that in the given visual definitions of oak decline severity, firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).
 
 The high contribution of bleed prevalence related descriptors to $M_{DAI}$ is also expected as this is one of the defining features of AOD [@denman2014]. 
-There was no contribution of *A.
-biguttatus* exit hole occurrence with with the DAI and therefore stem bleeding.
+There was no contribution of *A. biguttatus* exit hole occurrence to the DAI and therefore stem bleeding. 
 This could partly be due to smaller area surveyed compared to the stem bleed descriptors but it has also been identified previously that trees with bleeds do not always have exit holes, which has been attributed to the fact that not all *A.
@@
 The relationship of relative tree stature and the acuteness of oak decline has not previously been shown and likely reflects the period of time under which these trees have been afflicted. 
-AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.
+AOD occurs over a much shorter period compared to COD and therefore has less time in which to inhibit growth of the tree as a whole.
 
@@
 
-An unsupervised approach was favoured in this context due to the concession that there known difficulties with and inaccuracy in manual decline status classification in the field [@ferretti1997].
-This is highlighted in Figure 1 where a numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g.
+An unsupervised approach was favoured in this context due to the concession that there are known difficulties with and inaccuracy in manual decline status classification in the field [@ferretti1997].
+This is highlighted in Figure 1 where numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g.
 COD trees with DAI values > 0.5).
@@
 Machine learning approaches are usually seen as black box and difficulty in model interpretation is a concession in this approach; however, there are substantial gains in terms of predictive capability for model application in real world situations [@touw2012].
-The assessment descriptor contributions to the index models are an integral part of this approach.
+The assessment of descriptor contributions to the index models are an integral part of this approach.
 Input of knowledge of the biological problem by the investigator at this stage is essential in ensuring that the models suitably learn the patterns relevant biological problem to which they have been applied.

### 2020-02-17 manuscript/manuscript.Rmd
few alterations based on Mani comments

@@
 
-**Authors:** Jasen Finch^1\*^, Nathan Brown^2^, Manfred Beckmann^1^, Sandra Denman^2^ & John Draper^1^
+**Authors:** Jasen P. Finch^1\*^, Nathan Brown^2^, Manfred Beckmann^1^, Sandra Denman^2^ & John Draper^1^
 
@@
 Input of knowledge of the biological problem by the investigator at this stage is essential in ensuring that the models suitably learn the patterns relevant biological problem to which they have been applied.
-The descriptors contributing to these should be sensible and that the algorithm (random forest) is not learning spurious or artificial patterns within the data.
+The descriptors contributing to these should be sensible and that the algorithm (random forest) is not learning spurious or artificial patterns within the data such as site differences in tree size and age.
 Approaches such as LIME and model response surfaces enable investigators to query the contributions that phenotypic descriptors have in particular cases of interest.
@@
 The phenotypic descriptors used to calculate these indexes are easily measurable in the field, requiring no specialist equipment and are based on well established tree condition assessment protocols [@innes1990].
-Refinement of the protocols used here could make could reduce survey time and complexity and potentially be incorporated into citizen science programmes as a resource to build general UK oak training set [@brown2017b].
+Refinement of the protocols used here could reduce survey time and complexity and potentially be incorporated into citizen science programmes as a resource to build general UK oak training set [@brown2017b].
 Alternatively, the machine learning approach used to derive these decline indexes could also be applied to long established oak health monitoring programmes to aid sensitivity in detecting overall trends in oak health.

### 2020-02-17 manuscript/manuscript.Rmd
few more alterations based on Mani comments

@@
 The descriptors collected were based on @icpforests2016.
-Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
-Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m
-The timber height was measured at either 7cm top stem diameter or where no main stem could be distinguished.
+Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 1mm.
+Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m.
+The timber height was measured at either were the top stem diameter was 70mm or where no main stem could be distinguished.
 For the crown, crown contact, missing crown and absolute crown transparency percentages were visually estimated within 5% classes
@@
 This ensures that the subsequent comparisons of these descriptors in the context of oak decline are not confounded by these identified site differences.
-After the application of the tree size adjustment, the average random forest margin value was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)`.
+After the application of the tree size adjustment, the average random forest *margin* value was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)`.
 This negative value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were on average unable to correctly classify the site locations of the trees.
@@
 
-The phenotypic descriptors used to calculate these indexes are easily measurable in the field, requiring no specialist equipment and are based on well established tree condition assessment protocols [@innes1990].
+The phenotypic descriptors used to calculate these indexes are easily measurable in the field and are based on well established tree condition assessment protocols [@innes1990].
 Refinement of the protocols used here could reduce survey time and complexity and potentially be incorporated into citizen science programmes as a resource to build general UK oak training set [@brown2017b].
@@
 
-The decline indexes presented here provide a new framework in which to quantitatively describe the phenotypic oak decline spectrum using intuitive measures severity (PDI) and type (DAI).
+The decline indexes presented here provide a new framework to quantitatively describe the phenotypic oak decline spectrum using intuitive measures severity (PDI) and type (DAI).
 Their basis is on simple phenotypic descriptor measurements and a utilisation of a data driven, unsupervised random forest based machine learning approach.
@@
 Strategies have been presented such as the use of LIME and simulated model response surfaces that can aid the interpretation of the complex descriptor interactions that underpin these simple index measures and  greatly reduce the black box nature that is often associated with these approaches.
-It is encouraged that this approach of deriving descriptive index measures using machine learning is applied in other tree health phenotyping contexts and it is hoped that this will have an impact on future the consistency and comparability of oak health monitoring.
+It is encouraged that this approach of deriving descriptive index measures using machine learning is applied in other tree health phenotyping contexts and it is hoped that this will have an impact on the future consistency and comparability of oak health monitoring.
 

### 2020-02-17 R/functions/descriptorScatterPlots.R

@@ -6,6 +6,9 @@ descriptorScatterPlots <- function(site_corrected_analysis_suitable_data,decline
       bind_cols(select(decline_indexes,PDI)) %>%
       select_if(is.numeric) %>%
       gather(Descriptor,Value,-ID,-PDI) %>%
+      mutate(Descriptor = factor(Descriptor) %>%
+               fct_recode(`Agrilus exit hole density (m<sup>-2</sup>)` = 'Agrilus exit hole density (m^-2)',
+                          `Crown volume (m<sup>3</sup>)` = 'Crown volume (m^3)')) %>%
       ggplot(aes(x = Value,y = PDI)) +
       geom_point(shape = 21,fill = ptol_pal()(1)) +
       theme_bw(base_size = 10) +
@@ -13,13 +16,17 @@ descriptorScatterPlots <- function(site_corrected_analysis_suitable_data,decline
       labs(title = 'PDI against continuous descriptors',
            x = 'Descriptor value') +
       theme(plot.title = element_text(face = 'bold'),
-            axis.title = element_text(face = 'bold')),
+            axis.title = element_text(face = 'bold'),
+            strip.text = element_markdown()),
     
     PDI_factor =  site_corrected_analysis_suitable_data %>%
       select_if(is.factor) %>%
       rowid_to_column(var = 'ID') %>%
       bind_cols(select(decline_indexes,PDI)) %>%
       gather(Descriptor,Value,-ID,-PDI) %>%
+      mutate(Descriptor = factor(Descriptor) %>%
+               fct_recode(`Agrilus exit hole density (m<sup>-2</sup>)` = 'Agrilus exit hole density (m^-2)',
+                          `Crown volume (m<sup>3</sup>)` = 'Crown volume (m^3)')) %>%
       ggplot(aes(x = Value,y = PDI)) +
       geom_point(shape = 21,fill = ptol_pal()(1)) +
       theme_bw() +
@@ -27,13 +34,17 @@ descriptorScatterPlots <- function(site_corrected_analysis_suitable_data,decline
       labs(title = 'PDI versus categorical descriptors',
            x = 'Descriptor value') +
       theme(plot.title = element_text(face = 'bold'),
-            axis.title = element_text(face = 'bold')),
+            axis.title = element_text(face = 'bold'),
+            strip.text = element_markdown()),
     
     DAI_numeric = site_corrected_analysis_suitable_data %>%
       rowid_to_column(var = 'ID') %>%
[…]

### 2020-02-17 manuscript/manuscript.Rmd
alterations by and based on comments from Nathans

@@
 Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
-Abiotic factors can include both predisposing and inciting factors such as drought.
+Abiotic factors can include both long-term predisposing and one-off inciting factors.
 Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].
@@
 Machine learning is finding increasing utility in ecological and environmental applications due to it's ability to learn complex patterns within and between variables [@cutler2007;@belgiu2016].
-A popular algorithm is that of random forest due to it's versatility, relatively trivial training requirements and ease of use. 
+A popular algorithm is random forest due to it's versatility, relatively trivial training requirements and ease of use. 
 It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag sample prediction accuracies and variable importance [@breiman2001a;@biau2016].
@@
 At the COD sites trees were either COD (trees with clear signs of poor crown condition and canopy dieback) or healthy (dominant or co-dominant trees with full crowns). 
-At Big Wood trees with stronger crowns were categorised as COD when *Armillaria* fruiting bodies had been observed.
+At Big Wood trees with stronger crowns were also categorised as COD when *Armillaria* fruiting bodies had been observed.
 At the AOD sites, annual monitoring data collected since 2009 was used to guide tree selection.
@@
 The remission group included trees that had shown stem bleeds associated with AOD within the previous three years but were not showing signs of stem bleeds at the time of assessment.
-Thus at all sites, sampling attempted to include the unhealthiest and healthiest trees.
+Thus, at all sites, selection attempted to create a balanced sample of condition categories including both the unhealthiest and healthiest trees.
 
@@
 The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors were divided into 100 increments.
-For a given case, each combination of theses simulated descriptor values were used to calculate a PDI value from $M_{PDI}$, building a model response surface.
+For a given case, each combination of these simulated descriptor values were used to calculate a PDI value from $M_{PDI}$, building a model response surface.
 
@@
 The decline indexes presented here provide single intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity.
-These two indexes can be represented as a two dimensional oak decline continuum under which the spectrum of decline severity and type can be assessed (Figure 5).
+These two indexes can be represented as a two-dimensional oak decline continuum under which the spectrum of decline severity and type can be assessed (Figure 5).
 The index thresholds for category definitions (healthy, declining, severe decline etc.) shown currently serve only as a guide to aid human interpretation as opposed to definitive thresholds of disease.
@@
 
-Although significance was obtained between the decline status groups on both the PDI and DAI scales, Figure 1 highlights the inaccuracies in the manual decline status classifications when viewed in an individual tree context.
+Although significant differences were obtained between the decline status groups on both the PDI and DAI scales, Figure 1 highlights the inaccuracies in the manual decline status classifications when viewed in an individual tree context.
 It can be seen that a number of the trees, when compared quantitatively using these measures, could in fact be classified with a different decline state.
+This is likely due, at least in part, to the inclusion of trees in early stages of decline.
+For example, trees with visible fruiting bodies of root pathogens were manually classed as COD, even if crown health was still relatively good.
 
@@
 The unsupervised approach has worked in this context in aptly describing decline severity as the phenotypic descriptors selected were highly specific and well suited to describing the decline severity problem.
-In situations were descriptors may well be less suited, perhaps as a result of technical limitations, the use of a semi-supervised approach may be favourable.
+In situations where descriptors may well be less suited, perhaps as a result of technical limitations, the use of a semi-supervised approach may be favourable.
 The unsupervised approach to derive the decline indexes would be applied to a subset of the observations on which the descriptors are known to perform well and that would be considered benchmark observations by the investigators.
@@
 
-Machine learning approaches are usually seen as black box and difficulty in model interpretation is a concession in this approach; however, there are substantial gains in terms of predictive capability for model application in real world situations [@touw2012].
-The assessment of descriptor contributions to the index models are an integral part of this approach.
-Input of knowledge of the biological problem by the investigator at this stage is essential in ensuring that the models suitably learn the patterns relevant biological problem to which they have been applied.
-The descriptors contributing to these should be sensible and that the algorithm (random forest) is not learning spurious or artificial patterns within the data such as site differences in tree size and age.
-Approaches such as LIME and model response surfaces enable investigators to query the contributions that phenotypic descriptors have in particular cases of interest.
-This in turn promotes trust in the indexes derived from these models and reduces their black box nature [@ribeiro2016].
+Machine learning approaches are usually seen as black box and hard to interpret; however, there are substantial gains in terms of predictive capability for applying these models in real world situations [@touw2012].
+The assessment of descriptor contributions to the index models are an integral part of this approach and helps to overcome these limitations.
+Knowledge input of the biological mechanisms, by the investigator at this stage, is essential in ensuring that the models suitably learn the patterns relevant to the biological problem to which it is being applied.
+The descriptors contributing to these should be sensible and it is important to check that the algorithm (random forest) is not learning spurious or artificial patterns within the data such as site differences in tree size and age.
+Approaches such as LIME and model response surfaces enable investigators to query the contributions and interactions of individual phenotypic descriptors.
+This in turn promotes trust in the indexes derived from these models by the investigator and potential model end-users [@ribeiro2016].
 
@@
 
-The current context in which these decline indexes are presented is study specific and is based on a relatively small sample size that may not be generally representative of the UK oak population as a whole.
-Subsequent studies could incorporate the data from this study with additional data to train random forest models if comparable phenotyping protocols are used.
+While it is important to note that these decline indexes are study specific and based on a relatively small sample size, they do have the potential for wider use.
+Subsequent studies could easily build on this foundation by collecting additional data and incorporating it with the data from this study to train random forest models from which to derive decline index values.
+This would only be viable so long as comparable phenotyping protocols are used.
 Alternatively, study specific decline index values could be generated using the unsupervised random forest approach.
 
-The representation of the decline severity spectrum provided by these decline indexes provides greatly enhanced sensitivity over categorical decline status assignments.
-This could enhance integration with large scale data such as those from genome wide association studies or remote sensing [@stocks2019; @evans2012].
+By defining decline severity objectively across a spectrum, these decline indexes provide greatly enhanced sensitivity over previous categorical decline status assignments.
+This approach could enhance integration with large scale data such as those from genome wide association studies or remote sensing [@stocks2019; @evans2012].
 These index measures allow sensitive comparisons of visual health to be made for the same tree between years as well as comparisons of trees between different locations.
 
-Scaling this technique to enable generalised naive computation of decline indexes for any tree sampled in the UK would require index models trained using a substantially larger data set that incorporates the descriptive diversity of oak decline in the population as whole.
+Scaling this technique to enable computation of decline indexes for any tree sampled in the UK would require index models trained using a substantially larger data set that incorporates the full descriptive diversity of oak health in the population as whole.
 Scaling to this level would also require an appropriate method of tree size correction that could be applied on an individual tree basis.
 A potential solution to this would be to correct trees based on estimated age from DBH as proposed by @white1998.
-These indexes could have great utility in long term forest health monitoring by providing comparability over time that could be utilised for both forestry management and research [@denman2010].
+These indexes could have great utility in long term forest health monitoring by providing consistent scores which can be compared over time to inform both forestry management and research [@denman2010].
 
 The phenotypic descriptors used to calculate these indexes are easily measurable in the field and are based on well established tree condition assessment protocols [@innes1990].
-Refinement of the protocols used here could reduce survey time and complexity and potentially be incorporated into citizen science programmes as a resource to build general UK oak training set [@brown2017b].
-Alternatively, the machine learning approach used to derive these decline indexes could also be applied to long established oak health monitoring programmes to aid sensitivity in detecting overall trends in oak health.
+Further refinement of these protocols could be designed to reduce survey time and complexity and potentially be incorporated into citizen science programmes [@brown2017b].
+Alternatively, the machine learning approach used to derive these decline indexes could also be applied to long established oak health monitoring programmes to aid sensitivity in detecting overall trends in tree health.
 An example programme includes The International Co-operative Programme on Assessment and Monitoring of Air Pollution Effects on Forests (ICP Forests) [@lorenz1995].

### 2020-02-18 manuscript/manuscript.Rmd
added size adjusted site differences figure to supplementary. Altered figure references accordingly

@@
 
-A scatter plot (Figure 1) of the calculated PDI and DAI scores (see Supplementary information Figures 4 to 7) for all `r nrow(site_corrected_analysis_suitable_data)` trees for which phenotypic descriptor data was collected shows that the healthy trees group towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
+A scatter plot (Figure 1) of the calculated PDI and DAI scores (see Supplementary information Figures 5 to 8) for all `r nrow(site_corrected_analysis_suitable_data)` trees for which phenotypic descriptor data was collected shows that the healthy trees group towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
 This gave a significant difference when PDI of the control and symptomatic (AOD and COD) decline statuses were compared using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
@@
 
-Plots of the individual descriptors against the decline indexes are available in Supplementary Figures 4-7. 
+Plots of the individual descriptors against the decline indexes are available in Supplementary Figures 5-8. 
 The following sections identify how the descriptors contribute and interact within the decline index models in specific decline cases.

### 2020-03-02 manuscript/figures.Rmd
fixed x axis labels of descriptor contribution plots

@@
 ```{r descriptor_contributions, fig.height = 6,fig.cap = 'Descriptor mean decrease in accuracy importance scores from the random forest models $M_{PDI}$ and $M_{DAI}$ for relative contributions to the decline index measures.
-Descriptors are ordered by descending mean decrease in accuracy and therefore from highest contribution to lowest contribution to the given decline index.'}
+Descriptors are ordered by descending mean decrease in accuracy and therefore from highest contribution to lowest contribution to the given decline index.
+MSE = mean squared error.'}
 loadd(descriptor_contribution_plots)

### 2020-03-02 R/functions/descriptorImportancePlots.R

@@ -20,7 +20,7 @@ descriptorImportancePlots <- function(PDI_descriptor_importance,DAI_descriptor_i
         theme(plot.title = element_text(face = 'bold'),
               axis.title = element_text(face = 'bold',size = 10)) +
         labs(title = 'a) PDI',
-             y = 'Mean decrease in\naccuracy',
+             y = '% increase in\nMSE',
              x = NULL) +
         scale_x_discrete(labels = descriptorLabels)
     },
@@ -44,7 +44,7 @@ descriptorImportancePlots <- function(PDI_descriptor_importance,DAI_descriptor_i
         theme(plot.title = element_text(face = 'bold'),
               axis.title = element_text(face = 'bold',size = 10)) +
         labs(title = 'b) DAI',
-             y = 'Mean decrease in\naccuracy',
+             y = '% increase in\nMSE',
              x = NULL) +
         scale_x_discrete(labels = descriptorLabels)
     }

### 2020-03-03 manuscript/manuscript.Rmd
Few small tweaks based on previous comments

@@
 A chronically declining oak can also subsequently begin to develop the stem bleeds associated with AOD.
+
 The visual, manual assignment of decline status and severity in the field is variable and error prone due to the contribution of many factors.
@@
 
-As a result of this, there are many visual phenotypic descriptors that contribute to categorizing an oak tree as being in decline. 
+As a result of the variety of factors contributing to oak decline, there are many visual phenotypic descriptors that contribute to categorizing an oak tree as being in decline. 
 When the decline is severe there are clear symptoms visible in the crown, such as low foliage density and numerous dead branches. 
@@
 The phenotypic descriptors collected related to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
-The descriptors collected were based on @icpforests2016.
+The descriptors collected were based on @innes1990 and @icpforests2016.
 Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 1mm.
@@
 
-The AOD stem symptoms (active bleeds, black staining, callused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem.
+The AOD stem symptoms (active bleeds, black staining, calloused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem.
 The frequency of *A.
@@
 The canopy closure descriptor was converted from yes and no to present and absent.
-Finally, the stem symptoms lengths for active bleeds, black staining and callused wounds were averaged and a frequency per tree totalled.
+Finally, the stem symptoms lengths for active bleeds, black staining and calloused wounds were averaged and a frequency per tree totalled.
 After preparation there were a total of 31 raw descriptors for analysis.

### 2020-03-03 manuscript/manuscript.Rmd
added progression of AOD stem symptoms to introduction along with a restructure of paragraph order

@@
 Oak decline is becoming increasingly prevalent along with the need for accurate and quantitative oak health monitoring.
-The Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI), were developed to quantitatively describe and differentiate acute and chronic oak decline severity using a multitude of easily measurable phenotypic descriptors.
+The Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI) were developed to quantitatively describe and differentiate acute and chronic oak decline severity using a multitude of easily measurable phenotypic descriptors.
 Measured phenotypic descriptors included aspects of tree shape and size, crown condition and the presence of biotic agents.
 The decline indexes were derived using proximity values from unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
-These indexes are simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally, providing a potentially useful tool for forest monitoring and management.
+These indexes give simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally, providing a potentially useful tool for forest monitoring and management.
 
@@
 
-Trees and forests across the globe face threats from a changing climate, notably warming and an increase in drought episodes affect the health status of individual trees, causing vegetation changes at the landscape scale [@choat2012; @millar2015; @allen2015].
+Trees and forests across the globe face threats from a changing climate.
+Notably warming and an increase in drought episodes affect the health status of individual trees, causing vegetation changes at the landscape scale [@choat2012; @millar2015; @allen2015]. 
 Within this context, biotic agents (pests and diseases) can flourish and often drive mortality [@kautz2017; @jactel2012]. 
@@
 Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].
+The bacterial stem bleeds associated with AOD initially start as active dark weeping patches.
+Over time, as lesion development continues beneath the bark, external cracks develop between the bark plates that are surrounded by blackening stains.
+Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of calloused tissue on the exterior [@denman2014].
 
@@
 
-The visual, manual assignment of decline status and severity in the field is variable and error prone due to the contribution of many factors.
-Doing this accurately requires much expertise, experience, and training [@ferretti1997].
-
 As a result of the variety of factors contributing to oak decline, there are many visual phenotypic descriptors that contribute to categorizing an oak tree as being in decline. 
@@
 It is these characteristics that many studies use as the primary indicator of tree decline severity [@zarnoch2004], however these symptoms can develop slowly (especially in the case of COD) and simple classes are unlikely to capture the full detail of the subtle transition from being healthy to declined.
+
+The visual, manual assignment of decline status and severity in the field is variable and error prone due to the many factors contributing to oak decline.
+Doing this accurately requires much expertise, experience, and training [@ferretti1997].
 
@@
 
-* The phenotypic decline index (PDI) - a measure of decline severity scoring trees between 0 and 1.
+* The Phenotypic Decline Index (PDI) - a measure of decline severity scoring trees between 0 and 1.
 More severely declining oak trees have a score closer to 1.
 
-* The decline acuteness index (DAI) - a measure differentiating between chronically and acutely declining oak trees with a score between -1 and 1. 
+* The Decline Acuteness Index (DAI) - a measure differentiating between chronically and acutely declining oak trees with a score between -1 and 1. 
 Acutely declining trees have a score closer to 1 and chronically declining trees have a score closer to -1.

### 2020-03-03 manuscript/manuscript.Rmd
added detail on armillaria bleeds to introduction

@@
 Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of calloused tissue on the exterior [@denman2014].
+Stem bleeds can also, but not always, be associated with *Armillaria* attack on the inner stem tissues of oak.
+Active stem weeping and black staining can be found along the leading edge of the mycelial front of affected stem tissues [@denman2017].
 
@@
 Mortality in trees suffering from COD can occur over a period of decades whereas AOD can occur in 5-10 years [@denman2009].
-A chronically declining oak can also subsequently begin to develop the stem bleeds associated with AOD.
+A chronically declining oak can also subsequently begin to develop the stem bleeds associated with AOD [@denman2017].
 
@@
 
-Assigning discrete decline status labels neglects the true spectrum of decline severity encountered in the field.
+Assigning discrete decline status classes neglects the true spectrum of decline severity encountered in the field.
 There is a growing need for sensitive, continuous measures of tree decline that accurately reflect the decline severity spectrum [@pontius2014].

### 2020-03-03 manuscript/manuscript.Rmd
corrections to intro and methods

@@
 
-Machine learning is finding increasing utility in ecological and environmental applications due to it's ability to learn complex patterns within and between variables [@cutler2007;@belgiu2016].
+Machine learning is finding increasing utility in ecological and environmental applications due to its ability to learn complex patterns within and between variables [@cutler2007;@belgiu2016].
 A popular algorithm is random forest due to it's versatility, relatively trivial training requirements and ease of use. 
@@
 
-Here, we propose a framework to enable quantitative visible oak decline severity measurement that utilises a range of targeted phenotypic descriptors observed in oak forests across England and unsupervised random forest machine learning.
+Here, we propose a framework using unsupervised random forest machine learning to enable quantitative, visible oak decline severity measurement that utilises a range of targeted phenotypic descriptors observed in oak forests across England.
 
@@
 All trees were manually assigned a decline status category; healthy, AOD, COD or remission prior to collection of phenotypic descriptors.
-At the COD sites trees were either COD (trees with clear signs of poor crown condition and canopy dieback) or healthy (dominant or co-dominant trees with full crowns). 
-At Big Wood trees with stronger crowns were also categorised as COD when *Armillaria* fruiting bodies had been observed.
+At the COD sites, trees were either COD (trees with clear signs of poor crown condition and canopy dieback) or healthy (dominant or co-dominant trees with full crowns). 
+At Big Wood, trees with stronger crowns were also categorised as COD when *Armillaria* fruiting bodies had been observed.
 At the AOD sites, annual monitoring data collected since 2009 was used to guide tree selection.
@@
 
-The phenotypic descriptors collected related to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
+The collected phenotypic descriptors related to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
 The descriptors collected were based on @innes1990 and @icpforests2016.
@@
 Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m.
-The timber height was measured at either were the top stem diameter was 70mm or where no main stem could be distinguished.
-For the crown, crown contact, missing crown and absolute crown transparency percentages were visually estimated within 5% classes
+The timber height was measured at either where the top stem diameter was 70mm or where no main stem could be distinguished.
+For crown contact, missing crown and absolute crown transparency, percentages were visually estimated within 5% classes
 
@@
 The crown radius was measured using a vertex transponder attached to the relevant side of the stem at a height of 1.3m.
-Canopy closure recorded as yes or no depending on whether any other crowns are touching the sampling tree.
+Canopy closure was recorded as yes or no depending on whether any other crowns are touching the sampling tree.
 A high resolution photo was taken of the base to the top of the crown and of the stem from the base to a height of 3m for each cardinal point for later reference.
@@
 The canopy closure descriptor was converted from yes and no to present and absent.
-Finally, the stem symptoms lengths for active bleeds, black staining and calloused wounds were averaged and a frequency per tree totalled.
+Finally, the stem symptoms lengths for active bleeds, black staining and calloused wounds were averaged and a frequency per tree given by the total observed.
+The application and effect of this adjustment is described and discussed in Sections 3.1 and 4.3 respectively.
 After preparation there were a total of 31 raw descriptors for analysis.
@@
 This dimension was named the Decline Acuteness Index, DAI.
-The code used to generate the decline indexes is available as an R package pdi version `r packageVersion('pdi') %>% as.character()` at https://github.com/jasenfinch/pdi.
+The code used to generate the decline indexes is available as an R package pdi version `r packageVersion('pdi') %>% as.character()`, available at https://github.com/jasenfinch/pdi.
 
@@
 These models will subsequently be referred to as $M_{PDI}$ and $M_{DAI}$ for the PDI and DAI respectively. 
-The parameters number of trees ($ntree$) and number of randomly assigned variables ($mtry$) for random forest model were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
-The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
-Overall descriptor importance contributions to each of the indexes using the mean decrease in accuracy measure as described in Section \@ref(method-inter-site-differences).
+The parameters number of trees ($ntree$) and number of randomly assigned variables ($mtry$), for each random forest model, were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
+The combination parameters giving the lowest mean absolute percentage error (MAPE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, MAPE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, MAPE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
+Overall descriptor importance contributions to each of the indexes was assessed using the percentage increase in mean squared error measure with higher values indicating a greater overall contribution to the given model and therefore greater importance.
 
@@
 The interactions of key raw descriptors identified with high mean decrease in accuracy importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model ($M_{PDI}$,$M_{DAI}$) response surfaces for specific example cases.
-The same example cases were used as described for the LIME analyses above.
+The same example cases were used as described for the LIME analyses in Section 2.8.
 This approach was similar to response surface methodology without the need for probabilistic approximation due to the low computation time required for index calculation [@abbasi2012].
@@
 For $M_{PDI}$, missing crown and crown transparency were compared for example cases of a healthy, a moderate and a severely declining tree.
-The value ranges for the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors were divided into 100 increments.
+The value ranges were based on the observed range in the training data.
+For the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors, the ranges were divided into 100 increments.
 For a given case, each combination of these simulated descriptor values were used to calculate a PDI value from $M_{PDI}$, building a model response surface.
@@
 The same approach was used for $M_{DAI}$ using a comparison between active bleed length and black staining length as well as missing crown and crown transparency.
-The upper boundary for the ranges of active bleed length and black stain length was taken as the centre value between the maximum and minimum and reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enable more effective visualisation.
+The upper boundary for the ranges of active bleed length and black stain length was taken as the centre value between the maximum and minimum values in the training data sets.
+These were reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enabled more effective visualisation.
 

### 2020-03-04 manuscript/manuscript.Rmd
corrections to results section

@@
 After the application of the tree size adjustment, the average random forest *margin* value was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)`.
-This negative value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were on average unable to correctly classify the site locations of the trees.
+The reduction in this value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were no longer able accurately classify the site locations of the trees (Supplementary figure 4a) with almost no contribution of the corrected tree size descriptors (Supplementary figure 4b).
 
@@
 
-A scatter plot (Figure 1) of the calculated PDI and DAI scores (see Supplementary information Figures 5 to 8) for all `r nrow(site_corrected_analysis_suitable_data)` trees for which phenotypic descriptor data was collected shows that the healthy trees group towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
-This gave a significant difference when PDI of the control and symptomatic (AOD and COD) decline statuses were compared using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
+The calculated PDI and DAI scores (see Supplementary information #) derived from MDS of proximity values from unsupervised random forest for all `r nrow(site_corrected_analysis_suitable_data)` trees for which phenotypic descriptor data was collected is shown in Figure 1.
+The healthy trees grouped towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
+There was a significant difference when the PDI values of the healthy (non-symptomatic) and symptomatic (AOD and COD) decline statuses were compared using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
 
-On the DAI scale, the AOD trees group at the upper end of the scale (> 0) and the COD trees at the lower end (< 0).
+On the DAI scale, the AOD trees grouped at the upper end of the scale (> 0) and the COD trees at the lower end (< 0).
 This was confirmed with significance found in the DAI between these groups (`r glue("t({DAI_groups_ttest$parameter}) = {signif(DAI_groups_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(DAI_groups_ttest$estimate1,digits = 3)` and `r signif(DAI_groups_ttest$estimate2,digits = 3)` for the AOD and COD groups respectively.
@@
 $M_{DAI}$ was more complex with many more descriptors contributing more equally than for the PDI (Figure 2b).
-The overall tree size descriptors live crown ratio, DBH and crown volume contributed almost equally to $M_{DAI}$.
+The overall tree size descriptors such as live crown ratio, DBH and crown volume contributed almost equally to $M_{DAI}$.
 These were followed closely by the stem staining and bleeding descriptors such as bleed prevalence. 

### 2020-03-04 manuscript/figures.Rmd
additions and corrections to figure captions

--- 2020-02-09 manuscript/figures.Rmd
@@
 ```{r descriptor_contributions, fig.height = 6,fig.cap = 'Descriptor mean decrease in accuracy importance scores from the random forest models $M_{PDI}$ and $M_{DAI}$ for relative contributions to the decline index measures.
-Descriptors are ordered by descending mean decrease in accuracy and therefore from highest contribution to lowest contribution to the given decline index.'}
+Descriptors are ordered by descending percentage increase in mean squared error (MSE) and therefore from highest contribution to lowest contribution to the given decline index.'}
 loadd(descriptor_contribution_plots)
@@
 
-```{r PDI_response_surfaces,fig.height = 6,fig.cap='PDI response surfaces for simulated descriptor values from trained random forest models.'}
+```{r PDI_response_surfaces,fig.height = 6,fig.cap='PDI response surfaces from $M_{PDI}$ for simulated values of key contributing descriptors for example healthy, moderate and severly declining cases.
+The values of additional tree size descriptors are shown for the individual cases to indicate the relative stature of the example cases.'}
 loadd(PDI_response_surfaces)
@@
 
-```{r DAI_response_surfaces,fig.height = 8,fig.cap='DAI response surfaces for simulated descriptor values from trained random forest models.'}
+```{r DAI_response_surfaces,fig.height = 8,fig.cap='DAI response surfaces from $M_{DAI}$ for simulated values of key contributing descriptors for example cases.
+The neutral case is a tree that can be neither distinguished as AOD or COD.
+All cases have PDI values > 0.5.'}
 loadd(DAI_response_surfaces)

### 2020-03-04 manuscript/manuscript.Rmd

@@
 The index thresholds for category definitions (healthy, declining, severe decline etc.) shown currently serve only as a guide to aid human interpretation as opposed to definitive thresholds of disease.
-Further work is needed that includes phenotypic data with greater spatial and temporal distribution to establish definitive thresholds.
+Further work is needed that includes phenotypic data with greater spatial and temporal distribution to establish these definitive thresholds.
 
@@
 Crown condition and volume were shown to be the key contributors to the $M_{PDI}$. 
-This is to be expected and are an indicator of model suitability for the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007].
+This is to be expected and are indicators of model suitability using the unsupervised approach as crown condition and crown dimensions are well established as the key components of assessing overall tree health and condition [@schomaker2007].
 However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
-It has been shown here that in the given visual definitions of oak decline severity, firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).
+It has been shown here that in the given visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).
+
+It was shown that tree size, reflected by the live crown ratio descriptor, contributes most to the DAI in discriminating between COD and AOD trees.
+The relationship of relative tree stature and the acuteness of oak decline has not previously been shown and likely reflects the period of time under which these trees have been afflicted. 
+AOD occurs over a much shorter period compared to COD and therefore has less time in which to inhibit growth of the tree as a whole.
 
@@
 There was no contribution of *A. biguttatus* exit hole occurrence to the DAI and therefore stem bleeding. 
-This could partly be due to smaller area surveyed compared to the stem bleed descriptors but it has also been identified previously that trees with bleeds do not always have exit holes, which has been attributed to the fact that not all *A.
+This could partly be due to the low number of trees on which *A.
+biguttatus* exit holes were detected (`r which(deframe(pheno_data_with_additional_descriptors[,'Agrilus exit holes']) > 0) %>% length()` out `r nrow(pheno_data_with_additional_descriptors)`) but it has also been shown previously that trees with bleeds do not always have exit holes, attributed to the fact that not all *A.
 biguttatus* larvae complete their life cycle [@brown2017a;@vansteenkiste2004].
@@
 The association of crown condition and AOD severity has previously be shown by @brown2016.
-
-It was shown that tree size, reflected by the live crown ratio descriptor, contributes most to the DAI in discriminating between COD and AOD trees.
-The relationship of relative tree stature and the acuteness of oak decline has not previously been shown and likely reflects the period of time under which these trees have been afflicted. 
-AOD occurs over a much shorter period compared to COD and therefore has less time in which to inhibit growth of the tree as a whole.
 
@@
 
-A machine learning approach was used to derive the phenotypic index measures as it allows the models to be moulded to the complex interactions and trends between phenotypic descriptors from which the decline indexes are calculated.
+A machine learning approach was used to derive the phenotypic index measures as it allows the models to be moulded to the complex interactions that exist between the phenotypic descriptors from which the decline indexes are calculated.
 This would likely be oversimplified using conventional modelling strategies [@breiman2001b].
@@
 Indexes could then be attained for the rest of the observations by prediction based on supervised models of the benchmark subset [@chapelle2010].
-This approach would be more suitable than a fully supervised approach as it would limit the inaccuracies that would be introduced by manual classification and still allow the data driven flexibility of the unsupervised approach.
+This approach would be more suitable than a fully supervised approach as it would limit the inaccuracies that would be introduced by manual classification while still allowing the data driven flexibility of the unsupervised approach.
 
@@
 These indexes could have great utility in long term forest health monitoring by providing consistent scores which can be compared over time to inform both forestry management and research [@denman2010].
+For instance, the increase of the PDI values of trees within a woodland, beyond what is considered the healthy region of the PDI scale, could indicate that mangement intervention is required.
 
@@
 ```{r check,comment='',cache=FALSE}
-# manuscript_check(checks = c('word count','spelling'))
+manuscript_check(checks = c('word count')) # ,'spelling'
 ```

### 2020-03-04 manuscript/supplementary.Rmd

@@
 
+```{r site_differences_corrected,fig.cap='Differences between sites in phenotypic descriptors after adjustment of tree size descriptors analysed by supervised random forest classification.
+a) Multidimensional scaling (MDS) scatter plot of random forest proximity values of individual trees scaled to two dimensions.
+b) Descriptor mean decrease in accuracy importances measures.
+Descriptors ordered in descending importance.',fig.height=6}
+loadd(site_differences_post_correction_mds_plot)
+grid.arrange(site_differences_post_correction_mds_plot$a,site_differences_post_correction_mds_plot$b,nrow = 1)
+```
+
 ```{r PDInumeric,fig.height=9,fig.cap='Continuous descriptors plotted against the PDI.'}

### 2020-03-04 R/functions/DAIresponseSurfaces.R
Added additional Missing crown and crown transparency descriptors to COD response surface plots. Corrected caption

@@ -62,8 +62,6 @@ DAIresponseSurfaces <- function(DAIrf,DAI_example_cases,site_corrected_analysis_
         expand.grid() %>%
         as_tibble() %>%
         mutate(ID = 1) %>%
-        left_join(spectrumTrees_DAI[[type]] %>%
-                    select(-`Total height (m)`,-`Lower crown height (m)`) %>%
         left_join(DAI_example_cases[[type]] %>%
                     select(-`Total height (m)`,-`Lower crown height (m)`) %>%
                     mutate(ID = 1),
@@ -135,7 +133,9 @@ DAIresponseSurfaces <- function(DAIrf,DAI_example_cases,site_corrected_analysis_
         labs(title = type,
              caption = str_c(
                'Crown radius (m) = ',plotRanges_DAI_def[[type]]$`Crown radius (m)`[1] %>% round(3),'\n',
-               'Diameter at breast height (m) = ',plotRanges_DAI_def[[type]]$`Diameter at breast height (m)`[1] %>% round(3)
+               'Diameter at breast height (m) = ',plotRanges_DAI_def[[type]]$`Diameter at breast height (m)`[1] %>% round(3),'\n',
+               'Missing crown (%) = ',plotRanges_DAI_def[[type]]$`Missing crown (%)`[1] %>% round(3),'\n',  
+               'Crown transparency (%) = ',plotRanges_DAI_def[[type]]$`Crown transparency (%)`[1] %>% round(3)
              )
         )
       return(pl)

### 2020-03-04 manuscript/manuscript.Rmd
altered methods to reflect use of total height and lower crown height in DAI response surfaces. Lower bound of total height now upper bound of lower crown height to avoid impossible situations

@@
 
-The same approach was used for $M_{DAI}$ using a comparison between active bleed length and black staining length as well as missing crown and crown transparency.
+The same approach was used for $M_{DAI}$ using a comparison between active bleed length and black staining length as well as total height and lower crown height.
 The upper boundary for the ranges of active bleed length and black stain length was taken as the centre value between the maximum and minimum values in the training data sets.
 These were reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enabled more effective visualisation.
+For the total height (`r glue("{max(site_corrected_analysis_suitable_data[['Lower crown height (m)']]) %>% signif(3)} - {max(site_corrected_analysis_suitable_data[['Total height (m)']]) %>% signif(3)}m")`) and lower crown height (`r glue("{min(site_corrected_analysis_suitable_data[['Lower crown height (m)']]) %>% signif(3)} - {max(site_corrected_analysis_suitable_data[['Lower crown height (m)']]) %>% signif(3)}m")`) descriptors, the ranges were divided into 100 increments.
 

### 2020-03-04 R/functions/DAIresponseSurfaces.R

@@ -56,7 +56,7 @@ DAIresponseSurfaces <- function(DAIrf,DAI_example_cases,site_corrected_analysis_
     map(~{
       type <- .
       list(
-        `Total height (m)` = seq(ranges$min[ranges$Descriptor == 'Total height (m)'],ranges$max[ranges$Descriptor == 'Total height (m)'],length.out = 100),
+        `Total height (m)` = seq(ranges$max[ranges$Descriptor == 'Lower crown height (m)'],ranges$max[ranges$Descriptor == 'Total height (m)'],length.out = 100),
         `Lower crown height (m)` = seq(ranges$min[ranges$Descriptor == 'Lower crown height (m)'],ranges$max[ranges$Descriptor == 'Lower crown height (m)'],length.out = 100)
       ) %>%
         expand.grid() %>%

### 2020-03-04 manuscript/manuscript.Rmd
fixed discussion according to use of live crown ration descriptors for DAI response surfaces

@@
 
-The DAI ($M_{DAI}$) is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, DBH and crown condition descriptors. 
+The DAI ($M_{DAI}$) is a much more complex index with more descriptors contributing to it than the PDI such as live crown ration, bleed prevalence, DBH descriptors. 
 Simulated response surfaces for the DAI are shown in Figure 4. 
-These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and crown condition descriptors (Figures 4d, e & f) change for non-symptomatic AOD or COD trees (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively).
+These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and live crown ratio descriptors (Figures 4d, e & f) change for non-symptomatic AOD or COD trees (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively).
 Increasing active bleed, staining length and frequency contributed to greater positive values of the DAI and more severe AOD. 
 Higher DAI values could also be obtained from trees with greater stature (Figure 4c). 
-Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure 4d).
+Total height and lower crown height have no influence on the DAI for trees of neutral decline type (Figure 4d).
 Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD; however, only narrow ranges of DAI values could be obtained for each case when varying the ranges of these descriptors.

### 2020-03-04 manuscript/manuscript.Rmd
added reference to the need for tree size standardisation in intro. Spell check

@@
 The visual, manual assignment of decline status and severity in the field is variable and error prone due to the many factors contributing to oak decline.
-Doing this accurately requires much expertise, experience, and training [@ferretti1997].
+For instance, accurate and reproducible crown condition assessments often require rigorous quality assurance steps that include training and regular calibration among surveyors [@ferretti1997; @redfern2004].
+Additionally, differences in woodland age and individual tree sizes at different geographical locations can add difficulty to objectively comparing health status between sites.
+These factors require standardisation to reduce the potential confounding comparisons for tree health status [@zarnoch2004].
 
@@
 
-## Phenotypic descriptor preparation {#method-decriptor-preparation}
+## Phenotypic descriptor preparation {#method-descriptor-preparation}
 
-R version `r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and subsequent analysis of phenotypic data [@R2019].
+R version `r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and subsequent analysis of phenotypic data [@R2020].
 All raw data, data preparation routines and analysis code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.
@@
 
-To account for age, size and site specific tree difference, an adjustment of site means to the overall mean was applied to each of the size descriptors (DBH, lower crown height, timber height, total height, crown radius).
+To account for age, size and site specific tree differences, the tree size descriptors (DBH, lower crown height, timber height, total height, crown radius) were standardised using an adjustment of site means to the overall mean.
 An example of this adjustment is shown for DBH in Supplementary Figure 1 and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table 2.  
@@
 
-## Key descriptors contributing to the decline indexes {#results-descriptor-contirbutions}
+## Key descriptors contributing to the decline indexes {#results-descriptor-contributions}
 
@@
 The index values generated by the models from these simulated phenotypic data were then visualised as model response surface plots, to allow interpretation of descriptor interactions between the indexes.
-Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for the key phenotypic descriptors described in Section \@ref(results-descriptor-contirbutions) that contribute highly to the PDI and DAI.
+Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for the key phenotypic descriptors described in Section \@ref(results-descriptor-contributions) that contribute highly to the PDI and DAI.
 
@@
 These indexes could have great utility in long term forest health monitoring by providing consistent scores which can be compared over time to inform both forestry management and research [@denman2010].
-For instance, the increase of the PDI values of trees within a woodland, beyond what is considered the healthy region of the PDI scale, could indicate that mangement intervention is required.
+For instance, the increase of the PDI values of trees within a woodland, beyond what is considered the healthy region of the PDI scale, could indicate that management intervention is required.
 
@@
 ```{r check,comment='',cache=FALSE}
-manuscript_check(checks = c('word count')) # ,'spelling'
+# manuscript_check(checks = c('word count')) # ,'spelling'
 ```

### 2020-03-04 manuscript/manuscript.Rmd
added discussion point on AOD classifed COD trees

@@
       surveyors_names,
-      pheno_data_with_additional_descriptors)
+      pheno_data_with_additional_descriptors,
+      decline_indexes)
 ```
@@
 
+There were `r decline_indexes %>% filter(ChosenGroup == 'COD',PDI > 0.5,DAI > 0.5) %>% nrow()` manually assigned COD trees that were scored with PDI and DAI values greater than 0.5, suggesting that they were in fact AOD trees.
+These trees were manually assigned COD status due to the presence of *Armillaria* attack and associated bleeds.
+While the phenotyping protocol used here made no distinction as to the biotic origin of the stem bleeds, it is interesting to note that not only do these trees have the presence of stem bleeds similar to AOD trees, but also share their relatively larger stature compared to the majority of COD trees.
+This likely reflects a more acute and aggressive nature of the *Armillaria* infection in these trees with deterioration in crown health having occurred over a shorter time period. 
+It therefore has less time in which to impact overall growth compared to the other COD trees. 
+In this sense the DAI scale can be seen to differentiate decline types based less on the actual biotic agents involved in the interaction but more on the time-scales over which the trees themselves have been afflicted.
+
 AOD severity was able to be associated with overall decline severity with trees identified that had a combination of both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure 1).

### 2020-03-04 manuscript/manuscript.Rmd
corrections across manuscript

@@
 
-* The Phenotypic Decline Index (PDI) - a measure of decline severity scoring trees between 0 and 1.
+* The Phenotypic Decline Index (PDI) - a measure of decline severity, scoring trees between 0 and 1.
 More severely declining oak trees have a score closer to 1.
@@
 The occurrence of AOD has previously been reported and studied at these sites [@brown2016; @brown2017a].
-These sites included a variety of woodland, management and soil types as well as tree sizes/ages and were made up of mainly *Q. robur*.
+These sites included a variety of woodlands, management histories and soil types as well as tree sizes/ages and were made up of mainly *Q. robur*.
 
@@
 
-The AOD stem symptoms (active bleeds, black staining, calloused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem.
+The AOD stem symptoms (active bleeds, black staining, calloused wounds) were collected by measuring longest axis of each individual bleed, stain or calloused wound found within the basal 3m of the stem.
 The frequency of *A.
@@
 This was then transformed to a dissimilarity matrix ($1 - P$) and scaled to two dimensions using  multidimensional scaling (MDS).
-The first MDS dimension ($D$) was inverted ($1 - D$) to ensure the correct dimensional orientation relative to decline severity.
+The first MDS dimension ($D_1$) was inverted ($1 - D_1$) to ensure the correct dimensional orientation relative to decline severity.
 It was then min-max scaled to give a scale between 0 and 1 with more severely declining trees having a score closer to 1.
 This dimension was named the Phenotypic Decline Index, PDI.
-The second MDS dimension was min-max scaled and then transformed ($2D - 1$) to give a scale between -1 and 1 with COD trees having a score closer to -1 and AOD having a score closer to 1. 
+The second MDS dimension ($D_2$) was min-max scaled and then transformed ($2D_2 - 1$) to give a scale between -1 and 1 with COD trees having a score closer to -1 and AOD having a score closer to 1. 
 This dimension was named the Decline Acuteness Index, DAI.
@@
 To identify which phenotypic descriptors contributed locally to specific example cases of each of the decline indexes, Local Interpretable Model-Agnostic Explanations (LIME) were computed for each of models $M_{PDI}$ and $M_{DAI}$ [@ribeiro2016].
-The example cases for the PDI used for $M_{PDI}$ included the trees with the minimum (0) and maximum (1) PDI values and the tree with the PDI value closest to the median for the healthy, severe and moderately declining trees respectively.
+The example cases for the PDI used for $M_{PDI}$ included the trees with the minimum (0) and maximum (1) PDI values and the tree with the PDI value closest to the median (`r median(decline_indexes$PDI) %>% signif(3)`) for the healthy, severe and moderately declining trees respectively.
 For the DAI ($M_{DAI}$), five example cases were selected from trees with PDI scores greater than 0.5.
@@
 
-The interactions of key raw descriptors identified with high mean decrease in accuracy importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model ($M_{PDI}$,$M_{DAI}$) response surfaces for specific example cases.
+The interactions of key raw descriptors identified with high percentage increase in mean squared error importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model ($M_{PDI}$, $M_{DAI}$) response surfaces for specific example cases.
 The same example cases were used as described for the LIME analyses in Section 2.8.
@@
 
-The relative contributions of the phenotypic descriptors to the decline index models is shown in Figure 2.
+The relative contributions of the phenotypic descriptors to the decline index models ($M_{PDI}$,$M_{DAI}$) is shown in Figure 2.
 
@@
 The presence of *A.
-biguttatus* exit holes had very little contribution to the DAI with exit hole density ranked only `r which(DAI_descriptor_importance$Feature == 'Agrilus exit hole density (m^-2)')`th.
+biguttatus* exit holes had very little contribution to the DAI with exit hole density ranked only `r which(DAI_descriptor_importance$Feature == 'Agrilus exit hole density (m^-2)')`^th^.
 
 Plots of the individual descriptors against the decline indexes are available in Supplementary Figures 5-8. 
-The following sections identify how the descriptors contribute and interact within the decline index models in specific decline cases.
+The following sections (3.4, 3.5) identify how the descriptors contribute and interact within the decline index models in specific decline cases.
 
@@
 
-LIME identified crown volume, condition transparency and radius as the top four descriptors contributing to $M_{PDI}$ in each of the healthy, moderate and severe decline cases (Table 1).
+LIME identified crown volume, condition, transparency and radius as the top four descriptors contributing to $M_{PDI}$ in each of the healthy, moderate and severe decline cases (Table 1).
 These descriptors were also among the same top descriptors in the overall model importance lists for the PDI (Figure 2a).
@@
 This likely reflects the greater number of descriptors contributing to the DAI with different descriptors relevant to particular decline contexts.
-The important descriptors identified were mainly associated with tree size such as live crown ratio and lower crown radius.
+The important descriptors identified were mainly associated with tree size such as live crown ratio and lower crown height.
 Bleed prevalence and stem symptom descriptors were not found to be locally important to any of the decline type cases.
@@
 The complex descriptor interactions within the index models ($M_{PDI}$, $M_{DAI}$) were investigated by simulating phenotypic data and generating the index values from the supervised index random forest regression models.
-The index values generated by the models from these simulated phenotypic data were then visualised as model response surface plots, to allow interpretation of descriptor interactions between the indexes.
+The index values, generated by the models from these simulated phenotypic data, were then visualised as model response surface plots, to allow interpretation of descriptor interactions between the indexes.
 Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for the key phenotypic descriptors described in Section \@ref(results-descriptor-contributions) that contribute highly to the PDI and DAI.
@@
 
-The DAI ($M_{DAI}$) is a much more complex index with more descriptors contributing to it than the PDI such as live crown ration, bleed prevalence, DBH descriptors. 
+The DAI ($M_{DAI}$) is a much more complex index with more descriptors contributing to it than the PDI such as live crown ratio, bleed prevalence, and DBH. 
 Simulated response surfaces for the DAI are shown in Figure 4. 
 These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and live crown ratio descriptors (Figures 4d, e & f) change for non-symptomatic AOD or COD trees (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively).
-Increasing active bleed, staining length and frequency contributed to greater positive values of the DAI and more severe AOD. 
+Increasing active bleed, black staining length and frequency contributed to greater positive values of the DAI and more severe AOD. 
 Higher DAI values could also be obtained from trees with greater stature (Figure 4c). 

### 2020-03-04 manuscript/figures.Rmd
Added additional Missing crown and crown transparency descriptors to COD response surface plots. Corrected caption
altered methods to reflect use of total height and lower crown height in DAI response surfaces. Lower bound of total height now upper bound of lower crown height to avoid impossible situations
corrections to tables,figures and supplementary

@@
 
-```{r decline_index_plot,fig.cap='Decline indexes for all 174 trees.
+```{r decline_index_plot,fig.cap='Decline indexes for all 174 surveyed trees.
 The chosen group categories were selected in the field prior to data collection.
@@
 
-```{r descriptor_contributions, fig.height = 6,fig.cap = 'Descriptor mean decrease in accuracy importance scores from the random forest models $M_{PDI}$ and $M_{DAI}$ for relative contributions to the decline index measures.
+```{r descriptor_contributions, fig.height = 6,fig.cap = 'Descriptor importance scores from the random forest models $M_{PDI}$ and $M_{DAI}$ for relative contributions to the decline index measures.
 Descriptors are ordered by descending percentage increase in mean squared error (MSE) and therefore from highest contribution to lowest contribution to the given decline index.'}
@@
 The neutral case is a tree that can be neither distinguished as AOD or COD.
-All cases have PDI values > 0.5.'}
+All cases have PDI values > 0.5.
+The values of additional descriptors related to bleed prevalence and relative stature/crown condition are shown for the AOD and COD example cases respectively.'}
 loadd(DAI_response_surfaces)

### 2020-03-04 manuscript/supplementary.Rmd
corrections to tables,figures and supplementary

@@
 site_correction_factors %>%
+  rename(Adjustment = Correction) %>%
   mutate_if(is.double,signif,digits = 3) %>%

### 2020-03-04 manuscript/tables.Rmd
corrections to tables,figures and supplementary

@@
         escape = FALSE,
-        caption = 'LIME results for oak decline severity cases in the PDI index model. Top four important descriptors show. Descriptors ordered by descending absolute weight within each case.') %>%
+        caption = 'LIME results for oak decline severity cases in the PDI index model $M_{PDI}$.
+Top four important descriptors show.
+Descriptors ordered by descending absolute weight within each case.') %>%
   collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
@@
         escape = FALSE,
-        caption = 'LIME results of oak decline type and severity in the DAI index model. Top four important descriptors shown. Descriptors ordered by descending absolute weight within each case.') %>%
+        caption = 'LIME results of oak decline type and severity in the DAI index model $M_{DAI}$.
+Top four important descriptors shown.
+Descriptors ordered by descending absolute weight within each case.') %>%
   column_spec(5,width = '4cm') %>%

### 2020-03-09 R/functions/surveyors.R
Nathan additions

@@ -1,4 +1,4 @@
 
 surveyors <- function(){
-  c('S. Coventry','I. Craig','J. McMinn','Stephen Whall')
+  c('Steve Coventry','Ian Craig','Joe McMinn','Stephen Whall')
 }

### 2020-03-09 manuscript/manuscript.Rmd
Nathan additions
Sandra in text alterations

@@
 Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of calloused tissue on the exterior [@denman2014].
-Stem bleeds can also, but not always, be associated with *Armillaria* attack on the inner stem tissues of oak.
+Stem bleeds can also be associated with a range of root pathogens including *Armillaria* attack on the inner stem tissues of oak.
 Active stem weeping and black staining can be found along the leading edge of the mycelial front of affected stem tissues [@denman2017].
@@
 
-As a result of the variety of factors contributing to oak decline, there are many visual phenotypic descriptors that contribute to categorizing an oak tree as being in decline. 
+As a result of the variety of factors contributing to oak decline and oak responses to these, there are many visual phenotypic descriptors that contribute to categorizing an oak tree as being in decline.
 When the decline is severe there are clear symptoms visible in the crown, such as low foliage density and numerous dead branches. 
-It is these characteristics that many studies use as the primary indicator of tree decline severity [@zarnoch2004], however these symptoms can develop slowly (especially in the case of COD) and simple classes are unlikely to capture the full detail of the subtle transition from being healthy to declined.
+It is these characteristics that many studies use as the primary indicator of tree decline severity [@zarnoch2004], however symptoms can develop slowly (especially in COD) and simple classes are unlikely to capture the full detail of the subtle transition from being healthy to declined.
 
-The visual, manual assignment of decline status and severity in the field is variable and error prone due to the many factors contributing to oak decline.
+The visual, manual assignment of decline status and severity in the field is variable and error prone due to the many factors contributing to oak decline and surveyor bias.
 For instance, accurate and reproducible crown condition assessments often require rigorous quality assurance steps that include training and regular calibration among surveyors [@ferretti1997; @redfern2004].
@@
 Machine learning is finding increasing utility in ecological and environmental applications due to its ability to learn complex patterns within and between variables [@cutler2007;@belgiu2016].
-A popular algorithm is random forest due to it's versatility, relatively trivial training requirements and ease of use. 
+A popular algorithm is random forest due to it's versatility, relatively simple model training requirements and ease of use. 
 It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag sample prediction accuracies and variable importance [@breiman2001a;@biau2016].
 
-Here, we propose a framework using unsupervised random forest machine learning to enable quantitative, visible oak decline severity measurement that utilises a range of targeted phenotypic descriptors observed in oak forests across England.
+Here, we propose a framework using unsupervised random forest machine learning to enable quantitative, visible oak decline severity assessments that utilises a range of oak decline targeted phenotypic descriptors observed in oak forests across England.
 
@@
 All trees were manually assigned a decline status category; healthy, AOD, COD or remission prior to collection of phenotypic descriptors.
-At the COD sites, trees were either COD (trees with clear signs of poor crown condition and canopy dieback) or healthy (dominant or co-dominant trees with full crowns). 
-At Big Wood, trees with stronger crowns were also categorised as COD when *Armillaria* fruiting bodies had been observed.
+At the COD sites, trees were either COD (trees with clear signs of poor crown condition and dieback) or healthy (dominant or co-dominant trees with full crowns). 
+At Big Wood, trees with stronger crowns were also categorised as COD when *Armillaria* (*Desarmillaria tabescens*) fruiting bodies had been observed on their buttress roots.
 At the AOD sites, annual monitoring data collected since 2009 was used to guide tree selection.
-Healthy trees had no known history of AOD and fruiting bodies of root pathogens had not been observed; all trees had canopy condition of 4 or 5, see @brown2016 for details of this assessment.
+Healthy trees had no known history of AOD and fruiting bodies of root pathogens had not been observed; all trees had crown condition of 4 or 5, see @brown2016 for details of this assessment.
 AOD trees had current stem bleeds but a variety of crown conditions. 
@@
 
-The collected phenotypic descriptors related to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
+The collected phenotypic descriptors related to tree shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi are listed and described in Supplementary Table 1.
+Surveyors involved in phenotyping the trees used in this study had prior experience conducting ICP-Forests level I crown condition surveys. 
+In each year of sampling surveyors met to undergo refresher training and “calibration” assessments in oak woodland. 
 The descriptors collected were based on @innes1990 and @icpforests2016.
@@
 The timber height was measured at either where the top stem diameter was 70mm or where no main stem could be distinguished.
-For crown contact, missing crown and absolute crown transparency, percentages were visually estimated within 5% classes
+For crown contact, missing crown and absolute crown transparency, percentages were visually estimated between 0 and 100% in 5% interval classes.
 
@@
 
-Crown radius, canopy closure and tap tests for dead stem tissue were measured for each cardinal point.
+Crown radius, canopy closure and bark tap tests for dead stem tissue were measured for each cardinal point.
 The crown radius was measured using a vertex transponder attached to the relevant side of the stem at a height of 1.3m.
@@
 The frequency of *A.
-biguttatus* exit holes were counted along with the presence or absence of each of oval shaped exit holes, small circular shaped exit holes and ground level fungal fruiting bodies which were recorded from the basal 2m of the stem.
-The collected descriptors are all included in Supplementary Table 1.
+biguttatus* exit holes were counted along with the presence or absence of other oval shaped insect exit holes, small circular shaped exit holes and ground level fungal fruiting bodies which were recorded around the stem circumference from the basal tree collar to a radial distance of 2m of the stem.
 
@@
 All descriptors collected in cm were converted to mm except for DBH which was converted to m.
-The tap tests that were positive for hollow stem tissue were converted to present and absent then renamed dead stem tissue.
+The bark tap tests that were positive for hollow stem tissue were converted to present and absent then renamed dead stem tissue.
 The canopy closure descriptor was converted from yes and no to present and absent.
-Finally, the stem symptoms lengths for active bleeds, black staining and calloused wounds were averaged and a frequency per tree given by the total observed.
+Finally, the stem symptoms lengths for active bleeds, black staining and callused wounds were averaged and a frequency per tree given by the total observed.
 The application and effect of this adjustment is described and discussed in Sections 3.1 and 4.3 respectively.
@@
 
-Composite crown volume (m^3^) - estimated volume of the crown as an paraboloid proportional to the crown condition.
+Composite crown volume (m^3^) - estimated volume of the crown as a paraboloid proportional to the crown condition.
 Adapted from the composite crown volume from @zarnoch2004 using crown condition instead of crown density alone.
@@
 These models will subsequently be referred to as $M_{PDI}$ and $M_{DAI}$ for the PDI and DAI respectively. 
-The parameters number of trees ($ntree$) and number of randomly assigned variables ($mtry$), for each random forest model, were tuned prior to training using the values of 10, 100, 1,000, 10,000 for ntree and a range of 2-26 at intervals of 4 for the mtry parameter.
+The parameters number of trees ($ntree$) and number of randomly assigned variables ($mtry$), for each random forest model, were tuned prior to training using the values of 10, 100, 1,000, 10,000 for $ntree$ and a range of 2-26 at intervals of 4 for the $mtry$ parameter.
 The combination parameters giving the lowest mean absolute percentage error (MAPE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, MAPE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, MAPE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
@@
 
-The DAI ($M_{DAI}$) is a much more complex index with more descriptors contributing to it than the PDI such as live crown ratio, bleed prevalence, and DBH. 
+The DAI ($M_{DAI}$) is a more complex index with a greater number of descriptors contributing to it than the PDI such as live crown ratio, bleed prevalence, and DBH. 
 Simulated response surfaces for the DAI are shown in Figure 4. 
-These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and live crown ratio descriptors (Figures 4d, e & f) change for non-symptomatic AOD or COD trees (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively).
+These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and live crown ratio descriptors (Figures 4d, e & f) change for neutral decline type (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively).
 Increasing active bleed, black staining length and frequency contributed to greater positive values of the DAI and more severe AOD. 
@@
 Thanks to Liz Richardson and Mark Oram for coordinating surveying and the surveyors: `r str_c(surveyors_names,collapse = ', ')`.
-This work was supported by Woodland Heritage and the Future Proofing Plant Health (FPPH) Oak Resilience project.
+This work was supported by Woodland Heritage, the Forestry Commission and the Defra Future Proofing Plant Health (FPPH) Oak Resilience project.
 

### 2020-03-13 manuscript/manuscript.Rmd
package cache update. Couple of small manuscript adjustments

@@
 
-Machine learning is finding increasing utility in ecological and environmental applications due to its ability to learn complex patterns within and between variables [@cutler2007;@belgiu2016].
+Machine learning is finding increasing utility in ecological and environmental applications due to it's ability to learn complex patterns within and between variables [@cutler2007;@belgiu2016].
 A popular algorithm is random forest due to it's versatility, relatively simple model training requirements and ease of use. 
-It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both out of bag sample prediction accuracies and variable importance [@breiman2001a;@biau2016].
+It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both 'out of bag' sample prediction accuracies and variable importance [@breiman2001a;@biau2016].
 
@@
 
-The collected phenotypic descriptors related to tree shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi are listed and described in Supplementary Table 1.
+The collected phenotypic descriptors related to tree size, crown condition and the presence of biotic agents including insects, bacteria and fungi are listed and described in Supplementary Table 1.
 Surveyors involved in phenotyping the trees used in this study had prior experience conducting ICP-Forests level I crown condition surveys. 

### 2020-07-16 R/functions/siteTable.R
added site information overview table

@@ -1,0 +1,15 @@
+
+siteTable <- function(){
+  tribble(
+    ~Site, ~`UK Region`, ~`Site description`, ~`Decline types`, ~`Survey date`,
+    1, 'West Midlands', 'Oak dominated amenity woodland, parkland plantation.', 'AOD/COD', 'June 2017',
+    2, 'East of England', 'Oak dominated woodland managed for timber.', 'AOD/COD', 'September 2016',
+    3, 'South West', 'Oak dominated plantation managed for timber.', 'COD', 'August 2016',
+    4, 'East of England', 'Oak dominated woodland managed for timber, with lapsed lime coppice.', 'AOD/COD', 'June 2017',
+    5, 'South East', 'Parkland with shelterbelts.', 'AOD/COD', 'July 2017',
+    6, 'South West', 'Open high forest, mown grass below.', 'AOD/COD', 'June 2017',
+    7, 'London', 'Oak dominated urban woodland, boundary belt of park.', 'AOD/COD', 'August 2017',
+    8, 'South West', 'Oak dominated amenity woodland.', 'AOD/COD', 'August 2016',
+    9, 'East of England', 'Oak dominated woodland managed for timber.', 'AOD/COD', 'June 2017',
+  )
+}

### 2020-07-16 R/plan.R

@@ -28,6 +28,9 @@ plan <- drake_plan(
   ## create site location map
   site_location_map = locationMap(),
   
+  ## Site information table
+  site_information = siteTable(),
+  
   ## Analyse site differences using supervised random forest
   site_differences_rf = analysis_suitable_data %>%
     stratifiedRF(pheno_data_with_additional_descriptors$Location %>% factor(),

### 2020-07-16 manuscript/manuscript.Rmd
fixed keywords
fixed summarise message outputs

@@
 
-**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning, unsupervised learning, random forest, decline spectrum
+**Keywords:** Acute Oak Decline, Chronic Oak Decline, phenotyping, machine learning, unsupervised learning, random forest, decline spectrum
 
@@
             maximum = max(value),
-             center = minimum + (maximum - minimum) / 2
+            center = minimum + (maximum - minimum) / 2,
+            .groups = 'drop'
   )

### 2020-07-16 R/functions/margins.R

@@ -9,10 +9,12 @@ margins <- function(rfModels){
       margin(d) %>%
         tibble(Class = d$y, Margin = .) %>%
         group_by(Class) %>%
-        summarise(Margin = mean(Margin))
+        summarise(Margin = mean(Margin),
+                  .groups = 'drop')
     }) %>%
     bind_rows(.id = 'Repeat') %>%
     group_by(Class) %>%
     summarise(Margin = mean(Margin) %>%
-                round(3))
+                round(3),
+              .groups = 'drop')
 }

### 2020-07-16 R/functions/siteDifferencesRFplot.R

@@ -26,7 +26,8 @@ siteDifferencesRFplot <- function(site_differences_rf,phenoData){
     bind_rows(.id = 'Rep') %>%
     group_by(Descriptor) %>%
     summarise(MeanDecreaseAccuracy = mean(MeanDecreaseAccuracy),
-              GiniImportance = mean(MeanDecreaseGini)) %>%
+              GiniImportance = mean(MeanDecreaseGini),
+              .groups = 'drop') %>%
     arrange(MeanDecreaseAccuracy) %>%
     mutate(Descriptor = factor(Descriptor,levels = Descriptor))
   

### 2020-07-16 R/functions/varSummary.R

@@ -4,5 +4,5 @@ varSummary <- function(dat){
     select(Location,ID,`Diameter at breast height (m)`) %>%
     group_by(Location) %>%
     summarise(Mean = mean(`Diameter at breast height (m)`),
-              SE = sd(`Diameter at breast height (m)`)/ sqrt(n()))
+              SE = sd(`Diameter at breast height (m)`)/ sqrt(n()),.groups = 'drop')
 }

### 2020-07-16 R/plan.R

@@ -71,7 +71,7 @@ plan <- drake_plan(
   ## Calculate margin for site differences post adjustment
   site_rf_post_correction_margin = site_differences_rf_post_correction %>%
     margins() %>%
-    summarise(Margin = mean(Margin)),
+    summarise(Margin = mean(Margin),.groups = 'drop'),
   
   ## Calculate AUC for site differences post adjustment
   site_rf_post_correction_auc = site_differences_rf_post_correction %>%
@@ -115,7 +115,8 @@ plan <- drake_plan(
     filter(ChosenGroup %in% c('AOD','COD')) %>% 
     group_by(ChosenGroup) %>% 
     summarise(median = median(DAI),
-              mean = mean(DAI)),
+              mean = mean(DAI),
+              .groups = 'drop'),
   
   ## Remission PDI t-test
   remission_PDI_group_ttest = decline_indexes %>%

### 2020-07-16 manuscript/manuscript.Rmd
added site surveyed tree totals to site information table. Referenced site information table in text

@@
 
-The phenotypic data were collected over two summers between June and September from a total of 174 trees at nine sites across the south of England.
-In 2016 three COD sites were visited (Big Wood,Chestnuts and Speculation) and in 2017 six AOD sites were added to the dataset (Attingham, Great Monks Wood, Hatchlands, Langdale Wood, Richmond Park and Winding Wood).
+The phenotypic data were collected over two summers between June and September from a total of 174 trees at nine sites across the south of England (Table 1).
 The occurrence of AOD has previously been reported and studied at these sites [@brown2016; @brown2017a].
-These sites included a variety of woodlands, management histories and soil types as well as tree sizes/ages and were made up of mainly *Q. robur*.
+These sites included a variety of woodlands, management histories as well as tree sizes/ages and were made up of mainly *Q. robur*.
 

### 2020-07-16 R/functions/siteTable.R

@@ -1,7 +1,7 @@
 
 siteTable <- function(){
   tribble(
-    ~Site, ~`UK Region`, ~`Site description`, ~`Decline types`, ~`Survey date`,
+    ~Site, ~`UK Region`, ~`Site description`, ~`Oak decline types present`, ~`Survey date`,
     1, 'West Midlands', 'Oak dominated amenity woodland, parkland plantation.', 'AOD/COD', 'June 2017',
     2, 'East of England', 'Oak dominated woodland managed for timber.', 'AOD/COD', 'September 2016',
     3, 'South West', 'Oak dominated plantation managed for timber.', 'COD', 'August 2016',
@@ -9,7 +9,7 @@ siteTable <- function(){
     5, 'South East', 'Parkland with shelterbelts.', 'AOD/COD', 'July 2017',
     6, 'South West', 'Open high forest, mown grass below.', 'AOD/COD', 'June 2017',
     7, 'London', 'Oak dominated urban woodland, boundary belt of park.', 'AOD/COD', 'August 2017',
-    8, 'South West', 'Oak dominated amenity woodland.', 'AOD/COD', 'August 2016',
+    8, 'South West', 'Oak dominated amenity woodland.', 'COD', 'August 2016',
     9, 'East of England', 'Oak dominated woodland managed for timber.', 'AOD/COD', 'June 2017',
   )
 }

### 2020-07-16 R/plan.R

@@ -26,7 +26,14 @@ plan <- drake_plan(
     descriptorInfo(),
   
   ## Site information table
-  site_information = siteTable(),
+  site_information = siteTable() %>%
+    left_join(pheno_data %>%
+                select(Site = Location) %>%
+                mutate(Site = str_remove_all(Site,'Site ') %>%
+                         as.numeric()) %>%
+                group_by(Site) %>%
+                summarise(`Trees surveyed` = n(),
+                          .groups = 'drop'),by = 'Site'),
   
   ## Analyse site differences using supervised random forest
   site_differences_rf = analysis_suitable_data %>%

### 2020-07-16 manuscript/manuscript.Rmd
updated in-text table references

@@
 
-LIME identified crown volume, condition, transparency and radius as the top four descriptors contributing to $M_{PDI}$ in each of the healthy, moderate and severe decline cases (Table 1).
+LIME identified crown volume, condition, transparency and radius as the top four descriptors contributing to $M_{PDI}$ in each of the healthy, moderate and severe decline cases (Table 2).
 These descriptors were also among the same top descriptors in the overall model importance lists for the PDI (Figure 2a).
@@
 
-For $M_{DAI}$, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table 2).
+For $M_{DAI}$, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table 3).
 This likely reflects the greater number of descriptors contributing to the DAI with different descriptors relevant to particular decline contexts.

### 2020-07-24 manuscript/manuscript.Rmd
renamed correction objects to adjustment

@@
 loadd(site_rf_margin,
-      site_rf_post_correction_margin,
+      site_rf_post_adjustment_margin,
       PDI_status_ttest,
@@
       analysis_suitable_data,
-      site_corrected_analysis_suitable_data,
+      site_adjusted_analysis_suitable_data,
       remission_PDI_group_ttest,
@@
 The value ranges were based on the observed range in the training data.
-For the missing crown (`r glue("{min(site_corrected_analysis_suitable_data[['Missing crown (%)']])}-{max(site_corrected_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_corrected_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors, the ranges were divided into 100 increments.
+For the missing crown (`r glue("{min(site_adjusted_analysis_suitable_data[['Missing crown (%)']])}-{max(site_adjusted_analysis_suitable_data[['Missing crown (%)']])}%")`) and crown transparency (`r glue("{min(site_adjusted_analysis_suitable_data[['Crown transparency (%)']])}-{max(site_adjusted_analysis_suitable_data[['Crown transparency (%)']])}%")`) descriptors, the ranges were divided into 100 increments.
 For a given case, each combination of these simulated descriptor values were used to calculate a PDI value from $M_{PDI}$, building a model response surface.
@@
 ```{r stem_symptom_ranges}
-symptom_ranges <- site_corrected_analysis_suitable_data %>%
+symptom_ranges <- site_adjusted_analysis_suitable_data %>%
   select(`Active bleed length (mm)`,`Black staining length (mm)`) %>%
@@
 These were reduced by a factor of eight to give ranges of `r glue("{symptom_ranges$minimum[1]}-{signif(symptom_ranges$center[1] / 8,digits = 3)}mm")` and `r glue("{symptom_ranges$minimum[2]}-{signif(symptom_ranges$center[2] / 8,digits = 3)}mm")` respectively and enabled more effective visualisation.
-For the total height (`r glue("{max(site_corrected_analysis_suitable_data[['Lower crown height (m)']]) %>% signif(3)} - {max(site_corrected_analysis_suitable_data[['Total height (m)']]) %>% signif(3)}m")`) and lower crown height (`r glue("{min(site_corrected_analysis_suitable_data[['Lower crown height (m)']]) %>% signif(3)} - {max(site_corrected_analysis_suitable_data[['Lower crown height (m)']]) %>% signif(3)}m")`) descriptors, the ranges were divided into 100 increments.
+For the total height (`r glue("{max(site_adjusted_analysis_suitable_data[['Lower crown height (m)']]) %>% signif(3)} - {max(site_adjusted_analysis_suitable_data[['Total height (m)']]) %>% signif(3)}m")`) and lower crown height (`r glue("{min(site_adjusted_analysis_suitable_data[['Lower crown height (m)']]) %>% signif(3)} - {max(site_adjusted_analysis_suitable_data[['Lower crown height (m)']]) %>% signif(3)}m")`) descriptors, the ranges were divided into 100 increments.
 
@@
 This ensures that the subsequent comparisons of these descriptors in the context of oak decline are not confounded by these identified site differences.
-After the application of the tree size adjustment, the average random forest *margin* value was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)`.
+After the application of the tree size adjustment, the average random forest *margin* value was substantially reduced to `r site_rf_post_adjustment_margin$Margin %>% signif(digits = 3)`.
 The reduction in this value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were no longer able accurately classify the site locations of the trees (Supplementary figure 4a) with almost no contribution of the corrected tree size descriptors (Supplementary figure 4b).
@@
 
-The calculated PDI and DAI scores (see Supplementary information #) derived from MDS of proximity values from unsupervised random forest for all `r nrow(site_corrected_analysis_suitable_data)` trees for which phenotypic descriptor data was collected is shown in Figure 1.
+The calculated PDI and DAI scores (see Supplementary information #) derived from MDS of proximity values from unsupervised random forest for all `r nrow(site_adjusted_analysis_suitable_data)` trees for which phenotypic descriptor data was collected is shown in Figure 1.
 The healthy trees grouped towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.

### 2020-07-24 manuscript/supplementary.Rmd
renamed correction objects to adjustment

--- 2020-02-10 manuscript/supplementary.Rmd
@@
 ```{r descriptor_adjustment_factors}
-loadd(site_correction_factors)
+loadd(site_adjustment_factors)
 
-site_correction_factors %>%
+site_adjustment_factors %>%
   mutate_if(is.double,signif,digits = 3) %>%
@@
 
+```{r site_differences_corrected,fig.cap='Differences between sites in phenotypic descriptors after adjustment of tree size descriptors analysed by supervised random forest classification.
+a) Multidimensional scaling (MDS) scatter plot of random forest proximity values of individual trees scaled to two dimensions.
+b) Descriptor mean decrease in accuracy importances measures.
+Descriptors ordered in descending importance.',fig.height=6}
+loadd(site_differences_post_adjustment_mds_plot)
+grid.arrange(site_differences_post_adjustment_mds_plot$a,site_differences_post_adjustment_mds_plot$b,nrow = 1)
+```
+
 ```{r PDInumeric,fig.height=9,fig.cap='Continuous descriptors plotted against the PDI.'}

### 2020-07-24 R/plan.R

@@ -53,49 +53,49 @@ plan <- drake_plan(
   site_differences_mds_plot = site_differences_rf %>%
     siteDifferencesRFplot(pheno_data_with_additional_descriptors),
   
-  ## calculate descriptor correction factors
-  site_correction_factors = pheno_data_with_additional_descriptors %>%
-    siteCorrectionFactors(),
+  ## calculate descriptor adjustment factors
+  site_adjustment_factors = pheno_data_with_additional_descriptors %>%
+    siteAdjustmentFactors(),
   
-  ## apply site corrections and recalculate additional descriptors
-  site_corrected_pheno_data = pheno_data %>%
-    siteCorrection() %>%
+  ## apply site adjustments and recalculate additional descriptors
+  site_adjusted_pheno_data = pheno_data %>%
+    siteAdjustment() %>%
     calcAdditionalDescriptors(),
   
   ## plot descriptor adjustment example using dbh
   descriptor_adjustment_example = pheno_data_with_additional_descriptors %>%
-    descriptorAdjustmentPlot(site_corrected_pheno_data),
+    descriptorAdjustmentPlot(site_adjusted_pheno_data),
   
-  ## make site corrected analysis suitable table
-  site_corrected_analysis_suitable_data = site_corrected_pheno_data %>%
+  ## make site adjusted analysis suitable table
+  site_adjusted_analysis_suitable_data = site_adjusted_pheno_data %>%
     makeAnalysisTable(),
   
-  ## re-analyse site differences after site correction
-  site_differences_rf_post_correction = site_corrected_analysis_suitable_data %>%
+  ## re-analyse site differences after site adjustment
+  site_differences_rf_post_adjustment = site_adjusted_analysis_suitable_data %>%
     stratifiedRF( pheno_data_with_additional_descriptors$Location %>% factor(),
        n = 100),
   
   ## Calculate margin for site differences post adjustment
-  site_rf_post_correction_margin = site_differences_rf_post_correction %>%
+  site_rf_post_adjustment_margin = site_differences_rf_post_adjustment %>%
     margins() %>%
     summarise(Margin = mean(Margin),.groups = 'drop'),
   
   ## Calculate AUC for site differences post adjustment
-  site_rf_post_correction_auc = site_differences_rf_post_correction %>%
+  site_rf_post_adjustment_auc = site_differences_rf_post_adjustment %>%
[…]

### 2020-07-24 manuscript/manuscript.Rmd
added pdi citation

@@
 This dimension was named the Decline Acuteness Index, DAI.
-The code used to generate the decline indexes is available as an R package pdi version `r packageVersion('pdi') %>% as.character()`, available at https://github.com/jasenfinch/pdi.
+The code used to generate the decline indexes is available as an R package pdi version `r packageVersion('pdi') %>% as.character()` [@finch2020].
 

### 2020-07-27 manuscript/manuscript.Rmd
added reference to the decliner deployment
Finished comments from Nathan and Sandra

@@
 A popular algorithm is random forest due to it's versatility, relatively simple model training requirements and ease of use. 
-It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both 'out of bag' sample prediction accuracies and variable importance [@breiman2001a;@biau2016].
+It is an ensemble machine learning method where forests of bootstrap sampled decision trees are grown, yielding both 'out of bag' sample prediction accuracy and variable importance [@breiman2001a;@biau2016].
 
-Here, we propose a framework using unsupervised random forest machine learning to enable quantitative, visible oak decline severity assessments that utilises a range of oak decline targeted phenotypic descriptors observed in oak forests across England.
+Here, we propose a framework using unsupervised random forest machine learning to enable quantitative, visible oak decline severity assessments that utilise a range of oak decline informative phenotypic descriptors observed in oak forests across England.
 
@@
 All trees were manually assigned a decline status category; healthy, AOD, COD or remission prior to collection of phenotypic descriptors.
-At the COD sites, trees were either COD (trees with clear signs of poor crown condition and dieback) or healthy (dominant or co-dominant trees with full crowns). 
-At Big Wood, trees with stronger crowns were also categorised as COD when *Armillaria* (*Desarmillaria tabescens*) fruiting bodies had been observed on their buttress roots.
-At the AOD sites, annual monitoring data collected since 2009 was used to guide tree selection.
+At sites with only COD present, trees were either COD (trees with clear signs of poor crown condition and dieback) or healthy (dominant or co-dominant trees with full crowns). 
+At Site 2, trees with stronger crowns were also categorised as COD when *Armillaria* (*Desarmillaria tabescens*) fruiting bodies had been observed on their buttress roots.
+At sites with AOD present, annual monitoring data collected since 2009 was used to guide tree selection.
 Healthy trees had no known history of AOD and fruiting bodies of root pathogens had not been observed; all trees had crown condition of 4 or 5, see @brown2016 for details of this assessment.
@@
 The timber height was measured at either where the top stem diameter was 70mm or where no main stem could be distinguished.
-For crown contact, missing crown and absolute crown transparency, percentages were visually estimated between 0 and 100% in 5% interval classes.
+For crown contact, missing crown and absolute crown transparency, percentages were visually estimated between 0 and 100% in 5% interval ranges.
 
@@
 Plots of the individual descriptors against the decline indexes are available in Supplementary Figures 5-8. 
+A web application is also available at https://jasenfinch.shinyapps.io/decliner/ that allows interactive input of phenotypic descriptor values into the $M_PDI$ and $M_DAI$ models.
 The following sections (3.4, 3.5) identify how the descriptors contribute and interact within the decline index models in specific decline cases.
@@
 Simulated response surfaces for the DAI are shown in Figure 4. 
-These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and live crown ratio descriptors (Figures 4d, e & f) change for neutral decline type (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively).
+These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and live crown ratio descriptors (Figures 4d, e & f) change for neutral decline type (trees neither distinctively AOD or COD; Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively).
 Increasing active bleed, black staining length and frequency contributed to greater positive values of the DAI and more severe AOD. 
@@
 
-The decline indexes presented here provide single intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity.
+The decline indexes presented here provide simple intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity.
 These two indexes can be represented as a two-dimensional oak decline continuum under which the spectrum of decline severity and type can be assessed (Figure 5).

### 2020-08-11 manuscript/manuscript.Rmd
removed proposed journals
added tables and figures to manuscript

@@
 header-includes:
+  - \usepackage{booktabs}
+  - \usepackage{longtable}
+  - \usepackage{array}
+  - \usepackage{multirow}
+  - \usepackage{wrapfig}
+  - \usepackage{float}
+  - \usepackage{colortbl}
+  - \usepackage{pdflscape}
+  - \usepackage{tabu}
+  - \usepackage{threeparttable}
+  - \usepackage{threeparttablex}
+  - \usepackage[normalem]{ulem}
+  - \usepackage{makecell}
+  - \usepackage{xcolor}
+  - \usepackage{lineno}
   - \linespread{2}
-  - \usepackage{lineno}
   - \linenumbers
@@
 ```
-
-**Proposed journal:** (1) Forest Ecology & Management (2) Forestry (3) European Journal of Forest Research (4) Plant Methods
 
@@
 
+# Tables
+
+```{r site_information}
+readd(site_information) %>%
+  kable(format = 'latex',
+        booktabs = TRUE,
+        caption = 'Overview of the surveyed site information.') %>%
+  column_spec(3,width = '4cm') %>%
+  column_spec(4,width = '3cm')
+```
+
+```{r PDI_lime_results}
+loadd(PDI_lime_analysis_table)
+
+PDI_lime_analysis_table %>%
+  kable(format = 'latex',
+        booktabs = TRUE,
+        escape = FALSE,
+        caption = 'LIME results for oak decline severity cases in the PDI index model $M_{PDI}$.
+Top four important descriptors show.
+Descriptors ordered by descending absolute weight within each case.') %>%
+  collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
+```
+
+```{r DAI_lime_results}
+loadd(DAI_lime_analysis_table)
+
+DAI_lime_analysis_table %>%
+  kable(format = 'latex',
+        booktabs = TRUE,
+        escape = FALSE,
+        caption = 'LIME results of oak decline type and severity in the DAI index model $M_{DAI}$.
+Top four important descriptors shown.
+Descriptors ordered by descending absolute weight within each case.') %>%
+  column_spec(5,width = '4cm') %>%
+  collapse_rows(columns = 1:4, latex_hline = "custom", valign = "middle",custom_latex_hline = 1:4)
+```
+
+# Figures
+
+```{r decline_index_plot,fig.cap='Decline indexes for all 174 surveyed trees.
+The chosen group categories were selected in the field prior to data collection.
+AOD = acute oak decline; COD = chronic oak decline.
+Remisson tress were those that were known to have had active bleeds within the previous 3 years.'}
+loadd(decline_indexes_plot)
+
+decline_indexes_plot
+```
+
+```{r descriptor_contributions, fig.height = 6,fig.cap = 'Descriptor importance scores from the random forest models $M_{PDI}$ and $M_{DAI}$ for relative contributions to the decline index measures.
+Descriptors are ordered by descending percentage increase in mean squared error (MSE) and therefore from highest contribution to lowest contribution to the given decline index.'}
+loadd(descriptor_contribution_plots)
+
+descriptor_contribution_plots$a + descriptor_contribution_plots$b
+```
+
+```{r PDI_response_surfaces,fig.height = 6,fig.cap='PDI response surfaces from $M_{PDI}$ for simulated values of key contributing descriptors for example healthy, moderate and severly declining cases.
+The values of additional tree size descriptors are shown for the individual cases to indicate the relative stature of the example cases.'}
+loadd(PDI_response_surfaces)
+
+grid.arrange(PDI_response_surfaces[[1]],
+             PDI_response_surfaces[[2]],
+             PDI_response_surfaces[[3]],
+             PDI_response_surfaces$legend,
+             nrow = 2,newpage = F)
+```
+
+```{r DAI_response_surfaces,fig.height = 8,fig.cap='DAI response surfaces from $M_{DAI}$ for simulated values of key contributing descriptors for example cases.
+The neutral case is a tree that can be neither distinguished as AOD or COD.
+All cases have PDI values > 0.5.
+The values of additional descriptors related to bleed prevalence and relative stature/crown condition are shown for the AOD and COD example cases respectively.'}
+loadd(DAI_response_surfaces)
+
+grid.arrange(textGrob(expression(bold('AOD')),hjust = -0.1),
+             textGrob(expression(bold('COD')),hjust = -0.1),
+             textGrob(expression(bold('Neutral')),rot = 90,hjust = -0.1),
+             textGrob(expression(bold('Moderate')),rot = 90,hjust = -0.1),
+             textGrob(expression(bold('Severe')),rot = 90,hjust = -0.1),
+             DAI_response_surfaces$`a)`,
+             DAI_response_surfaces$`b)`,
+             DAI_response_surfaces$`c)`,
+             DAI_response_surfaces$`d)`,
+             DAI_response_surfaces$`e)`,
+             DAI_response_surfaces$`f)`,
+             DAI_response_surfaces$legend,
+             ncol = 4,
+             layout_matrix = rbind(c(NA,1,2,NA),c(3,6,9,12),c(4,7,10,12),c(5,8,11,12)),
+             widths = c(1,10,10,2),
+             heights = c(1,10,10,10),
+             newpage = F)
+```
+
+```{r DI_space,fig.cap='Two dimensional continuum of the phenotypic decline indexes for oak decline severity.
+Decline category boundaries intended as a guide to aid interpretation only.
+Grey area indicates area for which the decline indexes are currently undefined.' ,fig.height=3,fig.width=5}
+DIspace()
+```
+
 # References

### 2020-08-11 R/plan.R

@@ -260,12 +260,6 @@ plan <- drake_plan(
   ## Surveyor names
   surveyors_names = surveyors(),
   
-  ## render tables
-  tables = render(knitr_in('manuscript/tables.Rmd'),quiet = T),
-  
-  ## render figures
-  figures = render(knitr_in('manuscript/figures.Rmd'),quiet = T),
-  
   ## render manuscript 
   manuscript = render(knitr_in('manuscript/manuscript.Rmd'),
                       quiet = T,

### 2020-08-12 manuscript/manuscript.Rmd
Placed figures and tables alongside relevant text. Cross referenced figures and tables in text

@@
 
-The phenotypic data were collected over two summers between June and September from a total of 174 trees at nine sites across the south of England (Table 1).
+The phenotypic data were collected over two summers between June and September from a total of 174 trees at nine sites across the south of England (Table \@ref(tab:site-information)).
 The occurrence of AOD has previously been reported and studied at these sites [@brown2016; @brown2017a].
 These sites included a variety of woodlands, management histories as well as tree sizes/ages and were made up of mainly *Q. robur*.
+
+```{r site-information}
+readd(site_information) %>%
+  kable(format = 'latex',
+        booktabs = TRUE,
+        caption = 'Overview of the surveyed site information.') %>%
+  column_spec(3,width = '4cm') %>%
+  column_spec(4,width = '3cm')
+```
 
@@
 The parameters number of trees ($ntree$) and number of randomly assigned variables ($mtry$), for each random forest model, were tuned prior to training using the values of 10, 100, 1,000, 10,000 for $ntree$ and a range of 2-26 at intervals of 4 for the $mtry$ parameter.
-The combination parameters giving the lowest mean absolute percentage error (MAPE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, MAPE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, MAPE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure 2).
+The combination parameters giving the lowest mean absolute percentage error (MAPE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, MAPE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, MAPE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary Figure 2).
 Overall descriptor importance contributions to each of the indexes was assessed using the percentage increase in mean squared error measure with higher values indicating a greater overall contribution to the given model and therefore greater importance.
@@
 
-```{r stem_symptom_ranges}
+```{r stem-symptom-ranges}
 symptom_ranges <- site_adjusted_analysis_suitable_data %>%
@@
 
-A preliminary analysis using supervised random forest classification could separate trees based on their location (Supplementary  figure 3a) with an average *margin* value of `r site_rf_margin$Margin %>% signif(digits = 3)`.
-The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary figure 3b).
+A preliminary analysis using supervised random forest classification could separate trees based on their location (Supplementary Figure 3a) with an average *margin* value of `r site_rf_margin$Margin %>% signif(digits = 3)`.
+The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary Figure 3b).
 To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in Section \@ref(method-adjustment).
@@
 After the application of the tree size adjustment, the average random forest *margin* value was substantially reduced to `r site_rf_post_adjustment_margin$Margin %>% signif(digits = 3)`.
-The reduction in this value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were no longer able accurately classify the site locations of the trees (Supplementary figure 4a) with almost no contribution of the corrected tree size descriptors (Supplementary figure 4b).
+The reduction in this value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were no longer able accurately classify the site locations of the trees (Supplementary Figure 4a) with almost no contribution of the corrected tree size descriptors (Supplementary Figure 4b).
 
@@
 
-The calculated PDI and DAI scores (see Supplementary information #) derived from MDS of proximity values from unsupervised random forest for all `r nrow(site_adjusted_analysis_suitable_data)` trees for which phenotypic descriptor data was collected is shown in Figure 1.
+The calculated PDI and DAI scores (see Supplementary information #) derived from MDS of proximity values from unsupervised random forest for all `r nrow(site_adjusted_analysis_suitable_data)` trees for which phenotypic descriptor data was collected is shown in Figure \@ref(fig:decline-index-plot).
 The healthy trees grouped towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
 There was a significant difference when the PDI values of the healthy (non-symptomatic) and symptomatic (AOD and COD) decline statuses were compared using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
+
+```{r decline-index-plot,fig.cap='Decline indexes for all 174 surveyed trees.
+The chosen group categories were selected in the field prior to data collection.
+AOD = acute oak decline; COD = chronic oak decline.
+Remisson tress were those that were known to have had active bleeds within the previous 3 years.'}
+loadd(decline_indexes_plot)
+
+decline_indexes_plot
+```
 
@@
 
-The relative contributions of the phenotypic descriptors to the decline index models ($M_{PDI}$,$M_{DAI}$) is shown in Figure 2.
+The relative contributions of the phenotypic descriptors to the decline index models ($M_{PDI}$,$M_{DAI}$) is shown in Figure \@ref(fig:descriptor-contributions).
 
-The top four ranked descriptors found to contribute to $M_{PDI}$ were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure 2a).
+The top four ranked descriptors found to contribute to $M_{PDI}$ were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure \@ref(fig:descriptor-contributions)a).
 Other tree size related descriptors such as bleed prevalence and live crown ratio stem bleeding descriptors contributed secondarily.
 
-$M_{DAI}$ was more complex with many more descriptors contributing more equally than for the PDI (Figure 2b).
+```{r descriptor-contributions, fig.height = 6,fig.cap = 'Descriptor importance scores from the random forest models $M_{PDI}$ and $M_{DAI}$ for relative contributions to the decline index measures.
+Descriptors are ordered by descending percentage increase in mean squared error (MSE) and therefore from highest contribution to lowest contribution to the given decline index.'}
+loadd(descriptor_contribution_plots)
+
+descriptor_contribution_plots$a + descriptor_contribution_plots$b
+```
+
+$M_{DAI}$ was more complex with many more descriptors contributing more equally than for the PDI (Figure \@ref(fig:descriptor-contributions)b).
 The overall tree size descriptors such as live crown ratio, DBH and crown volume contributed almost equally to $M_{DAI}$.
@@
 
-LIME identified crown volume, condition, transparency and radius as the top four descriptors contributing to $M_{PDI}$ in each of the healthy, moderate and severe decline cases (Table 2).
-These descriptors were also among the same top descriptors in the overall model importance lists for the PDI (Figure 2a).
+LIME identified crown volume, condition, transparency and radius as the top four descriptors contributing to $M_{PDI}$ in each of the healthy, moderate and severe decline cases (Table \@ref(tab:PDI-lime-results)).
+These descriptors were also among the same top descriptors in the overall model importance lists for the PDI (Figure \@ref(fig:descriptor-contributions)a).
 Increasing decline severity was associated with a reduction in the size and condition of the crown.
@@
 
-For $M_{DAI}$, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table 3).
+```{r PDI-lime-results}
+loadd(PDI_lime_analysis_table)
+
+PDI_lime_analysis_table %>%
+  kable(format = 'latex',
+        booktabs = TRUE,
+        escape = FALSE,
+        caption = 'LIME results for oak decline severity cases in the PDI index model $M_{PDI}$.
+Top four important descriptors show.
+Descriptors ordered by descending absolute weight within each case.') %>%
+  collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
+```
+
+For $M_{DAI}$, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table \@ref(tab:DAI-lime-results)).
 This likely reflects the greater number of descriptors contributing to the DAI with different descriptors relevant to particular decline contexts.
@@
 
+```{r DAI-lime-results}
+loadd(DAI_lime_analysis_table)
+
+DAI_lime_analysis_table %>%
+  kable(format = 'latex',
+        booktabs = TRUE,
+        escape = FALSE,
+        caption = 'LIME results of oak decline type and severity in the DAI index model $M_{DAI}$.
+Top four important descriptors shown.
+Descriptors ordered by descending absolute weight within each case.') %>%
+  column_spec(5,width = '4cm') %>%
+  collapse_rows(columns = 1:4, latex_hline = "custom", valign = "middle",custom_latex_hline = 1:4)
+```
+
 ## Key descriptor interactions within the decline index models
@@
 The index values, generated by the models from these simulated phenotypic data, were then visualised as model response surface plots, to allow interpretation of descriptor interactions between the indexes.
-Figures 3 and 4 show the simulated index value response surfaces under a variety of different circumstances for the key phenotypic descriptors described in Section \@ref(results-descriptor-contributions) that contribute highly to the PDI and DAI.
+Figures \@ref(fig:PDI-response-surfaces) and 4 show the simulated index value response surfaces under a variety of different circumstances for the key phenotypic descriptors described in Section \@ref(results-descriptor-contributions) that contribute highly to the PDI and DAI.
 
-Crown condition descriptors such as composite crown volume and crown condition contributed highly to $M_{PDI}$ (Figure 2a).
+```{r PDI-response-surfaces,fig.height = 6,fig.cap='PDI response surfaces from $M_{PDI}$ for simulated values of key contributing descriptors for example healthy, moderate and severly declining cases.
+The values of additional tree size descriptors are shown for the individual cases to indicate the relative stature of the example cases.'}
+loadd(PDI_response_surfaces)
+
+grid.arrange(PDI_response_surfaces[[1]],
+             PDI_response_surfaces[[2]],
+             PDI_response_surfaces[[3]],
+             PDI_response_surfaces$legend,
+             nrow = 2,newpage = F)
+```
+
+Crown condition descriptors such as composite crown volume and crown condition contributed highly to $M_{PDI}$ (Figure \@ref(fig:descriptor-contributions)a).
 The phenotypic measures central to this were percentage missing crown and percentage crown transparency from which crown condition and crown volume are calculated. 
-It can be seen in Figure 3 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and composite crown volume decrease, the PDI increases. 
+It can be seen in Figure \@ref(fig:PDI-response-surfaces) that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and composite crown volume decrease, the PDI increases.
 A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g.
-greater crown radius, total height, lower crown height and DBH; Figures 3a & b), while relatively small stature trees have PDIs of 0.6 or greater even with 100% crown condition.
+greater crown radius, total height, lower crown height and DBH; Figures \@ref(fig:PDI-response-surfaces)a & b), while relatively small stature trees have PDIs of 0.6 or greater even with 100% crown condition.
 
 The DAI ($M_{DAI}$) is a more complex index with a greater number of descriptors contributing to it than the PDI such as live crown ratio, bleed prevalence, and DBH. 
-Simulated response surfaces for the DAI are shown in Figure 4. 
-These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and live crown ratio descriptors (Figures 4d, e & f) change for neutral decline type (trees neither distinctively AOD or COD; Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively).
+Simulated response surfaces for the DAI are shown in Figure \@ref(fig:DAI-response-surfaces). 
+These response surfaces show circumstances under which bleed prevalence related descriptors (Figures \@ref(fig:DAI-response-surfaces)a, b & c) and live crown ratio descriptors (Figures \@ref(fig:DAI-response-surfaces)d, e & f) change for neutral decline type (trees neither distinctively AOD or COD; Figures \@ref(fig:DAI-response-surfaces)a & d respectively), moderate AOD or COD (Figures \@ref(fig:DAI-response-surfaces)b & e respectively) and severe AOD or COD (Figures \@ref(fig:DAI-response-surfaces)c & f respectively).
 Increasing active bleed, black staining length and frequency contributed to greater positive values of the DAI and more severe AOD. 
-Higher DAI values could also be obtained from trees with greater stature (Figure 4c). 
-Total height and lower crown height have no influence on the DAI for trees of neutral decline type (Figure 4d).
-Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD; however, only narrow ranges of DAI values could be obtained for each case when varying the ranges of these descriptors.
+Higher DAI values could also be obtained from trees with greater stature (Figure \@ref(fig:DAI-response-surfaces)c). 
+Total height and lower crown height have no influence on the DAI for trees of neutral decline type (Figure \@ref(fig:DAI-response-surfaces)d).
+Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures \@ref(fig:DAI-response-surfaces)e & f) and therefore more severe COD; however, only narrow ranges of DAI values could be obtained for each case when varying the ranges of these descriptors.
+
+```{r DAI-response-surfaces,fig.height = 8,fig.cap='DAI response surfaces from $M_{DAI}$ for simulated values of key contributing descriptors for example cases.
+The neutral case is a tree that can be neither distinguished as AOD or COD.
+All cases have PDI values > 0.5.
+The values of additional descriptors related to bleed prevalence and relative stature/crown condition are shown for the AOD and COD example cases respectively.'}
+loadd(DAI_response_surfaces)
+
+grid.arrange(textGrob(expression(bold('AOD')),hjust = -0.1),
+             textGrob(expression(bold('COD')),hjust = -0.1),
+             textGrob(expression(bold('Neutral')),rot = 90,hjust = -0.1),
+             textGrob(expression(bold('Moderate')),rot = 90,hjust = -0.1),
+             textGrob(expression(bold('Severe')),rot = 90,hjust = -0.1),
+             DAI_response_surfaces$`a)`,
+             DAI_response_surfaces$`b)`,
+             DAI_response_surfaces$`c)`,
+             DAI_response_surfaces$`d)`,
+             DAI_response_surfaces$`e)`,
+             DAI_response_surfaces$`f)`,
+             DAI_response_surfaces$legend,
+             ncol = 4,
+             layout_matrix = rbind(c(NA,1,2,NA),c(3,6,9,12),c(4,7,10,12),c(5,8,11,12)),
+             widths = c(1,10,10,2),
+             heights = c(1,10,10,10),
+             newpage = F)
+```
 
@@
 The decline indexes presented here provide simple intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity.
-These two indexes can be represented as a two-dimensional oak decline continuum under which the spectrum of decline severity and type can be assessed (Figure 5).
+These two indexes can be represented as a two-dimensional oak decline continuum under which the spectrum of decline severity and type can be assessed (Figure \@ref(fig:DI-space)).
 The index thresholds for category definitions (healthy, declining, severe decline etc.) shown currently serve only as a guide to aid human interpretation as opposed to definitive thresholds of disease.
@@
 
-Although significant differences were obtained between the decline status groups on both the PDI and DAI scales, Figure 1 highlights the inaccuracies in the manual decline status classifications when viewed in an individual tree context.
+```{r DI-space,fig.cap='Two dimensional continuum of the phenotypic decline indexes for oak decline severity.
+Decline category boundaries intended as a guide to aid interpretation only.
+Grey area indicates area for which the decline indexes are currently undefined.' ,fig.height=3,fig.width=5}
+DIspace()
+```
+
+Although significant differences were obtained between the decline status groups on both the PDI and DAI scales, Figure \@ref(fig:decline-index-plot) highlights the inaccuracies in the manual decline status classifications when viewed in an individual tree context.
 It can be seen that a number of the trees, when compared quantitatively using these measures, could in fact be classified with a different decline state.
@@
 However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
-It has been shown here that in the given visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure 2a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure 3a).
+It has been shown here that in the given visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure \@ref(descriptor-contributions)a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure \@ref(fig:PDI-response-surfaces)a).
 
@@
 
-AOD severity was able to be associated with overall decline severity with trees identified that had a combination of both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure 1).
+AOD severity was able to be associated with overall decline severity with trees identified that had a combination of both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure \@ref(fig:decline-index-plot)).
 The association of crown condition and AOD severity has previously be shown by @brown2016.
@@
 An unsupervised approach was favoured in this context due to the concession that there are known difficulties with and inaccuracy in manual decline status classification in the field [@ferretti1997].
-This is highlighted in Figure 1 where numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g.
+This is highlighted in Figure \@ref(fig:decline-index-plot) where numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (e.g.
 COD trees with DAI values > 0.5).
@@
 Thanks to Liz Richardson and Mark Oram for coordinating surveying and the surveyors: `r str_c(surveyors_names,collapse = ', ')`.
-This work was supported by Woodland Heritage, the Forestry Commission and the Defra Future Proofing Plant Health (FPPH) Oak Resilience project.
+This work was funded by Woodland Heritage, the Forestry Commission and the Defra Future Proofing Plant Health (FPPH) Oak Resilience project.
 
@@
 
-# Tables
-
-```{r site_information}
-readd(site_information) %>%
-  kable(format = 'latex',
-        booktabs = TRUE,
-        caption = 'Overview of the surveyed site information.') %>%
-  column_spec(3,width = '4cm') %>%
-  column_spec(4,width = '3cm')
-```
-
-```{r PDI_lime_results}
-loadd(PDI_lime_analysis_table)
-
-PDI_lime_analysis_table %>%
-  kable(format = 'latex',
-        booktabs = TRUE,
-        escape = FALSE,
-        caption = 'LIME results for oak decline severity cases in the PDI index model $M_{PDI}$.
-Top four important descriptors show.
-Descriptors ordered by descending absolute weight within each case.') %>%
-  collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
-```
-
-```{r DAI_lime_results}
-loadd(DAI_lime_analysis_table)
-
-DAI_lime_analysis_table %>%
-  kable(format = 'latex',
-        booktabs = TRUE,
-        escape = FALSE,
-        caption = 'LIME results of oak decline type and severity in the DAI index model $M_{DAI}$.
-Top four important descriptors shown.
-Descriptors ordered by descending absolute weight within each case.') %>%
-  column_spec(5,width = '4cm') %>%
-  collapse_rows(columns = 1:4, latex_hline = "custom", valign = "middle",custom_latex_hline = 1:4)
-```
-
-# Figures
-
-```{r decline_index_plot,fig.cap='Decline indexes for all 174 surveyed trees.
-The chosen group categories were selected in the field prior to data collection.
-AOD = acute oak decline; COD = chronic oak decline.
-Remisson tress were those that were known to have had active bleeds within the previous 3 years.'}
-loadd(decline_indexes_plot)
-
-decline_indexes_plot
-```
-
-```{r descriptor_contributions, fig.height = 6,fig.cap = 'Descriptor importance scores from the random forest models $M_{PDI}$ and $M_{DAI}$ for relative contributions to the decline index measures.
-Descriptors are ordered by descending percentage increase in mean squared error (MSE) and therefore from highest contribution to lowest contribution to the given decline index.'}
-loadd(descriptor_contribution_plots)
-
-descriptor_contribution_plots$a + descriptor_contribution_plots$b
-```
-
-```{r PDI_response_surfaces,fig.height = 6,fig.cap='PDI response surfaces from $M_{PDI}$ for simulated values of key contributing descriptors for example healthy, moderate and severly declining cases.
-The values of additional tree size descriptors are shown for the individual cases to indicate the relative stature of the example cases.'}
-loadd(PDI_response_surfaces)
-
-grid.arrange(PDI_response_surfaces[[1]],
-             PDI_response_surfaces[[2]],
-             PDI_response_surfaces[[3]],
-             PDI_response_surfaces$legend,
-             nrow = 2,newpage = F)
-```
-
-```{r DAI_response_surfaces,fig.height = 8,fig.cap='DAI response surfaces from $M_{DAI}$ for simulated values of key contributing descriptors for example cases.
-The neutral case is a tree that can be neither distinguished as AOD or COD.
-All cases have PDI values > 0.5.
-The values of additional descriptors related to bleed prevalence and relative stature/crown condition are shown for the AOD and COD example cases respectively.'}
-loadd(DAI_response_surfaces)
-
-grid.arrange(textGrob(expression(bold('AOD')),hjust = -0.1),
-             textGrob(expression(bold('COD')),hjust = -0.1),
-             textGrob(expression(bold('Neutral')),rot = 90,hjust = -0.1),
-             textGrob(expression(bold('Moderate')),rot = 90,hjust = -0.1),
-             textGrob(expression(bold('Severe')),rot = 90,hjust = -0.1),
-             DAI_response_surfaces$`a)`,
-             DAI_response_surfaces$`b)`,
-             DAI_response_surfaces$`c)`,
-             DAI_response_surfaces$`d)`,
-             DAI_response_surfaces$`e)`,
-             DAI_response_surfaces$`f)`,
-             DAI_response_surfaces$legend,
-             ncol = 4,
-             layout_matrix = rbind(c(NA,1,2,NA),c(3,6,9,12),c(4,7,10,12),c(5,8,11,12)),
-             widths = c(1,10,10,2),
-             heights = c(1,10,10,10),
-             newpage = F)
-```
-
-```{r DI_space,fig.cap='Two dimensional continuum of the phenotypic decline indexes for oak decline severity.
-Decline category boundaries intended as a guide to aid interpretation only.
-Grey area indicates area for which the decline indexes are currently undefined.' ,fig.height=3,fig.width=5}
-DIspace()
-```
-
 # References

### 2020-08-12 manuscript/manuscript.Rmd
removed word document render. Fixed figure captions

@@
 output: 
-  bookdown::word_document2:
-    reference_docx: template.docx
   bookdown::pdf_document2:
+    keep_tex: true
     toc: false
@@
 
+\linespread{1}
+
 ```{r site-information}
@@
 ```
+
+\linespread{2}
 
@@
 
+\linespread{1}
+
 ```{r stem-symptom-ranges}
@@
 
+\linespread{1}
+
 ```{r decline-index-plot,fig.cap='Decline indexes for all 174 surveyed trees.
@@
 
+\linespread{1}
+
 ```{r descriptor-contributions, fig.height = 6,fig.cap = 'Descriptor importance scores from the random forest models $M_{PDI}$ and $M_{DAI}$ for relative contributions to the decline index measures.
@@
 
+\linespread{1}
+
 ```{r PDI-lime-results}
@@
 
+\linespread{2}
+
 For $M_{DAI}$, a number of different descriptors were found to be locally important to the decline type cases compared to the overall model importance such as crown contact (Table \@ref(tab:DAI-lime-results)).
@@
 COD trees were found to have greater crown contact (greater than 50%) compared to AOD trees.
+
+\linespread{1}
 
@@
 
+\linespread{2}
+
 ## Key descriptor interactions within the decline index models
@@
 Figures \@ref(fig:PDI-response-surfaces) and 4 show the simulated index value response surfaces under a variety of different circumstances for the key phenotypic descriptors described in Section \@ref(results-descriptor-contributions) that contribute highly to the PDI and DAI.
+
+\linespread{1}
 
@@
 Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures \@ref(fig:DAI-response-surfaces)e & f) and therefore more severe COD; however, only narrow ranges of DAI values could be obtained for each case when varying the ranges of these descriptors.
+
+\linespread{1}
 
@@
 
+\linespread{2}
+
 # Discussion
@@
 
+\linespread{1}
+
 ```{r DI-space,fig.cap='Two dimensional continuum of the phenotypic decline indexes for oak decline severity.

### 2020-08-12 manuscript/manuscript.Rmd
fixed table formatting

@@
   column_spec(3,width = '4cm') %>%
-  column_spec(4,width = '3cm')
+  column_spec(4,width = '3cm') %>%
+  column_spec(5,width = '2cm') %>%
+  column_spec(6,width = '2cm')
 ```
@@
 Descriptors ordered by descending absolute weight within each case.') %>%
+  column_spec(1,width = '2cm') %>%
+  column_spec(3,width = '2cm') %>%
   collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
@@
 Descriptors ordered by descending absolute weight within each case.') %>%
-  column_spec(5,width = '4cm') %>%
+  column_spec(1,width = '1.5cm') %>%
+  column_spec(2,width = '1.5cm') %>%
+  column_spec(4,width = '2cm') %>%
+  column_spec(5,width = '3.5cm') %>%
+  column_spec(6,width = '3cm') %>%
   collapse_rows(columns = 1:4, latex_hline = "custom", valign = "middle",custom_latex_hline = 1:4)

### 2020-08-12 manuscript/supplementary.Rmd
added zipped directory of supplementary material

@@
 ---
-title: Supplementary information
 output: 
@@
   - \usepackage{xcolor}
+  - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
 ---
@@
 ```
+
+\beginsupplement
 

### 2020-08-12 R/plan.R

@@ -13,10 +13,6 @@ plan <- drake_plan(
   pheno_data_with_additional_descriptors = pheno_data %>%
     calcAdditionalDescriptors(),
   
-  ## export phenotypic data
-  export_pheno_data_with_additional_descriptors = pheno_data_with_additional_descriptors %>%
-    write_csv(file_out('data/exports/pheno_data_with_additional_descriptors.csv')),
-  
   ## make analysis suitable table
   analysis_suitable_data = pheno_data_with_additional_descriptors %>%
     makeAnalysisTable(),
@@ -38,7 +34,7 @@ plan <- drake_plan(
   ## Analyse site differences using supervised random forest
   site_differences_rf = analysis_suitable_data %>%
     stratifiedRF(pheno_data_with_additional_descriptors$Location %>% factor(),
-       n = 100),
+                 n = 100),
   
   ## Calculate margin for site differences
   site_rf_margin = site_differences_rf %>%
@@ -73,7 +69,7 @@ plan <- drake_plan(
   ## re-analyse site differences after site adjustment
   site_differences_rf_post_adjustment = site_adjusted_analysis_suitable_data %>%
     stratifiedRF( pheno_data_with_additional_descriptors$Location %>% factor(),
-       n = 100),
+                  n = 100),
   
   ## Calculate margin for site differences post adjustment
   site_rf_post_adjustment_margin = site_differences_rf_post_adjustment %>%
@@ -97,10 +93,6 @@ plan <- drake_plan(
     calcDIs(invertPDI = TRUE,invertDAI = FALSE) %>%
     bind_cols(site_adjusted_pheno_data %>%
                 select(Location,ID,`Tree No`,Status,ChosenGroup)),
-  
-  ## export decline indexes
-  export_decline_indexes = decline_indexes %>%
-    write_csv(file_out('data/exports/decline_indexes.csv')),
   
   ## plot decline indexes
   decline_indexes_plot = declineIndexesPlot(decline_indexes),
@@ -265,6 +257,43 @@ plan <- drake_plan(
                       quiet = T,
                       output_format = 'all'),
   
-  ## render supplementary
-  supplementary = render(knitr_in('manuscript/supplementary.Rmd'),quiet = T)
+  ## create directory for supplementary material
+  supplementary_directory = {
+    if (!dir.exists('manuscript/supplementary_material')){
+      dir.create('manuscript/supplementary_material')
+    }
+  },
+  
+  ## render supplementary figures
[…]

### 2020-08-12 manuscript/supplementary_figures.Rmd
rearrange of supplementary materials

--- 2020-07-24 manuscript/supplementary.Rmd
@@
 header-includes:
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{wrapfig}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{pdflscape}
-  - \usepackage{tabu}
-  - \usepackage{threeparttable}
-  - \usepackage{threeparttablex}
-  - \usepackage[normalem]{ulem}
-  - \usepackage{makecell}
-  - \usepackage{xcolor}
+  - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
 ---
@@
 
-```{r phenotypic_descriptors,cache=FALSE}
-loadd(phenotypic_descriptor_info)
-
-phenotypic_descriptor_info %>%
-  select(-`Descriptor Type`,-`Data Type`) %>%
-  kable(format = 'latex',
-        booktabs = TRUE,
-        longtable = TRUE,
-        caption = 'Phenotypic descriptors collected for oak decline severity.') %>%
-  kable_styling(latex_options = c('striped','repeat_header'),repeat_header_continued = TRUE) %>%
-  # column_spec(1,width = '1.5cm') %>%
-  column_spec(1,width = '3cm') %>%
-  column_spec(2:3,width = '6cm') #%>%
-  # collapse_rows(columns = 1,latex_hline = "major", valign = "middle")
-```
+\beginsupplement
 
@@
   wrap_plots()
-```
-
-```{r descriptor_adjustment_factors}
-loadd(site_adjustment_factors)
-
-site_adjustment_factors %>%
-  mutate_if(is.double,signif,digits = 3) %>%
-  kable(format = 'latex',booktabs = TRUE,caption = 'Descriptor adjustment factors at each site.') %>%
-  collapse_rows(columns = 1,latex_hline = "major", valign = "middle")
 ```

### 2020-08-12 manuscript/supplementary_tables.Rmd
rearrange of supplementary materials

--- 2020-08-12 manuscript/supplementary.Rmd
@@
   kable_styling(latex_options = c('striped','repeat_header'),repeat_header_continued = TRUE) %>%
-  # column_spec(1,width = '1.5cm') %>%
   column_spec(1,width = '3cm') %>%
-  column_spec(2:3,width = '6cm') #%>%
-  # collapse_rows(columns = 1,latex_hline = "major", valign = "middle")
-```
-
-```{r descriptor_adjustment,fig.cap='Exapmle of overall mean adjustment applied to the diameter at breast height (m) descriptor.
-Error bars show ±1SE of the mean.
-Dashed red line shows the overall mean.'}
-loadd(descriptor_adjustment_example)
-
-descriptor_adjustment_example %>%
-  wrap_plots()
+  column_spec(2:3,width = '6cm') 
 ```
@@
 ```
-
-```{r modelTuningResultsPlot,fig.cap='Random forest parameter tuning results for decline index models.'}
-loadd(rf_tune_plot)
-
-rf_tune_plot
-```
-
-```{r site_differences,fig.cap='Differences between sites in phenotypic descriptors analysed by supervised random forest classification.
-a) Multidimensional scaling (MDS) scatter plot of random forest proximity values of individual trees scaled to two dimensions.
-b) Descriptor mean decrease in accuracy importances measures.
-Descriptors ordered in descending importance.',fig.height=6}
-loadd(site_differences_mds_plot)
-grid.arrange(site_differences_mds_plot$a,site_differences_mds_plot$b,nrow = 1)
-```
-
-```{r site_differences_corrected,fig.cap='Differences between sites in phenotypic descriptors after adjustment of tree size descriptors analysed by supervised random forest classification.
-a) Multidimensional scaling (MDS) scatter plot of random forest proximity values of individual trees scaled to two dimensions.
-b) Descriptor mean decrease in accuracy importances measures.
-Descriptors ordered in descending importance.',fig.height=6}
-loadd(site_differences_post_adjustment_mds_plot)
-grid.arrange(site_differences_post_adjustment_mds_plot$a,site_differences_post_adjustment_mds_plot$b,nrow = 1)
-```
-
-```{r PDInumeric,fig.height=9,fig.cap='Continuous descriptors plotted against the PDI.'}
-loadd(descriptor_scatter_plots)
-
-descriptor_scatter_plots$PDI_numeric
-```
-
-```{r PDIfactor,fig.height=8,fig.cap='Categorical descriptors plotted against the PDI.'}
-descriptor_scatter_plots$PDI_factor
-```
-
-```{r DAInumeric,fig.height=9,fig.cap='Continuous descriptors plotted against the DAI.'}
-descriptor_scatter_plots$DAI_numeric
-```
-
-```{r DAIfactor,fig.height=8,fig.cap='Categorical descriptors plotted against the DAI.'}
-descriptor_scatter_plots$DAI_factor
-```

### 2020-08-12 R/plan.R

@@ -259,41 +259,59 @@ plan <- drake_plan(
   
   ## create directory for supplementary material
   supplementary_directory = {
-    if (!dir.exists('manuscript/supplementary_material')){
-      dir.create('manuscript/supplementary_material')
+    if (!dir.exists('manuscript/supplementary_materials')){
+      dir.create('manuscript/supplementary_materials')
     }
   },
   
   ## render supplementary figures
   supplememtary_figures = supplementary_directory %>%
     {
-      render(knitr_in('manuscript/supplementary.Rmd'),
+      render(knitr_in('manuscript/supplementary_figures.Rmd'),
              quiet = T,
-             output_dir = 'manuscript/supplementary_material/supplementary_figures.pdf')
+             output_file = 'supplementary_materials/supplementary_figures.pdf')
     },
   
-  ## Supplementary table S1
-  table_S1 =  supplementary_directory %>%
+  ## render supplementary tables S1 & 2
+  supplememtary_tables = supplementary_directory %>%
+    {
+      render(knitr_in('manuscript/supplementary_tables.Rmd'),
+             quiet = T,
+             output_file = 'supplementary_materials/supplementary_tables.pdf')
+    },
+  
+  ## Supplementary table S3
+  table_S3 =  supplementary_directory %>%
     {
       pheno_data_with_additional_descriptors %>%
-        write_csv(file_out('manuscript/supplementary_material/Table_S1.csv'))
+        write_csv(file_out('manuscript/supplementary_materials/Table_S3.csv'))
     },
   
-  ## Supplementary table S2
-  table_S2 =  supplementary_directory %>%
+  
+  
+  ## Supplementary table S4
+  table_S4 =  supplementary_directory %>%
     {
       decline_indexes %>%
-        write_csv(file_out('manuscript/supplementary_material/Table_S2.csv'))
+        write_csv(file_out('manuscript/supplementary_materials/Table_S4.csv'))
     },
   
   ## zip supplementary material
   supplementary_zip = {
     invisible(supplementary_directory)
     invisible(supplememtary_figures)
-    invisible(table_S1)
-    invisible(table_S2)
+    invisible(table_S3)
+    invisible(table_S4)
     
-    zip(file_out('manuscript/supplementary_material.zip'),
-                          list.files('manuscript/supplementary_material',full.names = TRUE))
+    zipfile <- 'manuscript/supplementary_materials.zip'
+    
[…]

### 2020-08-12 manuscript/manuscript.Rmd
Moved abstract to own page
Added reference to supplementary table S3 in manuscript

@@
 **Keywords:** Acute Oak Decline, Chronic Oak Decline, phenotyping, machine learning, unsupervised learning, random forest, decline spectrum
+
+\newpage
 
@@
 
-The collected phenotypic descriptors related to tree size, crown condition and the presence of biotic agents including insects, bacteria and fungi are listed and described in Supplementary Table 1.
+The collected phenotypic descriptors related to tree size, crown condition and the presence of biotic agents including insects, bacteria and fungi are listed and described in Supplementary Table S1.
 Surveyors involved in phenotyping the trees used in this study had prior experience conducting ICP-Forests level I crown condition surveys. 
@@
 Other categorical descriptors included social class, dieback type, main crown dieback location, stem epicormic growth, branch epicormic growth, extent of insect damage, insect damage type, extent of crown mildew and the presence or absence of pruning damage or branch loss from the main stem.
-See Supplementary Table 1 for category details.
+See Supplementary Table S1 for category details.
 
@@
 This then gave a total of `r ncol(analysis_suitable_data)` raw and composite phenotypic descriptors.
+The data for these phenotypic descriptors for all `r nrow(analysis_suitable_data)` trees can be found in Supplementary Table S3.
 The composite descriptors are defined as follows:
@@
 To account for age, size and site specific tree differences, the tree size descriptors (DBH, lower crown height, timber height, total height, crown radius) were standardised using an adjustment of site means to the overall mean.
-An example of this adjustment is shown for DBH in Supplementary Figure 1 and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table 2.  
+An example of this adjustment is shown for DBH in Supplementary Figure S1 and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table S2.  
 All composite descriptors utilising these adjusted descriptors were recalculated after adjustment.
@@
 The parameters number of trees ($ntree$) and number of randomly assigned variables ($mtry$), for each random forest model, were tuned prior to training using the values of 10, 100, 1,000, 10,000 for $ntree$ and a range of 2-26 at intervals of 4 for the $mtry$ parameter.
-The combination parameters giving the lowest mean absolute percentage error (MAPE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, MAPE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, MAPE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary Figure 2).
+The combination parameters giving the lowest mean absolute percentage error (MAPE) were selected and set as follows: for the PDI ($M_{PDI}$): `r glue('$ntree$ = {prettyNum(PDI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {PDI_rf_tune_params$mtry}, MAPE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; for the DAI ($M_{DAI}$): `r glue('$ntree$ = {prettyNum(DAI_rf_tune_params$ntree,big.mark = ',')}, $mtry$ = {DAI_rf_tune_params$mtry}, MAPE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary Figure S2).
 Overall descriptor importance contributions to each of the indexes was assessed using the percentage increase in mean squared error measure with higher values indicating a greater overall contribution to the given model and therefore greater importance.
@@
 
-A preliminary analysis using supervised random forest classification could separate trees based on their location (Supplementary Figure 3a) with an average *margin* value of `r site_rf_margin$Margin %>% signif(digits = 3)`.
-The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary Figure 3b).
+A preliminary analysis using supervised random forest classification could separate trees based on their location (Supplementary Figure S3a) with an average *margin* value of `r site_rf_margin$Margin %>% signif(digits = 3)`.
+The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary Figure S3b).
 To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in Section \@ref(method-adjustment).
@@
 After the application of the tree size adjustment, the average random forest *margin* value was substantially reduced to `r site_rf_post_adjustment_margin$Margin %>% signif(digits = 3)`.
-The reduction in this value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were no longer able accurately classify the site locations of the trees (Supplementary Figure 4a) with almost no contribution of the corrected tree size descriptors (Supplementary Figure 4b).
+The reduction in this value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were no longer able accurately classify the site locations of the trees (Supplementary Figure S4a) with almost no contribution of the corrected tree size descriptors (Supplementary Figure S4b).
 
@@
 
-The calculated PDI and DAI scores (see Supplementary information #) derived from MDS of proximity values from unsupervised random forest for all `r nrow(site_adjusted_analysis_suitable_data)` trees for which phenotypic descriptor data was collected is shown in Figure \@ref(fig:decline-index-plot).
+The calculated PDI and DAI scores (see Supplementary Table S4) derived from MDS of proximity values from unsupervised random forest for all `r nrow(site_adjusted_analysis_suitable_data)` trees for which phenotypic descriptor data was collected is shown in Figure \@ref(fig:decline-index-plot).
 The healthy trees grouped towards the lower end of the PDI scale and the symptomatic COD and AOD groups towards the higher end.
@@
 
-Plots of the individual descriptors against the decline indexes are available in Supplementary Figures 5-8. 
+Plots of the individual descriptors against the decline indexes are available in Supplementary Figures S5-S8. 
 A web application is also available at https://jasenfinch.shinyapps.io/decliner/ that allows interactive input of phenotypic descriptor values into the $M_PDI$ and $M_DAI$ models.

### 2020-08-12 R/plan.R

@@ -284,6 +284,7 @@ plan <- drake_plan(
   table_S3 =  supplementary_directory %>%
     {
       pheno_data_with_additional_descriptors %>%
+        select(-`Tree No`) %>%
         write_csv(file_out('manuscript/supplementary_materials/Table_S3.csv'))
     },
   

### 2020-08-17 manuscript/manuscript.Rmd
proofing

@@
 Over time, as lesion development continues beneath the bark, external cracks develop between the bark plates that are surrounded by blackening stains.
-Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of calloused tissue on the exterior [@denman2014].
+Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of callused tissue on the exterior [@denman2014].
 Stem bleeds can also be associated with a range of root pathogens including *Armillaria* attack on the inner stem tissues of oak.
@@
 
-Here, we propose a framework using unsupervised random forest machine learning to enable quantitative, visible oak decline severity assessments that utilise a range of oak decline informative phenotypic descriptors observed in oak forests across England.
+Here, we propose a framework using unsupervised random forest machine learning to enable quantitative, visible oak decline severity assessments that utilise a range of oak decline informative phenotypic descriptors, observed in oak forests across England.
 
@@
 Surveyors involved in phenotyping the trees used in this study had prior experience conducting ICP-Forests level I crown condition surveys. 
-In each year of sampling surveyors met to undergo refresher training and “calibration” assessments in oak woodland. 
+In each year of sampling, surveyors met to undergo refresher training and “calibration” assessments in oak woodland. 
 The descriptors collected were based on @innes1990 and @icpforests2016.
@@
 
-The AOD stem symptoms (active bleeds, black staining, calloused wounds) were collected by measuring longest axis of each individual bleed, stain or calloused wound found within the basal 3m of the stem.
+The AOD stem symptoms (active bleeds, black staining, callused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem.
 The frequency of *A.
@@
 To account for age, size and site specific tree differences, the tree size descriptors (DBH, lower crown height, timber height, total height, crown radius) were standardised using an adjustment of site means to the overall mean.
-An example of this adjustment is shown for DBH in Supplementary Figure S1 and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table S2.  
+An example of this adjustment is shown for DBH in Supplementary Figure S1 and the adjustment factors for each of the descriptors at each site are given in Supplementary Table S2.  
 All composite descriptors utilising these adjusted descriptors were recalculated after adjustment.
@@
 The remission trees were also found to be significantly different to the AOD (`r glue("t({remission_DAI_group_ttest$parameter[1]}) = {signif(remission_DAI_group_ttest$statistic[1],digits = 3) %>% abs()}, p = < 0.001")`) and COD (`r glue("t({remission_DAI_group_ttest$parameter[2]}) = {signif(remission_DAI_group_ttest$statistic[2],digits = 3) %>% abs()}, p = {round(remission_DAI_group_ttest$p.value[2],3)}")`) groups on the DAI scale with group means of `r signif(remission_DAI_group_ttest$estimate1[1],digits = 3)`, `r signif(remission_DAI_group_ttest$estimate2[1],digits = 3)` and `r signif(remission_DAI_group_ttest$estimate1[2],digits = 3)` for the AOD, remission and COD groups respectively.
-As a group, the remission trees were found to be neither AOD or COD.
+As a group, the remission trees were found to be neither distinctively AOD or COD.
 
@@
 Plots of the individual descriptors against the decline indexes are available in Supplementary Figures S5-S8. 
-A web application is also available at https://jasenfinch.shinyapps.io/decliner/ that allows interactive input of phenotypic descriptor values into the $M_PDI$ and $M_DAI$ models.
+A web application is also available at https://jasenfinch.shinyapps.io/decliner/ that allows interactive input of phenotypic descriptor values into the $M_{PDI}$ and $M_{DAI}$ models.
 The following sections (3.4, 3.5) identify how the descriptors contribute and interact within the decline index models in specific decline cases.
@@
         caption = 'LIME results for oak decline severity cases in the PDI index model $M_{PDI}$.
-Top four important descriptors show.
+Top four important descriptors shown.
 Descriptors ordered by descending absolute weight within each case.') %>%
@@
 However, previous studies such as @evans2012 and @pontius2014 have treated these descriptors as linear progressions of oak health that contribute equally to overall tree condition.
-It has been shown here that in the given visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure \@ref(descriptor-contributions)a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure \@ref(fig:PDI-response-surfaces)a).
+It has been shown here that in the given visual definitions of oak decline severity; firstly, factors of overall crown condition and size are unlikely to contribute equally (Figure \@ref(fig:descriptor-contributions)a) and secondly, it is unlikely that these aspects of crown condition contribute linearly (Figure \@ref(fig:PDI-response-surfaces)a).
 

### 2020-08-17 R/functions/descriptorInfo.R

@@ -42,8 +42,8 @@ descriptorInfo <- function(analysisTable){
     'Active bleed length (mm)','Average active bleed length in the 3m section from the base of the tree.','mm',
     'Black staining','Number of black stains in the 3m section from the base of the tree. Inactive bleeds.','fequency',
     'Black staining length (mm)','Average black stain length in the 3m section from the base of the tree.','mm',
-    'Calloused wound','Number of callused wounds in the 3m section from the base of the tree. Scars.','frequency',
-    'Calloused wound length (mm)','Average callouse wound length in the 3m section from the base of the tree.','mm',
+    'Callused wound','Number of callused wounds in the 3m section from the base of the tree. Scars.','frequency',
+    'Callused wound length (mm)','Average callouse wound length in the 3m section from the base of the tree.','mm',
   ) %>%
     mutate(`Descriptor Type` = c(rep('tree size',5),
                                  rep('crown',13),

### 2020-08-17 manuscript/submission/author_contributions.Rmd
author contributions now output as word document

---
output: word_document
---

### CRediT authorship contribution statement

**Jasen Finch:** Conceptualization, Methodology, Software, Formal analysis, Data Curation, Visualization, Writing - Original Draft, Writing - review & editing.
**Nathan Brown:** Methodology, Investigation, Data Curation, Writing - Review & Editing. 
**Manfred Beckmann:** Writing - Review & Editing, Supervision.
**Sandra Denman:** Methodology, Writing - Review & Editing, Project administration, Funding acquisition.
**John Draper:** Methodology, Supervision, Project administration.

### 2020-08-17 R/plan.R
added highlights

@@ -318,5 +318,9 @@ plan <- drake_plan(
   
   ## Author contributions
   author_contributions = render(knitr_in('manuscript/submission/author_contributions.Rmd'),
-                                quiet = TRUE)
+                                quiet = TRUE),
+  
+  ## Highlights
+  highlights = render(knitr_in('manuscript/submission/highlights.Rmd'),
+                      quiet = TRUE)
 )

### 2020-08-18 manuscript/manuscript.Rmd
Abstract rewrite

@@
 
-Oak decline is becoming increasingly prevalent along with the need for accurate and quantitative oak health monitoring.
-The Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI) were developed to quantitatively describe and differentiate acute and chronic oak decline severity using a multitude of easily measurable phenotypic descriptors.
-Measured phenotypic descriptors included aspects of tree shape and size, crown condition and the presence of biotic agents.
-The decline indexes were derived using proximity values from unsupervised random forest of the phenotypic descriptors, multidimensional scaling (MDS) and min-max scaling.
-These indexes give simple but sensitive measures of tree decline severity allowing easy comparisons of oak trees both spatially and temporally, providing a potentially useful tool for forest monitoring and management.
+Oak decline syndromes such as Acute Oak Decline (AOD) and Chronic Oak Decline (COD) are becoming increasingly prevalent and with this comes the need for more quantitative and sensitive visual oak health monitoring.
+
+Phenotyping protocols were developed to specifically measure oak decline severity and were based on a multitude of simple to measure phenotypic descriptors. 
+There were a total of 36 phenotypic measurements describing oak decline status which included aspects of tree size, crown condition, the presence of biotic agents and a number of derived composite descriptors.
+
+Phenotypic measurements were collected from a total of 174 *Quercus robur*, surveyed from 9 sites across England and included healthy, AOD, COD and AOD trees in remission.
+Using these data, the Phenotypic Decline Index (PDI) and the Decline Acuteness Index (DAI) were developed to quantitatively describe and differentiate the acute and chronic oak decline severity spectrum.
+These decline indexes were derived from unsupervised random forest machine learning models, trained using the collected phenotypic data.
+
+The suitability of the decline indexes for describing decline severity and type were assessed by comparing decline index scores to manual decline status classifications along with an assessment of descriptor importance and contribution to the decline index models.
+Crown condition and trees size descriptors such as composite crown volume contributed positively to the PDI.
+Trees with smaller crowns in poor condition had greater PDI values. 
+Tree stature and the presence of stem bleeding contributed highly to the DAI, allowing differentiation between trees with AOD and COD syndromes. 
+AOD trees had relatively larger stature and the presence of stem bleeding where as COD trees had small stature and stem bleeding absent.
+
+The oak decline indexes are simple but sensitive measures of tree decline severity and allow easy comparisons of oak trees both spatially and temporally. 
+These have the potential to provide useful tools for forest monitoring and management as well application to remote sensing and omics research.
 

### 2020-08-18 manuscript/submission/cover_letter.Rmd
added manuscript cover letter

---
output: word_document
---

Dear Editor

Please find attached our manuscript entitled **“Index measures for oak decline severity using phenotypic descriptors” ** to be considered for publication in Forest Ecology and Management. 
I am submitting this on behalf of my co-authors who are happy for me to act as corresponding author for this work.

This manuscript outlines the development of two indexes for measuring the visual health and severity of oak decline in *Quercus robur* and to quantitatively differentiate Acute Oak Decline and Chronic Oak Decline syndromes.
The methodology for deriving these indexes is based on an unsupervised machine learning approach and enables the objective comparison of visual oak health over large geographical areas and over the course of numerous growing seasons.

This work is a progression of work on Acute Oak Decline symptom expression by my co-authors Nathan Brown and Sandra Denman, previously published in this journal.
We believe it will provide benefit to both remote sensing and omics based approaches for forest health monitoring and have utility as a decision making aid for forest management strategies.
It is for this reason that we feel that Forest Ecology and Management is the ideal journal for presenting this novel methodology.
Further work is already underway to build upon the models presented in the manuscript as well as the development of web based tools to improve their accessibility and use in the forestry community.

We thank you for your time in considering this submission.

Best wishes,

Jasen Finch

### 2020-08-18 R/plan.R

@@ -322,5 +322,9 @@ plan <- drake_plan(
   
   ## Highlights
   highlights = render(knitr_in('manuscript/submission/highlights.Rmd'),
-                      quiet = TRUE)
+                      quiet = TRUE),
+  
+  ## Cover letter
+  cover_letter = render(knitr_in('manuscript/submission/cover_letter.Rmd'),
+                        quiet = TRUE)
 )

### 2020-08-18 manuscript/submission/cover_letter.Rmd
corrections to cover letter

@@
 This manuscript outlines the development of two indexes for measuring the visual health and severity of oak decline in *Quercus robur* and to quantitatively differentiate Acute Oak Decline and Chronic Oak Decline syndromes.
-The methodology for deriving these indexes is based on an unsupervised machine learning approach and enables the objective comparison of visual oak health over large geographical areas and over the course of numerous growing seasons.
+The methodology for deriving these indexes is based on an unsupervised machine learning approach.
+The indexes enable the objective comparison of visual oak health over large geographical areas and over the course of numerous growing seasons.
 
-This work is a progression of work on Acute Oak Decline symptom expression by my co-authors Nathan Brown and Sandra Denman, previously published in this journal.
+This work is a progression of previous research on Acute Oak Decline symptom expression by my co-authors Nathan Brown and Sandra Denman, published in this journal.
 We believe it will provide benefit to both remote sensing and omics based approaches for forest health monitoring and have utility as a decision making aid for forest management strategies.
-It is for this reason that we feel that Forest Ecology and Management is the ideal journal for presenting this novel methodology.
+For this reason, we feel that Forest Ecology and Management is the ideal journal for presenting this novel methodology.
 Further work is already underway to build upon the models presented in the manuscript as well as the development of web based tools to improve their accessibility and use in the forestry community.

### 2020-10-27 manuscript/manuscript.Rmd
added Sandra edits, fixed referencing

@@
     toc: false
+    citation_package: natbib
 fontsize: 12pt
@@
 bibliography: references.bib
+biblio-style: abbrvnat
+natbiboptions: round
 ---
@@
 
-Oak decline syndromes such as Acute Oak Decline (AOD) and Chronic Oak Decline (COD) are becoming increasingly prevalent and with this comes the need for more quantitative and sensitive visual oak health monitoring.
+Oak decline syndromes such as Acute Oak Decline (AOD) and Chronic Oak Decline (COD) are becoming increasingly prevalent and with this comes the need for more quantitative, sensitive and standardised visual oak health monitoring.
 
@@
 Trees and forests across the globe face threats from a changing climate.
-Notably warming and an increase in drought episodes affect the health status of individual trees, causing vegetation changes at the landscape scale [@choat2012; @millar2015; @allen2015]. 
+Notably warming and an increase in drought episodes can adversely affect the health status of individual trees, causing vegetation changes at the landscape scale [@choat2012; @millar2015; @allen2015].
 Within this context, biotic agents (pests and diseases) can flourish and often drive mortality [@kautz2017; @jactel2012]. 
@@
 Abiotic factors can include both long-term predisposing and one-off inciting factors.
-Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].
+Biotic agents can include insect infestation such as the larvae of the bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].
 The bacterial stem bleeds associated with AOD initially start as active dark weeping patches.

### 2020-10-27 manuscript/manuscript.Rmd
manuscript corrections

@@
 The bacterial stem bleeds associated with AOD initially start as active dark weeping patches.
-Over time, as lesion development continues beneath the bark, external cracks develop between the bark plates that are surrounded by blackening stains.
+Over time, as lesion development continues beneath the bark, external cracks form between the bark plates that are surrounded by blackening stains.
 Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of callused tissue on the exterior [@denman2014].
@@
 AOD trees had current stem bleeds but a variety of crown conditions. 
-COD trees either had poor crowns (category 2) or the fruiting bodies of *Gymnopus* or *Armillaria* had been observed in previous years.
+COD trees either had poor crowns (category 2) or the fruiting bodies of *Gymnopus* or *Armillaria* species had been observed in previous years.
 The remission group included trees that had shown stem bleeds associated with AOD within the previous three years but were not showing signs of stem bleeds at the time of assessment.
@@
 Surveyors involved in phenotyping the trees used in this study had prior experience conducting ICP-Forests level I crown condition surveys. 
-In each year of sampling, surveyors met to undergo refresher training and “calibration” assessments in oak woodland. 
+In each year of sampling, surveyors underwent refresher training and “calibration” assessments in oak woodland. 
 The descriptors collected were based on @innes1990 and @icpforests2016.
@@
 Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m.
-The timber height was measured at either where the top stem diameter was 70mm or where no main stem could be distinguished.
+The timber height was measured at either where the upper stem diameter was 70mm or where no main stem could be distinguished.
 For crown contact, missing crown and absolute crown transparency, percentages were visually estimated between 0 and 100% in 5% interval ranges.
@@
 The crown radius was measured using a vertex transponder attached to the relevant side of the stem at a height of 1.3m. 
-Canopy closure was recorded as yes or no depending on whether any other crowns are touching the sampling tree.
+Canopy closure was recorded as yes or no depending on whether any other crowns were touching the sampling tree.
 A high resolution photo was taken of the base to the top of the crown and of the stem from the base to a height of 3m for each cardinal point for later reference.
@@
 The frequency of *A.
-biguttatus* exit holes were counted along with the presence or absence of other oval shaped insect exit holes, small circular shaped exit holes and ground level fungal fruiting bodies which were recorded around the stem circumference from the basal tree collar to a radial distance of 2m of the stem.
+biguttatus* exit holes was counted along with the presence or absence of other oval shaped insect exit holes, small circular shaped exit holes and ground level fungal fruiting bodies which were recorded around the stem circumference from the basal tree collar to a radial distance of 2m of the stem.
 
@@
 All cardinal assessments were averaged to give a single value per tree.
-The crown density ($CD$) values collected at Attingham and Langdale were converted to crown transparency ($t = 100 - CD$)
+The crown density ($CD$) values collected at Site 1 and 6 were converted to crown transparency ($t = 100 - CD$)
 All descriptors collected in cm were converted to mm except for DBH which was converted to m.
@@
 
-Agrilus exit hole density (m^-2^) - density of observed *A. biguttatus* exit holes across the surveyed 2m basal trunk area.
+*Agrilus* exit hole density (m^-2^) - density of observed *A. biguttatus* exit holes across the surveyed 2m basal trunk area.
 
@@
 The descriptors that contributed most to this discrimination were those that were related to trees size such as DBH, total height and crown radius and likely reflect the varied tree age and planting density across the nine sites (Supplementary Figure S3b).
-To counter this, effect sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in Section \@ref(method-adjustment).
+To counter this effect, sizes were adjusted based on the site means for each tree size descriptor to remove inter-site variability but maintain intra-site variability as described in Section \@ref(method-adjustment).
 This ensures that the subsequent comparisons of these descriptors in the context of oak decline are not confounded by these identified site differences.

### 2020-10-28 manuscript/manuscript.Rmd
manuscript corrections

@@
 
-Phenotyping protocols were developed to specifically measure oak decline severity and were based on a multitude of simple to measure phenotypic descriptors. 
-There were a total of 36 phenotypic measurements describing oak decline status which included aspects of tree size, crown condition, the presence of biotic agents and a number of derived composite descriptors.
+Phenotyping protocols were developed to specifically measure oak decline severity and were based on a comprehensive set of simple to measure phenotypic descriptors. 
+A total of 36 phenotypic measurements describing oak decline status included aspects of tree size, crown condition, the presence of biotic agents and a number of derived composite descriptors.
 
@@
 The suitability of the decline indexes for describing decline severity and type were assessed by comparing decline index scores to manual decline status classifications along with an assessment of descriptor importance and contribution to the decline index models.
-Crown condition and trees size descriptors such as composite crown volume contributed positively to the PDI.
+Crown condition and trees size descriptors such as 'composite crown volume' contributed positively to the PDI.
 Trees with smaller crowns in poor condition had greater PDI values. 
 Tree stature and the presence of stem bleeding contributed highly to the DAI, allowing differentiation between trees with AOD and COD syndromes. 
-AOD trees had relatively larger stature and the presence of stem bleeding where as COD trees had small stature and stem bleeding absent.
+AOD trees had relatively larger stature and the presence of stem bleeding where as COD trees had small stature and stem bleeding was absent.
 
@@
 Monitoring vegetation changes and understanding the underlying processes affecting tree health are important first steps to predicting the scale and extent of these threats [@hartmann2018] and plan future policies for planting and mitigation efforts [@bradford2017].
-Current ‘omics and remote sensing approaches provide powerful tools to document and understand these changes across multiple scales (from cellular to tree to landscape scales), but an important prerequisite is the ability to accurately describe tree health status, so that even subtle changes can be documented [@dungey2018; @hornero2020].
+Current omics and remote sensing approaches provide powerful tools to document and understand these changes across multiple scales (from cellular to tree to landscape scales), but an important prerequisite is the ability to accurately describe tree health status, so that even subtle changes can be documented [@dungey2018; @hornero2020].
 Phenotyping, documenting the physical traits of an individual or group, is thus a crucial first step relevant to studies across disciplines and further, it is necessary to understand how these measurements combine to describe health status.
@@
 Abiotic factors can include both long-term predisposing and one-off inciting factors.
-Biotic agents can include insect infestation such as the larvae of the bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].
+Biotic agents can include insect infestation such as larvae of the bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated with Chronic Oak Decline (COD) [@denman2017].
 The bacterial stem bleeds associated with AOD initially start as active dark weeping patches.
@@
 The occurrence of AOD has previously been reported and studied at these sites [@brown2016; @brown2017a].
-These sites included a variety of woodlands, management histories as well as tree sizes/ages and were made up of mainly *Q. robur*.
+These sites included a variety of woodlands with diverse management histories, tree sizes/ages and were made up of mainly *Q. robur*.
 
@@
 
-The collected phenotypic descriptors related to tree size, crown condition and the presence of biotic agents including insects, bacteria and fungi are listed and described in Supplementary Table S1.
+The phenotypic descriptors related to tree size, crown condition and the presence of biotic agents including insects, bacteria and fungi are listed and described in Supplementary Table S1.
 Surveyors involved in phenotyping the trees used in this study had prior experience conducting ICP-Forests level I crown condition surveys. 
@@
 Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 1mm.
-Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m.
+Total height, timber height and lowest live crown were measured using a Vertex IV Digital Hypsometer (Haglof, Sweden) to the nearest 0.1m.
 The timber height was measured at either where the upper stem diameter was 70mm or where no main stem could be distinguished.
@@
 Crown radius, canopy closure and bark tap tests for dead stem tissue were measured for each cardinal point.
-The crown radius was measured using a vertex transponder attached to the relevant side of the stem at a height of 1.3m. 
-Canopy closure was recorded as yes or no depending on whether any other crowns were touching the sampling tree.
+The crown radius was measured using a Vertex IV Digital Hypsometer attached to the relevant side of the stem at a height of 1.3m. 
+Canopy closure was recorded as yes or no depending on whether any other crowns were touching the sampled tree.
 A high resolution photo was taken of the base to the top of the crown and of the stem from the base to a height of 3m for each cardinal point for later reference.
@@
 The frequency of *A.
-biguttatus* exit holes was counted along with the presence or absence of other oval shaped insect exit holes, small circular shaped exit holes and ground level fungal fruiting bodies which were recorded around the stem circumference from the basal tree collar to a radial distance of 2m of the stem.
+biguttatus* exit holes was counted along with the presence or absence of other oval shaped insect exit holes, small circular shaped exit holes and ground level fungal fruiting bodies which were recorded around the stem circumference from the basal tree collar to a radial distance of 2m from the stem.
 
@@
 All cardinal assessments were averaged to give a single value per tree.
-The crown density ($CD$) values collected at Site 1 and 6 were converted to crown transparency ($t = 100 - CD$)
+The crown density ($CD$) values collected at Site 1 and 6 were converted to crown transparency ($t = 100 - CD$).
 All descriptors collected in cm were converted to mm except for DBH which was converted to m.
@@
 After the application of the tree size adjustment, the average random forest *margin* value was substantially reduced to `r site_rf_post_adjustment_margin$Margin %>% signif(digits = 3)`.
-The reduction in this value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were no longer able accurately classify the site locations of the trees (Supplementary Figure S4a) with almost no contribution of the corrected tree size descriptors (Supplementary Figure S4b).
+The reduction in this value indicates that the adjustment was able to remove the majority of the between site variance and that the random forest models were no longer able to accurately classify the site locations of the trees (Supplementary Figure S4a) with almost no contribution of the corrected tree size descriptors (Supplementary Figure S4b).
 
@@
 ```{r decline-index-plot,fig.cap='Decline indexes for all 174 surveyed trees.
-The chosen group categories were selected in the field prior to data collection.
+The decline status categories were manually assigned in the field prior to data collection.
 AOD = acute oak decline; COD = chronic oak decline.
@@
 The top four ranked descriptors found to contribute to $M_{PDI}$ were all crown size and condition related descriptors with composite crown volume the top ranked descriptor overall (Figure \@ref(fig:descriptor-contributions)a).
-Other tree size related descriptors such as bleed prevalence and live crown ratio stem bleeding descriptors contributed secondarily.
+Other tree size related descriptors such as bleed prevalence, live crown ratio and stem bleeding descriptors contributed secondarily.
 
@@
 
-AOD severity was able to be associated with overall decline severity with trees identified that had a combination of both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure \@ref(fig:decline-index-plot)).
-The association of crown condition and AOD severity has previously be shown by @brown2016.
+AOD severity was associated with overall decline severity in trees having both PDI values greater than 0.7 and DAI values greater than 0.5 (Figure \@ref(fig:decline-index-plot)).
+The association of crown condition and AOD severity has previously been shown by @brown2016.
 
@@
 However, this was also the most varied of the decline status groups.
-It is possible that this could reflect the potential for trees to recover from AOD but this is speculative and it is not possible to attribute this further with these observations.
+It is possible that this could reflect the potential for trees to recover from AOD but this is speculative and it is not possible to test this suggestion further with current observations.
 Using these decline indexes, remission trees could be more objectively defined in terms of their temporal trajectory within the decline index continuum.
@@
 
-The decline indexes presented here provide a new framework to quantitatively describe the phenotypic oak decline spectrum using intuitive measures severity (PDI) and type (DAI).
+The decline indexes presented here provide a new framework to quantitatively describe the phenotypic oak decline spectrum using intuitive measures of severity (PDI) and type (DAI).
 Their basis is on simple phenotypic descriptor measurements and a utilisation of a data driven, unsupervised random forest based machine learning approach.

### 2021-01-04 manuscript/supplementary_figures.Rmd
box plots added to categorical descriptor supplementary figures

@@
 
-```{r PDIfactor,fig.height=8,fig.cap='Categorical descriptors plotted against the PDI.'}
+```{r PDIfactor,fig.height=9,fig.cap='PDI box plots of the categorical descriptors.'}
 descriptor_scatter_plots$PDI_factor
@@
 
-```{r DAIfactor,fig.height=8,fig.cap='Categorical descriptors plotted against the DAI.'}
+```{r DAIfactor,fig.height=9,fig.cap='DAI box plots of the categorical descriptors.'}
 descriptor_scatter_plots$DAI_factor

### 2021-01-04 manuscript/manuscript.Rmd
descriptor contribution rank links now shown between indexes

@@
 
-```{r descriptor-contributions, fig.height = 6,fig.cap = 'Descriptor importance scores from the random forest models $M_{PDI}$ and $M_{DAI}$ for relative contributions to the decline index measures.
+```{r descriptor-contributions, fig.height = 6,fig.width=8,fig.cap = 'Descriptor importance scores from the random forest models $M_{PDI}$ and $M_{DAI}$ for relative contributions to the decline index measures.
 Descriptors are ordered by descending percentage increase in mean squared error (MSE) and therefore from highest contribution to lowest contribution to the given decline index.'}
-loadd(descriptor_contribution_plots)
+loadd(descriptor_contribution_plot)
 
-descriptor_contribution_plots$a + descriptor_contribution_plots$b
+descriptor_contribution_plot
 ```

### 2021-01-04 R/functions/descriptorImportancePlots.R

@@ -1,10 +1,12 @@
 
 descriptorImportancePlots <- function(PDI_descriptor_importance,DAI_descriptor_importance){
-  list(
+  importance_plots <- list(
     a = {
       dat <- PDI_descriptor_importance %>%
         arrange(`%IncMSE`) %>%
-        mutate(Feature = factor(Feature,levels = Feature)) 
+        mutate(Feature = factor(Feature,levels = Feature),
+               rev_Rank = Feature %>%
+                 seq_along())
       descriptorLabels <- dat$Feature %>%
         as.character() %>%
         {
@@ -13,21 +15,31 @@ descriptorImportancePlots <- function(PDI_descriptor_importance,DAI_descriptor_i
           return(.)
         }
       
-      ggplot(dat,aes(x = Feature,y = `%IncMSE`)) +
+      ggplot(dat,aes(y = Feature,x = `%IncMSE`)) +
+        geom_segment(aes(y = rev_Rank,yend = rev_Rank,x = -3,xend = `%IncMSE`)) +
         geom_point(fill = ptol_pal()(1),shape = 21,size = 2) +
-        coord_flip() +
         theme_bw() +
-        theme(plot.title = element_text(face = 'bold'),
-              axis.title = element_text(face = 'bold',size = 10)) +
+        theme(plot.title = element_text(face = 'bold',hjust = 0.5),
+              axis.title = element_text(face = 'bold',size = 10),
+              panel.grid = element_blank(),
+              panel.border = element_blank(),
+              plot.margin = unit(c(5.5, 0, 5.5, 5.5), "pt"),
+              axis.line.x = element_line(),
+              axis.text.y = element_text(colour = 'black'),
+              axis.ticks.y = element_blank()) +
         labs(title = 'a) PDI',
-             y = '% increase in\nMSE',
-             x = NULL) +
-        scale_x_discrete(labels = descriptorLabels)
+             x = '% increase in\nMSE',
+             y = NULL) +
+        scale_y_discrete(labels = descriptorLabels)  +
+        scale_x_reverse(limits = c(112,-5),
+                        expand = c(0,0))
     },
     b = {
       dat <- DAI_descriptor_importance %>%
         arrange(`%IncMSE`) %>%
-        mutate(Feature = factor(Feature,levels = Feature)) 
+        mutate(Feature = factor(Feature,levels = Feature),
+               rev_Rank = Feature %>%
+                 seq_along())
       
[…]

### 2021-01-05 manuscript/manuscript.Rmd
Forest ecology and management .csl now used for reference format

@@
     toc: false
-    citation_package: natbib
 fontsize: 12pt
@@
 bibliography: references.bib
-biblio-style: abbrvnat
-natbiboptions: round
+csl: forest-ecology-and-management.csl
 ---

### 2021-01-05 manuscript/manuscript.Rmd
elsartical-harv natbib citation style now used for reference format

@@
 bibliography: references.bib
-csl: forest-ecology-and-management.csl
+biblio-style: elsarticle-harv
+# csl: forest-ecology-and-management.csl
 ---
@@
 By defining decline severity objectively across a spectrum, these decline indexes provide greatly enhanced sensitivity over previous categorical decline status assignments.
-This approach could enhance integration with large scale data such as those from genome wide association studies or remote sensing [@stocks2019; @evans2012].
+This approach could enhance integration with large scale data such as those from genome wide association studies or for providing ground-truth data for remote sensing [@stocks2019; @evans2012].
 These index measures allow sensitive comparisons of visual health to be made for the same tree between years as well as comparisons of trees between different locations.

### 2021-01-05 manuscript/manuscript.Rmd
removed redundant manuscript header entry

@@
 biblio-style: elsarticle-harv
-# csl: forest-ecology-and-management.csl
 ---
